---
title: "Victoria Police Search Data and Hit Rates Overview (descriptive statistics)"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

```
# Data Overview  

## About the dataset 

This dataset contains records of **62,350 searches** warrantless searches conducted by Victoria Police in the years **2018, 2019, 2022, 2023 and 2024**. Data for the years **2020 and 2021** is excluded due to the atypical policing environment associated with the COVID-19 pandemic.
 
Descriptive analysis was undertaken to understand whether the data is suitable for racial profiling analysis, specifically using hit-rate methodology. This document also serves as a general data overview for researches interested in using this data to answer other questions about warrantless searches in Victoria. 

📁 **Download the cleaned dataset:**  
[**VicPol Search Data Clean.xlsx**](https://github.com/monikasarder/VictoriaPoliceRacialSearch/raw/main/VicPol%20Search%20Data%20Clean/VicPol%20Search%20Data%20Clean.xlsx)

The dataset comprises records of **warrantless stop and search events** conducted in public places across Victoria. It was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/).

For a warrantless search to be lawful, police must **suspect on reasonable grounds** that the person is carrying a prohibited item, such as drugs, weapons, or graffiti implements.

Each search is recorded under the relevant **legal power** (e.g. *Drugs, Poisons and Controlled Substances Act 1981*, s82) and includes one of three possible **outcomes**:

- **Nothing found**  
- **Search item found**: an item matching the type of contraband suspected (e.g. drugs found during a drug search)  
- **Non-search item found**: a different type of contraband was found (e.g. a weapon found during a drug search)

This dataset has been cleaned and transformed to enable analysis of find rates - more commonly described as 'hit rates' - by unit type and geographical area, where this information is available. Racial categories have also been cleaned and transformed to allow meaningful comparison, as the naming conventions frequently changed in the seven years from 2018 to 2024. All methods are documented at [in the corresponding GitHub](https://github.com/monikasarder/VictoriaPoliceRacialSearch/. The specific racial mapping/cleaning undertaken is summarised in the Appendix below.

## Variables

The following variables are available in this dataset. 

- **Date**: Date of the stop and search  
- **Area.type**: Metropolitan or regional classification  
- **Region**, **Division**, **PSA**, **Station**: Police geographic hierarchy  
- **Racial.appearance.original**: The original officer-perceived racial appearance category as recorded by Victoria Police  
- **Racial.appearance.transformed**: Racial appearance category transformed to enable comparison, adjusting for multiple category name changes across the period 
- **Gender**: Officer-perceived gender of the person searched  
- **Age**: Age of the person searched  
- **Age group**: Age of the person searched  
- **Search.type - Drugs**: Search conducted under *Drugs, Poisons and Controlled Substances Act 1981 s82* — can be carried out anywhere in public with reasonable suspicion of possession of a controlled drug.  
- **Search.type - Weapons**: Search conducted under *Control of Weapons Act 1990 s10* — can be carried out anywhere in public with reasonable suspicion of possession of a controlled weapon, dangerous article, or prohibited weapon.  
- **Search.type - Graffiti**: Search conducted under *Graffiti Prevention Act 2007 s13* — only in specified circumstances (e.g. person aged 14+ in a prescribed area such as public transport property or trespassing). 
- **Search.type - Firearms**: Search conducted under *Firearms Act 1996 s112Q* — only in *specified circumstances* (person subject to a Firearm Prohibition Order).  
- **Search.type - Volatile.sub.U18**: Search conducted under *Drugs, Poisons and Controlled Substances Act 1981 ss60E–60F* — only in *specified circumstances* (person under 18 suspected of possessing/inhaling volatile substances, or supplying them to a child).  
- **Search.type - Volatile.sub.adult**: Search conducted under the same provisions as above — applies to adults only when suspected of supplying volatile substances to a child.  
- **Found**: Whether any prohibited item was found — regardless of whether it was the item searched for under the power invoked (e.g. both “drug power invoked, drugs found” and “drug power invoked, weapon found” count as a 'hit').  
- **Contact.Type**: Whether the search was conducted on foot (`P`) or during a vehicle stop (`V`).  
- **Unit.type**: Broad category of the police unit conducting the search (e.g. uniform, highway patrol).  
- **Station.uniform**: The police station linked to the uniformed officer conducting the search (available for uniform police only).  
- **Rank**: Rank of the officer responsible for conducting the search (e.g. Constable, Senior Constable, Sergeant).  

Note that multiple powers can apply in a single search - for example where police have reasonable grounds for suspecting both weapons and drugs are present.

## Data limitations  

### Racial Appearance Categories

Victoria Police records the **perceived racial appearance** of individuals at the time of the search. These categories reflect how the officer perceives the person’s racial appearance — they are *not* based on self-identification or ancestry and should be understood as indicators of how race is *perceived* in the policing context.

In addition, **2,745 records** are classified under the merged **‘Mediterranean/Middle Eastern’** category. Prior to 2022 these were separate categories. It is now not possible to analyse Victoria Police data for racial profiling of people of Middle Eastern or Arab appearance.

### Missing Data

A substantial proportion of searches were also missing demographic information. In particular, **42% of all searches (25,851 records)** had no recorded racial appearance, while a further **2.5% (924 records)** were recorded as **‘Other’** — a category that may reflect situations where officers were unable or unwilling to assign a racial appearance category.

Missing racial appearance data was not evenly distributed across policing contexts. Searches involving **vehicles** were substantially more likely to have missing racial appearance information than searches of **pedestrians**. 

Uniform police searches of **pedestrians** had the **lowest proportion of missing racial appearance data among the major policing contexts (21.6%)**, as well as the largest number of records (26,394). This makes them a more reliable subset for racial profiling analysis. Restricting the analysis to this operational context supports more meaningful “like with like” comparisons by reducing variation introduced by different unit types, policing functions, and vehicle-based enforcement activity, while also reducing the impact of missing racial appearance data.

**Table 1 Percentage of searches with missing racial appearance information by unit type and contact type**

```{r missing-race-unit-contact, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

library(tidyverse)
library(scales)
library(flextable)

missing_race_tbl <- dat %>%
  mutate(
    Missing_race = case_when(
      is.na(Racial.appearance.transformed) ~ "Missing",
      Racial.appearance.transformed %in% c("Missing", "Other") ~ "Missing",
      TRUE ~ "Recorded"
    ),
    Contact.Type = recode(
      Contact.Type,
      "P" = "Pedestrian",
      "V" = "Vehicle",
      .default = NA_character_
    )
  ) %>%
  group_by(Unit.type, Contact.Type) %>%
  summarise(
    Total = n(),
    Missing_n = sum(Missing_race == "Missing", na.rm = TRUE),
    Missing_pct = Missing_n / Total,
    .groups = "drop"
  ) %>%
  mutate(
    cell = paste0(
      percent(Missing_pct, accuracy = 0.1),
      " (",
      comma(Total),
      ")"
    )
  ) %>%
  select(Unit.type, Contact.Type, cell) %>%
  pivot_wider(
    names_from = Contact.Type,
    values_from = cell
  )

overall_missing_race <- dat %>%
  mutate(
    Missing_race = case_when(
      is.na(Racial.appearance.transformed) ~ "Missing",
      Racial.appearance.transformed %in% c("Missing", "Other") ~ "Missing",
      TRUE ~ "Recorded"
    )
  ) %>%
  group_by(Unit.type) %>%
  summarise(
    Total = n(),
    Missing_n = sum(Missing_race == "Missing", na.rm = TRUE),
    Missing_pct = Missing_n / Total,
    .groups = "drop"
  ) %>%
  mutate(
    Overall = paste0(
      percent(Missing_pct, accuracy = 0.1),
      " (",
      comma(Total),
      ")"
    )
  ) %>%
  select(Unit.type, Overall)

missing_race_tbl <- missing_race_tbl %>%
  left_join(overall_missing_race, by = "Unit.type") %>%
  mutate(
    Unit.type = fct_reorder(Unit.type, Overall)
  ) %>%
  arrange(Unit.type) %>%
  rename(`Unit type` = Unit.type)

flextable(missing_race_tbl) %>%
  set_header_labels(
    Pedestrian = "Pedestrian",
    Vehicle = "Vehicle",
    Overall = "Overall"
  ) %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  align(j = "Unit type", align = "left", part = "all") %>%
  bold(part = "header") %>%
  add_footer_lines(
    values = "Cells show percentage missing racial appearance information, with total number of searches in brackets. ‘Other’ racial appearance is treated as missing/undetermined for this table."
  )
```
# Descriptive statistics  

##  Search types  

Table 2 below summarises the total search powers invoked during searches. A single search could relate to several powers. For example - a person in a vehicle could be searched for both weapons and drugs if police reasonably suspect both types of contraband to be present.  The power most often relied for any given search is [Section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*](https://classic.austlii.edu.au/au/legis/vic/consol_act/dpacsa1981422/s82.html), followed by [Section 10 of the *Control of Weapons Act 1990*](https://classic.austlii.edu.au/au/legis/vic/consol_act/cowa1990217/s10.html).

**Table 2 All search powers that applied in searches by search count**
```{r searched for, echo = FALSE, warning = FALSE}

finds <- dat %>%
  select(starts_with("Search.type"))%>%
  pivot_longer(
    cols = starts_with("Search.type"),
    names_to = "Search type",
    values_to = "Total",
    values_drop_na = TRUE
  ) %>%
  mutate(`Search type` = forcats::fct_infreq(`Search type`))%>%
  select(-Total)


 
print(dfSummary(finds, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Persons searched 

This section summarises the characteristics of individuals searched. Note that some individuals were searched on multiple occasions.

Most searches (67%) involved people perceived by police as White. People perceived as Asian or African each accounted for more than 5% of searches. The second-largest category was “Mediterranean/Middle Eastern”; however, this category cannot be reliably used for racial profiling analysis, as it combines both White-appearing and non-White-appearing individuals.

More than four-fifths of those searched were male. Over half of all searches involved people aged between 18 and 35 years.

Searches of pedestrians (67%) were more common than vehicle stops.

**Table 3 Characteristics of persons searched**
```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group", "Contact.Type")


dat$`Age group`[dat$`Age group` == "Missing"] <- NA  

dat$Racial.appearance.transformed[dat$Racial.appearance.transformed == "Missing"] <- NA  

dat.per <- dat%>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.per <- dat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(dat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police searching 

This section summarises the characteristics of police members and units recorded as conducting searches.  

Over the five years examined, search numbers were relatively consistent, with a slight peak in 2022.

Most searches were conducted by **Uniform police** (72%). **Public Order Response Team (PORT)** members - a specialist Victoria Police unit trained and equipped for large-scale public order operations such as protests and rallies - and **Transit police** both carried out **8%** of warrantless searches.

**Constables** and **Senior Constables** were the ranks most frequently involved, reflecting the predominance of uniformed officers in search activity.


**Table 4 Police and units undertaking searches**
```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Unit.type","Rank.of.Member","Area.type", "Region", "Year")

dat.pol <- dat %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023","2024")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```


# Hit rate statistics

This section summarises the proportion of searches which are 'successful' and result in a find. It should be noted that there are two types of finds - finds of the specific contraband invoked in the search power, and searches of a different contraband. For the majority of this summary we will count *any* prohibited item find as successful. 

##  Hit rates by search type 

Table 4 shows clear variation in hit rates across search types. Searches for **graffiti items** have the highest success rate (**51%**), followed by **Firearms searches** (**21%**). It should be noted that these search types are slightly different from other three as graffiti contraband searches can take place only in a prescribed area (such as a train stations) whereas firearms searches require a Firearm Prohibition order to be in place. So while warrantless, these search types are more focused by design.

Meanwhile, **weapons-related searches** and **drug-related searches**, can be carried out on any person in any public place, as can **volatile substances** (often referred to as “huffing”) searches. The hit rates for searches conducted under these powers are **33% vs 15%** respectively. Volatile substance searches are not reported due to the small sample size(**n < 100**).


**Table 5 Hit rates by search type**
```{r Finds search type, echo=FALSE, warning=FALSE, results='asis'}

finds <- dat %>%
  select(Found, starts_with("Search.type")) %>%
  pivot_longer(
    cols = starts_with("Search.type"),
    names_to = "Search type",
    values_to = "Detailed find",
    values_drop_na = TRUE
  )%>%
  mutate(
    # <-- Ensure Yes before No everywhere downstream
    Found = fct_relevel(as.factor(Found), "Yes", "No")
  )

# 1) Stabilise categories (optional but recommended)
finds2 <- finds %>%
  filter(!is.na(Found)) %>%
  mutate(
    `Search type` = fct_lump_min(`Search type`, min = 50, other_level = "Other"),
    Found = fct_drop(Found)
  )

# 2) Build contingency table and run a safe test
tbl <- with(finds2, table(`Search type`, Found))
tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]

if (all(dim(tbl) == c(2, 2))) {
  test_obj <- fisher.test(tbl, workspace = 2e8)
  test_label <- "Fisher's exact"
} else {
  test_obj <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
  test_label <- "Chi-square (simulated p)"
}

pval_str <- format.pval(test_obj$p.value, digits = 4, eps = .Machine$double.eps)

# 3) Make the crosstable (no internal test) and append the p-value
ct <- crosstable(
  finds2, `Search type`,
  by = Found,
  total = TRUE,
  test = FALSE,                
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct) |>
  add_footer_lines(values = paste0("Overall association: ", test_label, ", p = ", pval_str))

ft

```

## Hit rates by person searched   

Hit rates vary across racial appearance, gender, and age. Compared with people perceived as **White** (**18.2%**), most other groups had lower hit rates—for example, **Middle Eastern** (**10.3%**),  **African** (**13.9%**) and Asian (**13.7%**). The only group not following this pattern is **Pacific Islander (19.2%)**, although the search success rate is higher only by one percentage point.

**Males** recorded higher hit rates than **females** (**18% vs 14%**).

By age, **teenagers (≤17)** generally had higher hit rates than adults.

**Table 6 Hit rates by person characteristics**

```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}

# Recode and reorder factors
dat2 <- dat %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),  # Yes before No
    Year  = fct_relevel(as.character(Year), "2018", "2019", "2022", "2023", "2024"),
    `Age group` = fct_relevel(`Age group`,
                              "Under 12", "12 to 17", "18 to 25",
                              "26 to 35", "36 to 45", "46 and over") |> fct_drop(),
    # Use freq order for the others ONLY
    across(
      .cols = all_of(c("Racial.appearance.transformed", "Gender", "Age group")),
      .fns  = ~ fct_infreq(.) |> fct_drop()
    )
  )

# Variables to tabulate (exclude 'Found' here; it's the 'by' variable)
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Contact.Type")

# Build crosstable WITHOUT internal tests (prevents Fisher blow-ups)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2, else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```

## Hit rates by police member and operating context  

**By member type:** Hit rates were higher for **Protective Services Officers (PSOs)** than for other members. This likely reflects deployment: PSOs concentrate at train stations and nearby precincts—often in the evening—where offending (and therefore the base rate of finds) is higher. Across **ranks**, hit rates were otherwise broadly similar.

**By unit:** **Highway Patrol**, **PSO units**, and **Transit** recorded hit rates in the **25–35%** range, again consistent with searches concentrated around travel corridors and transport hubs. **Uniform police**—who conduct the largest share of searches—had a hit rate of **~17%**. 

Find rates for searches conducted by the **Family Violence Investigation** unit were also high (one in four), however these were largely searches for firearms of an individual under an FPO (Firearms Prohibition Order). 

**By area:** Overall hit rates were similar in **metropolitan vs regional** areas. Within Melbourne, the **North West Metro Region** (covering Melbourne’s socio-economically less advantaged and more culturally diverse northern and western suburbs) had a **significantly lower** hit rate than other regions. That is, individuals living in these locations are searched more often (additional analysis revealed high search rate per capita), and with less success.

**Over time:** Hit rates varied only **modestly** across 2018, 2019, 2022, 2023 and 2024.

**Table 6 Hit rates by police member and operating context**
```{r Finds pol, echo=FALSE, warning=FALSE, results='asis'}


# Recode and reorder factors
dat2 <- dat %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),  # Yes before No
    Year  = fct_relevel(as.character(Year), "2018", "2019", "2022", "2023", "2024"),
    # Frequency order for operating-context variables
    across(
      .cols = all_of(c("Unit.type", "Rank.of.Member", "Area.type", "Region")),
      .fns  = ~ as_factor(.) |> fct_infreq() |> fct_drop()
    )
  )

# Variables to tabulate (exclude 'Found' here; it's the 'by' variable)
vars_for_table <- c("Unit.type", "Rank.of.Member", "Area.type", "Region", "Year")

# Build crosstable WITHOUT internal tests (prevents Fisher blow-ups)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2, else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft


```


# Conclusion


The descriptive results highlight both the promise and the limits of raw hit-rate comparisons as a method for identifying racial profiling. Search type, unit type and geography all influence the find rate, and must be controlled for in order to determine whether differences in find rates across racial groupings are driven by racism, or some other operational or socio-economic factor. 

A lower find rate in neighbourhoods with greater numbers of migrants of any given appearance may point to racial inequality of society at large, but it does not necessarily point to racism in the exercise of the police discretion to search. It is a well established phenomenon that police tend to conduct more searches in locations with higher crime rates, with less success. 

The high rate of missing racial appearance fields in search entries for many units also supports a restriction of analysis to those units where racial completeness is nearer to 80%.

The preferred approach is to restrict analysis of racial profiling to pedestrian searches conducted by uniform officers, looking at drug and weapon searches in separate analyses.

---

[^data]: Dataset size, years, and warrantless scope; “any item found” definition for hit rate. :contentReference[oaicite:0]{index=0} :contentReference[oaicite:1]{index=1} :contentReference[oaicite:2]{index=2}

[^types]: Hit rates by search type (graffiti 50.8%; firearms (FPO) 20.5%; weapons 32.4% vs drugs 15.4%). :contentReference[oaicite:3]{index=3}

[^unit]: Higher PSO hit rates; unit band 25–35%; Uniform ~17%. :contentReference[oaicite:4]{index=4}

[^area]: Metro vs regional similar overall; **NWMR** significantly lower; NWMR includes diverse/marginalised LGAs (Brimbank, Melbourne). :contentReference[oaicite:5]{index=5}

[^person]: Racial, gender, and age hit-rate differences (White 18%, Middle Eastern 10%, African 13.9%; males 18% vs females 14%; ≤17 higher). :contentReference[oaicite:6]{index=6}

[^missing]: Missing racial appearance (**42%**) and **2.5% ‘Other’**. :contentReference[oaicite:7]{index=7}


# Appendix – Ethnic Appearance Mapping

```{r mapping-categories-2022-2023, echo = FALSE}
dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```

