---
title: "Racial profiling - drug searches on foot by uniform police in Melbourne"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---
# Abstract {-}

This study examines warrantless drug searches as a mechanism through which racial profiling may operate in Victoria Police pedestrian policing. It analyses 7,534 warrantless drug searches conducted by uniformed, station-based officers during pedestrian encounters in metropolitan Melbourne (2018, 2019, 2022, 2023). The analysis focuses on people perceived by police as White or African, who together account for more than half of all recorded searches.

The purpose of the analysis is to assess whether search outcomes are consistent with racial profiling, using a hit-rate framework commonly applied in empirical studies of police stop-and-search practices. Under race-neutral policing, searches based on objective indicators of offending should yield similar evidentiary hit rates across racial groups. Systematic differences in hit rates are therefore consistent with differential thresholds for suspicion, a core feature of racial profiling.

First, descriptive cross-tabulations show that across metropolitan Melbourne, hit rates (contraband found) are lower for African-appearing individuals (11.4%) than for White-appearing individuals (14.1%), with a statistically significant association between perceived race and search outcomes.

Second, a city-wide fixed-effects logistic regression (Model A1) compares search outcomes within the same station and year, holding constant patrol area and time period. Restricting the sample to stations with at least five searches of African-appearing individuals, and controlling for officer gender and rank, African-appearing individuals have approximately 30 per cent lower odds of a hit than White-appearing individuals (odds ratio = 0.72, 95% confidence interval 0.54–0.95, p = 0.022). This disparity remains substantively and statistically similar under a stricter restriction requiring at least ten African-appearing searches per station (Model A2), indicating that the result is not driven by low-volume stations.

Third, a multilevel logistic regression (Model B) with random intercepts for stations models the full metropolitan sample (7,534 searches across 84 stations), allowing baseline hit rates to vary across stations. This model produces a near-identical estimate of racial disparity (odds ratio = 0.73, 95% confidence interval 0.57–0.93, p = 0.011), while again showing no clear association between search outcomes and officer gender or rank.

Finally, division-specific fixed-effects models and station-level cross-tabulations examine the spatial distribution of these disparities. The most pronounced and statistically robust disparity appears in North West Metro Region Division 3 (Brimbank–Melton), with suggestive but non-significant patterns in several other north-western divisions, and little evidence of disparity in most southern and eastern divisions. Although station-level estimates are often under-powered, they point to a small number of patrol teams where searches of African-appearing individuals are consistently less likely to yield contraband.

Taken together, these findings are consistent with racial profiling in the use of warrantless search powers, particularly as exercised by specific station-level patrol teams rather than uniformly across the organisation. While hit-rate analysis cannot establish individual officer intent or motive, the persistence and spatial concentration of lower evidentiary yields for African-appearing individuals indicate systematic differences in the threshold for initiating searches. These results underscore the need for ongoing station-level monitoring, transparency, and accountability in Victoria Police stop-and-search practices.

# Introduction

This analysis examines **7,534 warrantless drug searches** conducted by **uniformed officers during pedestrian encounters** in **metropolitan Melbourne** during **2018, 2019, 2022, and 2023**, involving people perceived as having *White* or *African* appearance by Victoria Police.  

These encounters represent the most routine and discretionary form of public-place policing, where outcomes depend heavily on individual officer judgement rather than intelligence-led or warrant-based operations.  

## Purpose and scope

The aim is to investigate whether search outcomes are consistent with **racial profiling** — that is, whether data indicates that people African are subjected to searches on a **lower evidentiary threshold** than people perceived as White.  

This focus reflects both **data quality and historical context**. The **Haile-Michael case** and **recurring allegations of racial profiling of African-Australians** in Victoria have drawn sustained public and policy attention to policing practices affecting this group. Concentrating on this comparison allows a direct test of whether such disparities are observable in routine operational data.  

The two racial appearance categories analysed — *African* and *White* — represent two of the largest and most consistently recorded groupings in the dataset, enabling robust and interpretable comparison of outcomes across time and space.  

Restricting the analysis to **uniform pedestrian encounters** and **drug-related searches** further reduces confounding. These encounters occur in public settings under a single legislative authority — *section 82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)* — which permits searches without a warrant when an officer has *reasonable grounds to suspect* drug possession.  

By excluding **vehicle stops**, **warrant searches**, and **specialist or Transit Police operations**, the analysis isolates the most discretionary form of search activity. In particular, removing **Transit Police** and **Protective Services Officers (PSOs)** excludes searches around train stations and transport hubs — areas with distinct patterns of offending and demographic composition — ensuring the analysis reflects **community-based policing contexts** rather than **transport-enforcement settings**.  

## VicPol organisational structure

A key feature of this dataset is that each search is linked to the officer’s home police station, allowing station-level controls to capture local operational context.

The data include 84 uniform police stations. Each station has its own defined patrol zone, which sits within a Police Service Area (PSA). PSAs are nested within one or more Local Government Areas (LGAs), which are in turn nested within Victoria Police Divisions.

At the operational level, each uniform station is responsible for its patrol zone as the core area for first response and routine patrol. Patrol zones are set by Victoria Police regional commands to balance population coverage, workload, and response times, and represent the smallest consistent unit of geographic responsibility for general-duties policing. Specialist units, such as Transit Police or Highway Patrol, operate across broader areas and are not tied to these station-level zones.


## Legal context: section 82 powers

Under **section 82** of the *Drugs, Poisons and Controlled Substances Act 1981 (Vic)*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.  

This power represents the majority of search activity in the dataset and embodies the high-discretion environment most relevant to testing for racial bias: officers must decide, often rapidly, who appears “suspicious enough” to search.  

## The hit-rate test framework

The analysis applies the outcomes-based or “hit-rate” test introduced by **Knowles, Persico, and Todd (2001)** and formalised by **Persico and Todd (2004)**.  

In their economic model, police officers are assumed to allocate **pedestrian searches** across groups to maximise the probability of discovering contraband. If officers act without bias, equilibrium requires that the hit rates — the proportion of searches yielding contraband — are equal across racial groups. Conversely, if officers apply a lower suspicion threshold to one group (searching them more frequently or with less cause), that group will exhibit a lower hit rate, as more searches are conducted on weaker grounds.  

### Limitations and critiques

The hit-rate framework has attracted criticism in the policing and econometric literature, particularly around its sensitivity to contextual differences across encounters. Analysts including Ridgeway and MacDonald (2010) note that hit-rate disparities can be misattributed if situational factors vary systematically across racial groups.

However, these limitations primarily arise when analyses fail to account for local operational context. When station-level variation, officer assignment patterns, and temporal effects are ignored, racial disparities may be overstated or obscured. By contrast, when hit-rate analysis is embedded within fixed-effects models and internal benchmarking frameworks that compare officers or stations operating in similar environments, it provides a meaningful and policy-relevant lens on discretionary policing.

In this setting, the hit-rate framework is a valuable tool for identifying patterns consistent with racial profiling, particularly where lower evidentiary yields persist for specific racial groups within the same patrol areas and time periods. While it does not establish individual intent, it is well suited to detecting systematic differences in how suspicion thresholds are applied, and to flagging stations or teams where further scrutiny, monitoring, or intervention may be warranted.


```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(scales)
library(stringr)
library(tidyr)
library(fixest)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )


is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}


make_hit_table <- function(
  data,
  group_var,                   # e.g. "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    dplyr::filter(!is.na(!!grp), !is.na(Found)) %>%
    dplyr::filter(Racial.appearance.transformed %in% c("African","White")) %>%
    dplyr::mutate(FoundYes = is_yes(Found)) %>%
    dplyr::count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    dplyr::group_by(!!grp, Racial.appearance.transformed) %>%
    dplyr::summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    dplyr::filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble::tibble())

  out <- tab %>%
    dplyr::mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    dplyr::select(-test) %>%
    tidyr::unnest(tidied) %>%
    dplyr::mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = dplyr::if_else(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    # Important: use !!rlang::sym(group_var) on LHS, and avoid Unicode minus in names
    dplyr::transmute(
      !!rlang::sym(group_var) := !!grp,
      `African (succ/total)`   = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`     = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`       = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`         = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr - Wh)`        = scales::percent(diff_est,         accuracy = 0.1),  # <-- ASCII hyphen
      `Risk ratio (Afr/Wh)`    = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                      = sprintf("%.3f", p.value),
      `p (BH)`                 = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                    = stars
    )

  if (filter_sig_only) {
    out <- out %>% dplyr::filter(as.numeric(stringr::str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% dplyr::arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))
}

print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) display <- display %>% dplyr::select(-`p (BH)`)

  display %>%
    knitr::kable(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    kableExtra::column_spec(1, width = "24em") %>%
    kableExtra::add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}

```

# Dataset

## Dataset scope and coverage

The dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in **2018, 2019, 2022, and 2023**.  

Data for **2020 and 2021** were excluded due to the atypical policing environment during the COVID-19 pandemic, when search activity and mobility patterns were highly irregular.

The data were released through a **Freedom of Information (FOI)** request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and cover **public-place searches** carried out under a range of legal powers.

For this analysis, we restrict to:

- **Geography:** Metropolitan Melbourne  
- **Contact type:** Pedestrian contact only  
- **Unit type:** Uniform police only  
- **Search type:** Drug searches conducted under *s82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)*  
- **Station linkage:** Station-based officers with a station patrol area excluding train stations
- **Racial appearance:** Restricted to *White* and *African* appearance  


The original FOI dataset was **enhanced with additional contextual data** to improve the precision and interpretability of geographic comparisons.  

Each search record was linked to:

- The **Victoria Police organisational hierarchy**, allowing each record to be associated with its corresponding **Region, Division, Police Service Area (PSA), and Station**.  

- **Crime Statistics Agency (CSA) data** on the **drug offence rate** and **total offence rate** in the **Local Government Area (LGA)** in which each police station is situated, for the corresponding **year of the search**.  

## Data limitations

### Racial appearance classification

The original dataset for drug searches conducted by uniform officers on foot patrol included 13,896 records.

Approximately **20% of these records lacked racial appearance information**, and several smaller categories had minimal counts or inconsistent coding across years.

For analytical reliability and interpretability, disparity testing focuses on **White** and **African** appearance categories. Although the African group is not the second-largest category by count, these two groupings together represent over half (54%) of all relevant records and are consistently coded across years. The original coded categories, mapped to the cleaned racial categories utilised in this analysis, appears in the Appendix.

**A note on 'Missing' and 'White-appearing'**

Descriptively, the hit rate among cases with missing racial appearance (14.5%) was very close to that of White-appearing individuals (14.1%), suggesting that many blank or unrecorded entries are likely to be White-appearing rather than missing at random. 

### Age information

Age data were missing for **42.9%** of all pedestrian searches in the restricted subset.  Among cases with recorded ages, the distribution was heavily concentrated among younger people — **individuals aged 18 to 45 years accounted for more than half of all identified persons** — with relatively few searches involving older adults (see Table 2 below).  

Because the age profile of those searched is already strongly youthful, the omission of age as a covariate is **unlikely to distort results substantially**, provided that missingness is not systematically related to racial appearance. 


```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddatr <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45","46 to 55", 
                                                         "56 to 65","66 and over")))
 
```

# Hit Rates — Descriptive Cross-Tabulations

Before estimating formal models, it is important to examine the *raw relationship* between racial appearance and search outcomes, and how contextual variables may shape these differences.  

The following tables report:  

- Hit rates by racial appearance, gender, year, rank, and police division.  
- Statistical tests (χ² or Fisher’s exact) assessing whether differences are significant.  

Subsequent sections will model these same relationships using fixed-effects and multilevel logistic regression frameworks, controlling for the confounding influences identified here.

**Table 1 Hit rates by person characteristics**
```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}

# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))%>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop(),
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Rank  = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    )
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Year", "Rank", "Division")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(
    p_fmt = ifelse(p < 0.001, "<0.001", sprintf("%.5f", p)),
    p_str = sprintf("%s — %s, p = %s", variable, test, p_fmt)
  )

# Append p-values to the table footer (each on its own line)
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft


```


### Overview of results

The cross-tabulations reveal **modest but potentially meaningful variation in hit rates across racial appearance**.  The **Fisher’s exact test for racial appearance** (p = 0.039) indicates a statistically significant association between race and search outcomes. Searches of **African-appearing individuals** yielded a lower evidentiary return (11.4%) than those of **White-appearing individuals** (14.1%). 

By contrast, **gender (p = 0.33)** and **age group (p = 0.64)** show no significant differences in hit rates. Given that most searches involve young adults, this reduces concern that age composition alone explains the observed racial disparity. We do however recognise that missingness in the data is a limitation - we cannot definitively rule out a relevent systemic relationship.


**Year effects** (p ≈ 0.17) and police officer **Rank effects** are minor, suggesting overall stability in hit rates across the observed period, and some increase in search efficacy relating to experience.  

Finally, the **strong association by police division (p < 0.001)** highlights the importance of community context: hit rates vary considerably between divisions, indicating that **local operational culture and context** may play a key role in shaping search outcomes.


# Statistical modelling  

Model A provides the primary test of within-station racial disparity under comparable operational conditions, while Model B assesses whether these disparities persist across the metropolitan policing system and vary between stations.

## Model A — Fixed-Effects Logistic Regression (Station × Year)


While descriptive cross-tabulations highlight consistent racial differences in hit rates, they cannot determine whether these disparities reflect **unequal treatment** or simply differences in **where and when** searches occur.  

To separate these effects, we estimate a **fixed-effects logistic regression** that isolates racial disparities *within the same police station and year*, holding constant local operational context and temporal variation.  This approach allows us to test whether, under comparable conditions, discretionary searches of African-appearing individuals were systematically less likely than White-appearing individuals to yield contraband.

### Model specification

- **Dependent variable:** whether a search resulted in contraband being found (`Found_yes = 1`).  
- **Key predictor:** perceived racial appearance (*African* vs *White*).  
- **Covariates:** gender and officer rank.  
- **Fixed effects:** Station × Year, absorbing all between-station and between-year variation.  
- **Standard errors:** clustered by station to account for repeated observations within sites.  

Formally, for search \(i\) conducted at station \(s\) in year \(t\), the model is:

\[
Y_{ist} \sim \text{Bernoulli}(p_{ist}),
\]

\[
\log\left(\frac{p_{ist}}{1 - p_{ist}} \right)
= \beta_1 \,\text{African}_i
+ \boldsymbol{\gamma}^\top \mathbf{G}_i
+ \boldsymbol{\delta}^\top \mathbf{R}_i
+ \alpha_{st},
\]

where \(Y_{ist}=1\) if contraband was found; \(\text{African}_i\) is an indicator for perceived racial appearance “African” (with “White” as the reference); \(\mathbf{G}_i\) and \(\mathbf{R}_i\) are vectors of officer gender and officer rank indicators; and \(\alpha_{st}\) is a Station × Year fixed effect. Standard errors are clustered at the station level, allowing arbitrary correlation among searches within the same station.

### Data restriction and rationale  

To support within-station racial comparisons, the analysis is restricted to stations with more than five pedestrian searches of African-appearing individuals. This yields 5,535 of the 7,534 searches and includes 40 of the 84 uniform stations in metropolitan Melbourne.

This restriction is applied to avoid mechanically unstable estimates arising from stations with very few African-appearing searches, where hit rates and odds ratios can be driven by one or two observations rather than by systematic policing patterns. The threshold is therefore intended as a minimum inclusion rule to ensure that within-station comparisons are meaningful, rather than to imply precise station-level estimation.

Although this restriction reduces the number of stations included, it improves the reliability of the fixed-effects estimates by limiting the influence of sparse cells and quasi-separation. Substantively, the estimand remains a pooled within-station effect, with identification driven by comparisons between racial groups operating in the same station and year. Results are robust to reasonable alternative minimum-count thresholds.

**Table 2 Model A — Fixed-effects logistic regression (Station×Year)**
```{r Model A1, echo=FALSE,message=FALSE, warning=FALSE}

# ===============================================================
# Model A — Fixed-effects logistic regression (Station×Year)
# Restrict to stations with ≥5 African-appearance searches (aggregated across all years)
# Includes Gender and Rank covariates; clustered SEs by Station
# ===============================================================


# --- 1. Define eligible stations (≥5 African across all years) ---
elig_stations <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  count(Station.uniform, Racial.appearance.transformed, name = "n") %>%
  tidyr::pivot_wider(
    names_from  = Racial.appearance.transformed,
    values_from = n,
    values_fill = 0
  ) %>%
  filter(African >= 5) %>%  # stations with at least 10 African searches overall
  pull(Station.uniform)

# --- 2. Prepare dataset restricted to those stations ---
dat1 <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Station.uniform %in% elig_stations
  ) %>%
  mutate(
    Race        = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes   = is_yes(Found),
    StationYear = interaction(Station.uniform, Year, drop = TRUE),
    Gender      = factor(Gender),

    # Clean and collapse ranks
    Rank_clean = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    ),
    Rank_clean = forcats::fct_drop(Rank_clean),
    Rank_clean = forcats::fct_relevel(Rank_clean, "CONST")
  )

# --- 3. Fit Station×Year fixed-effects logistic model ---
m_fe_af5_all <- fixest::feglm(
  Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") | StationYear,
  family  = "binomial",
  data    = dat1,
  cluster = ~ Station.uniform
)

# --- 4. Tidy results (exclude FE dummies) ---
model_table <- broom::tidy(m_fe_af5_all, conf.int = TRUE) %>%
  dplyr::filter(!grepl("StationYear", term)) %>%
  dplyr::mutate(
    OR        = exp(estimate),
    CI_low    = exp(conf.low),
    CI_high   = exp(conf.high),
    p_str     = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  dplyr::transmute(
    Term         = term,
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = p_str
  )

# --- 5. Display results ---
kbl(
  model_table,
  caption = paste0(
    "Fixed-effects logistic regression (Station×Year) — restricted to stations with ≥5 African searches (clustered by Station.uniform; N = ",
    nrow(dat1),
    " searches)."
  ),
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")


```

As a robustness check, the fixed-effects model was re-estimated using a stricter inclusion threshold of at least ten African-appearing searches per station. This restriction reduces the sample to 4,366 searches across 27 stations but substantially increases within-station support for racial comparisons.

Under this specification, the estimated racial disparity remains substantively and statistically similar to the main model. African-appearing individuals have approximately 30 per cent lower odds of a hit than White-appearing individuals (odds ratio = 0.70, 95% confidence interval 0.51–0.96, p = 0.028). Estimates for officer gender and rank remain small and statistically indistinguishable from zero.

The persistence of the racial disparity under a stricter minimum-count threshold indicates that the main findings are not driven by sparse cells or low-volume stations, and strengthens the interpretation of the results as reflecting systematic differences in search outcomes within comparable policing contexts.

**Table 3 Model A2 — Fixed-effects logistic regression (Station×Year)**
```{r Model A2, echo=FALSE,message=FALSE, warning=FALSE}

# --- 1. Define eligible stations (≥5 African across all years) ---
elig_stations <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  count(Station.uniform, Racial.appearance.transformed, name = "n") %>%
  tidyr::pivot_wider(
    names_from  = Racial.appearance.transformed,
    values_from = n,
    values_fill = 0
  ) %>%
  filter(African >= 10) %>%  # stations with at least 10 African searches overall
  pull(Station.uniform)

# --- 2. Prepare dataset restricted to those stations ---
dat1 <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Station.uniform %in% elig_stations
  ) %>%
  mutate(
    Race        = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes   = is_yes(Found),
    StationYear = interaction(Station.uniform, Year, drop = TRUE),
    Gender      = factor(Gender),

    # Clean and collapse ranks
    Rank_clean = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    ),
    Rank_clean = forcats::fct_drop(Rank_clean),
    Rank_clean = forcats::fct_relevel(Rank_clean, "CONST")
  )

# --- 3. Fit Station×Year fixed-effects logistic model ---
m_fe_af5_all <- fixest::feglm(
  Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") | StationYear,
  family  = "binomial",
  data    = dat1,
  cluster = ~ Station.uniform
)

# --- 4. Tidy results (exclude FE dummies) ---
model_table <- broom::tidy(m_fe_af5_all, conf.int = TRUE) %>%
  dplyr::filter(!grepl("StationYear", term)) %>%
  dplyr::mutate(
    OR        = exp(estimate),
    CI_low    = exp(conf.low),
    CI_high   = exp(conf.high),
    p_str     = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  dplyr::transmute(
    Term         = term,
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = p_str
  )

# --- 5. Display results ---
kbl(
  model_table,
  caption = paste0(
    "Fixed-effects logistic regression (Station×Year) — restricted to stations with ≥10 African searches (clustered by Station.uniform; N = ",
    nrow(dat1),
    " searches)."
  ),
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")


```

### Interpretation  


Within the same station and year, searches of African-appearing individuals were significantly less likely to result in contraband being found than searches of White-appearing individuals (odds ratio = 0.72, 95% confidence interval 0.54–0.95, p = 0.022). This corresponds to approximately 30 per cent lower odds of a “hit” for African-appearing individuals under comparable operational contexts.

This disparity remains substantively and statistically similar under a stricter restriction requiring at least ten African-appearing searches per station (odds ratio = 0.70, 95% confidence interval 0.51–0.96, p = 0.028), indicating that the result is not driven by low-volume stations or sparse cells.

By contrast, officer gender and rank were not statistically significant predictors of search success, and their inclusion did not materially affect the estimated race coefficient. Searches conducted by Constables, Senior Constables, and Sergeants yielded broadly similar outcomes once station- and year-level variation were held constant.


## Model B — Multilevel Logistic Regression (Random Intercepts for Station)

To complement the fixed-effects analysis, we estimated a **multilevel logistic regression** with **random intercepts for police stations**.  

Whereas Models A1 and A2 isolates within-station differences by removing all station-level variation, Model B explicitly models that variation, treating each station’s baseline likelihood of a “hit” as a random effect.

This approach recognises that stations differ systematically in operational practices, local contexts, and levels of policing intensity.  

Rather than controlling away this variation, the multilevel model captures it directly, enabling more **city-wide generalisation** while still accounting for the clustering of searches within stations.  

The dependent variable again indicates whether contraband was found (*Found_yes = 1*), with predictors for perceived race (*African* vs *White*), gender, and local drug-crime rate.

No data restrictions were applied.

### Model specification  

Model B was estimated using `lme4::glmer` with a logit link:

\[
\text{logit}\big(P(\text{Found}_{i})\big)
  = \beta_0 
  + \beta_1(\text{Race}_{i})
  + \beta_2(\text{Gender}_{i})
  + \beta_3(\text{DrugCrimeRate}_{\text{LGA,Year}})
  + u_{\text{Station}},
\]

where \(u_{\text{Station}} \sim N(0, \sigma^2_{\text{Station}})\) captures station-level random intercepts.

The dependent variable is whether a search resulted in contraband being found (`Found_yes = 1`). Drug-crime rates are included as a **contextual covariate measured at the LGA–year level**, assigned to each search based on the LGA in which the station is located.

Standard errors are clustered by **station** to account for repeated observations within sites.

**Table 4 Model B — Multilevel logistic regression (random intercepts for Station)**
```{r modelB multilevel logistic, echo = FALSE, warning = FALSE, message = FALSE}
# ===============================================================
# Model B — Multilevel logistic regression (random intercepts for Station)
# Adds Rank as a fixed effect (CONST as reference; SSGT→SGT, PSO→CONST)
# ===============================================================


dat2 <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  mutate(
    Found_yes = is_yes(Found),
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Gender    = factor(Gender),
    Station   = factor(Station.uniform),

    # Clean Rank: collapse small/rare levels; set CONST as ref
    Rank_clean = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    ),
    Rank_clean = fct_drop(Rank_clean),
    Rank_clean = fct_relevel(Rank_clean, "CONST"),

    # Ensure crime rate is numeric (if present)
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate))
  )

# Optional stability filter: exclude very small stations
#station_counts <- dat2 %>% count(Station)
#dat2 <- dat2 %>% filter(Station %in% station_counts$Station[station_counts$n >= 20])

# Store N for table captions
N_obs <- nrow(dat2)
N_stations <- length(unique(dat2$Station))

# Fit multilevel logistic model: random intercepts by Station
m_station <- glmer(
  Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") + scale(Drug.crime.rate) + (1 | Station),
  family  = binomial,
  data    = dat2,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# ---- Table A: Race effect (OR with 95% CI) ----
b_results <- broom.mixed::tidy(m_station, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
  mutate(
    OR      = exp(estimate),
    CI_low  = exp(conf.low),
    CI_high = exp(conf.high)
  ) %>%
  filter(term == "RaceAfrican") %>%
  transmute(
    Term         = "African vs White",
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = signif(p.value, 3)
  )

kbl(
  b_results,
  caption = paste0(
    "Multilevel logistic regression (random intercepts for Station) — odds of contraband found (Race effect). ",
    "N = ", format(N_obs, big.mark = ","), 
    " searches across ", N_stations, " stations."
  ),
  align = "lrrrr"
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")

# ---- Table B (optional): all fixed effects nicely formatted ----
coef_table <- broom.mixed::tidy(m_station, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
  mutate(
    OR      = exp(estimate),
    CI_low  = exp(conf.low),
    CI_high = exp(conf.high),
    Term    = case_when(
      term == "RaceAfrican"                ~ "Race: African vs White",
      term == "GenderM"                    ~ "Gender: Male vs Female",
      grepl("^i\\(Rank_clean.*SCONST", term) ~ "Rank: Senior Constable vs Constable",
      grepl("^i\\(Rank_clean.*SGT", term)    ~ "Rank: Sergeant vs Constable",
      term == "scale(Drug.crime.rate)"     ~ "Drug crime rate (z)",
      TRUE                                 ~ term
    )
  ) %>%
  transmute(
    Term,
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  )

kbl(
  coef_table,
  caption = paste0(
    "Multilevel logistic regression (random intercepts for Station) — all fixed effects. ",
    "N = ", format(N_obs, big.mark = ","), 
    " searches across ", N_stations, " stations."
  ),
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")
```

### Interpretation

The multilevel model, estimated on **7,534 searches across 84 police stations**, shows that even after accounting for officer rank, gender, and variation between stations, **African-appearing individuals remained significantly less likely to yield contraband** than White-appearing individuals searched under similar conditions.

The odds of a “hit” were approximately **27% lower** for African-appearing individuals (*OR* = 0.73, 95% CI 0.57–0.93, *p* = 0.011).  This disparity persists after adjusting for station-level context and individual-level covariates.

Gender and rank effects were not statistically significant at the 5 percent level, though searches by **Sergeants** trended toward a higher hit rate (*OR* = 1.35, *p* = 0.087) and **men** were marginally more likely to be found with contraband (*OR* = 1.16, *p* = 0.083).  

Local drug-crime rates showed no meaningful association with search success - this may be due to the coarseness of the LGA level unit of measurement, or the fact that the station level variation is already capturing this effect.

Together, these findings indicate a **robust racial disparity in evidentiary yield** that is not explained by officer characteristics or local policing environments.


### Geographic concentration of disparities

Taken together, the two modelling approaches indicate that racial disparities in search outcomes are persistent but unevenly distributed across metropolitan Melbourne.

The multilevel model (Model B) shows that, on average across all stations, searches of individuals perceived as African are significantly less likely to yield contraband than those of White individuals, even after adjusting for officer rank, gender, local drug-crime rates, and between-station differences in baseline hit rates.

The fixed-effects model (Model A), which compares searches conducted within the same station and year, demonstrates that this disparity also operates within tightly comparable operational contexts. The stability of the race effect under increasingly strict station-level restrictions indicates that the observed disparity is not driven by sparse data or compositional differences across stations.

At the same time, the presence of significant station-level variance in the multilevel model (σ²_station > 0) indicates substantial heterogeneity in baseline hit rates across stations. This implies that while the racial disparity is evident on average, its magnitude and practical impact vary across local policing environments.

Taken together, these results suggest that racial disparities in evidentiary yield reflect a combination of system-wide patterns and locally concentrated practices. Rather than being evenly distributed across all stations, the disparity appears to be shaped by specific operational settings, pointing to the role of local patrol cultures, leadership practices, and forms of oversight in how discretionary search powers are exercised.


# Division and station disparities

Division-level cross-tabulations and modelling are used to examine how racial disparities in search outcomes vary across local policing environments, providing insight into the spatial concentration of patterns consistent with racial profiling.


## Division - Cross tabs

Differences in hit rates between African- and White-appearing individuals were assessed using two-sample tests for equality of proportions (chi-square tests without continuity correction), conducted separately for each division.

**Table 5 Division cross-tabs**
```{r division cross tabs, echo = FALSE, warning=FALSE, message=FALSE}

# ===============================================================

# Division-level hit-rate disparity (African vs White)

# - Uses existing `ddatr`, `is_yes()`, loaded packages

# - Keeps numeric p for sorting; formats labels cleanly

# ===============================================================

# ---- parameters (can reuse; shown here for clarity) ----

MIN_AFRICAN <- 0
MIN_WHITE   <- 0

# ---- build analysis frame ----

dat_div_long <- ddatr %>%
dplyr::filter(
!is.na(Found),
!is.na(Division),
Racial.appearance.transformed %in% c("White", "African")
) %>%
dplyr::transmute(
Division  = factor(Division),
Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
Found_yes = is_yes(Found)
)

# ---- summarise counts by Division × Race ----

dat_div_wide <- dat_div_long %>%
dplyr::count(Division, Race, name = "n") %>%
tidyr::pivot_wider(names_from = Race, values_from = n, values_fill = 0, names_prefix = "n_") %>%
dplyr::left_join(
dat_div_long %>%
dplyr::group_by(Division, Race) %>%
dplyr::summarise(found_yes = sum(Found_yes, na.rm = TRUE), .groups = "drop") %>%
tidyr::pivot_wider(names_from = Race, values_from = found_yes, values_fill = 0, names_prefix = "fy_"),
by = "Division"
) %>%
dplyr::filter(n_African >= MIN_AFRICAN, n_White >= MIN_WHITE)

# ---- bail out if nothing passes thresholds ----

if (nrow(dat_div_wide) == 0L) {
cat("No divisions met the thresholds (≥", MIN_AFRICAN, " African and ≥", MIN_WHITE, " White searches).\n", sep = "")
} else {

# ---- compute proportions + prop.test per division ----

div_tests <- dat_div_wide %>%
dplyr::mutate(
prop_African = fy_African / n_African,
prop_White   = fy_White   / n_White,
test = purrr::pmap(
list(fy_African, fy_White, n_African, n_White),
~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
)
) %>%
dplyr::mutate(tidy = purrr::map(test, broom::tidy)) %>%
tidyr::unnest(tidy) %>%
dplyr::mutate(
risk_ratio = ifelse(prop_White > 0, prop_African / prop_White, NA_real_),
p_val_num  = p.value,
p_val_lab  = ifelse(p_val_num < 0.001, "<0.001", sprintf("%.3f", p_val_num)),
sig        = dplyr::case_when(
p_val_num < 0.001 ~ "***",
p_val_num < 0.010 ~ "**",
p_val_num < 0.050 ~ "*",
TRUE              ~ ""
)
) %>%
dplyr::arrange(p_val_num) %>%
dplyr::transmute(
Division,
`African n`  = n_African,
`White n`    = n_White,
`African %`  = scales::percent(prop_African, accuracy = 0.1),
`White %`    = scales::percent(prop_White,   accuracy = 0.1),
`Risk ratio` = round(risk_ratio, 2),
`95% CI (L)` = round(conf.low, 2),
`95% CI (U)` = round(conf.high, 2),
`p-value`    = p_val_lab,
sig
)

# ---- pretty table ----

kbl(
div_tests,
caption = paste0(
"Hit-rate disparity by Division (African vs White; only divisions with ≥",
MIN_AFRICAN, " African & ≥", MIN_WHITE, " White searches)"
),
align = "lrrrrrrrrr"
) |>
kable_paper(full_width = FALSE) |>
kable_styling(
bootstrap_options = c("striped", "hover"),
font_size = 12,
position = "left"
) |>
add_header_above(c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2))
}

```

## Divisions - Fixed-effects logistic regressions (Station × Year)  

To assess whether the within-station racial disparity identified in Model A is evenly distributed across metropolitan Melbourne or concentrated within particular operational contexts, we re-estimate the station–year fixed-effects model separately for each police division.

This approach preserves the core identification strategy—comparing searches conducted within the same station and year—while allowing the magnitude and direction of racial disparities to vary across divisions. Division-specific estimates therefore provide insight into how patterns consistent with racial profiling differ across regional policing environments.

For each Division \(d\), For each division d, we estimate a station–year fixed-effects logistic regression of the form:

\[
Y_{istd} \sim \text{Bernoulli}(p_{istd}),
\]

\[
\log\left( \frac{p_{istd}}{1 - p_{istd}} \right)
= \beta_{0d}
+ \beta_{1d}\,\text{African}_{i}
+ \boldsymbol{\gamma}_d^{\top}\mathbf{G}_{i}
+ \boldsymbol{\delta}_d^{\top}\mathbf{R}_{i}
+ \alpha_{st},
\]

where:  
- \(Y_{istd} = 1\) if contraband was found on search \(i\) at station \(s\) in year \(t\) in Division \(d\), and \(0\) otherwise;  
- \(\text{African}_{i}\) is an indicator for perceived racial appearance “African” (reference: “White”);  
- \(\mathbf{G}_{i}\) and \(\mathbf{R}_{i}\) are dummy variables for officer gender and rank, controlling for differences in officer composition across stations;  
- \(\alpha_{st}\) is a Station×Year fixed effect, included to ensure comparisons are made **within the same station and year**, thereby removing confounding from location-specific or time-varying enforcement patterns;  
- \(\beta_{1d}\) is the Division-specific log-odds coefficient for African-appearing individuals, with \(\exp(\beta_{1d})\) interpreted as the corresponding odds ratio.

Standard errors are clustered at the station level to account for repeated searches conducted within the same policing environment.

**Table 6 Division models**
```{r division station year, echo=FALSE, message=FALSE, warning = FALSE}

# ===============================================================
# Division-split models
# Each model: Fixed-effects logistic regression (Station×Year)
# Subset sample by Division; cluster SEs by Station
# ===============================================================

# --- Parameters ---
MIN_AFRICAN <- 0   # minimum African searches per station (across all years)

# --- Auto-detect Division column ---
div_col <- names(ddat)[str_detect(names(ddat), regex("division", ignore_case = TRUE))]
if (length(div_col) == 0) stop("No Division column found in ddat.")
DIV_COL <- div_col[1]
DIV_SYM <- rlang::sym(DIV_COL)

# --- Eligible stations (≥ MIN_AFRICAN African total) ---
elig_stations <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  count(Station.uniform, Racial.appearance.transformed, name = "n") %>%
  tidyr::pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = n,
    values_fill = 0
  ) %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(Station.uniform)

# --- Prepare base dataset ---
dat <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Station.uniform %in% elig_stations
  ) %>%
  mutate(
    Race        = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes   = is_yes(Found),
    StationYear = interaction(Station.uniform, Year, drop = TRUE),
    Gender      = factor(Gender),
    Rank_clean  = recode(Rank.of.Member, "SSGT" = "SGT", "PSO" = "CONST", .default = Rank.of.Member),
    Rank_clean  = forcats::fct_relevel(factor(Rank_clean), "CONST")
  )

# --- Run model per Division ---
divisions <- sort(unique(dat[[DIV_COL]]))
results_list <- list()

for (div in divisions) {
  d_sub <- dat %>% filter(!!DIV_SYM == div)
  if (nrow(d_sub) < 50) next  # skip small samples
  
  # Fit FE model
  mod <- tryCatch(
    fixest::feglm(
      Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") | StationYear,
      family = "binomial",
      data = d_sub,
      cluster = ~ Station.uniform
    ),
    error = function(e) NULL
  )
  if (is.null(mod)) next
  
  # Tidy table
  tidy_tab <- broom::tidy(mod, conf.int = TRUE) %>%
    filter(!str_detect(term, "StationYear")) %>%
    mutate(
      OR     = exp(estimate),
      CI_low = exp(conf.low),
      CI_high= exp(conf.high),
      p_str  = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
    ) %>%
    transmute(
      Term = term,
      `Odds Ratio` = round(OR, 3),
      `95% CI (L)` = round(CI_low, 3),
      `95% CI (U)` = round(CI_high, 3),
      `p-value` = p_str
    )
  
  caption <- paste0(
    "Division: ", div,
    " — Fixed-effects logistic regression (Station×Year), clustered by Station.uniform (N = ",
    format(nrow(d_sub), big.mark=","), ")."
  )
  
  results_list[[div]] <- kbl(tidy_tab, caption = caption, align = c("l","r","r","r","r")) |>
    kable_paper(full_width = FALSE) |>
    kable_styling(bootstrap_options = c("striped","hover"), position = "left")
}

# --- Print results sequentially ---
for (div in names(results_list)) {
  cat("\n\n")
  print(results_list[[div]])
  cat("\n\n")
}
```

```{r div model summary, echo=FALSE, warning=FALSE, message=FALSE}


# Helper for p-value formatting (unchanged)
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

div_summary <- tribble(
  ~Division, ~OR,   ~CI_L,   ~CI_U,    ~p,     ~Interpretation,

  # Eastern
  "Eastern Div 1 (Boroondara – Monash – Whitehorse – Manningham)",
    0.612,   0.258,   1.448,   0.263,
    "No statistically reliable difference; point estimate suggests lower yield for African-appearing people, but with substantial uncertainty.",

  "Eastern Div 2 (Knox – Maroondah – Yarra Ranges)",
    0.970,   0.209,   4.506,   0.969,
    "No statistically reliable difference; estimates are highly imprecise due to small effective sample size.",

  # NWMR
  "NWMR Div 1 (Melbourne – Yarra)",
    1.127,   0.513,   2.477,   0.767,
    "Comparable hit rates for African- and White-appearing individuals.",

  "NWMR Div 2 (Wyndham – Hobsons Bay – Maribyrnong)",
    0.608,   0.313,   1.181,   0.142,
    "Lower odds of contraband finds for African-appearing people (≈ 40% lower), though not conventionally significant.",

  "NWMR Div 3 (Brimbank – Melton)",
    0.144,   0.033,   0.636,   0.011,
    "Substantially lower evidentiary yield for African-appearing individuals; disparity statistically significant.",

  "NWMR Div 4 (Hume – Moonee Valley – Merri-bek)",
    0.639,   0.377,   1.083,   0.096,
    "Moderately lower yield for African-appearing people; suggestive but not conventionally significant (p ≈ 0.10).",

  "NWMR Div 5 (Banyule – Darebin – Whittlesea – Nillumbik)",
    0.717,   0.503,   1.021,   0.065,
    "Suggestive lower yield for African-appearing people (≈ 30% lower odds), but not conventionally significant.",

  # SMR
  "SMR Div 1 (Port Phillip – Stonnington)",
    0.960,   0.554,   1.664,   0.884,
    "No difference detected; odds ratio close to 1 with wide confidence interval.",

  "SMR Div 2 (Glen Eira – Kingston)",
    1.084,   0.139,   8.433,   0.939,
    "No statistically reliable difference; estimates highly imprecise.",

  "SMR Div 3 (Dandenong – Casey – Cardinia)",
    0.838,   0.330,   2.130,   0.710,
    "No difference detected.",

  "SMR Div 4 (Frankston – Mornington)",
    1.190,   0.128,  11.047,   0.879,
    "No statistically reliable difference; estimates highly imprecise."
) %>%
  mutate(
    `OR (African vs White)` = round(OR, 2),
    `95% CI` = paste0("[", sprintf("%.2f", CI_L), ", ", sprintf("%.2f", CI_U), "]"),
    `p` = fmt_p(p)
  ) %>%
  select(
    Division,
    `OR (African vs White)`,
    `95% CI`,
    `p`,
    Interpretation
  )

kbl(
  div_summary,
  caption = "4.2.1 Summary of racial disparity patterns",
  align = c("l","r","l","r","l")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")


```
Taken together, the division-specific fixed-effects models indicate substantial spatial variation in both the magnitude and direction of racial disparities in search outcomes.

The clearest and most statistically robust disparity is observed in NWMR Division 3 (Brimbank–Melton), where searches of African-appearing individuals were markedly less likely to yield contraband even after controlling for officer rank, gender, and station–year context.

Several other divisions—most notably NWMR Division 2 (Wyndham–Maribyrnong–Hobsons Bay) and NWMR Division 5 (Darebin–Whittlesea–Banyule–Nillumbik)—exhibit similar directional patterns toward lower evidentiary yields for African-appearing individuals, although these estimates do not reach conventional levels of statistical significance.

By contrast, southern and eastern divisions show no consistent racial pattern, with point estimates clustered around unity and wide confidence intervals reflecting smaller numbers of African-appearing searches and limited within-division variation.

Overall, these results suggest that racial disparities in evidentiary yield are not evenly distributed across metropolitan Melbourne, but are concentrated within particular operational contexts, most prominently in parts of Melbourne’s north-western growth corridors.

The following section examines station-level results to identify the specific local areas contributing to these divisional patterns.

## Station-level - Cross tabs  

Station-level cross-tabulations provide a more granular, descriptive view of how racial disparities in search outcomes are distributed across local policing teams. While these comparisons are not intended to establish station-level causal effects, they serve as a useful screening tool for identifying stations where patterns consistent with racial profiling may warrant closer examination.

To limit mechanically unstable comparisons, the analysis is restricted to stations with at least ten searches of African-appearing individuals. Differences in hit rates between African- and White-appearing individuals were assessed using two-sample tests for equality of proportions (chi-square tests without continuity correction), conducted separately for each station. 

*Table 7 Station cross tabs* 
```{r station table, echo = FALSE, warning = FALSE, message=FALSE}
# ===============================================================

# Station-level hit rates by race

# • Whites searched, White finds, % White finds

# • Africans searched, African finds, % African finds

# • Total searches

# ===============================================================

# ===============================================================

# Station-level hit-rate disparity (African vs White)

# - Cleans up logic, avoids duplicate libraries, keeps numeric p for sorting

# - Uses existing objects: `ddatr`, helper `is_yes()`, and loaded packages

# ===============================================================

# ---- parameters ----

MIN_AFRICAN <- 10
MIN_WHITE   <- 0

# ---- choose station column (prefer Station.uniform if present) ----

if ("Station.uniform" %in% names(ddatr)) {
STATION_COLNAME <- "Station.uniform"
} else {
station_like <- names(ddatr)[stringr::str_detect(names(ddatr), regex("station", ignore_case = TRUE))]
if (length(station_like) == 0) stop("No station-like column found in `ddatr`.")
if (length(station_like) > 1) warning("Multiple station-like columns detected: ",
paste(station_like, collapse = ", "),
"\nUsing the first: ", station_like[1])
STATION_COLNAME <- station_like[1]
}

# ---- build analysis frame ----

dat_station_long <- ddatr %>%
dplyr::filter(
!is.na(Found),
Racial.appearance.transformed %in% c("White", "African"),
!is.na(.data[[STATION_COLNAME]])
) %>%
dplyr::transmute(
Station   = factor(.data[[STATION_COLNAME]]),
Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
Found_yes = is_yes(Found)
)

# ---- summarise counts by Station × Race ----

dat_station_wide <- dat_station_long %>%
dplyr::count(Station, Race, name = "n") %>%
tidyr::pivot_wider(names_from = Race, values_from = n, values_fill = 0, names_prefix = "n_") %>%
dplyr::left_join(
dat_station_long %>%
dplyr::group_by(Station, Race) %>%
dplyr::summarise(found_yes = sum(Found_yes, na.rm = TRUE), .groups = "drop") %>%
tidyr::pivot_wider(names_from = Race, values_from = found_yes, values_fill = 0, names_prefix = "fy_"),
by = "Station"
) %>%

# apply thresholds

dplyr::filter(n_African >= MIN_AFRICAN, n_White >= MIN_WHITE)

# ---- bail out gracefully if nothing passes thresholds ----

if (nrow(dat_station_wide) == 0L) {
cat("No stations met the thresholds (≥", MIN_AFRICAN, " African and ≥", MIN_WHITE, " White searches).\n", sep = "")
} else {

# ---- compute proportions + prop.test per station (vectorised via pmap) ----

tests_tbl <- dat_station_wide %>%
dplyr::mutate(
prop_African = fy_African / n_African,
prop_White   = fy_White   / n_White,
test = purrr::pmap(
list(fy_African, fy_White, n_African, n_White),
~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
)
) %>%
dplyr::mutate(tidy = purrr::map(test, broom::tidy)) %>%
tidyr::unnest(tidy) %>%
dplyr::mutate(
risk_ratio = ifelse(prop_White > 0, prop_African / prop_White, NA_real_),
p_val_num  = p.value,                                      # keep numeric for sorting
p_val_lab  = ifelse(p_val_num < 0.001, "<0.001", sprintf("%.3f", p_val_num)),
sig        = dplyr::case_when(
p_val_num < 0.001 ~ "***",
p_val_num < 0.010 ~ "**",
p_val_num < 0.050 ~ "*",
TRUE              ~ ""
)
) %>%
dplyr::arrange(p_val_num)

# ---- format for display ----

station_tests_fmt <- tests_tbl %>%
dplyr::transmute(
Station,
`African n`  = n_African,
`White n`    = n_White,
`African %`  = scales::percent(prop_African, accuracy = 0.1),
`White %`    = scales::percent(prop_White,   accuracy = 0.1),
`Risk ratio` = round(risk_ratio, 2),
`95% CI (L)` = round(conf.low, 2),
`95% CI (U)` = round(conf.high, 2),
`p-value`    = p_val_lab,
sig
)

# ---- pretty table ----

kbl(
station_tests_fmt,
caption = paste0(
"Hit-rate disparity by Station (African vs White; Station column = '",
STATION_COLNAME, "'; only stations with ≥", MIN_AFRICAN, " African & ≥", MIN_WHITE, " White searches)"
),
align = "lrrrrrrrrr"
) |>
kable_paper(full_width = FALSE) |>
kable_styling(
bootstrap_options = c("striped", "hover"),
font_size = 12,
position = "left"
) |>
add_header_above(c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2))
}

```
Across stations with at least ten searches of African-appearing individuals, the cross-tabulations show substantial variation in hit rates across locations. A number of stations—including Werribee, Caroline Springs, Melbourne East, Sunshine, Dandenong, and Mill Park—display lower observed hit rates for African-appearing individuals, with risk ratios below one.

Only one station shows a difference that reaches conventional levels of statistical significance. For most stations, confidence intervals are wide and p-values non-significant, reflecting the limited number of African-appearing searches available at the individual station level. These estimates should therefore be interpreted as descriptive rather than inferential.

Nonetheless, when considered collectively, the pattern of results is informative. The fact that more stations exhibit lower rather than higher hit rates for African-appearing individuals suggests that disparities observed at the division and metropolitan levels are not driven by a single outlier location.


# Appendix

## Hit rates all racial categories  

**Table 8 Hit rates all racial appearances**
```{r Finds racial all, echo=FALSE, warning=FALSE, results='asis'}

ct <- crosstable::crosstable(
  ddat,
  by = "Found",
  cols = "Racial.appearance.transformed",
  total = TRUE,
  test = FALSE,          # we'll add our own robust test
  showNA = "ifany",
  percent_digits = 1
)

ft <- crosstable::as_flextable(ct)

# 3) Safe association test (Fisher for 2x2, else Chi-square with simulated p)
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tres <- safe_test(ddat$Racial.appearance.transformed, ddat$Found)
p_line <- sprintf("Racial.appearance.transformed — %s, p = %s",
                  tres$label, format.pval(tres$p, digits = 4))

# 4) Add footer and a clear title
ft <- flextable::add_footer_lines(ft, values = c("Overall association test:", p_line))
ft <- flextable::set_caption(ft, caption = "Racial appearance vs Found (with totals)")
ft
```

## Racial categories mapped

The below table shows how the original recorded data maps to the dataset used in this analysis. Categories changed multiple times across the time period in scope.

**Table 9 Racial appearance mapping**
```{r mapping-categories-2022-2023, echo = FALSE}
ddat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```

