---
title: "Racial profiling - drug searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---
# Abstract {-}

This study analyses 7,534 warrantless drug searches conducted by uniform, station-based officers during pedestrian encounters in metropolitan Melbourne (2018, 2019, 2022, 2023). It focuses on people perceived by police as White or African, who together comprise over half of all recorded searches.

First, descriptive cross-tabulations show that, across Melbourne, “hit rates” (contraband found) are lower for African-appearing individuals (11.4%) than for White-appearing individuals (14.1%), with a significant association between race and search outcomes.

Second, a city-wide fixed-effects logistic regression (Model A) is estimated on stations with at least five searches of African-appearing people, comparing outcomes within the same station and year, i.e. within comparable patrol areas and time periods. After controlling for officer gender and rank, African-appearing individuals have about 30% lower odds of a hit than White-appearing individuals (OR = 0.72, 95% CI 0.54–0.95, p = 0.022).

Third, a multilevel logistic regression (Model B) with random intercepts for stations (N = 7,534 searches across 84 stations) models the full metropolitan sample, allowing baseline hit rates to vary across stations. This model finds a similar Melbourne-wide disparity (OR = 0.73, 95% CI 0.57–0.93, p = 0.011), while again showing no clear effect of officer gender or rank.

Finally, division-specific fixed-effects models and station-level cross-tabs examine how these disparities are distributed spatially. The largest and most statistically robust disparity appears in NWMR Division 3 (Brimbank–Melton), with suggestive but non-significant patterns in several other north-western divisions, and little evidence of disparity in most southern and eastern divisions. Station-level results are often under-powered but point to a small number of teams where searches of African-appearing people are consistently less likely to yield contraband.

Taken together, these findings are consistent with the idea that, despite Victoria Police’s formal discipline and hierarchy, local station patrol teams operate with distinct cultures, management styles, and oversight practices that shape how search powers are exercised in the communities they serve. In this context, racial attitudes and informal norms within particular teams can influence search decisions, potentially more in some community settings than others. While the hit-rate framework cannot establish motive, the persistence and spatial concentration of lower evidentiary yields for African-appearing individuals underscore the need for ongoing, station-level monitoring, transparency, and accountability in Victoria Police’s stop-and-search practices.

# Introduction

This analysis examines **7,534 warrantless drug searches** conducted by **uniformed officers during pedestrian encounters** in **metropolitan Melbourne** during **2018, 2019, 2022, and 2023**, involving people perceived as having *White* or *African* appearance by Victoria Police.  

These encounters represent the most routine and discretionary form of public-place policing, where outcomes depend heavily on individual officer judgement rather than intelligence-led or warrant-based operations.  

## Purpose and scope

The aim is to investigate whether search outcomes are consistent with **racial profiling** — that is, whether data indicates that people African are subjected to searches on a **lower evidentiary threshold** than people perceived as White.  

This focus reflects both **data quality and historical context**. The **Haile-Michael case** and **recurring allegations of racial profiling of African-Australians** in Victoria have drawn sustained public and policy attention to policing practices affecting this group. Concentrating on this comparison allows a direct test of whether such disparities are observable in routine operational data.  

The two racial appearance categories analysed — *African* and *White* — represent two of the largest and most consistently recorded groupings in the dataset, enabling robust and interpretable comparison of outcomes across time and space.  

Restricting the analysis to **uniform pedestrian encounters** and **drug-related searches** further reduces confounding. These encounters occur in public settings under a single legislative authority — *section 82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)* — which permits searches without a warrant when an officer has *reasonable grounds to suspect* drug possession.  

By excluding **vehicle stops**, **warrant searches**, and **specialist or Transit Police operations**, the analysis isolates the most discretionary form of search activity. In particular, removing **Transit Police** and **Protective Services Officers (PSOs)** excludes searches around train stations and transport hubs — areas with distinct patterns of offending and demographic composition — ensuring the analysis reflects **community-based policing contexts** rather than **transport-enforcement settings**.  

## VicPol organisational structure

A key feature of this dataset is that every search is linked to the officer’s **home police station**, enabling the use of **station-level controls** to account for local operational contexts.  

There are **84 uniform stations** represented in the data, each associated with a **Police Service Area (PSA)** and nested within one or more **Local Government Areas (LGAs)** and **Victoria Police Divisions**.  

Within each PSA, every uniform station maintains a **geographic response or patrol zone** that defines its core area of responsibility for first response and routine patrol. These zones are set operationally by Victoria Police’s regional commands to balance population, workload, and response times. They represent the smallest consistent unit of geographic responsibility for **general-duties (uniform) police**, excluding specialist units such as Transit Police or Highway Patrol that operate across broader areas.  


## Legal context: section 82 powers

Under **section 82** of the *Drugs, Poisons and Controlled Substances Act 1981 (Vic)*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.  

This power represents the majority of search activity in the dataset and embodies the high-discretion environment most relevant to testing for racial bias: officers must decide, often rapidly, who appears “suspicious enough” to search.  

## The hit-rate test framework

The analysis applies the outcomes-based or “hit-rate” test introduced by **Knowles, Persico, and Todd (2001)** and formalised by **Persico and Todd (2004)**.  

In their economic model, police officers are assumed to allocate **pedestrian searches** across groups to maximise the probability of discovering contraband. If officers act without bias, equilibrium requires that the hit rates — the proportion of searches yielding contraband — are equal across racial groups. Conversely, if officers apply a lower suspicion threshold to one group (searching them more frequently or with less cause), that group will exhibit a lower hit rate, as more searches are conducted on weaker grounds.  

### Limitations and critiques

Despite its appeal, the hit-rate framework has attracted substantial criticism in the policing and econometric literature. Analysts including Ridgeway and MacDonald (2010) note that hit-rate analyses can misattribute disparities if situational contexts differ systematically across racial groups.

Similarly, failing to control for station-level variation or officer assignment patterns can confound results, overstating or obscuring racial effects. As such, the hit-rate test should be interpreted as an indicator of potential bias, not as a definitive measure. When combined with station-level controls, fixed-effects modelling, and internal benchmarking (comparing officers or stations operating in similar contexts), it remains a valuable tool for identifying areas where discretionary decision-making may be unequally applied.

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(scales)
library(stringr)
library(tidyr)
library(fixest)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Weapons`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )


is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}


make_hit_table <- function(
  data,
  group_var,                   # e.g. "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    dplyr::filter(!is.na(!!grp), !is.na(Found)) %>%
    dplyr::filter(Racial.appearance.transformed %in% c("African","White")) %>%
    dplyr::mutate(FoundYes = is_yes(Found)) %>%
    dplyr::count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    dplyr::group_by(!!grp, Racial.appearance.transformed) %>%
    dplyr::summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    dplyr::filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble::tibble())

  out <- tab %>%
    dplyr::mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    dplyr::select(-test) %>%
    tidyr::unnest(tidied) %>%
    dplyr::mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = dplyr::if_else(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    # Important: use !!rlang::sym(group_var) on LHS, and avoid Unicode minus in names
    dplyr::transmute(
      !!rlang::sym(group_var) := !!grp,
      `African (succ/total)`   = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`     = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`       = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`         = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr - Wh)`        = scales::percent(diff_est,         accuracy = 0.1),  # <-- ASCII hyphen
      `Risk ratio (Afr/Wh)`    = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                      = sprintf("%.3f", p.value),
      `p (BH)`                 = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                    = stars
    )

  if (filter_sig_only) {
    out <- out %>% dplyr::filter(as.numeric(stringr::str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% dplyr::arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))
}

print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) display <- display %>% dplyr::select(-`p (BH)`)

  display %>%
    knitr::kable(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    kableExtra::column_spec(1, width = "24em") %>%
    kableExtra::add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}

```

# Dataset

## Dataset scope and coverage

The dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in **2018, 2019, 2022, and 2023**.  

Data for **2020 and 2021** were excluded due to the atypical policing environment during the COVID-19 pandemic, when search activity and mobility patterns were highly irregular.

The data were released through a **Freedom of Information (FOI)** request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and cover **public-place searches** carried out under a range of legal powers.

For this analysis, we restrict to:

- **Geography:** Metropolitan Melbourne  
- **Contact type:** Pedestrian contact only  
- **Unit type:** Uniform police only  
- **Search type:** Drug searches conducted under *s82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)*  
- **Station linkage:** Station-based officers with a station patrol area excluding train stations
- **Racial appearance:** Restricted to *White* and *African* appearance  


The original FOI dataset was **enhanced with additional contextual data** to improve the precision and interpretability of geographic comparisons.  

Each search record was linked to:

- The **Victoria Police organisational hierarchy**, allowing each record to be associated with its corresponding **Region, Division, Police Service Area (PSA), and Station**.  

- **Crime Statistics Agency (CSA) data** on the **drug offence rate** and **total offence rate** in the **Local Government Area (LGA)** in which each police station is situated, for the corresponding **year of the search**.  

## Data limitations

### Racial appearance classification

The original dataset for drug searches conducted by uniform officers on foot patrol included 13,896 records.

Approximately **20% of these records lacked racial appearance information**, and several smaller categories had minimal counts or inconsistent coding across years.

For analytical reliability and interpretability, disparity testing focuses on **White** and **African** appearance categories. Although the African group is not the second-largest category by count, these two groupings together represent over half (54%) of all relevant records and are consistently coded across years. The original coded categories, mapped to the cleaned racial categories utilised in this analysis, appears in the Appendix.

**A note on 'Missing' and 'White-appearing'**

Descriptively, the hit rate among cases with missing racial appearance (14.5%) was very close to that of White-appearing individuals (14.1%), suggesting that many blank or unrecorded entries are likely to be White-appearing rather than missing at random. 

### Age information

Age data were missing for **42.9%** of all pedestrian searches in the restricted subset.  Among cases with recorded ages, the distribution was heavily concentrated among younger people — **individuals aged 18 to 45 years accounted for more than half of all identified persons** — with relatively few searches involving older adults (see Table 2 below).  

Because the age profile of those searched is already strongly youthful, the omission of age as a covariate is **unlikely to distort results substantially**, provided that missingness is not systematically related to racial appearance. 


```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddatr <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45","46 to 55", 
                                                         "56 to 65","66 and over")))
 
```

# Hit Rates — Descriptive Cross-Tabulations

Before estimating formal models, it is important to examine the *raw relationship* between racial appearance and search outcomes, and how contextual variables may shape these differences.  

The following tables report:  

- Hit rates by racial appearance, gender, year, rank, and police division.  
- Statistical tests (χ² or Fisher’s exact) assessing whether differences are significant.  

Subsequent sections will model these same relationships using fixed-effects and multilevel logistic regression frameworks, controlling for the confounding influences identified here.

**Table 1 Hit rates by person characteristics**
```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}

# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))%>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop(),
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Rank  = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    )
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Year", "Rank", "Division")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(
    p_fmt = ifelse(p < 0.001, "<0.001", sprintf("%.5f", p)),
    p_str = sprintf("%s — %s, p = %s", variable, test, p_fmt)
  )

# Append p-values to the table footer (each on its own line)
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft


```


### Overview of results

The cross-tabulations reveal **modest but potentially meaningful variation in hit rates across racial appearance**.  The **Fisher’s exact test for racial appearance** (p = 0.039) indicates a statistically significant association between race and search outcomes. Searches of **African-appearing individuals** yielded a lower evidentiary return (11.4%) than those of **White-appearing individuals** (14.1%). 

By contrast, **gender (p = 0.33)** and **age group (p = 0.64)** show no significant differences in hit rates. Given that most searches involve young adults, this reduces concern that age composition alone explains the observed racial disparity. We do however recognise that missingness in the data is a limitation - we cannot definitively rule out a relevent systemic relationship.


**Year effects** (p ≈ 0.17) and police officer **Rank effects** are minor, suggesting overall stability in hit rates across the observed period, and some increase in search efficacy relating to experience.  

Finally, the **strong association by police division (p < 0.001)** highlights the importance of community context: hit rates vary considerably between divisions, indicating that **local operational culture and context** may play a key role in shaping search outcomes.


# Statistical modelling  

## Model A — Fixed-Effects Logistic Regression (Station × Year)


While descriptive cross-tabulations highlight consistent racial differences in hit rates, they cannot determine whether these disparities reflect **unequal treatment** or simply differences in **where and when** searches occur.  

To separate these effects, we estimate a **fixed-effects logistic regression** that isolates racial disparities *within the same police station and year*, holding constant local operational context and temporal variation.  This approach allows us to test whether, under comparable conditions, discretionary searches of African-appearing individuals were systematically less likely than White-appearing individuals to yield contraband.

### Model specification

- **Dependent variable:** whether a search resulted in contraband being found (`Found_yes = 1`).  
- **Key predictor:** perceived racial appearance (*African* vs *White*).  
- **Covariates:** gender and officer rank.  
- **Fixed effects:** Station × Year, absorbing all between-station and between-year variation.  
- **Standard errors:** clustered by station to account for repeated observations within sites.  

Formally, for search \(i\) conducted at station \(s\) in year \(t\), the model is:

\[
Y_{ist} \sim \text{Bernoulli}(p_{ist}),
\]

\[
\log\left(\frac{p_{ist}}{1 - p_{ist}} \right)
= \beta_1 \,\text{African}_i
+ \boldsymbol{\gamma}^\top \mathbf{G}_i
+ \boldsymbol{\delta}^\top \mathbf{R}_i
+ \alpha_{st},
\]

where \(Y_{ist}=1\) if contraband was found; \(\text{African}_i\) is an indicator for perceived racial appearance “African” (with “White” as the reference); \(\mathbf{G}_i\) and \(\mathbf{R}_i\) are vectors of officer gender and officer rank indicators; and \(\alpha_{st}\) is a Station × Year fixed effect. Standard errors are clustered at the station level, allowing arbitrary correlation among searches within the same station.

### Data restriction and rationale  

**Only stations with more than 5 pedestrian searches of African-appearing individuals** were retained (5,535 of 7,534 observations).  This threshold ensures that comparisons between racial groups are statistically meaningful *within each station* while maintaining adequate sample size for estimation. As a result, the analysis is restricted to **40 of the 84 uniform stations** in metropolitan Melbourne.  

Although this restriction reduces the overall number of observations, it improves reliability and avoids unstable or spurious effects from stations with very few African searches.

**Table 2 Model A — Fixed-effects logistic regression (Station×Year)**
```{r Model A, echo=FALSE,message=FALSE, warning=FALSE}

# ===============================================================
# Model A — Fixed-effects logistic regression (Station×Year)
# Restrict to stations with ≥5 African-appearance searches (aggregated across all years)
# Includes Gender and Rank covariates; clustered SEs by Station
# ===============================================================


# --- 1. Define eligible stations (≥5 African across all years) ---
elig_stations <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  count(Station.uniform, Racial.appearance.transformed, name = "n") %>%
  tidyr::pivot_wider(
    names_from  = Racial.appearance.transformed,
    values_from = n,
    values_fill = 0
  ) %>%
  filter(African >= 5) %>%  # stations with at least 10 African searches overall
  pull(Station.uniform)

# --- 2. Prepare dataset restricted to those stations ---
dat1 <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Station.uniform %in% elig_stations
  ) %>%
  mutate(
    Race        = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes   = is_yes(Found),
    StationYear = interaction(Station.uniform, Year, drop = TRUE),
    Gender      = factor(Gender),

    # Clean and collapse ranks
    Rank_clean = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    ),
    Rank_clean = forcats::fct_drop(Rank_clean),
    Rank_clean = forcats::fct_relevel(Rank_clean, "CONST")
  )

# --- 3. Fit Station×Year fixed-effects logistic model ---
m_fe_af5_all <- fixest::feglm(
  Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") | StationYear,
  family  = "binomial",
  data    = dat1,
  cluster = ~ Station.uniform
)

# --- 4. Tidy results (exclude FE dummies) ---
model_table <- broom::tidy(m_fe_af5_all, conf.int = TRUE) %>%
  dplyr::filter(!grepl("StationYear", term)) %>%
  dplyr::mutate(
    OR        = exp(estimate),
    CI_low    = exp(conf.low),
    CI_high   = exp(conf.high),
    p_str     = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) %>%
  dplyr::transmute(
    Term         = term,
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = p_str
  )

# --- 5. Display results ---
kbl(
  model_table,
  caption = paste0(
    "Fixed-effects logistic regression (Station×Year) — restricted to stations with ≥5 African searches (clustered by Station.uniform; N = ",
    nrow(dat1),
    " searches)."
  ),
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")


```

### Interpretation  

Within the same station and year, searches of African-appearing people were **significantly less likely** to result in contraband being found (OR = 0.72, 95% CI [0.54, 0.95], p = 0.022).  

This corresponds to roughly a **30 percent lower odds** of a “hit” for African-appearing individuals compared with White-appearing individuals under like-for-like operational contexts.  

By contrast, gender and officer rank were not statistically significant predictors of search success, and their inclusion did not materially affect the race coefficient. Searches conducted by Constables, Senior Constables, and Sergeants yielded broadly similar outcomes once station- and year-level variation were held constant.  

## Model B — Multilevel Logistic Regression (Random Intercepts for Station)

To complement the fixed-effects analysis, we estimated a **multilevel logistic regression** with **random intercepts for police stations**.  

Whereas Model A isolates within-station differences by removing all station-level variation, Model B explicitly models that variation, treating each station’s baseline likelihood of a “hit” as a random effect.

This approach recognises that stations differ systematically in operational practices, local contexts, and levels of policing intensity.  

Rather than controlling away this variation, the multilevel model captures it directly, enabling more **city-wide generalisation** while still accounting for the clustering of searches within stations.  

The dependent variable again indicates whether contraband was found (*Found_yes = 1*), with predictors for perceived race (*African* vs *White*), gender, and local drug-crime rate.

No data restrictions were applied.

### Model specification  

Model B was estimated using `lme4::glmer` with a logit link:

\[
\text{logit}\big(P(\text{Found}_{i})\big)
  = \beta_0 
  + \beta_1(\text{Race}_{i})
  + \beta_2(\text{Gender}_{i})
  + \beta_3(\text{DrugCrimeRate}_{\text{LGA,Year}})
  + u_{\text{Station}},
\]

where \(u_{\text{Station}} \sim N(0, \sigma^2_{\text{Station}})\) captures station-level random intercepts.

The dependent variable is whether a search resulted in contraband being found (`Found_yes = 1`). Drug-crime rates are included as a **contextual covariate measured at the LGA–year level**, assigned to each search based on the LGA in which the station is located.

Standard errors are clustered by **station** to account for repeated observations within sites.

**Table 3 Model B — Multilevel logistic regression (random intercepts for Station)**
```{r modelB multilevel logistic, echo = FALSE, warning = FALSE, message = FALSE}
# ===============================================================
# Model B — Multilevel logistic regression (random intercepts for Station)
# Adds Rank as a fixed effect (CONST as reference; SSGT→SGT, PSO→CONST)
# ===============================================================


dat2 <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  mutate(
    Found_yes = is_yes(Found),
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Gender    = factor(Gender),
    Station   = factor(Station.uniform),

    # Clean Rank: collapse small/rare levels; set CONST as ref
    Rank_clean = dplyr::recode(
      Rank.of.Member,
      "SSGT" = "SGT",
      "PSO"  = "CONST",
      .default = Rank.of.Member
    ),
    Rank_clean = fct_drop(Rank_clean),
    Rank_clean = fct_relevel(Rank_clean, "CONST"),

    # Ensure crime rate is numeric (if present)
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate))
  )

# Optional stability filter: exclude very small stations
#station_counts <- dat2 %>% count(Station)
#dat2 <- dat2 %>% filter(Station %in% station_counts$Station[station_counts$n >= 20])

# Store N for table captions
N_obs <- nrow(dat2)
N_stations <- length(unique(dat2$Station))

# Fit multilevel logistic model: random intercepts by Station
m_station <- glmer(
  Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") + scale(Drug.crime.rate) + (1 | Station),
  family  = binomial,
  data    = dat2,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# ---- Table A: Race effect (OR with 95% CI) ----
b_results <- broom.mixed::tidy(m_station, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
  mutate(
    OR      = exp(estimate),
    CI_low  = exp(conf.low),
    CI_high = exp(conf.high)
  ) %>%
  filter(term == "RaceAfrican") %>%
  transmute(
    Term         = "African vs White",
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = signif(p.value, 3)
  )

kbl(
  b_results,
  caption = paste0(
    "Multilevel logistic regression (random intercepts for Station) — odds of contraband found (Race effect). ",
    "N = ", format(N_obs, big.mark = ","), 
    " searches across ", N_stations, " stations."
  ),
  align = "lrrrr"
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")

# ---- Table B (optional): all fixed effects nicely formatted ----
coef_table <- broom.mixed::tidy(m_station, effects = "fixed", conf.int = TRUE, conf.method = "Wald") %>%
  mutate(
    OR      = exp(estimate),
    CI_low  = exp(conf.low),
    CI_high = exp(conf.high),
    Term    = case_when(
      term == "RaceAfrican"                ~ "Race: African vs White",
      term == "GenderM"                    ~ "Gender: Male vs Female",
      grepl("^i\\(Rank_clean.*SCONST", term) ~ "Rank: Senior Constable vs Constable",
      grepl("^i\\(Rank_clean.*SGT", term)    ~ "Rank: Sergeant vs Constable",
      term == "scale(Drug.crime.rate)"     ~ "Drug crime rate (z)",
      TRUE                                 ~ term
    )
  ) %>%
  transmute(
    Term,
    `Odds Ratio` = round(OR, 3),
    `95% CI (L)` = round(CI_low, 3),
    `95% CI (U)` = round(CI_high, 3),
    `p-value`    = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  )

kbl(
  coef_table,
  caption = paste0(
    "Multilevel logistic regression (random intercepts for Station) — all fixed effects. ",
    "N = ", format(N_obs, big.mark = ","), 
    " searches across ", N_stations, " stations."
  ),
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")
```

### Interpretation

The multilevel model, estimated on **7,534 searches across 84 police stations**, shows that even after accounting for officer rank, gender, and variation between stations, **African-appearing individuals remained significantly less likely to yield contraband** than White-appearing individuals searched under similar conditions.

The odds of a “hit” were approximately **27% lower** for African-appearing individuals (*OR* = 0.73, 95% CI 0.57–0.93, *p* = 0.011).  This disparity persists after adjusting for station-level context and individual-level covariates.

Gender and rank effects were not statistically significant at the 5 percent level, though searches by **Sergeants** trended toward a higher hit rate (*OR* = 1.35, *p* = 0.087) and **men** were marginally more likely to be found with contraband (*OR* = 1.16, *p* = 0.083).  

Local drug-crime rates showed no meaningful association with search success - this may be due to the coarseness of the LGA level unit of measurement, or the fact that the station level variation is already capturing this effect.

Together, these findings indicate a **robust racial disparity in evidentiary yield** that is not explained by officer characteristics or local policing environments.


### Geographic concentration of disparities

Taken together, the two models indicate that racial disparities in search outcomes are **not evenly distributed across metropolitan Melbourne**.

The **multilevel model (Model B)** shows that, on average across all stations, searches of individuals perceived as *African* are significantly less likely to yield contraband than those of *White* individuals—even after adjusting for officer rank, gender, and station-level differences.  

By contrast, the **fixed-effects model (Model A)**—which compares searches *within the same station and year*—shows that much of this disparity arises from **specific local contexts** rather than a uniform pattern across the city. The fact that the race effect persists overall (Model B) but weakens when controlling tightly for Station×Year (Model A) indicates that the disparity is **geographically concentrated**, driven by particular operational environments rather than evenly distributed across all stations.  

The significant variance in random station intercepts (σ²ₛₜₐₜᵢₒₙ > 0) further confirms heterogeneity in baseline hit rates, implying that some stations systematically produce lower evidentiary yields for African-appearing individuals.  

This pattern suggests that the disparity likely reflects **local operational cultures, leadership practices, or community-policing dynamics** rather than a single, city-wide bias. It underscores the importance of **spatial accountability**—identifying and addressing the specific stations where discretionary searches of African-appearing people are consistently less evidence-based.

# Division and station disparities

Division level cross tabulations and modelling add further insight into racial profiling at the community level.


## Division - Cross tabs

Differences in hit rates between African- and White-appearing individuals were assessed using two-sample tests for equality of proportions (chi-square, no continuity correction), conducted separately for each division.

**Table 4 Division cross-tabs**
```{r division cross tabs, echo = FALSE, warning=FALSE, message=FALSE}

library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(scales)
library(kableExtra)

MIN_AFRICAN <- 1
MIN_WHITE   <- 1

dat_div_long <- ddatr %>%
  filter(
    !is.na(Found),
    !is.na(Division),
    Racial.appearance.transformed %in% c("White", "African")
  ) %>%
  transmute(
    Division  = factor(Division),
    Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
    race_col  = tolower(as.character(Race)),   # "white", "african"
    Found_yes = is_yes(Found)
  )

dat_div_wide <- dat_div_long %>%
  count(Division, race_col, name = "n") %>%
  pivot_wider(names_from = race_col, values_from = n, values_fill = 0, names_prefix = "n_") %>%
  left_join(
    dat_div_long %>%
      group_by(Division, race_col) %>%
      summarise(fy = sum(Found_yes, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = race_col, values_from = fy, values_fill = 0, names_prefix = "fy_"),
    by = "Division"
  ) %>%
  filter(n_african >= MIN_AFRICAN, n_white >= MIN_WHITE) %>%
  # extra guard: prop.test requires strictly positive denominators
  filter(n_african > 0, n_white > 0)

if (nrow(dat_div_wide) == 0L) {
  cat("No divisions met thresholds.\n")
} else {

  # safe prop.test (returns NULL on any failure)
  safe_prop <- purrr::possibly(
    function(fyA, fyW, nA, nW) stats::prop.test(x = c(fyA, fyW), n = c(nA, nW), correct = FALSE),
    otherwise = NULL
  )

  div_tests_raw <- dat_div_wide %>%
    mutate(
      prop_african = fy_african / n_african,
      prop_white   = fy_white   / n_white,
      test = pmap(
        list(fy_african, fy_white, n_african, n_white),
        ~ safe_prop(..1, ..2, ..3, ..4)
      ),
      tidy = map(test, ~ if (is.null(.x)) NULL else broom::tidy(.x))
    )

  # drop divisions where test failed so p.value/conf.low/conf.high exist
  div_tests <- div_tests_raw %>%
    filter(!map_lgl(tidy, is.null)) %>%
    unnest(tidy, keep_empty = FALSE) %>%
    mutate(
      risk_ratio = ifelse(prop_white > 0, prop_african / prop_white, NA_real_),
      p_val_num  = p.value,
      p_val_lab  = ifelse(p_val_num < 0.001, "<0.001", sprintf("%.3f", p_val_num)),
      sig        = case_when(
        p_val_num < 0.001 ~ "***",
        p_val_num < 0.010 ~ "**",
        p_val_num < 0.050 ~ "*",
        TRUE              ~ ""
      )
    ) %>%
    arrange(p_val_num) %>%
    transmute(
      Division,
      `African n`  = n_african,
      `White n`    = n_white,
      `African %`  = scales::percent(prop_african, accuracy = 0.1),
      `White %`    = scales::percent(prop_white,   accuracy = 0.1),
      `Risk ratio` = round(risk_ratio, 2),
      `Δ hit-rate (L)` = round(conf.low, 3),
      `Δ hit-rate (U)` = round(conf.high, 3),
      `p-value`    = p_val_lab,
      sig
    )

  # optional message if any divisions were skipped
  n_skipped <- sum(map_lgl(div_tests_raw$tidy, is.null))
  if (n_skipped > 0) {
    message("Note: skipped ", n_skipped, " division(s) where prop.test failed (e.g., degenerate counts).")
  }

  if (nrow(div_tests) == 0L) {
    cat("All divisions meeting thresholds had prop.test failures; no results to display.\n")
  } else {
    kbl(
      div_tests,
      caption = paste0(
        "Hit-rate disparity by Division (African vs White; divisions with ≥",
        MIN_AFRICAN, " African & ≥", MIN_WHITE, " White searches)"
      ),
      align = "lrrrrrrrrr"
    ) |>
      kable_paper(full_width = FALSE) |>
      kable_styling(bootstrap_options = c("striped","hover"), font_size = 12, position = "left") |>
      add_header_above(c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2))
  }
}


```

## Divisions - Fixed-effects logistic regressions (Station × Year)  

To estimate racial disparities within each Division \(d\), we fit a station–year fixed-effects logistic regression of the form:

\[
Y_{istd} \sim \text{Bernoulli}(p_{istd}),
\]

\[
\log\left( \frac{p_{istd}}{1 - p_{istd}} \right)
= \beta_{0d}
+ \beta_{1d}\,\text{African}_{i}
+ \boldsymbol{\gamma}_d^{\top}\mathbf{G}_{i}
+ \boldsymbol{\delta}_d^{\top}\mathbf{R}_{i}
+ \alpha_{st},
\]

where:  
- \(Y_{istd} = 1\) if contraband was found on search \(i\) at station \(s\) in year \(t\) in Division \(d\), and \(0\) otherwise;  
- \(\text{African}_{i}\) is an indicator for perceived racial appearance “African” (reference: “White”);  
- \(\mathbf{G}_{i}\) and \(\mathbf{R}_{i}\) are dummy variables for officer gender and rank, controlling for differences in officer composition across stations;  
- \(\alpha_{st}\) is a Station×Year fixed effect, included to ensure comparisons are made **within the same station and year**, thereby removing confounding from location-specific or time-varying enforcement patterns;  
- \(\beta_{1d}\) is the Division-specific log-odds coefficient for African-appearing individuals, with \(\exp(\beta_{1d})\) interpreted as the corresponding odds ratio.

Standard errors are clustered at the station level to account for repeated searches conducted within the same policing environment.

**Table 5 Division models**
```{r division station year, echo=FALSE, message=FALSE, warning = FALSE}

# ===============================================================
# Division-split models
# Each model: Fixed-effects logistic regression (Station×Year)
# Subset sample by Division; cluster SEs by Station
# ===============================================================

# --- Parameters ---
MIN_AFRICAN <- 0   # minimum African searches per station (across all years)

# --- Auto-detect Division column ---
div_col <- names(ddat)[str_detect(names(ddat), regex("division", ignore_case = TRUE))]
if (length(div_col) == 0) stop("No Division column found in ddat.")
DIV_COL <- div_col[1]
DIV_SYM <- rlang::sym(DIV_COL)

# --- Eligible stations (≥ MIN_AFRICAN African total) ---
elig_stations <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African", "White")) %>%
  count(Station.uniform, Racial.appearance.transformed, name = "n") %>%
  tidyr::pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = n,
    values_fill = 0
  ) %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(Station.uniform)

# --- Prepare base dataset ---
dat <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Station.uniform %in% elig_stations
  ) %>%
  mutate(
    Race        = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes   = is_yes(Found),
    StationYear = interaction(Station.uniform, Year, drop = TRUE),
    Gender      = factor(Gender),
    Rank_clean  = recode(Rank.of.Member, "SSGT" = "SGT", "PSO" = "CONST", .default = Rank.of.Member),
    Rank_clean  = forcats::fct_relevel(factor(Rank_clean), "CONST")
  )

# --- Run model per Division ---
divisions <- sort(unique(dat[[DIV_COL]]))
results_list <- list()

for (div in divisions) {
  d_sub <- dat %>% filter(!!DIV_SYM == div)
  if (nrow(d_sub) < 50) next  # skip small samples
  
  # Fit FE model
  mod <- tryCatch(
    fixest::feglm(
      Found_yes ~ Race + Gender + i(Rank_clean, ref = "CONST") | StationYear,
      family = "binomial",
      data = d_sub,
      cluster = ~ Station.uniform
    ),
    error = function(e) NULL
  )
  if (is.null(mod)) next
  
  # Tidy table
  tidy_tab <- broom::tidy(mod, conf.int = TRUE) %>%
    filter(!str_detect(term, "StationYear")) %>%
    mutate(
      OR     = exp(estimate),
      CI_low = exp(conf.low),
      CI_high= exp(conf.high),
      p_str  = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
    ) %>%
    transmute(
      Term = term,
      `Odds Ratio` = round(OR, 3),
      `95% CI (L)` = round(CI_low, 3),
      `95% CI (U)` = round(CI_high, 3),
      `p-value` = p_str
    )
  
  caption <- paste0(
    "Division: ", div,
    " — Fixed-effects logistic regression (Station×Year), clustered by Station.uniform (N = ",
    format(nrow(d_sub), big.mark=","), ")."
  )
  
  results_list[[div]] <- kbl(tidy_tab, caption = caption, align = c("l","r","r","r","r")) |>
    kable_paper(full_width = FALSE) |>
    kable_styling(bootstrap_options = c("striped","hover"), position = "left")
}

# --- Print results sequentially ---
for (div in names(results_list)) {
  cat("\n\n")
  print(results_list[[div]])
  cat("\n\n")
}
```

```{r div model summary, echo=FALSE, warning=FALSE, message=FALSE}


# Helper for p-value formatting (unchanged)
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

div_summary <- tribble(
  ~Division, ~OR,   ~CI_L,   ~CI_U,    ~p,     ~Interpretation,

  # Eastern
  "Eastern Div 1 (Boroondara – Monash – Whitehorse – Manningham)",
    0.612,   0.258,   1.448,   0.263,
    "No statistically reliable difference; point estimate suggests lower yield for African-appearing people, but with substantial uncertainty.",

  "Eastern Div 2 (Knox – Maroondah – Yarra Ranges)",
    0.970,   0.209,   4.506,   0.969,
    "No statistically reliable difference; estimates are highly imprecise due to small effective sample size.",

  # NWMR
  "NWMR Div 1 (Melbourne – Yarra)",
    1.127,   0.513,   2.477,   0.767,
    "Comparable hit rates for African- and White-appearing individuals.",

  "NWMR Div 2 (Wyndham – Hobsons Bay – Maribyrnong)",
    0.608,   0.313,   1.181,   0.142,
    "Lower odds of contraband finds for African-appearing people (≈ 40% lower), though not conventionally significant.",

  "NWMR Div 3 (Brimbank – Melton)",
    0.144,   0.033,   0.636,   0.011,
    "Substantially lower evidentiary yield for African-appearing individuals; disparity statistically significant.",

  "NWMR Div 4 (Hume – Moonee Valley – Merri-bek)",
    0.639,   0.377,   1.083,   0.096,
    "Moderately lower yield for African-appearing people; suggestive but not conventionally significant (p ≈ 0.10).",

  "NWMR Div 5 (Banyule – Darebin – Whittlesea – Nillumbik)",
    0.717,   0.503,   1.021,   0.065,
    "Suggestive lower yield for African-appearing people (≈ 30% lower odds), but not conventionally significant.",

  # SMR
  "SMR Div 1 (Port Phillip – Stonnington)",
    0.960,   0.554,   1.664,   0.884,
    "No difference detected; odds ratio close to 1 with wide confidence interval.",

  "SMR Div 2 (Glen Eira – Kingston)",
    1.084,   0.139,   8.433,   0.939,
    "No statistically reliable difference; estimates highly imprecise.",

  "SMR Div 3 (Dandenong – Casey – Cardinia)",
    0.838,   0.330,   2.130,   0.710,
    "No difference detected.",

  "SMR Div 4 (Frankston – Mornington)",
    1.190,   0.128,  11.047,   0.879,
    "No statistically reliable difference; estimates highly imprecise."
) %>%
  mutate(
    `OR (African vs White)` = round(OR, 2),
    `95% CI` = paste0("[", sprintf("%.2f", CI_L), ", ", sprintf("%.2f", CI_U), "]"),
    `p` = fmt_p(p)
  ) %>%
  select(
    Division,
    `OR (African vs White)`,
    `95% CI`,
    `p`,
    Interpretation
  )

kbl(
  div_summary,
  caption = "4.2.1 Summary of racial disparity patterns",
  align = c("l","r","l","r","l")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover"), position = "left")


```
Taken together, the division-specific fixed-effects models reveal **clear spatial variation** in the strength and direction of racial disparities.  

The most pronounced and statistically robust disparity appears in **NWMR Division 3 (Brimbank–Melton)**, where searches of African-appearing individuals were far less likely to yield contraband even after controlling for officer rank, gender, and station-year context.  

Several other Divisions—particularly **NWMR 2 (Wyndham–Maribyrnong–Hobsons Bay)** and **NWMR 5 (Darebin–Whittlesea–Banyule–Nillumbik)**—show similar but non-significant trends toward lower evidentiary yields.  

In contrast, **southern and eastern Divisions** show no consistent racial pattern, with odds ratios close to unity and wide confidence intervals reflecting smaller numbers of African-appearing searches.  

Overall, these findings suggest that **racial profiling is not uniformly distributed across metropolitan Melbourne**, but rather concentrated within certain policing environments, notably in Melbourne’s **north-western growth corridors**.  

The following section examines **station-level results** to identify which local areas drive these divisional patterns.

## Station-level - Cross tabs  

Station level cross tabulations add further granular insight into racial profiling at the organisational team level. This analysis was only performed for stations with at least 10 searches of African appearing people.

Differences in hit rates between African- and White-appearing individuals were assessed using two-sample tests for equality of proportions (chi-square, no continuity correction), conducted separately for each division. 

*Table 6 Station cross tabs* 
```{r station table, echo = FALSE, warning = FALSE, message=FALSE}
# ===============================================================

# Station-level hit rates by race

# • Whites searched, White finds, % White finds

# • Africans searched, African finds, % African finds

# • Total searches

# ===============================================================

# ===============================================================

# Station-level hit-rate disparity (African vs White)

# - Cleans up logic, avoids duplicate libraries, keeps numeric p for sorting

# - Uses existing objects: `ddatr`, helper `is_yes()`, and loaded packages

# ===============================================================

# ---- parameters ----

MIN_AFRICAN <- 10
MIN_WHITE   <- 0

# ---- choose station column (prefer Station.uniform if present) ----

if ("Station.uniform" %in% names(ddatr)) {
STATION_COLNAME <- "Station.uniform"
} else {
station_like <- names(ddatr)[stringr::str_detect(names(ddatr), regex("station", ignore_case = TRUE))]
if (length(station_like) == 0) stop("No station-like column found in `ddatr`.")
if (length(station_like) > 1) warning("Multiple station-like columns detected: ",
paste(station_like, collapse = ", "),
"\nUsing the first: ", station_like[1])
STATION_COLNAME <- station_like[1]
}

# ---- build analysis frame ----

dat_station_long <- ddatr %>%
dplyr::filter(
!is.na(Found),
Racial.appearance.transformed %in% c("White", "African"),
!is.na(.data[[STATION_COLNAME]])
) %>%
dplyr::transmute(
Station   = factor(.data[[STATION_COLNAME]]),
Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
Found_yes = is_yes(Found)
)

# ---- summarise counts by Station × Race ----

dat_station_wide <- dat_station_long %>%
dplyr::count(Station, Race, name = "n") %>%
tidyr::pivot_wider(names_from = Race, values_from = n, values_fill = 0, names_prefix = "n_") %>%
dplyr::left_join(
dat_station_long %>%
dplyr::group_by(Station, Race) %>%
dplyr::summarise(found_yes = sum(Found_yes, na.rm = TRUE), .groups = "drop") %>%
tidyr::pivot_wider(names_from = Race, values_from = found_yes, values_fill = 0, names_prefix = "fy_"),
by = "Station"
) %>%

# apply thresholds

dplyr::filter(n_African >= MIN_AFRICAN, n_White >= MIN_WHITE)

# ---- bail out gracefully if nothing passes thresholds ----

if (nrow(dat_station_wide) == 0L) {
cat("No stations met the thresholds (≥", MIN_AFRICAN, " African and ≥", MIN_WHITE, " White searches).\n", sep = "")
} else {

# ---- compute proportions + prop.test per station (vectorised via pmap) ----

tests_tbl <- dat_station_wide %>%
dplyr::mutate(
prop_African = fy_African / n_African,
prop_White   = fy_White   / n_White,
test = purrr::pmap(
list(fy_African, fy_White, n_African, n_White),
~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
)
) %>%
dplyr::mutate(tidy = purrr::map(test, broom::tidy)) %>%
tidyr::unnest(tidy) %>%
dplyr::mutate(
risk_ratio = ifelse(prop_White > 0, prop_African / prop_White, NA_real_),
p_val_num  = p.value,                                      # keep numeric for sorting
p_val_lab  = ifelse(p_val_num < 0.001, "<0.001", sprintf("%.3f", p_val_num)),
sig        = dplyr::case_when(
p_val_num < 0.001 ~ "***",
p_val_num < 0.010 ~ "**",
p_val_num < 0.050 ~ "*",
TRUE              ~ ""
)
) %>%
dplyr::arrange(p_val_num)

# ---- format for display ----

station_tests_fmt <- tests_tbl %>%
dplyr::transmute(
Station,
`African n`  = n_African,
`White n`    = n_White,
`African %`  = scales::percent(prop_African, accuracy = 0.1),
`White %`    = scales::percent(prop_White,   accuracy = 0.1),
`Risk ratio` = round(risk_ratio, 2),
`95% CI (L)` = round(conf.low, 2),
`95% CI (U)` = round(conf.high, 2),
`p-value`    = p_val_lab,
sig
)

# ---- pretty table ----

kbl(
station_tests_fmt,
caption = paste0(
"Hit-rate disparity by Station (African vs White; Station column = '",
STATION_COLNAME, "'; only stations with ≥", MIN_AFRICAN, " African & ≥", MIN_WHITE, " White searches)"
),
align = "lrrrrrrrrr"
) |>
kable_paper(full_width = FALSE) |>
kable_styling(
bootstrap_options = c("striped", "hover"),
font_size = 12,
position = "left"
) |>
add_header_above(c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2))
}

```
Across stations with at least ten searches of African-appearing individuals, the cross-tab results show substantial variability in hit rates across locations. A number of stations—including Werribee, Caroline Springs, Melbourne East, Sunshine, Dandenong and Mill Park—display lower observed hit rates for African-appearing individuals, with risk ratios below 1. Only one station (Pakenham) showed a higher observed hit rate for African-appearing people, though this difference was not statistically significant.

Importantly, only Werribee’s result reached conventional statistical significance. For all other stations, confidence intervals were wide and p-values non-significant, reflecting the relatively limited number of African searches available at individual stations. As a result, most station-level differences should be interpreted cautiously and as descriptive rather than inferential findings.

Nonetheless, when viewed collectively, the distribution of results shows more stations with lower rather than higher hit rates for African-appearing individuals. This directional pattern is not, on its own, evidence of a station-level effect, but it is consistent with the broader division-level regression findings.

Taken together, the station-level cross tabs provide localised descriptive insight: although few individual stations have sufficient sample size to produce statistically reliable estimates, the overall pattern does not contradict—and in several cases mirrors—the disparities observed in the higher-powered division-level models.

# Appendix

## Hit rates all racial categories  

**Table 7 Hit rates all racial appearances**
```{r Finds racial all, echo=FALSE, warning=FALSE, results='asis'}

ct <- crosstable::crosstable(
  ddat,
  by = "Found",
  cols = "Racial.appearance.transformed",
  total = TRUE,
  test = FALSE,          # we'll add our own robust test
  showNA = "ifany",
  percent_digits = 1
)

ft <- crosstable::as_flextable(ct)

# 3) Safe association test (Fisher for 2x2, else Chi-square with simulated p)
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tres <- safe_test(ddat$Racial.appearance.transformed, ddat$Found)
p_line <- sprintf("Racial.appearance.transformed — %s, p = %s",
                  tres$label, format.pval(tres$p, digits = 4))

# 4) Add footer and a clear title
ft <- flextable::add_footer_lines(ft, values = c("Overall association test:", p_line))
ft <- flextable::set_caption(ft, caption = "Racial appearance vs Found (with totals)")
ft
```


```{r missingness-metro-station-drugs-vs-weapons, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(Area.type == "Metro")

# ---- Drugs ---------------------------------------------------------------

drugs_station <- dat_metro %>%
  filter(!is.na(`Search.type - Drugs`)) %>%
  group_by(Station) %>%
  summarise(
    `Drugs: Total searches` = n(),

    drugs_race_missing_total_n =
      sum(is.na(Racial.appearance.transformed)) +
      sum(Racial.appearance.transformed %in% c("Other", "Unknown"), na.rm = TRUE),

    drugs_age_missing_total_n = sum(is.na(Age)),

    `Drugs: Race missing total (%)` =
      drugs_race_missing_total_n / `Drugs: Total searches`,

    `Drugs: Age missing total (%)` =
      drugs_age_missing_total_n / `Drugs: Total searches`,

    .groups = "drop"
  ) %>%
  select(
    Station,
    `Drugs: Total searches`,
    `Drugs: Race missing total (%)`,
    `Drugs: Age missing total (%)`
  )

# ---- Weapons -------------------------------------------------------------

weapons_station <- dat_metro %>%
  filter(!is.na(`Search.type - Weapons`)) %>%
  group_by(Station) %>%
  summarise(
    `Weapons: Total searches` = n(),

    weapons_race_missing_total_n =
      sum(is.na(Racial.appearance.transformed)) +
      sum(Racial.appearance.transformed %in% c("Other", "Unknown"), na.rm = TRUE),

    weapons_age_missing_total_n = sum(is.na(Age)),

    `Weapons: Race missing total (%)` =
      weapons_race_missing_total_n / `Weapons: Total searches`,

    `Weapons: Age missing total (%)` =
      weapons_age_missing_total_n / `Weapons: Total searches`,

    .groups = "drop"
  ) %>%
  select(
    Station,
    `Weapons: Total searches`,
    `Weapons: Race missing total (%)`,
    `Weapons: Age missing total (%)`
  )

# ---- Combine -------------------------------------------------------------

station_side_by_side <- full_join(
  drugs_station,
  weapons_station,
  by = "Station"
) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  arrange(Station)

DT::datatable(
  station_side_by_side,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Race and age missing totals by Station and search type (Metro only)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 10,      # <-- scroll 10 at a time
    lengthChange = FALSE,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel"),
    order = list(list(1, "desc"))  # default: sort by Drugs total
  )
)

```
## Racial categories mapped

The below table shows how the original recorded data maps to the dataset used in this analysis. Categories changed multiple times across the time period in scope.

**Table 8 Racial appearance mapping**
```{r mapping-categories-2022-2023, echo = FALSE}
ddat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```

