---
title: "Missing racial appearance and age in warrantless searches"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

```

# Missingness of data  

The following tables examine the **completeness of key demographic fields** for **weapon and drug searches arising from pedestrian and vehicle encounters conducted by uniformed police in metropolitan areas**. These searches represent a substantial proportion of frontline police interactions and form the basis for later analyses of search outcomes and hit rates.

Across the tables, results are presented as the **total number of relevant searches** alongside the proportion of records with **missing demographic information**. “Race missing total” combines records where racial appearance is not recorded (NA) with those recorded as “Unclassified/Otherrace”, "Undetermined" or “Unknown”, reflecting the absence of usable racial appearance information. “Age missing total” captures records with no recorded age.

These tables are intended to provide an overview of **data quality and consistency in demographic recording** across search contexts and organisational levels. They do not assess individual officer behaviour or search outcomes, but instead establish the reliability of key demographic fields used in subsequent analysis.

In metropolitan areas, the dataset records **25,230 drug searches** and **1,800 weapon searches**, with **235 searches involving both drug and weapon powers**. Drug searches were conducted under *Drugs, Poisons and Controlled Substances Act 1981 s82*, and weapon searches under *Control of Weapons Act 1990 s10*.


This table reports demographic missingness for **metropolitan weapon and drug searches**, including total searches and the proportion with **missing race** and **missing age** information. 
```{r metro-drugs-weapons-overlap, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(Area.type == "Metro",  
            
         !is.na(Station.uniform))

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

overlap_table <- tibble(
  Category = c(
    "Drug searches",
    "Weapon searches",
    "Drug and weapon searches"
  ),
  `Number of searches` = c(
    sum(is_applied(dat_metro$`Search.type - Drugs`)),
    sum(is_applied(dat_metro$`Search.type - Weapons`)),
    sum(
      is_applied(dat_metro$`Search.type - Drugs`) &
      is_applied(dat_metro$`Search.type - Weapons`)
    )
  )
)

DT::datatable(
  overlap_table,
  rownames = FALSE,
  options = list(
    dom = "t",
    autoWidth = TRUE
  )
)

```

```{r legislativepower-missingness-metro-drugs-weapons, echo = FALSE, warning = FALSE, message = FALSE}
dat_metro <- dat %>%
  filter(Area.type == "Metro",  
            
         !is.na(Station.uniform))

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

OTHER_SET <- c("UNCLASSIFIED/OTHER RACE", "UNDETERMINED", "UNKNOWN")

# long data so records that are both Drugs & Weapons count in both rows
dat_long <- bind_rows(
  dat_metro %>%
    filter(is_applied(`Search.type - Drugs`)) %>%
    mutate(search_type = "Drugs"),
  dat_metro %>%
    filter(is_applied(`Search.type - Weapons`)) %>%
    mutate(search_type = "Weapons")
) %>%
  mutate(
    race_orig = as.character(Racial.Appearance.original),
    race_norm = toupper(trimws(race_orig)),
    # treat blanks as NA (harmless even if none exist)
    race_norm = na_if(race_norm, "")
  )

search_type_missing <- dat_long %>%
  group_by(search_type) %>%
  summarise(
    total = n(),

    race_na_n = sum(is.na(race_norm)),
    race_other_unknown_n = sum(race_norm %in% OTHER_SET, na.rm = TRUE),

    race_missing_n = sum(is.na(race_norm) | race_norm %in% OTHER_SET, na.rm = TRUE),

    age_missing_n = sum(is.na(Age)),

    `Race NA (%)` = race_na_n / total,
    `Race Unknown (%)` = race_other_unknown_n / total,
    `Race missing total (%)` = race_missing_n / total,
    `Age missing total (%)` = age_missing_n / total,

    .groups = "drop"
  ) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  rename(`Search type` = search_type) %>%
  select(
    `Search type`,
    `Total searches` = total,
    `Race NA (%)`,
    `Race Unknown (%)`,
    `Race missing total (%)`,
    `Age missing total (%)`
  )

DT::datatable(
  search_type_missing,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Missing race and age by search type (Metro only; both drugs & weapons counted in both rows)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)



```

## Division level  

This table presents the completeness of demographic information for **weapon and drug searches arising from pedestrian and vehicle stops in metropolitan areas carried out by uniform police**, disaggregated by **division**. 

```{r missingness-metro-division-drugs-or-weapons, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(
    Area.type == "Metro",
    
    !is.na(Station.uniform)
  )

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

RACE_UNKNOWN_SET <- c("UNCLASSIFIED/OTHER RACE", "UNDETERMINED", "UNKNOWN")

dat_dw <- dat_metro %>%
  filter(is_applied(`Search.type - Drugs`) | is_applied(`Search.type - Weapons`)) %>%
  mutate(
    # normalise Division so grouping behaves
    Division = trimws(as.character(Division)),
    Division = dplyr::na_if(Division, ""),
    Division = if_else(is.na(Division), "Division missing", Division),

    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  )

division_missing_combined <- dat_dw %>%
  group_by(Division) %>%
  summarise(
    `Total searches` = n(),
    `Race missing total (%)` =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE) / n(),
    `Age missing total (%)` =
      sum(is.na(Age)) / n(),
    .groups = "drop"
  ) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  arrange(desc(`Race missing total (%)`))

DT::datatable(
  division_missing_combined,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Race and age missing percentages by Division (Metro, pedestrian, uniformed; drug OR weapon searches, counted once even if both)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)




```

This table presents the completeness of demographic information for **weapon and drug searches arising from pedestrian and vehicle stops in metropolitan areas carried out by uniform police**, disaggregated by **division** and **search type**.
```{r missingness-metro-division-drugs-vs-weapons, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(Area.type == "Metro",  
            
         !is.na(Station.uniform))

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

RACE_UNKNOWN_SET <- c(
  "UNCLASSIFIED/OTHER RACE",
  "UNDETERMINED",
  "UNKNOWN"
)

# ---- Drugs ---------------------------------------------------------------

drugs_div <- dat_metro %>%
  filter(is_applied(`Search.type - Drugs`)) %>%
  mutate(
    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  ) %>%
  group_by(Division) %>%
  summarise(
    `Drugs: Total searches` = n(),

    drugs_race_missing_total_n =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE),

    drugs_age_missing_total_n = sum(is.na(Age)),

    `Drugs: Race missing total (%)` =
      drugs_race_missing_total_n / `Drugs: Total searches`,

    `Drugs: Age missing total (%)` =
      drugs_age_missing_total_n / `Drugs: Total searches`,

    .groups = "drop"
  ) %>%
  select(
    Division,
    `Drugs: Total searches`,
    `Drugs: Race missing total (%)`,
    `Drugs: Age missing total (%)`
  )

# ---- Weapons -------------------------------------------------------------

weapons_div <- dat_metro %>%
  filter(is_applied(`Search.type - Weapons`)) %>%
  mutate(
    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  ) %>%
  group_by(Division) %>%
  summarise(
    `Weapons: Total searches` = n(),

    weapons_race_missing_total_n =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE),

    weapons_age_missing_total_n = sum(is.na(Age)),

    `Weapons: Race missing total (%)` =
      weapons_race_missing_total_n / `Weapons: Total searches`,

    `Weapons: Age missing total (%)` =
      weapons_age_missing_total_n / `Weapons: Total searches`,

    .groups = "drop"
  ) %>%
  select(
    Division,
    `Weapons: Total searches`,
    `Weapons: Race missing total (%)`,
    `Weapons: Age missing total (%)`
  )

# ---- Combine + format ----------------------------------------------------

division_missing_simple <- full_join(drugs_div, weapons_div, by = "Division") %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  arrange(desc(`Weapons: Race missing total (%)`))%>%
  select(Division,  `Drugs: Total searches`,`Weapons: Total searches`,
    `Drugs: Race missing total (%)`,
    
    `Weapons: Race missing total (%)`,
    `Drugs: Age missing total (%)`,
    `Weapons: Age missing total (%)`)%>%
  arrange(desc(`Drugs: Race missing total (%)`))

DT::datatable(
  division_missing_simple,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Race and age missing totals by Division and search type (Metro only)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 15,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)



```

## Station level  

This table presents the completeness of demographic information for **weapon and drug searches arising from pedestrian and vehicle stops in metropolitan areas carried out by uniform police**, disaggregated by **station**. To ensure data is meaningful, only stations with **20 or more weapon and/or drug searches** are included.

```{r missingness-metro-station-drugs-or-weapons, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(
    Area.type == "Metro",
       
    !is.na(Station.uniform)
  )

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

RACE_UNKNOWN_SET <- c(
  "UNCLASSIFIED/OTHER RACE",
  "UNDETERMINED",
  "UNKNOWN"
)

# Metro pedestrian + uniformed searches where Drugs OR Weapons applies (counted once even if both)
dat_dw <- dat_metro %>%
  filter(
    is_applied(`Search.type - Drugs`) | is_applied(`Search.type - Weapons`)
  ) %>%
  mutate(
    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  )

station_missing_combined <- dat_dw %>%
  group_by(Station.uniform) %>%
  summarise(
    `Total searches` = n(),

    `Race missing total (%)` =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE) / n(),

    `Age missing total (%)` =
      sum(is.na(Age)) / n(),

    .groups = "drop"
  ) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  arrange(Station.uniform)%>%
  filter(`Total searches`>=20)%>%
  arrange(desc(`Race missing total (%)`))


DT::datatable(
  station_missing_combined,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Race and age missing percentages by Station (Metro, pedestrian, uniformed; drug OR weapon searches, counted once even if both)"
  ),
  rownames = FALSE,
  options = list(
    pageLength = 10,   # ten stations per page
    autoWidth = TRUE,
    dom = "tip"
  )
)

```
This table highlights the **extent of missing demographic information** in **weapon searches conducted during pedestrian and vehicle encounters by uniformed police in metropolitan areas**, shown by **station**. It reports the proportion of weapon search records with **missing racial appearance information** (“Race missing total”) and **missing age information** (“Age missing total”).

To ensure data is meaningful, only stations with **20 or more weapon searches** are included. Percentages are calculated using station-level weapon search totals as the denominator. 

```{r missingness-metro-station-weapons-min25, echo = FALSE, warning = FALSE, message = FALSE}

dat_metro <- dat %>%
  filter(
    Area.type == "Metro",
       
    !is.na(Station.uniform)
  )

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

RACE_UNKNOWN_SET <- c(
  "UNCLASSIFIED/OTHER RACE",
  "UNDETERMINED",
  "UNKNOWN"
)

weapons_station <- dat_metro %>%
  filter(is_applied(`Search.type - Weapons`)) %>%
  mutate(
    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  ) %>%
  group_by(Station.uniform) %>%
  summarise(
    `Total searches` = n(),

    `Race missing total (%)` =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE) / n(),

    `Age missing total (%)` =
      sum(is.na(Age)) / n(),

    .groups = "drop"
  ) %>%
  filter(`Total searches` >= 20) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  rename(Station = Station.uniform) %>%
  arrange(desc(`Total searches`), Station)%>%
  arrange(desc(`Race missing total (%)`))

DT::datatable(
  weapons_station,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Weapon searches: race and age missing percentages by station (Metro,  uniformed; stations shown only where ≥25 weapon searches)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)

```

This table highlights the **extent of missing demographic information** in **drug searches conducted during pedestrian and vehicle encounters by uniformed police in metropolitan areas**, shown by **station**. It reports the proportion of drug search records with **missing racial appearance information** (“Race missing total”) and **missing age information** (“Age missing total”).

To ensure data is meaningful, only stations with **20 or more drug searches** are included. Percentages are calculated using station-level drug search totals as the denominator. 

```{r missingness-metro-searchtype-min25-combined, echo = FALSE, warning = FALSE, message = FALSE}


dat_metro <- dat %>%
  filter(
    Area.type == "Metro",
       
    !is.na(Station.uniform)
  )

is_applied <- function(x) {
  !is.na(x) & !(x %in% c(0, "0", FALSE, "", "False", "FALSE", "No", "NO"))
}

RACE_UNKNOWN_SET <- c("UNCLASSIFIED/OTHER RACE", "UNDETERMINED", "UNKNOWN")

drugs_station <- dat_metro %>%
  filter(is_applied(`Search.type - Drugs`)) %>%
  mutate(
    race_norm = toupper(trimws(as.character(`Racial.Appearance.original`))),
    race_norm = dplyr::na_if(race_norm, "")
  ) %>%
  group_by(Station.uniform) %>%
  summarise(
    `Station` = first(Station.uniform),
    `Total searches` = n(),
    `Race missing total (%)` =
      sum(is.na(race_norm) | race_norm %in% RACE_UNKNOWN_SET, na.rm = TRUE) / n(),
    `Age missing total (%)` =
      sum(is.na(Age)) / n(),
    .groups = "drop"
  ) %>%
  filter(`Total searches` >= 20) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  select(Station, `Total searches`, `Race missing total (%)`, `Age missing total (%)`) %>%
  arrange(desc( `Race missing total (%)`))

DT::datatable(
  drugs_station,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Drug searches: race and age missing percentages by station (Metro, uniformed; stations shown only where ≥25 drug searches)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)



```
## Appendix - Missingness all unit types  

For completeness, this appendix presents missingness in key demographic fields for **all drug and weapon searches across all police unit types across Victoria**, including searches beyond uniformed pedestrian encounters.


```{r legislativepower-missingness-unit.type, echo = FALSE, warning = FALSE, message = FALSE}


OTHER_SET <- c("UNCLASSIFIED/OTHER RACE", "UNDETERMINED", "UNKNOWN")

unit_missing <- dat %>%
  mutate(
    race_orig = as.character(`Racial.Appearance.original`),
    race_norm = toupper(trimws(race_orig)),
    race_norm = dplyr::na_if(race_norm, "")
  ) %>%
  group_by(Unit.type) %>%
  summarise(
    `Total searches` = n(),

    `Race NA (%)` =
      sum(is.na(race_norm)) / n(),

    `Race Unknown (%)` =
      sum(race_norm %in% OTHER_SET, na.rm = TRUE) / n(),

    `Race missing total (%)` =
      sum(is.na(race_norm) | race_norm %in% OTHER_SET, na.rm = TRUE) / n(),

    `Age missing total (%)` =
      sum(is.na(Age)) / n(),

    .groups = "drop"
  ) %>%
  mutate(across(ends_with("(%)"), ~ scales::percent(.x, accuracy = 0.1))) %>%
  arrange(desc(`Total searches`), Unit.type)

DT::datatable(
  unit_missing,
  caption = htmltools::tags$caption(
    style = "caption-side: bottom; text-align: left;",
    "Missing race and age by Unit type (all searches)"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel")
  )
)



```
