---
title: "Victoria Police Search Data Overview and Hit Rates"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

```

# Short summary

This overview was carried out with the aim of understanding how Victoria Police stop-and-search data can be examined for racial profiling using hit-rate analysis—the share of searches that result in a find. The goal is twofold: (1) to assess whether the dataset is suitable for credible analysis of potential racial disparities in policing thresholds, and (2) to document its structure so others can use it appropriately.  

The dataset covers 52,654 warrantless public-place searches across 2018, 2019, 2022, and 2023, mostly by uniformed officers, with drug searches the dominant search type. Descriptive results highlight clear differences in outcomes:  

- **By racial appearance**: people perceived as White recorded higher hit rates (17.8%) than many racialised groups, including African (12.9%) and Middle Eastern (10%) individuals.  
- **By context**: PSO, Transit, and Highway Patrol units had higher hit rates than Uniform police; graffiti and weapons searches were more successful than drug searches.  
- **By geography**: the North West Metro Region, covering Melbourne’s diverse western and northern suburbs, showed significantly lower hit rates than other regions.  
- **By demographics**: males had higher hit rates than females, and teenagers (≤17) had lower hit rates than adults.  

These patterns suggest that disparities in search outcomes are real, but raw hit-rate comparisons cannot alone establish whether police apply different evidentiary thresholds across groups. To move beyond description, the analysis must be narrowed to a standardised setting (e.g. drug searches by uniformed officers on foot patrols) and then modelled to account for confounders such as gender, year, and geography.  

# About the Dataset


📁 **Download the cleaned dataset:**  
[**VicPol Search Data Clean.xlsx**](https://github.com/monikasarder/VictoriaPoliceRacialSearch/raw/main/VicPol%20Search%20Data%20Clean/VicPol%20Search%20Data%20Clean.xlsx)


This dataset contains records of **52,654 searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**. 

Data for the years **2020 and 2021** is excluded due to the atypical policing environment associated with the COVID-19 pandemic.

The dataset comprises records of **warrantless stop and search events** conducted in public places across Victoria. It was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and includes detailed information about each search, including location, officer-perceived characteristics of the person searched, search type, and outcome.

For a warrantless search to be lawful, police must **suspect on reasonable grounds** that the person is carrying a prohibited item, such as drugs, weapons, or graffiti implements.

Each search is recorded under the relevant **legal power** (e.g. *Drugs, Poisons and Controlled Substances Act 1981*, s82) and includes one of three possible **outcomes**:

- **Nothing found**  
- **Search item found**: an item matching the type of contraband suspected (e.g. drugs found during a drug search)  
- **Non-search item found**: a different type of contraband was found (e.g. a weapon found during a drug search)

This dataset has been cleaned and transformed to enable analysis of find rates - more commonly described as 'hit rates' - by unit type and geographical area, where this information is available. Racial categories have also been cleaned and transformed to allow meaningful comparison, as the naming conventions frequently changed in the six years from 2018 to 2023.

## Variables

The following variables are available in this dataset. 

- `Date`: Date of the stop and search  
- `Area.type`: Metropolitan or regional classification  
- `Region`, `Division`, `PSA`, `Station`: Police geographic hierarchy  
- `Racial.appearance.original`: The original officer-perceived racial appearance category as recorded by Victoria Police  
- `Racial.appearance.transformed`: Racial appearance category transformed to enable comparison, adjusting for multiple category name changes across the period 
- `Gender`: Officer-perceived gender of the person searched  
- `Age`: Age of the person searched  
- `Age group`: Age of the person searched  
- `Search.type - Drugs`: Search conducted under *Drugs, Poisons and Controlled Substances Act 1981 s82* — can be carried out anywhere in public with reasonable suspicion of possession of a controlled drug.  
- `Search.type - Weapons`: Search conducted under *Control of Weapons Act 1990 s10* — can be carried out anywhere in public with reasonable suspicion of possession of a controlled weapon, dangerous article, or prohibited weapon.  
- `Search.type - Graffiti`: Search conducted under *Graffiti Prevention Act 2007 s13* — only in specified circumstances (e.g. person aged 14+ in a prescribed area such as public transport property or trespassing). 
- `Search.type - Firearms`: Search conducted under *Firearms Act 1996 s112Q* — only in *specified circumstances* (person subject to a Firearm Prohibition Order).  
- `Search.type - Volatile.sub.U18`: Search conducted under *Drugs, Poisons and Controlled Substances Act 1981 ss60E–60F* — only in *specified circumstances* (person under 18 suspected of possessing/inhaling volatile substances, or supplying them to a child).  
- `Search.type - Volatile.sub.adult`: Search conducted under the same provisions as above — applies to adults only when suspected of supplying volatile substances to a child.  
- `Found`: Whether any prohibited item was found — regardless of whether it was the item searched for under the power invoked (e.g. both “drug power invoked, drugs found” and “drug power invoked, weapon found” count as a 'hit').  
- `Contact.Type`: Whether the search was conducted on foot (`P`) or during a vehicle stop (`V`).  
- `Unit.type`: Broad category of the police unit conducting the search (e.g. uniform, highway patrol).  
- `Station.uniform`: The police station linked to the uniformed officer conducting the search (available for uniform police only).  
- `Rank`: Rank of the officer responsible for conducting the search (e.g. Constable, Senior Constable, Sergeant).  

Note that multiple powers can apply in a single search - for example where police have reasonable grounds for suspecting both weapons and drugs are present.

## Data limitations  

### Racial Appearance Categories

Victoria Police records the **perceived racial appearance** of individuals at the time of the search. These categories reflect how the officer perceives the person’s racial appearance — they are *not* based on self-identification or ancestry and should be understood as indicators of how race is *perceived* in the policing context.

In addition, **2,097 records** are classified under the merged **‘Mediterranean/Middle Eastern’** category. Since at least 2022, Victoria Police has combined these two appearance labels, limiting the ability to meaningfully assess outcomes for people of Middle Eastern appearance. While individuals of Mediterranean descent are often perceived as non-racialised or White, those of Middle Eastern appearance may be racialised differently in police interactions. The merger obscures this distinction, rendering these records **unusable for racial disparity analysis** involving Middle Eastern individuals.

### Missing Data

A substantial proportion of records are missing demographic information about the person who was searched.

In particular, **racial appearance information is either missing or undetermined in a significant number of cases**. Specifically, **42% of all searches (22,117 records)** have no recorded racial appearance — despite this being a mandatory field since 2019. An additional **1.5% of searches (785 records)** are recorded as **‘Other’**, a category sometimes used when officers are unable or unwilling to assign a racial appearance. Together, these issues raise concerns about data quality and the consistency with which demographic details are recorded.

# Data Overview  

##  Search types  

Table 1 below summarises the total search powers invoked during searches. A single search could involve the application of multiple powers. For example - a person in a vehicle could be searched for both weapons and drugs.  The power most often relied upon by police when undertaking the warrantless searches recorded in this dataset is section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*.

**Table 1 All search powers that applied in searches by search count**
```{r searched for, echo = FALSE, warning = FALSE}

finds <- dat %>%
  select(starts_with("Search.type"))%>%
  pivot_longer(
    cols = starts_with("Search.type"),
    names_to = "Search type",
    values_to = "Total",
    values_drop_na = TRUE
  ) %>%
  mutate(`Search type` = forcats::fct_infreq(`Search type`))%>%
  select(-Total)


 
print(dfSummary(finds, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Persons searched 

This section summarises the characteristics of individuals searched in 2018, 2019, 2022, and 2023. Note that some individuals were searched on multiple occasions.

Most searches (67%) involved people perceived by police as White. People perceived as Asian or African each accounted for more than 5% of searches. The second-largest category was “Mediterranean/Middle Eastern”; however, this category cannot be reliably used for racial profiling analysis, as it combines both White-appearing and non-White-appearing individuals.

Men accounted for more than four-fifths of those searched. Over half of all searches involved people aged between 18 and 35 years.

Searches were more likely to occur while the person was on foot (66%) than during a vehicle stop.

**Table 2 Characteristics of persons searched**
```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group", "Contact.Type")


dat$`Age group`[dat$`Age group` == "Missing"] <- NA  

dat$Racial.appearance.transformed[dat$Racial.appearance.transformed == "Missing"] <- NA  

dat.per <- dat%>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.per <- dat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(dat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police undertaking searches 

This section summarises the characteristics of police members and units recorded as conducting searches in 2018, 2019, 2022, and 2023.  

Over the four years examined, search numbers were relatively consistent, with a slight peak in 2022.

Most searches were conducted by **Uniform police** (72%). **Transit police** were responsible for **7%** whilst **Public Order Response Team (PORT)** members - a specialist Victoria Police unit trained and equipped for large-scale public order operations such as protests and rallies - carried out **6%** of all searches.

**Constables** and **Senior Constables** were the ranks most frequently involved, reflecting the predominance of uniformed officers in search activity.


**Table 3 Police and units undertaking searches**
```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Unit.type","Rank.of.Member","Area.type", "Region", "Year")

dat.pol <- dat %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```


# Hit Rate Patterns

This section summarises whether the likelihood of a “hit” (i.e. discovery of a prohibited item). It should be noted that there are two types of finds - finds of the prohibited item searched for and finds of any prohibited item. For the majority of this summary we will count *any* prohibited item find as successful. 

##  Hit rates by search type 

Table 4 shows clear variation in hit rates across search types. Searches for **graffiti items in specified places** have the highest success rate (**50.1%**). **Firearms searches of people subject to a Firearms Prohibition Order** are also relatively successful (**22%**). Among warrantless searches that can be conducted on any person in public places, **weapons-related searches** have a higher hit rate than **drug-related searches** (**33% vs 15%**). Hit rates for **volatile substances** (often referred to as “huffing”) are not reported due to a small sample size (**n < 50**).


**Table 4 Hit rates by search type**
```{r Finds search type, echo=FALSE, warning=FALSE, results='asis'}

finds <- dat %>%
  select(Found, starts_with("Search.type")) %>%
  pivot_longer(
    cols = starts_with("Search.type"),
    names_to = "Search type",
    values_to = "Detailed find",
    values_drop_na = TRUE
  )%>%
  mutate(
    # <-- Ensure Yes before No everywhere downstream
    Found = fct_relevel(as.factor(Found), "Yes", "No")
  )

# 1) Stabilise categories (optional but recommended)
finds2 <- finds %>%
  filter(!is.na(Found)) %>%
  mutate(
    `Search type` = fct_lump_min(`Search type`, min = 50, other_level = "Other"),
    Found = fct_drop(Found)
  )

# 2) Build contingency table and run a safe test
tbl <- with(finds2, table(`Search type`, Found))
tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]

if (all(dim(tbl) == c(2, 2))) {
  test_obj <- fisher.test(tbl, workspace = 2e8)
  test_label <- "Fisher's exact"
} else {
  test_obj <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
  test_label <- "Chi-square (simulated p)"
}

pval_str <- format.pval(test_obj$p.value, digits = 4, eps = .Machine$double.eps)

# 3) Make the crosstable (no internal test) and append the p-value
ct <- crosstable(
  finds2, `Search type`,
  by = Found,
  total = TRUE,
  test = FALSE,                # <- prevent Fisher blow-ups
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct) |>
  add_footer_lines(values = paste0("Overall association: ", test_label, ", p = ", pval_str))

ft

```

## Hit rates by person searched   

Hit rates vary across racial appearance, gender, and age. Compared with people perceived as **White** (**18%**), most other groups had lower hit rates—for example, **Middle Eastern** (**10%**) and **African** (**12.9%**). The only group not following this pattern is **Pacific Islander**, but this category is small (**n = 776**) and should be interpreted with caution.

By gender, **males** recorded higher hit rates than **females** (**17% vs 14%**).

By age, **teenagers (≤17)** generally had lower hit rates than adults.

*Interpretation:* Lower hit rates mean police were less likely to find contraband when searching those groups, which can indicate a lower threshold for initiating searches for those populations.


**Table 5 Hit rates by person characteristics**

```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}

# Recode and reorder factors
dat2 <- dat %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),  # Yes before No
    Year  = fct_relevel(as.character(Year), "2018", "2019", "2022", "2023"),
    `Age group` = fct_relevel(`Age group`,
                              "Under 12", "12 to 17", "18 to 25",
                              "26 to 35", "36 to 45", "46 and over") |> fct_drop(),
    # Use freq order for the others ONLY
    across(
      .cols = all_of(c("Racial.appearance.transformed", "Gender", "Age group")),
      .fns  = ~ fct_infreq(.) |> fct_drop()
    )
  )

# Variables to tabulate (exclude 'Found' here; it's the 'by' variable)
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Contact.Type")

# Build crosstable WITHOUT internal tests (prevents Fisher blow-ups)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2, else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```

## Hit rates by police member and operating context  

**By member type.** Hit rates were higher for **Protective Services Officers (PSOs)** than for other members. This likely reflects deployment: PSOs concentrate at train stations and nearby precincts—often in the evening—where offending (and therefore the base rate of finds) is higher. Across **ranks**, hit rates were otherwise broadly similar.

**By unit.** **Highway Patrol**, **PSO units**, and **Transit** recorded hit rates in the **25–35%** range, again consistent with searches concentrated around travel corridors and transport hubs. **Uniform police**—who conduct the largest share of searches—had a hit rate of **~17%**.

**By area.** Overall hit rates were similar in **metropolitan vs regional** areas. However, the **North West Metro Region** (covering Melbourne’s northern and western suburbs) had a **significantly lower** hit rate than other regions. This region includes several culturally diverse and economically marginalised LGAs such as **Brimbank** and **Melbourne**.

**Over time.** Hit rates varied only **modestly** across 2018, 2019, 2022, and 2023.

**Table 6 Hit rates by police member and operating context**
```{r Finds pol, echo=FALSE, warning=FALSE, results='asis'}


# Recode and reorder factors
dat2 <- dat %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),  # Yes before No
    Year  = fct_relevel(as.character(Year), "2018", "2019", "2022", "2023"),
    # Frequency order for operating-context variables
    across(
      .cols = all_of(c("Unit.type", "Rank.of.Member", "Area.type", "Region")),
      .fns  = ~ as_factor(.) |> fct_infreq() |> fct_drop()
    )
  )

# Variables to tabulate (exclude 'Found' here; it's the 'by' variable)
vars_for_table <- c("Unit.type", "Rank.of.Member", "Area.type", "Region", "Year")

# Build crosstable WITHOUT internal tests (prevents Fisher blow-ups)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2, else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft


```


# Conclusion


The descriptive results highlight both the promise and the limits of raw hit-rate comparisons. Clear disparities appear across appearance groups and geographic divisions, but unadjusted figures cannot fully resolve whether these patterns reflect differences in policing thresholds or in underlying population composition.  

To move beyond description, we next narrow the scope to a more standardised setting — drug searches by uniformed officers on foot patrols. This setting captures the bulk of everyday encounters and removes confounding variation in search type and policing unit. Within that restricted frame, we can apply regression models to test whether racial disparities in hit rates remain once gender, year, and geography are accounted for.  

In this way, the descriptive overview provides the baseline picture, while the next section introduces the analytic tools needed to determine whether the disparities observed are localised anomalies or evidence of a broader, systematic pattern in Victoria Police stop-and-search practices.




---

[^data]: Dataset size, years, and warrantless scope; “any item found” definition for hit rate. :contentReference[oaicite:0]{index=0} :contentReference[oaicite:1]{index=1} :contentReference[oaicite:2]{index=2}

[^types]: Hit rates by search type (graffiti 50.1%; firearms (FPO) 22%; weapons 33% vs drugs 15%). :contentReference[oaicite:3]{index=3}

[^unit]: Higher PSO hit rates; unit band 25–35%; Uniform ~17%. :contentReference[oaicite:4]{index=4}

[^area]: Metro vs regional similar overall; **NWMR** significantly lower; NWMR includes diverse/marginalised LGAs (Brimbank, Melbourne). :contentReference[oaicite:5]{index=5}

[^person]: Racial, gender, and age hit-rate differences (White 18%, Middle Eastern 10%, African 12.9%; males 17% vs females 14%; ≤17 lower). :contentReference[oaicite:6]{index=6}

[^missing]: Missing racial appearance (**42%**) and **1.5% ‘Other’**. :contentReference[oaicite:7]{index=7}


# Appendix – Ethnic Appearance Mapping

```{r mapping-categories-2022-2023, echo = FALSE}
dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```

