---
title: "Racial profiling - drug searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(scales)
library(stringr)
library(tidyr)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

```

```{r hitrate-prop-helpers, echo=FALSE, message=FALSE, warning=FALSE}


is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}


make_hit_table <- function(
  data,
  group_var,                   # e.g. "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    dplyr::filter(!is.na(!!grp), !is.na(Found)) %>%
    dplyr::filter(Racial.appearance.transformed %in% c("African","White")) %>%
    dplyr::mutate(FoundYes = is_yes(Found)) %>%
    dplyr::count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    dplyr::group_by(!!grp, Racial.appearance.transformed) %>%
    dplyr::summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    dplyr::filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble::tibble())

  out <- tab %>%
    dplyr::mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    dplyr::select(-test) %>%
    tidyr::unnest(tidied) %>%
    dplyr::mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = dplyr::if_else(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    # Important: use !!rlang::sym(group_var) on LHS, and avoid Unicode minus in names
    dplyr::transmute(
      !!rlang::sym(group_var) := !!grp,
      `African (succ/total)`   = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`     = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`       = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`         = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr - Wh)`        = scales::percent(diff_est,         accuracy = 0.1),  # <-- ASCII hyphen
      `Risk ratio (Afr/Wh)`    = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                      = sprintf("%.3f", p.value),
      `p (BH)`                 = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                    = stars
    )

  if (filter_sig_only) {
    out <- out %>% dplyr::filter(as.numeric(stringr::str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% dplyr::arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))
}

print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) display <- display %>% dplyr::select(-`p (BH)`)

  display %>%
    knitr::kable(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    kableExtra::column_spec(1, width = "24em") %>%
    kableExtra::add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}

```

# Short summary

This analysis of 7,534 warrantless drug searches carried out by **uniform officers on foot patrol in metropolitan Melbourne (2018, 2019, 2022, 2023)** provides clear evidence that individuals perceived as **African** are searched on a **lower evidentiary threshold** compared to those perceived as White.

Key findings include:

-   **Descriptive results** showed lower hit rates for African-appearing individuals across most contexts.\
-   **Proportion tests** identified particularly strong disparities in Brimbank, Melton, and Werribee.\
-   **Division level models** controlling for gender, as well operating context including year and Local Government Area (LGA) revealed disparities in\
-   **Local government area level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in
-   **Statio level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in

These results are consistent with **racial profiling**: African-appearing individuals are more frequently subjected to searches that are less likely to yield contraband, implying that suspicion is applied more readily to them.

The magnitude of disparity varies locally, the overall pattern is systematic.

This highlights the need for greater police accountability and independent oversight, and provides an empirical foundation for community advocacy and policy reform aimed at fairer, more equitable policing in Melbourne.

### Melbourne-wide findings

-   Hit rates: White 14.1% (957/6,779) vs African 11.4% (86/755); Fisher’s exact p = 0.040.
-   Pooled logistic regression (controls: Division fixed effects, Year, Gender): OR = 0.771 (95% CI 0.596–0.985), p = 0.042 — African-appearing males had \~23% lower odds of a find than White-appearing males.
-   Multilevel logistic regression (random intercepts for Division): OR = 0.761 (95% CI 0.594–0.975), p = 0.031 — confirms the disparity after accounting for clustering by Division.

### Local hotspots

-   NWMR Division 3 (Brimbank & Melton): African 4.3% (3/70) vs White 14.2% (30/212); risk ratio = 0.30, p = 0.026.
-   Werribee Station: African 8.8% (5/57) vs White 21.4% (22/103); risk ratio = 0.41, p = 0.042.

### Limits of statistical significance

-   Most other divisions and stations did not reach statistical significance, primarily because of small African sample sizes (often only a few dozen searches). Wide confidence intervals and multiple-testing adjustments mean that many apparent disparities cannot be declared significant.
-   Non-significance does not equal absence of disparity — rather, it reflects the limited statistical power in areas with few recorded searches of African-appearing individuals.

### Interpretation

Across Melbourne, African-appearing people are consistently subjected to searches with lower hit rates and lower odds of contraband being found, even after adjusting for year and division. This pattern is strongest in Melbourne’s outer west (Brimbank, Melton, Werribee), but persists at a city-wide level. Together, these results are consistent with racial profiling: African-appearing individuals are more frequently searched on a weaker threshold of suspicion.

# Introduction

This analysis examines **7,534 drug searches conducted by uniformed officers on foot patrols in Melbourne, of white- and African-appearing people**. These encounters represent the most routine form of warrantless public-place searches, where outcomes are highly dependent on officer discretion. The racial appearance categories in scope represent two of the largest and consistently recorded racial groupings in the dataset.

The focus is on whether search outcomes suggest evidence of **racial profiling**. Specifically, we compare the **hit rates** — the proportion of searches that result in contraband being found — for two groups:

-   **White-appearing individuals**, who form the majority baseline\
-   **African-appearing individuals**, a racialised minority consistently highlighted in community concerns about disproportionate policing

Concerns about the over-policing of African-Australian communities are not new. People such as **Haile Michael**, who successfully challenged Victoria Police’s discriminatory stop-and-search practices in the landmark [**Haile Michael**](https://www.abc.net.au/news/2013-03-18/victoria-police-accused-of-racially-profiling-african-men/4578358) case, have documented and contested these patterns for more than a decade.

By restricting the dataset to this context and to these two appearance categories, we provide a clear test of whether African-appearing people are subjected to searches on a **lower evidentiary threshold**. Lower hit rates for one group, despite comparable or higher search intensity, would suggest that police suspicion is being applied unevenly — a hallmark of racial profiling.

# About the dataset (subset definition)

This dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**.

Data from **2020 and 2021** is excluded due to the atypical policing environment during the COVID-19 pandemic.

The dataset was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and covers public-place searches carried out under a range of legal powers.

For this analysis, we restrict to:

-   **Geography:** Melbourne metropolitan area\
-   **Contact type:** Foot patrol only\
-   **Unit type:** Uniform police only\
-   **Search type:** Drug searches conducted under s82 of the *Drugs, Poisons and Controlled Substances Act 1981*\
-   **Station linkage:** Only searches associated with a uniform station\
-   **Racial appearance:** White and African appearance

This restriction enables us to compare searches for the most common drug searches experienced by the public in Melbourne, and restrict the data to reliable racial appearance categories.

## What s82 allows

Under section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.\
This is the most frequently invoked search power in the dataset and represents a context where discretionary judgments by officers play a central role.

## Variables in this subset

The following variables are included in the analysis:

-   **Year** — year of the stop and search\
-   **Region / Division / PSA / Station** — police geographic hierarchy\
-   **Racial.appearance.transformed** — officer-perceived racial appearance (cleaned categories)\
-   **Gender** — officer-perceived gender of the person searched\
-   **Found** — whether any contraband was located (hit outcome)\
-   **Unit.type** — uniform officers only (subset restriction)\
-   **Rank** — officer rank (e.g. Constable, Senior Constable)\
-   **Year** — 2018, 2019, 2022, or 2023

Derived variables include:

-   **Found.Flag** — binary indicator (1 if contraband found, 0 otherwise)\
-   **Race** — restricted to White vs African

------------------------------------------------------------------------

## Data limitations

### Racial appearance categories

The original dataset for drug searches conducted by uniform officers on foot patrol numbered 13,896. Race data was missing from *20% of the records*. Several categories had very small numbers or had been recategorised over the observation period. As a result, disparity testing is restricted to **White vs African** searches, which are sufficiently large and analytically comparable. Together these racial categories account for over half (54%) of the relevant data.

```{r race prop, echo=FALSE, message=FALSE, warning=FALSE}
# Count + percent for all groups (no filtering)
tab_ra <- ddat %>%
  mutate(
    Racial.appearance.transformed =
      fct_explicit_na(Racial.appearance.transformed, "Missing")
  ) %>%
  count(Racial.appearance.transformed, name = "Count") %>%
  arrange(desc(Count)) %>%
  mutate(Percent = Count / sum(Count))

# Add total row
tab_ra_tot <- bind_rows(
  tab_ra,
  tibble(
    Racial.appearance.transformed = "Total",
    Count = sum(tab_ra$Count),
    Percent = 1
  )
)

# Format percent for display
tab_ra_tot <- tab_ra_tot %>%
  mutate(Percent = percent(Percent, accuracy = 0.1))

# Render table
kbl(
  tab_ra_tot,
  caption = "Racial appearance (all groups): count and percentage of searches",
  align = c("l", "r", "r"),
  col.names = c("Racial appearance (transformed)", "Count", "Percent")
) %>%
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Searches" = 2))

```

### Age data

Age data was missing for 42.9% of searches in the White vs African subset. Age is therefore excluded from modelling, which limits the ability to adjust for age differences across groups. We expect this would have an effect as African migrant populations in in local government areas tend to be younger than their white-appearing counterparts.

### Scope of coverage

-   The dataset covers only **warrantless stop and search events conducted in public places**. This focus is deliberate: such encounters are entirely reliant on officer discretion, since there is no warrant, prior intelligence, or judicial approval.\
-   The analysis does not include searches with warrants, searches in private settings, or informal/unrecorded encounters.\
-   Findings therefore apply only to **Melbourne metropolitan drug searches by uniform officers on foot patrol**. They do not generalise to specialist units, vehicle stops, or regional policing contexts.\
-   Because modelling is restricted to White vs African, results do not extend to other racialised groups with smaller or analytically unreliable sample sizes.

# Data Overview

## Persons searched

This section summarises the characteristics of people searched in warrantless public-place encounters by uniform officers on foot patrols (2018, 2019, 2022, 2023).\
Records with missing or unusable racial appearance are excluded to allow meaningful comparison.

-   **Racial appearance.** Most searches involved people perceived as White, with African-appearance individuals making up around one in ten of the analytic sample.\
-   **Gender.** Searches were concentrated on men, who accounted for about four-fifths of those stopped.\
-   **Age.** While children and adolescents appear in the dataset, most searches targeted adults aged 18 to 45. People aged 46 and over were searched relatively infrequently.

These patterns highlight that discretionary stop-and-search practices fall unevenly across the community, with young men and racialised groups particularly over-represented.

**Table 2 Characteristics of persons searched**

```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddatr <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(ddat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police undertaking searches

Search numbers were relatively consistent across the four years examined, with a slight peak in 2022. By rank, Constables and Senior Constables were the most frequently involved, reflecting the predominance of frontline officers in this type of activity.

**Table 3 Police and units undertaking searches**

```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Year", "Rank.of.Member", "Region",  "Division")

ddatr <- ddatr %>%
  mutate(
    Rank = case_when(
      Rank.of.Member == "SSGT" ~ "SGT",
      Rank.of.Member == "PSO"  ~ "CONST",
      TRUE           ~ Rank.of.Member
    )
  )

dat.pol <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')


```

# Hit Rate - cross tables

An important way to assess stop-and-search practices is through **hit rates** — the proportion of searches that result in contraband being found.

If searches are consistently based on *reasonable suspicion* in comparable spatial areas, hit rates should be similar across groups.

Where a group has consistently lower hit rates for a particular area, this may suggest that officers may be applying a **lower threshold of suspicion**, a core indicator of potential racial profiling.

This section examines hit rates in the restricted dataset of **7,534 Melbourne metropolitan drug searches by uniform police on foot patrol** (2018, 2019, 2022, 2023).

A “hit” is defined as the discovery of any prohibited item, not just the item specified under the search power.

## Hit rates - cross tables

An initial analysis of comparative hit rates across our sample of 7,453 searches was performed, looking both at:

-   whether hit rates varied by racial appearance to a degree that is statistically significant and

-   Whether hit rates varied by other factors (such as local government area, gender, etc) to a degree that may counfound our results.

-   **By racial appearance.**

    -   White-appearing people had a hit rate of **14.1%** (957 finds out of 6,779 searches).\
    -   African-appearing people had a lower hit rate of **11.4%** (86/755).\
    -   This indicates that African-appearance individuals were **less likely to be found with contraband**, despite being searched in substantial numbers.

-   **By gender.**

    -   Men had a hit rate of **14.1%** (837/5,956), compared with **13.1%** for women (206/1,578).\
    -   Searches were heavily concentrated on men, accounting for four in five cases.

-   **By age.**

    -   Differences in hit rates across ages were marginal, however there does appear to be a slight decrease in hit rates where older people are searched.

-   **By time.** Across 2018, 2019, 2022, and 2023, hit rates varied only modestly.

-   **By rank.** Constables and Senior Constables dominated search activity, and their hit rates were broadly similar.

-   **By geography.** Hit rates were lowest in the **North West Metro Region**, which includes diverse and disadvantaged areas such as Brimbank and Melbourne, compared with other parts of the city.

**Table 5 Hit rates by person characteristics**

```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}




# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddatr %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),                
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop(),
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Rank = dplyr::recode(Rank.of.Member,
                         "SSGT" = "SGT",
                         "PSO"  = "CONST",
                         .default = Rank.of.Member)
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Year","Rank.of.Member", "Division")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```

**Interpretation.**\
These preliminary results show that both racial appearance and where a search occurs have a statistically significant effect on hit rates.

## Hit rates and time

We determined to look at whether the disparity in hit rates persisted over the years in our data - 2018, 2019, 2022 and 2023.

```{r race trend, echo=FALSE, message=FALSE, warning=FALSE}

# --- Filter to White & African and prep ---
ddatm <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African")) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1L, 0L),
    Race = factor(Racial.appearance.transformed, levels = c("White","African")),
    Year = suppressWarnings(as.integer(as.character(Year)))
  ) %>%
  filter(!is.na(Year))

# --- Yearly hit rates ---
ddat_rate <- ddatm %>%
  group_by(Race, Year) %>%
  summarise(n = n(),
            hits = sum(Found.Flag, na.rm = TRUE),
            hit_rate = hits / n,
            .groups = "drop")

# Build gap metadata
ddat_rate2 <- ddat_rate %>%
  arrange(Race, Year) %>%
  group_by(Race) %>%
  mutate(
    next_year = lead(Year),
    year_gap  = next_year - Year,
    next_hit  = lead(hit_rate),
    # run id: increments when the year jump is not 1
    runid = cumsum(dplyr::lag(Year, default = first(Year)) != Year - 1)
  ) %>%
  ungroup()

# Keep only runs that have at least 2 points (so a line can be drawn)
solid_data <- ddat_rate2 %>%
  group_by(Race, runid) %>%
  filter(n() >= 2) %>%
  ungroup()

# Dotted gaps (e.g., 2019 -> 2022)
gap_data <- ddat_rate2 %>%
  filter(!is.na(next_year), year_gap > 1)

x_breaks <- seq(min(ddat_rate2$Year), max(ddat_rate2$Year), by = 1)

ggplot(ddat_rate2, aes(x = Year, y = hit_rate, colour = Race)) +
  # solid lines for contiguous sequences (2018–2019 and 2022–2023)
  geom_line(
    data = solid_data,
    aes(group = interaction(Race, runid)),
    linewidth = 1
  ) +
  # dotted neutral bridge across missing years (2019 → 2022)
  geom_segment(
    data = gap_data,
    aes(x = Year, y = hit_rate, xend = next_year, yend = next_hit),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    colour = "grey50"
  ) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +  # no fixed limits
  scale_x_continuous(breaks = x_breaks) +
  labs(
    title = "Drug Search Hit Rate by Year (Metro)",
    subtitle = "White vs African. Dotted grey line spans missing years (2020–2021).",
    x = "Year",
    y = "Hit rate (percent of searches with a find)",
    colour = "Race"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

Hit rates were persistently higher for white-appearing versus African-appearing people searched across all years, however this disparity did vary from year to year, with the most recent observations showing a small disparity.

```{r race trend year, echo=FALSE, message=FALSE, warning=FALSE}
# Summarise into wide format: one column per race
hit_table <- ddat_rate %>%
  select(Race, Year, hit_rate) %>%
  pivot_wider(names_from = Race, values_from = hit_rate) %>%
  mutate(
    Difference = (`African` - `White`) * 100,        # difference in percentage points
    across(c(`White`, `African`), ~ round(.x * 100, 2)),  # convert to percent
    Difference = round(Difference, 2)
  ) %>%
  rename(
    "Year" = Year,
    "White (%)" = `White`,
    "African (%)" = `African`,
    "Difference (African - White, pp)" = Difference
  )

# Pretty table for RMarkdown
kbl(hit_table, caption = "Hit Rate by Year and Race (Metro)") |>
  kable_paper(full_width = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "left"
  )

```


# Modelling racial disparities in search outcomes

To assess whether the lower hit rates observed for African-appearance individuals persist once broader contextual factors are accounted for, we estimated a sequence of multilevel (hierarchical) logistic regression models. Each successive model introduces additional controls for geography, time, and local policing context.

This stepwise approach isolates the component of the racial disparity that remains once variation across **stations**, **years**, and **policing environments** is held constant.

---

## Model 1 - Race, Station and Year

### Model 1a: Baseline multilevel model (random intercepts)

We first fit a model including **random intercepts for Station and Year**:

\[
\text{Found\_yes} \sim \text{Race} + (1 \mid \text{Station}) + (1 \mid \text{Year})
\]

This model accounts for overall differences in hit rates across places and years, while estimating a single statewide Race effect.

The results indicate that, after controlling for clustering by Station and Year, searches of African-appearance individuals were significantly **less likely** to find contraband than those of White-appearance individuals (odds ratio ≈ 0.74, *p* = 0.016).

```{r racial_hit_multilevel_nested 1a, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

# --- Helper for pretty p-values ---
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --- Build dat (self-contained) ---
dat <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("White", "African"),
    !is.na(Found)
  ) %>%
  mutate(
    Found_yes = as.integer(tolower(Found) %in% c("yes","y","true","1")),
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Station   = factor(Station.uniform),
    Year      = factor(Year)
  )



# --- Fit Model 1a: random intercepts for Station & Year ---
m1a_station <- glmer(
  Found_yes ~ Race + (1 | Station) + (1 | Year),
  data = dat,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

# --- Fixed-effects table (odds ratios) ---
fx_1a_station <- broom.mixed::tidy(m1a_station, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term %in% c("(Intercept)", "RaceAfrican")) %>%
  transmute(
    Term         = term,
    `Odds Ratio` = sprintf("%.3f", estimate),
    `95% CI (L)` = sprintf("%.3f", conf.low),
    `95% CI (U)` = sprintf("%.3f", conf.high),
    `p-value`    = fmt_p(p.value)
  )

# --- Print nicely ---
kableExtra::kbl(
  fx_1a_station,
  caption = "Model 1a — Race (fixed effect) with random intercepts for Station and Year"
) %>%
  kableExtra::kable_paper(full_width = FALSE) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("condensed", "hover"),
    position = "left"
  )

```

### Model 1b: Testing whether race effects vary across Stations

To examine whether the race disparity differs between stations, we fit an extended model that allows the **Race effect to vary by Station** (a random slope):

\[
\text{Found\_yes} \sim \text{Race} + (1 + \text{Race} \mid \text{Station}) + (1 \mid \text{Year})
\]


```{r racial_hit_randomslope_station, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}


# --- Helper for pretty p-values ---
fmt_p <- function(p) {
  ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

# --- Prepare data ---
dat <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("White", "African"),
    !is.na(Found)
  ) %>%
  mutate(
    Found_yes = ifelse(tolower(Found) == "yes", 1L, 0L),
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Station   = factor(Station.uniform),
    Year      = factor(Year)
  )

# --- Fit model with random slope for Race within Station ---
m_raceslope_station <- glmer(
  Found_yes ~ Race + (1 + Race | Station) + (1 | Year),
  data = dat,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

# --- Extract only fixed effects (average state-wide estimates) ---
fx_station <- broom.mixed::tidy(m_raceslope_station, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term %in% c("(Intercept)", "RaceAfrican")) %>%
  transmute(
    Term       = term,
    Odds_Ratio = sprintf("%.3f", estimate),
    CI_L       = sprintf("%.3f", conf.low),
    CI_U       = sprintf("%.3f", conf.high),
    p_value    = fmt_p(p.value)
  )

# --- Nicely formatted table ---
kbl(fx_station, caption = "Race (fixed effect) from model with random slope for Race by Station") %>%
  kable_paper(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"), position = "left")
```
A likelihood-ratio test comparing this model to the simpler random-intercept version found **no improvement in fit** (χ² = 0.24, *p* = 0.89).  This suggests that the racial disparity in hit rates is relatively consistent across Stations rather than varying meaningfully between them.
```{r lrtest_station_1a_1b, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

# --- Helper for pretty p-values ---
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --- Make sure models exist; refit if needed ---
if (!exists("m1a_station") || !inherits(m1a_station, "glmerMod")) {
  message("Refitting Model 1a (Station + Year random intercepts)...")
  dat <- ddatr %>%
    filter(
      !is.na(Station.uniform),
      Racial.appearance.transformed %in% c("White", "African"),
      !is.na(Found)
    ) %>%
    mutate(
      Found_yes = as.integer(tolower(Found) %in% c("yes","y","true","1")),
      Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
      Station   = factor(Station.uniform),
      Year      = factor(Year)
    )

  ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

  m1a_station <- glmer(
    Found_yes ~ Race + (1 | Station) + (1 | Year),
    data = dat, family = binomial, control = ctrl
  )
}

if (!exists("m_raceslope_station") || !inherits(m_raceslope_station, "glmerMod")) {
  message("Refitting Model 1b (random slope for Race by Station)...")
  m_raceslope_station <- glmer(
    Found_yes ~ Race + (1 + Race | Station) + (1 | Year),
    data = dat, family = binomial, control = ctrl
  )
}

# --- Likelihood-ratio test (Model 1a vs 1b) ---
cmp <- suppressWarnings(anova(m1a_station, m_raceslope_station, test = "LRT"))
cmp_df <- as.data.frame(cmp)

results <- tibble::tibble(
  Model = c("Model 1a: (1|Station)+(1|Year)",
            "Model 1b: (1+Race|Station)+(1|Year)"),
  logLik     = sprintf("%.1f", cmp_df$logLik),
  AIC        = sprintf("%.1f", cmp_df$AIC),
  BIC        = sprintf("%.1f", cmp_df$BIC),
  Delta_Chi2 = c("-", ifelse(nrow(cmp_df) >= 2, sprintf("%.3f", cmp_df$Chisq[2]), "-")),
  Delta_df   = c("-", ifelse(nrow(cmp_df) >= 2, as.character(cmp_df$`Chi Df`[2]), "-")),
  p_value    = c("-", ifelse(nrow(cmp_df) >= 2, fmt_p(cmp_df$`Pr(>Chisq)`[2]), "-"))
)

# --- Print nicely ---
kableExtra::kbl(
  results,
  caption = "Model 1a vs 1b (Station-level): Likelihood-Ratio Test and Fit Statistics",
  align = "lrrrrrr"
) %>%
  kableExtra::kable_paper(full_width = FALSE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") %>%
  print()
```
Consequently, the simpler random-intercept specification (Model 1a) was retained as the baseline for subsequent models, which add contextual covariates such as **local drug-crime rates** and **police station effects**.

This suggests that the racial disparity is **consistent statewide**, rather than concentrated in particular areas.\

Per-station “shrunken” estimates confirm that almost all LGAs show odds ratios below 1, indicating uniformly lower hit rates for African-appearance individuals across Victoria.

Overall, the multilevel results demonstrate that the racial disparity in search hit rates remains robust after adjusting for both geographic and temporal clustering, with no evidence that the magnitude of the disparity differs meaningfully between LGAs.

We also re-estimated these two models using a restricted sample limited to stations with sufficient data — at least 30 total searches, including at least 10 involving African-appearance individuals. This robustness check confirmed that there is no meaningful between-area variation in the racial disparity: adding a random slope for Race by Station did not improve model fit (p = 0.94). Across Stations with adequate sample sizes, searches of African-appearance individuals remained significantly less likely to result in a find, with an estimated hit rate around 0.75 × that of White-appearance searches. The lack of improvement in the random-slope model indicates that this racial disparity is systematic and consistent across Stations, rather than being driven by a few stations.

```{r restriction check, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

# ===== Helper =====
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# ===== Build base data (unrestricted) =====
dat <- ddatr %>%
  dplyr::filter(
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("White", "African"),
    !is.na(Found)
  ) %>%
  dplyr::mutate(
    Found_yes = as.integer(tolower(Found) %in% c("yes","y","true","1")),
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Station   = factor(Station.uniform),
    Year      = factor(Year)
  )

# ===== Filter to Stations with African >= 10 and total >= 30 =====
counts <- dat %>%
  dplyr::count(Station, Race, name = "n") %>%
  tidyr::pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  dplyr::mutate(total = White + African)

keep_stations <- counts %>%
  dplyr::filter(African >= 10, total >= 30) %>%
  dplyr::pull(Station)

dat_filtered <- dat %>%
  dplyr::filter(Station %in% keep_stations)

# Safety check
if (nrow(dat_filtered) < 2L || dplyr::n_distinct(dat_filtered$Found_yes) < 2L) {
  knitr::kable(
    data.frame(Message = "Not enough variation after filtering (African >= 10 & total >= 30) to fit models.")
  ) |> print()
} else {

  ctrl <- lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

  # ===== Model 1a (filtered): random intercepts for Station & Year =====
  m1a_f <- lme4::glmer(
    Found_yes ~ Race + (1 | Station) + (1 | Year),
    data = dat_filtered, family = binomial, control = ctrl
  )

  # ===== Model 1b (filtered): random slope for Race by Station =====
  m1b_f <- lme4::glmer(
    Found_yes ~ Race + (1 + Race | Station) + (1 | Year),
    data = dat_filtered, family = binomial, control = ctrl
  )

  # ===== Comparison table (AIC/BIC/LRT) =====
  cmp <- suppressWarnings(anova(m1a_f, m1b_f, test = "LRT"))
  cmp_df <- as.data.frame(cmp)

  results <- tibble::tibble(
    Model = c("Model 1a (filtered): (1|Station)+(1|Year)",
              "Model 1b (filtered): (1+Race|Station)+(1|Year)"),
    logLik     = sprintf("%.1f", cmp_df$logLik),
    AIC        = sprintf("%.1f", cmp_df$AIC),
    BIC        = sprintf("%.1f", cmp_df$BIC),
    Delta_Chi2 = c("-", ifelse(nrow(cmp_df) >= 2, sprintf("%.3f", cmp_df$Chisq[2]), "-")),
    Delta_df   = c("-", ifelse(nrow(cmp_df) >= 2, as.character(cmp_df$`Chi Df`[2]), "-")),
    p_value    = c("-", ifelse(nrow(cmp_df) >= 2, fmt_p(cmp_df$`Pr(>Chisq)`[2]), "-"))
  )

  if (requireNamespace("kableExtra", quietly = TRUE)) {
    kableExtra::kbl(results,
      caption = "Filtered Stations (African ≥ 10 & total ≥ 30) — Model 1a vs 1b comparison (AIC, BIC, LRT)",
      align = "lrrrrrr"
    ) |>
      kableExtra::kable_paper(full_width = FALSE) |>
      kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") |>
      print()
  } else {
    knitr::kable(results,
      caption = "Filtered Stations (African ≥ 10 & total ≥ 30) — Model 1a vs 1b comparison (AIC, BIC, LRT)",
      align = "lrrrrrr"
    ) |> print()
  }

  # ===== Fixed effects tables for each filtered model =====
  fx_tbl <- function(mod, cap) {
    out <- broom.mixed::tidy(mod, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
      dplyr::filter(term %in% c("(Intercept)", "RaceAfrican")) |>
      dplyr::transmute(
        Term         = term,
        `Odds Ratio` = sprintf("%.3f", estimate),
        `95% CI (L)` = sprintf("%.3f", conf.low),
        `95% CI (U)` = sprintf("%.3f", conf.high),
        `p-value`    = fmt_p(p.value)
      )
    if (requireNamespace("kableExtra", quietly = TRUE)) {
      kableExtra::kbl(out, caption = cap, align = c("l","r","r","r","r")) |>
        kableExtra::kable_paper(full_width = FALSE) |>
        kableExtra::kable_styling(bootstrap_options = c("condensed","hover"), position = "left") |>
        print()
    } else {
      knitr::kable(out, caption = cap, align = c("l","r","r","r","r")) |> print()
    }
  }

  fx_tbl(m1a_f, "Model 1a (filtered, Station) — Fixed effects (OR, 95% CI, p)")
  fx_tbl(m1b_f, "Model 1b (filtered, Station) — Fixed effects (OR, 95% CI, p)")
}

```

**Interpretation**

The estimated *RaceAfrican* odds ratio remains approximately **0.75** and is statistically significant in both models.  
This means that, even after adjusting for **Year** and **Station**, searches of African-appearance individuals are about **25 % less likely** to find contraband than searches of White-appearance individuals.

The likelihood-ratio test (χ² = 0.73, p = 0.70) shows that allowing the race effect to vary across stations does **not** improve model fit.

Together, these results indicate that the racial disparity in search hit rates is **systematic and consistent statewide**, rather than driven by a small number of stations.

### Model 1c: Testing whether racial disparities vary by year

The previous models established that the racial disparity in search outcomes is robust across locations (stations).  
We next test whether this disparity **varies over time** — that is, whether the difference between African- and White-appearance hit rates changes across years.

To do this, we fit a model that allows the *Race* effect to vary randomly by **Year**, while continuing to include **Station** and **Year** random intercepts.  
This specification (Model 1c) assesses whether temporal variation in the race effect improves model fit relative to the simpler random-intercept model (Model 1a).



```{r year check, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}


# --- helpers ---
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --- data ---
dat <- ddatr %>%
  dplyr::filter(
    !is.na(Station.uniform),
    !is.na(Found),
    Racial.appearance.transformed %in% c("White","African")
  ) %>%
  dplyr::mutate(
    Found_yes = as.integer(is_yes(Found)),
    Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
    Station   = factor(Station.uniform),
    Year      = factor(Year)
  )

# --- model: allow Race slope to vary by Year; Station random intercept ---
m1c_station <- lme4::glmer(
  Found_yes ~ Race + (1 + Race | Year) + (1 | Station),
  family  = binomial,
  data    = dat,
  control = lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# --- extract fixed effects ---
fx_1c_station <- broom.mixed::tidy(m1c_station, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
  dplyr::filter(term %in% c("(Intercept)", "RaceAfrican")) %>%
  dplyr::transmute(
    Term         = term,
    `Odds Ratio` = sprintf("%.3f", estimate),
    `95% CI (L)` = sprintf("%.3f", conf.low),
    `95% CI (U)` = sprintf("%.3f", conf.high),
    `p-value`    = fmt_p(p.value)
  )

# --- pretty print ---
if (requireNamespace("kableExtra", quietly = TRUE)) {
  kableExtra::kbl(fx_1c_station,
                  caption = "Model 1c — Race (fixed effect) with random slopes by Year and random intercept for Station") %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("condensed","hover"), position = "left")
} else {
  knitr::kable(fx_1c_station,
               caption = "Model 1c — Race (fixed effect) with random slopes by Year and random intercept for Station")
}

```


Allowing the race effect to vary by year did not improve model fit (Δχ² = 0.08, *p* = 0.96).  

The AIC and BIC were slightly higher for Model 1c (AIC = 6010.5 vs 6006.6; BIC = 6052.0 vs 6034.3), confirming that the added complexity provides no explanatory benefit.  

Therefore, the simpler Model&nbsp;1a — with fixed racial effect and random intercepts for **Station** and **Year** — is retained as the most parsimonious and interpretable specification.  

The racial disparity (OR ≈ 0.75) remains consistent across both years and stations, indicating a systematic statewide pattern rather than one driven by temporal or local effects.


```{r lrtest_station_1a_1c, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}


# --- helper ---
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --- build data (if needed) ---
if (!exists("dat")) {
  dat <- ddatr %>%
    filter(
      !is.na(Station.uniform),
      !is.na(Found),
      Racial.appearance.transformed %in% c("White","African")
    ) %>%
    mutate(
      Found_yes = as.integer(tolower(Found) %in% c("yes","y","true","1")),
      Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
      Station   = factor(Station.uniform),
      Year      = factor(Year)
    )
}

ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))

# --- Model 1a: Race + (1|Station) + (1|Year) ---
if (!exists("m1a_station") || !inherits(m1a_station, "glmerMod")) {
  m1a_station <- glmer(
    Found_yes ~ Race + (1 | Station) + (1 | Year),
    data = dat, family = binomial, control = ctrl
  )
}

# --- Model 1c: Race + (1+Race|Year) + (1|Station) ---
m1c_station <- glmer(
  Found_yes ~ Race + (1 + Race | Year) + (1 | Station),
  data = dat, family = binomial, control = ctrl
)

# --- LRT & fit table ---
cmp <- suppressWarnings(anova(m1a_station, m1c_station, test = "LRT"))
cmp_df <- as.data.frame(cmp)

results <- tibble::tibble(
  Model = c("Model 1a: (1|Station)+(1|Year)",
            "Model 1c: (1+Race|Year)+(1|Station)"),
  logLik     = sprintf("%.1f", cmp_df$logLik),
  AIC        = sprintf("%.1f", cmp_df$AIC),
  BIC        = sprintf("%.1f", cmp_df$BIC),
  Delta_Chi2 = c("-", ifelse(nrow(cmp_df) >= 2, sprintf("%.3f", cmp_df$Chisq[2]), "-")),
  Delta_df   = c("-", ifelse(nrow(cmp_df) >= 2, as.character(cmp_df$`Chi Df`[2]), "-")),
  p_value    = c("-", ifelse(nrow(cmp_df) >= 2, fmt_p(cmp_df$`Pr(>Chisq)`[2]), "-"))
)

if (requireNamespace("kableExtra", quietly = TRUE)) {
  kableExtra::kbl(
    results,
    caption = "Model 1a vs 1c (Station-level): Likelihood-Ratio Test and Fit Statistics",
    align = "lrrrrrr"
  ) |>
    kableExtra::kable_paper(full_width = FALSE) |>
    kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") |>
    print()
} else {
  knitr::kable(
    results,
    caption = "Model 1a vs 1c (Station-level): Likelihood-Ratio Test and Fit Statistics",
    align = "lrrrrrr"
  ) |> print()
}
```
## Model 2 - adding in the crime rate  

To test whether the observed racial disparities in search “hit rates” might simply reflect differences in local offending environments or policing intensity, we extended the baseline multilevel model to include the local drug-crime rate as a contextual covariate. This variable captures spatial and temporal variation in drug-related offences across LGAs and years, providing a proxy for policing environment and underlying crime prevalence. Model 2 therefore estimates whether the racial disparity in search outcomes persists after controlling for the local crime context, while continuing to account for clustering by LGA and Year through random intercepts. We also tested an interaction between Race and crime rate to examine whether the disparity widens or narrows in higher-crime areas, but this additional term did not improve model fit, indicating that the racial gap remains stable regardless of local drug-crime intensity.
```{r crime rate, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}


# ------- helpers -------
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
kbl_print <- function(x, cap, align=NULL){
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    kableExtra::kbl(x, caption = cap, align = align) |>
      kableExtra::kable_paper(full_width = FALSE) |>
      kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") |>
      print()
  } else {
    print(knitr::kable(x, caption = cap, align = align))
  }
}

# ------- ensure data (uses your existing `dat`) -------
# dat must contain: Found_yes (0/1), Race (factor: White, African), LGA (factor), Year (factor), Drug.crime.rate
stopifnot(all(c("Found_yes","Race","LGA","Year") %in% names(dat)))

# Coerce crime rate to numeric, build an explicit scaled column (avoids scale() inside formula)
dat <- dat %>%
  mutate(
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate)),
    z_crime = if (all(is.na(Drug.crime.rate))) NA_real_
              else as.numeric(scale(Drug.crime.rate))
  )

use_crime <- "z_crime" %in% names(dat) && !all(is.na(dat$z_crime))

if (!use_crime) {
  kbl_print(data.frame(Message = "Drug.crime.rate missing or all NA; fitting models without crime rate."), "Note")
}

ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))

# ------- Model 2 (additive) -------
# --- safer conditional formula assignment ---
form_add <- if (use_crime) {
  Found_yes ~ Race + z_crime + (1 | LGA) + (1 | Year)
} else {
  Found_yes ~ Race + (1 | LGA) + (1 | Year)
}

m2 <- glmer(form_add, family = binomial, data = dat, control = ctrl)

# ------- Model 2 with interaction (Race × crime) -------
if (use_crime) {
  # Try update() first (cleanest)
  m2_int <- try(update(m2, . ~ Race * z_crime + (1|LGA) + (1|Year)), silent = TRUE)
  if (inherits(m2_int, "try-error")) {
    # Fallback: direct fit
    m2_int <- glmer(Found_yes ~ Race * z_crime + (1|LGA) + (1|Year),
                    family = binomial, data = dat, control = ctrl)
  }

  # ------- LRT compare additive vs interaction -------
  cmp <- suppressWarnings(anova(m2, m2_int, test = "LRT"))
  cmp_df <- as.data.frame(cmp)

  # Defensive extraction
  has <- function(nm) nm %in% names(cmp_df)
  chisq_val <- if (has("Chisq")) cmp_df$Chisq[2] else NA_real_
  chidf_val <- if (has("Chi Df")) cmp_df$`Chi Df`[2] else NA_real_
  pval_val  <- if (has("Pr(>Chisq)")) cmp_df$`Pr(>Chisq)`[2] else NA_real_

  out_cmp <- tibble::tibble(
    Model      = c("Model 2 (additive): Race + z_crime + (1|LGA) + (1|Year)",
                   "Model 2 (+ interaction): Race * z_crime + (1|LGA) + (1|Year)"),
    logLik     = sprintf("%.1f", cmp_df$logLik),
    AIC        = sprintf("%.1f", cmp_df$AIC),
    BIC        = sprintf("%.1f", cmp_df$BIC),
    Delta_Chi2 = c("-", if (!is.na(chisq_val)) sprintf("%.3f", chisq_val) else "-"),
    Delta_df   = c("-", if (!is.na(chidf_val)) as.character(chidf_val) else "-"),
    p_value    = c("-", if (!is.na(pval_val)) fmt_p(pval_val) else "-")
  )
  kbl_print(out_cmp, "Model 2 comparison: additive vs Race × crime interaction", "lrrrrrr")

  # Fixed effects from preferred model (choose by AIC)
  m_pref <- if (AIC(m2_int) + 2 < AIC(m2)) m2_int else m2
  fx <- broom.mixed::tidy(m_pref, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
    dplyr::transmute(
      Term = term,
      `Odds Ratio` = sprintf("%.3f", estimate),
      `95% CI (L)` = sprintf("%.3f", conf.low),
      `95% CI (U)` = sprintf("%.3f", conf.high),
      `p-value`    = fmt_p(p.value)
    )
  kbl_print(fx, "Preferred Model 2 — Fixed effects (OR, 95% CI, p)", c("l","r","r","r","r"))

} else {
  # Crime rate unavailable: just report Model 2 (additive without crime)
  fx <- broom.mixed::tidy(m2, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
    dplyr::transmute(
      Term = term,
      `Odds Ratio` = sprintf("%.3f", estimate),
      `95% CI (L)` = sprintf("%.3f", conf.low),
      `95% CI (U)` = sprintf("%.3f", conf.high),
      `p-value`    = fmt_p(p.value)
    )
  kbl_print(fx, "Model 2 (no crime variable available) — Fixed effects (OR, 95% CI, p)", c("l","r","r","r","r"))
}

```
Introducing the local drug-crime rate does not materially alter the estimated racial disparity. The coefficient for Race = African remains statistically significant (OR ≈ 0.77, p = 0.036), while the crime-rate covariate itself is non-significant (p = 0.41). This suggests that area-level variation in drug offences does not account for the lower hit rate among African-appearance individuals. The modest random-effect variance at the LGA–Year level indicates that most of the remaining variability lies within rather than between area–year contexts.

## Model 4: LGA × Year + Station random-intercept model

While Model 3 controls for geographic and temporal clustering at the LGA–Year level, it does not yet account for differences between police stations operating within the same area.

Stations vary in staffing, priorities, and local enforcement practices, all of which can affect the likelihood that a search finds contraband.

To capture this layer of variation, we extend the model to include a **random intercept for each Station**, allowing baseline hit rates to differ between stations within the same LGA–Year combination.

The specification is:

$$
\text{logit}(P(\text{Hit})) =
\beta_0 + \beta_1(\text{RaceAfrican}) + \beta_2(\text{scale(Drug.crime.rate)}) +
u_{\text{LGA:Year}} + v_{\text{Station}},
$$ where $u_{\text{LGA:Year}}$ represents unobserved area–year differences, and $v_{\text{Station}}$ captures station-level deviations from those baselines.

This structure makes comparisons even more conservative: it contrasts searches of African- and White-appearance individuals **within the same year, same LGA, and same station context**, holding local crime rates constant.

```{r racial_hit_multilevel_station, echo=FALSE, warning=FALSE, message=FALSE}

# ---- Helpers ----
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
kbl_print <- function(x, cap, align=NULL){
  if (requireNamespace("kableExtra", quietly = TRUE)) {
    kableExtra::kbl(x, caption = cap, align = align) |>
      kableExtra::kable_paper(full_width = FALSE) |>
      kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") |>
      print()
  } else {
    print(knitr::kable(x, caption = cap, align = align))
  }
}

# ---- Prepare data ----
dat <- ddatr %>%
  filter(
    !is.na(Local.Government.Area),
    !is.na(Found),
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("White","African")
  ) %>%
  mutate(
    Found_yes = as.integer(tolower(Found) %in% c("yes","y","true","1")),
    Race = factor(Racial.appearance.transformed, levels = c("White","African")),
    LGA = factor(Local.Government.Area),
    Station = factor(Station.uniform),
    Year = factor(Year),
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate)),
    z_crime = as.numeric(scale(Drug.crime.rate))
  )

# ---- Fit model: includes nested random effects ----
m_station <- glmer(
  Found_yes ~ Race + z_crime + (1 | LGA/Station) + (1 | Year),
  family = binomial,
  data = dat,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# ---- Fixed effects ----
fx_station <- broom.mixed::tidy(m_station, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
  transmute(
    Term = term,
    `Odds Ratio` = sprintf("%.3f", estimate),
    `95% CI (L)` = sprintf("%.3f", conf.low),
    `95% CI (U)` = sprintf("%.3f", conf.high),
    `p-value` = fmt_p(p.value)
  )

kbl_print(fx_station,
  "Model 3 — Race and crime rate with nested random intercepts for Station (within LGA) and Year",
  c("l","r","r","r","r")
)

# ---- Random effects summary ----
re_var <- as.data.frame(VarCorr(m_station)) %>%
  transmute(
    Group = grp,
    Variance = round(vcov, 4),
    SD = round(sdcor, 4)
  )
kbl_print(re_var,
  "Model 3 — Random-effects variance components (LGA, Station[LGA], Year)",
  c("l","r","r")
)

```

Adding random intercepts for individual stations nested within LGAs substantially improved the model’s ability to account for local variation (SD ≈ 0.33 at the station level, compared with 0.15 across LGAs and 0.05 across years). Most spatial heterogeneity in search outcomes therefore occurs between stations rather than between broader regions. Importantly, the race effect remained significant (OR ≈ 0.75, p = 0.018), indicating that the lower hit rate for African-appearance individuals persists even after controlling for local crime rates, station-level clustering, and year-to-year fluctuations.

## Station-Level Findings

-   **Werribee Station** also displayed a significant disparity:
    -   African: **5/57 (8.8%)** vs White: **22/103 (21.4%)**\
    -   Risk difference = –12.6%, Risk ratio = 0.41, *p* = 0.042\
-   Werribee is in the same outer-western growth corridor as Brimbank/Melton, with a **substantial African community** and a history of **public concern about over-policing**.

```{r stations-hitrate-with-prop, echo=FALSE, message=FALSE, warning=FALSE}
alpha <- 0.05

sta_tbl <- make_hit_table(
  data = ddatr,
  group_var = "Station.uniform",
  alpha = alpha,
  min_african_total = 5,      # inclusion rule you requested
  filter_sig_only = TRUE,     # show only p < 0.05; set FALSE to see all included
  add_fdr = TRUE
)

print_hit_table(
  sta_tbl,
  caption = sprintf("Stations (≥ 5 African searches) — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
  show_bh = TRUE
)

```

## Interpretation

The proportion tests suggest that disparities are **geographically concentrated** in Melbourne’s **outer west**, where disadvantage and demographic change intersect with intensive policing.\
Other Divisions and Stations did not show statistically significant differences, but many had small African sample sizes. Absence of significance does not mean absence of disparity.

These results highlight the need for multivariate modelling to control for confounders such as **gender, year, and local policing context**.

# Statistical Modelling

Logistic regression estimates the odds that a search will result in contraband being found, while adjusting for other variables. This allows us to move beyond descriptive comparisons and ask whether observed differences reflect systematic disparities in how suspicion is applied.

## Division-Level Logistic Regression Models

The proportion tests showed that disparities in hit rates were not uniform across Melbourne, with some divisions and stations suggesting stronger evidence of racial disparity than others.

To explore this further, we estimated **logistic regression models separately for each Local Government Area.**

The outcome is whether **any contraband was found** (`Found = Yes`). Each model includes predictors for:

-   **Race** (African vs White)\
-   **Gender** (Female vs Male)\
-   An **interaction between race and gender**\
-   Local Government Area - recognising that
-   Year: recognising that search intensity may vary from year to year.

Summary tables models where the coefficient of racial appearance was lower than 0.05 appear below.

```{r lga-models-optionA-minAfr, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(broom.mixed)
library(knitr)
library(kableExtra)
library(lme4)

# ---------------- params ----------------
MIN_AFRICAN <- 10   # minimum African-appearance searches per Division

# ---------------- helpers ----------------
fmt_p <- function(p) ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
is_yes <- function(x) if (is.logical(x)) !is.na(x) & x else tolower(as.character(x)) %in% c("yes","y","true","1")

# ---------------- prepare data ----------------
dat <- ddatr %>%
  filter(
    !is.na(Division),
    !is.na(Station.uniform),
    !is.na(Found),
    Racial.appearance.transformed %in% c("White","African")
  ) %>%
  mutate(
    Found_yes = as.integer(is_yes(Found)),
    Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
    Division  = factor(Division),
    Station   = factor(Station.uniform),
    Year      = factor(Year),
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate)),
    z_crime   = as.numeric(scale(Drug.crime.rate))
  )

# ---------------- select eligible Divisions ----------------
counts <- dat %>%
  count(Division, Race, name = "n") %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0)

eligible_divs <- counts %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(Division)

# ---------------- fit per-Division models ----------------
data_by_div <- split(dat, dat$Division)

for (div_name in eligible_divs) {
  df <- data_by_div[[div_name]]
  if (n_distinct(df$Found_yes) < 2L) next

  # --- fit the model ---
  mdl <- tryCatch(
    glmer(
      Found_yes ~ Race + z_crime + (1 | Station) + (1 | Year),
      data = df,
      family = binomial,
      control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
    ),
    error = function(e) NULL
  )
  if (is.null(mdl)) next

  fx <- broom.mixed::tidy(mdl, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
    transmute(
      Term = term,
      `Odds Ratio` = sprintf("%.3f", estimate),
      `95% CI (L)` = sprintf("%.3f", conf.low),
      `95% CI (U)` = sprintf("%.3f", conf.high),
      `p-value` = fmt_p(p.value)
    )

  kbl(fx,
      caption = paste0("Division: ", div_name,
                       " — Model: Found_yes ~ Race + z_crime + (1|Station) + (1|Year)  (African ≥ ",
                       MIN_AFRICAN, ")"),
      align = c("l","r","r","r","r")) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","hover"), position = "left") %>%
    print()
}


```

```{r station-models-optionA-minAfr, echo=FALSE, warning=FALSE, message=FALSE}



# ---------------- params ----------------
MIN_AFRICAN <- 20   # adjust if needed

# -------------- helpers -----------------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --------- prepare data ----------
dat_sta <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("African","White"),
    !is.na(Found),
    Gender %in% c("M","F")
  ) %>%
  transmute(
    Station = factor(Station.uniform),
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F")),
    Year    = Year
  )

# Context for captions (counts by Station & race)
sta_counts <- dat_sta %>%
  group_by(Station, Race) %>%
  summarise(succ = sum(Success), tot = n(), .groups = "drop") %>%
  pivot_wider(names_from = Race, values_from = c(succ, tot), values_fill = 0) %>%
  mutate(n_total = tot_African + tot_White)

# Eligibility: enforce minimum African sample size per Station
eligible_stations <- dat_sta %>%
  count(Station, Race) %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(Station) %>%
  as.character()

# --------- fit per-Station models ----------
data_by_sta <- split(dat_sta, dat_sta$Station)

walk(eligible_stations, function(sta_name) {
  df <- data_by_sta[[sta_name]]

  # skip if only one outcome class
  if (n_distinct(df$Success) < 2L) return(invisible(NULL))

  # Model: Race * Gender + Year (fixed effects)
  mdl <- tryCatch(
    glm(Success ~ Race * Gender + factor(Year), data = df, family = binomial),
    error = function(e) NULL
  )
  if (is.null(mdl)) return(invisible(NULL))

  td_raw <- tryCatch(
    tidy(mdl, exponentiate = TRUE, conf.int = TRUE),
    error = function(e) {
      tidy(mdl, exponentiate = TRUE, conf.int = FALSE) %>%
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    }
  ) %>%
    select(term, estimate, conf.low, conf.high, p.value)

  # show only if any non-intercept term significant
  keep_if_sig <- any(td_raw$term != "(Intercept)" & td_raw$p.value < 0.05, na.rm = TRUE)
  if (!keep_if_sig) return(invisible(NULL))

  # Friendly labels
  td <- td_raw %>%
    mutate(
      Term = case_when(
        term == "(Intercept)" ~ "(Intercept)",
        term == "RaceAfrican" ~ "Race: African vs White (Male ref.)",
        term == "GenderF" ~ "Gender: Female vs Male (White ref.)",
        term == "RaceAfrican:GenderF" ~ "Interaction: African × Female",
        str_detect(term, "^factor\\(Year\\).+") ~
          paste0("Year: ", str_replace(term, "^factor\\(Year\\)", ""), " vs ref."),
        TRUE ~ term
      ),
      `Odds Ratio`      = estimate,
      `95% CI (Lower)`  = conf.low,
      `95% CI (Upper)`  = conf.high,
      `p-value`         = p.value
    ) %>%
    select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
    mutate(
      `p-value` = fmt_p(`p-value`),
      `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
      `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
      `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`))
    )

  # Caption: n and years
  cx <- filter(sta_counts, Station == sta_name)
  yrs <- sort(unique(stats::na.omit(df$Year)))
  years_txt <- if (length(yrs) == 0) NULL else if (length(yrs) == 1) {
    paste0("Year: ", yrs)
  } else {
    paste0("Years: ", min(yrs), "–", max(yrs))
  }

  cap_bits <- c(
    paste0("n=", ifelse(nrow(cx) == 1, cx$n_total, NA_integer_)),
    if (nrow(cx) == 1) paste0("Afr: ", cx$succ_African, "/", cx$tot_African) else NULL,
    if (nrow(cx) == 1) paste0("Wh: ",  cx$succ_White,   "/", cx$tot_White)   else NULL,
    years_txt
  )
  cap_ctx <- paste(cap_bits[!is.na(cap_bits)], collapse = " · ")

  # Print table
  td %>%
    kbl(
      caption = paste0("Station: ", sta_name,
                       " — Logistic regression of search success (any contraband). ",
                       cap_ctx),
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","condensed","hover"),
                  position = "left") %>%
    column_spec(1, width = "22em") %>%
    add_footnote(
      label = paste0(
        "Model: Success ~ Race * Gender + Year (fixed effects). ",
        "Odds ratios shown (95% CI). Reference group = White males and earliest Year. ",
        "Eligibility: African count per Station ≥ ", MIN_AFRICAN, "."
      ),
      notation = "symbol"
    ) %>%
    print()
})
```

# Appendix

**Table X Hit rates by largest racial grouping**

```{r Finds racial all, echo=FALSE, warning=FALSE, results='asis'}
ddats<-ddat %>%
  filter(Racial.appearance.transformed %in% c("Missing", "White","Mediterranean/Middle Eastern – Unusable",
                                              "Asian","African", "Aboriginal"))

ct <- crosstable::crosstable(
  ddats,
  by = "Found",
  cols = "Racial.appearance.transformed",
  total = TRUE,
  test = FALSE,          # we'll add our own robust test
  showNA = "ifany",
  percent_digits = 1
)

ft <- crosstable::as_flextable(ct)

# 3) Safe association test (Fisher for 2x2, else Chi-square with simulated p)
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tres <- safe_test(ddats$Racial.appearance.transformed, ddats$Found)
p_line <- sprintf("Racial.appearance.transformed — %s, p = %s",
                  tres$label, format.pval(tres$p, digits = 4))

# 4) Add footer and a clear title
ft <- flextable::add_footer_lines(ft, values = c("Overall association test:", p_line))
ft <- flextable::set_caption(ft, caption = "Racial appearance vs Found (with totals)")
ft
```

```{r Finds lga effect, echo=FALSE, warning=FALSE, results='asis'}
# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <-  ddat 
names(ddatr)
# Variables to tabulate against Found
vars_for_table <- c("Local.Government.Area")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft
```
