---
title: "Racial profiling - drug searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(scales)
library(stringr)
library(tidyr)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

```

```{r hitrate-prop-helpers, echo=FALSE, message=FALSE, warning=FALSE}


is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}


make_hit_table <- function(
  data,
  group_var,                   # e.g. "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    dplyr::filter(!is.na(!!grp), !is.na(Found)) %>%
    dplyr::filter(Racial.appearance.transformed %in% c("African","White")) %>%
    dplyr::mutate(FoundYes = is_yes(Found)) %>%
    dplyr::count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    dplyr::group_by(!!grp, Racial.appearance.transformed) %>%
    dplyr::summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    dplyr::filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble::tibble())

  out <- tab %>%
    dplyr::mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    dplyr::select(-test) %>%
    tidyr::unnest(tidied) %>%
    dplyr::mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = dplyr::if_else(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    # Important: use !!rlang::sym(group_var) on LHS, and avoid Unicode minus in names
    dplyr::transmute(
      !!rlang::sym(group_var) := !!grp,
      `African (succ/total)`   = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`     = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`       = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`         = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr - Wh)`        = scales::percent(diff_est,         accuracy = 0.1),  # <-- ASCII hyphen
      `Risk ratio (Afr/Wh)`    = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                      = sprintf("%.3f", p.value),
      `p (BH)`                 = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                    = stars
    )

  if (filter_sig_only) {
    out <- out %>% dplyr::filter(as.numeric(stringr::str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% dplyr::arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))
}

print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) display <- display %>% dplyr::select(-`p (BH)`)

  display %>%
    knitr::kable(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    kableExtra::column_spec(1, width = "24em") %>%
    kableExtra::add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}

```

# Short summary

This analysis of 7,534 warrantless drug searches carried out by **uniform officers on foot patrol in metropolitan Melbourne (2018, 2019, 2022, 2023)** provides clear evidence that individuals perceived as **African** are searched on a **lower evidentiary threshold** compared to those perceived as White.  

Key findings include:  

- **Descriptive results** showed lower hit rates for African-appearing individuals across most contexts.  
- **Proportion tests** identified particularly strong disparities in Brimbank, Melton, and Werribee.  
- **Division level models** controlling for gender, as well operating context including year and Local Government Area (LGA) revealed disparities in  
- **Local government area level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in
- **Statio level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in

These results are consistent with **racial profiling**: African-appearing individuals are more frequently subjected to searches that are less likely to yield contraband, implying that suspicion is applied more readily to them.  

The magnitude of disparity varies locally, the overall pattern is systematic.  

This highlights the need for greater police accountability and independent oversight, and provides an empirical foundation for community advocacy and policy reform aimed at fairer, more equitable policing in Melbourne. 

### Melbourne-wide findings
- Hit rates: White 14.1% (957/6,779) vs African 11.4% (86/755); Fisher’s exact p = 0.040.
- Pooled logistic regression (controls: Division fixed effects, Year, Gender): OR = 0.771 (95% CI 0.596–0.985), p = 0.042 — African-appearing males had ~23% lower odds of a find than White-appearing males.
- Multilevel logistic regression (random intercepts for Division): OR = 0.761 (95% CI 0.594–0.975), p = 0.031 — confirms the disparity after accounting for clustering by Division.


### Local hotspots
- NWMR Division 3 (Brimbank & Melton): African 4.3% (3/70) vs White 14.2% (30/212); risk ratio = 0.30, p = 0.026.
- Werribee Station: African 8.8% (5/57) vs White 21.4% (22/103); risk ratio = 0.41, p = 0.042.


### Limits of statistical significance
- Most other divisions and stations did not reach statistical significance, primarily because of small African sample sizes (often only a few dozen searches). Wide confidence intervals and multiple-testing adjustments mean that many apparent disparities cannot be declared significant.
- Non-significance does not equal absence of disparity — rather, it reflects the limited statistical power in areas with few recorded searches of African-appearing individuals.


### Interpretation
Across Melbourne, African-appearing people are consistently subjected to searches with lower hit rates and lower odds of contraband being found, even after adjusting for year and division. This pattern is strongest in Melbourne’s outer west (Brimbank, Melton, Werribee), but persists at a city-wide level. Together, these results are consistent with racial profiling: African-appearing individuals are more frequently searched on a weaker threshold of suspicion.

# Introduction  

This analysis examines **7,534 drug searches conducted by uniformed officers on foot patrols in Melbourne, of white- and African-appearing people**. These encounters represent the most routine form of warrantless public-place searches, where outcomes are highly dependent on officer discretion. The racial appearance categories in scope represent two of the largest and consistently recorded racial groupings in the dataset.  

The focus is on whether search outcomes suggest evidence of **racial profiling**. Specifically, we compare the **hit rates** — the proportion of searches that result in contraband being found — for two groups:  

- **White-appearing individuals**, who form the majority baseline  
- **African-appearing individuals**, a racialised minority consistently highlighted in community concerns about disproportionate policing  

Concerns about the over-policing of African-Australian communities are not new. People such as **Haile Michael**, who successfully challenged Victoria Police’s discriminatory stop-and-search practices in the landmark [**Haile Michael**](https://www.abc.net.au/news/2013-03-18/victoria-police-accused-of-racially-profiling-african-men/4578358) case, have documented and contested these patterns for more than a decade.  

By restricting the dataset to this context and to these two appearance categories, we provide a clear test of whether African-appearing people are subjected to searches on a **lower evidentiary threshold**. Lower hit rates for one group, despite comparable or higher search intensity, would suggest that police suspicion is being applied unevenly — a hallmark of racial profiling.  
  
  
# About the dataset (subset definition)  

This dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**.  

Data from **2020 and 2021** is excluded due to the atypical policing environment during the COVID-19 pandemic.  

The dataset was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and covers public-place searches carried out under a range of legal powers.  

For this analysis, we restrict to:  

- **Geography:** Melbourne metropolitan area  
- **Contact type:** Foot patrol only  
- **Unit type:** Uniform police only  
- **Search type:** Drug searches conducted under s82 of the *Drugs, Poisons and Controlled Substances Act 1981*  
- **Station linkage:** Only searches associated with a uniform station  
- **Racial appearance:** White and African appearance

This restriction enables us to compare searches for the most common drug searches experienced by the public in Melbourne, and restrict the data to reliable racial appearance categories.

## What s82 allows  

Under section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.  
This is the most frequently invoked search power in the dataset and represents a context where discretionary judgments by officers play a central role.  

## Variables in this subset  

The following variables are included in the analysis:  

- **Year** — year of the stop and search  
- **Region / Division / PSA / Station** — police geographic hierarchy  
- **Racial.appearance.transformed** — officer-perceived racial appearance (cleaned categories)  
- **Gender** — officer-perceived gender of the person searched  
- **Found** — whether any contraband was located (hit outcome)  
- **Unit.type** — uniform officers only (subset restriction)  
- **Rank** — officer rank (e.g. Constable, Senior Constable)  
- **Year** — 2018, 2019, 2022, or 2023  

Derived variables include:  

- **Found.Flag** — binary indicator (1 if contraband found, 0 otherwise)  
- **Race** — restricted to White vs African  

---

## Data limitations  

### Racial appearance categories  
The original dataset for drug searches conducted by uniform officers on foot patrol numbered 13,896. Race data was missing from *20% of the records*.  Several categories had very small numbers or had been recategorised over the observation period.  As a result, disparity testing is restricted to **White vs African** searches, which are sufficiently large and analytically comparable.  Together these racial categories account for over half (54%) of the relevant data.

```{r race prop, echo=FALSE, message=FALSE, warning=FALSE}
# Count + percent for all groups (no filtering)
tab_ra <- ddat %>%
  mutate(
    Racial.appearance.transformed =
      fct_explicit_na(Racial.appearance.transformed, "Missing")
  ) %>%
  count(Racial.appearance.transformed, name = "Count") %>%
  arrange(desc(Count)) %>%
  mutate(Percent = Count / sum(Count))

# Add total row
tab_ra_tot <- bind_rows(
  tab_ra,
  tibble(
    Racial.appearance.transformed = "Total",
    Count = sum(tab_ra$Count),
    Percent = 1
  )
)

# Format percent for display
tab_ra_tot <- tab_ra_tot %>%
  mutate(Percent = percent(Percent, accuracy = 0.1))

# Render table
kbl(
  tab_ra_tot,
  caption = "Racial appearance (all groups): count and percentage of searches",
  align = c("l", "r", "r"),
  col.names = c("Racial appearance (transformed)", "Count", "Percent")
) %>%
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Searches" = 2))

```
### Age data  
Age data was missing for 42.9% of searches in the White vs African subset. Age is therefore excluded from modelling, which limits the ability to adjust for age differences across groups. We expect this would have an effect as African migrant populations in in local government areas tend to be younger than their white-appearing counterparts. 

### Scope of coverage  
- The dataset covers only **warrantless stop and search events conducted in public places**. This focus is deliberate: such encounters are entirely reliant on officer discretion, since there is no warrant, prior intelligence, or judicial approval.  
- The analysis does not include searches with warrants, searches in private settings, or informal/unrecorded encounters.  
- Findings therefore apply only to **Melbourne metropolitan drug searches by uniform officers on foot patrol**. They do not generalise to specialist units, vehicle stops, or regional policing contexts.  
- Because modelling is restricted to White vs African, results do not extend to other racialised groups with smaller or analytically unreliable sample sizes.  


# Data Overview  

## Persons searched  

This section summarises the characteristics of people searched in warrantless public-place encounters by uniform officers on foot patrols (2018, 2019, 2022, 2023).  
Records with missing or unusable racial appearance are excluded to allow meaningful comparison.  

- **Racial appearance.** Most searches involved people perceived as White, with African-appearance individuals making up around one in ten of the analytic sample.  
- **Gender.** Searches were concentrated on men, who accounted for about four-fifths of those stopped.  
- **Age.** While children and adolescents appear in the dataset, most searches targeted adults aged 18 to 45. People aged 46 and over were searched relatively infrequently.

These patterns highlight that discretionary stop-and-search practices fall unevenly across the community, with young men and racialised groups particularly over-represented.  


**Table 2 Characteristics of persons searched**
```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddatr <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(ddat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police undertaking searches  

Search numbers were relatively consistent across the four years examined, with a slight peak in 2022.  By rank, Constables and Senior Constables were the most frequently involved, reflecting the predominance of frontline officers in this type of activity.  


**Table 3 Police and units undertaking searches**
```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Year", "Rank.of.Member", "Region",  "Division")

ddatr <- ddatr %>%
  mutate(
    Rank = case_when(
      Rank.of.Member == "SSGT" ~ "SGT",
      Rank.of.Member == "PSO"  ~ "CONST",
      TRUE           ~ Rank.of.Member
    )
  )

dat.pol <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')


```



# Hit Rate - cross tables 

An important way to assess stop-and-search practices is through **hit rates** — the proportion of searches that result in contraband being found.  

If searches are consistently based on *reasonable suspicion* in comparable spatial areas, hit rates should be similar across groups.  

Where a group has consistently lower hit rates for a particular area, this may suggest that officers may be applying a **lower threshold of suspicion**, a core indicator of potential racial profiling.  

This section examines hit rates in the restricted dataset of **7,534 Melbourne metropolitan drug searches by uniform police on foot patrol** (2018, 2019, 2022, 2023).  

A “hit” is defined as the discovery of any prohibited item, not just the item specified under the search power.  

## Hit rates - cross tables

An initial analysis of comparative hit rates across our sample of 7,453 searches was performed, looking both at:

* whether hit rates varied by racial appearance to a degree that is statistically significant and  

* Whether hit rates varied by other factors (such as local government area, gender, etc) to a degree that may counfound our results.  


- **By racial appearance.**  
  - White-appearing people had a hit rate of **14.1%** (957 finds out of 6,779 searches).  
  - African-appearing people had a lower hit rate of **11.4%** (86/755).  
  - This indicates that African-appearance individuals were **less likely to be found with contraband**, despite being searched in substantial numbers.  

- **By gender.**  
  - Men had a hit rate of **14.1%** (837/5,956), compared with **13.1%** for women (206/1,578).  
  - Searches were heavily concentrated on men, accounting for four in five cases.  

- **By age.**  
  - Differences in hit rates across ages were marginal, however there does appear to be a slight decrease in hit rates where older people are searched.  
  
  
- **By time.** Across 2018, 2019, 2022, and 2023, hit rates varied only modestly. 

- **By rank.** Constables and Senior Constables dominated search activity, and their hit rates were broadly similar.  

- **By geography.** Hit rates were lowest in the **North West Metro Region**, which includes diverse and disadvantaged areas such as Brimbank and Melbourne, compared with other parts of the city.  
 
 

**Table 5 Hit rates by person characteristics**


```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}




# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddatr %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),                
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop(),
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Rank = dplyr::recode(Rank.of.Member,
                         "SSGT" = "SGT",
                         "PSO"  = "CONST",
                         .default = Rank.of.Member)
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Year","Rank.of.Member", "Division")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```


**Interpretation.**  
These preliminary results show that both racial appearance and where a search occurs have a statistically significant effect on hit rates. 


# Modelling racial disparities in search outcomes


To more precisely assess whether the lower hit rates observed for African-appearance individuals persist after accounting for contextual differences, we fitted a sequence of multilevel (hierarchical) logistic regression models that progressively adjust for geography, time, and policing environment.  

We began by controlling for **LGA** and **Year** to account for where and when searches occurred, then sequentially added controls for **local drug-crime rates** and **police station**. 

This stepwise approach allows us to see how much of the apparent racial disparity is explained by contextual factors versus how much remains once those are held constant.

## Model 1: Multilevel model with random intercepts and slopes by LGA

To assess whether the lower hit rates for African-appearance individuals persist after adjusting for geography and time, we fitted a multilevel logistic regression model with **random intercepts for both LGA and Year** and a **random slope for Race within LGA**.  

This allows each LGA to have its own baseline hit rate and its own local race disparity, while estimating an overall statewide effect.  

The model therefore tests whether the racial disparity in search outcomes is consistent across LGAs or varies from place to place.



```{r racial_hit_multilevel_all, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

# ===================== Minimal helpers (self-contained) =====================
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes", "y", "true", "1")
}

fmt_p <- function(p) {
  ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

have_kx <- requireNamespace("kableExtra", quietly = TRUE)

kbl_print <- function(df, caption, align = NULL) {
  if (have_kx) {
    kableExtra::kbl(df, caption = caption, align = align) %>%
      kableExtra::kable_paper(full_width = FALSE) %>%
      kableExtra::kable_styling(
        bootstrap_options = c("condensed", "hover"),
        position = "left"
      ) %>%
      print()
  } else {
    print(knitr::kable(df, caption = caption, align = align))
  }
}

# ===================== Sanity checks =====================
needed_cols <- c("Local.Government.Area", "Racial.appearance.transformed", "Found", "Year")
missing <- setdiff(needed_cols, names(ddatr))

if (length(missing)) {
  kbl_print(
    data.frame(
      Error   = "Missing required columns in ddatr",
      Columns = paste(missing, collapse = ", ")
    ),
    caption = "Model could not be fitted"
  )

} else {

  # ===================== Prepare data =====================
  dat <- ddatr %>%
    dplyr::filter(
      !is.na(Local.Government.Area),
      Racial.appearance.transformed %in% c("White", "African"),
      !is.na(Found)
    ) %>%
    dplyr::mutate(
      Found_yes = as.integer(is_yes(Found)),
      Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
      LGA       = factor(Local.Government.Area),
      Year      = factor(Year)
    )

  if (nrow(dat) < 2L || dplyr::n_distinct(dat$Found_yes) < 2L) {

    kbl_print(
      data.frame(Message = "Not enough variation to fit the model."),
      caption = "Model could not be fitted"
    )

  } else {

    # ===================== Fit models =====================
    ctrl <- lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

    # Full model: random slope for Race by LGA
    m_best <- lme4::glmer(
      Found_yes ~ Race + (1 + Race | LGA) + (1 | Year),
      family  = binomial,
      data    = dat,
      control = ctrl
    )

    # Reduced model: no random slope for Race
    m_no_slope <- lme4::glmer(
      Found_yes ~ Race + (1 | LGA) + (1 | Year),
      family  = binomial,
      data    = dat,
      control = ctrl
    )

    # ===================== Fixed effects =====================
    fx <- broom.mixed::tidy(m_best, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) %>%
      dplyr::filter(term %in% c("(Intercept)", "RaceAfrican")) %>%
      dplyr::transmute(
        Term       = term,
        Odds_Ratio = sprintf("%.3f", estimate),
        CI_L       = sprintf("%.3f", conf.low),
        CI_U       = sprintf("%.3f", conf.high),
        p_value    = fmt_p(p.value)
      )

    cat("### City-wide race disparity (fixed effect)\n\n")

    kbl_print(
      fx,
      caption = "Fixed effects (odds ratios, 95% CI, p-value)",
      align   = c("l", "r", "r", "r", "r")
    )

    # ===================== Likelihood-ratio test =====================
   # =============== Likelihood-ratio test (random slope vs intercept) ===============
lrt <- suppressWarnings(anova(m_no_slope, m_best, test = "LRT"))
lrt_df <- as.data.frame(lrt)

has <- function(nm) nm %in% names(lrt_df)
npar   <- if (has("npar"))   lrt_df$npar   else rep(NA_real_, nrow(lrt_df))
logLik <- if (has("logLik")) lrt_df$logLik else rep(NA_real_, nrow(lrt_df))
AICv   <- if (has("AIC"))    lrt_df$AIC    else rep(NA_real_, nrow(lrt_df))
BICv   <- if (has("BIC"))    lrt_df$BIC    else rep(NA_real_, nrow(lrt_df))
chisq  <- if (has("Chisq"))  lrt_df$Chisq  else if (length(logLik) == 2) 2 * (logLik[2] - logLik[1]) else NA_real_
chidf  <- if (has("Chi Df")) lrt_df$`Chi Df` else if (length(npar) == 2 && !any(is.na(npar))) npar[2] - npar[1] else NA_real_
pval   <- if (has("Pr(>Chisq)")) lrt_df$`Pr(>Chisq)` else if (!any(is.na(c(chisq, chidf))) && chidf >= 0) stats::pchisq(chisq, df = chidf, lower.tail = FALSE) else NA_real_

# ---- take the full-model row as scalars ----
chisq_val <- if (length(chisq) >= 2) chisq[2] else chisq[1]
chidf_val <- if (length(chidf) >= 2) chidf[2] else chidf[1]
pval_val  <- if (length(pval)  >= 2) pval[2]  else pval[1]

dash <- "-"

lrt_out <- dplyr::tibble(
  Model      = c("Random intercepts only (Race fixed across LGAs)",
                 "Random slope for Race by LGA"),
  npar       = npar,
  logLik     = ifelse(is.na(logLik), "", sprintf("%.1f", logLik)),
  AIC        = ifelse(is.na(AICv),   "", sprintf("%.1f", AICv)),
  BIC        = ifelse(is.na(BICv),   "", sprintf("%.1f", BICv)),
  Delta_Chi2 = c(dash, if (!is.na(chisq_val)) sprintf("%.3f", chisq_val) else dash),
  Delta_df   = c(dash, if (!is.na(chidf_val)) as.character(chidf_val) else dash),
  p_value    = c(dash, fmt_p(pval_val))
)

    # ===================== Per-LGA shrunken ORs =====================
    re_list <- lme4::ranef(m_best, condVar = TRUE)
    re_LGA  <- re_list$LGA
    post    <- attr(re_LGA, "postVar")
    cn      <- colnames(re_LGA)
    j       <- which(cn == "RaceAfrican")

    if (length(j) == 0L || is.null(post)) {

      kbl_print(
        data.frame(
          Message = "Model did not retain a random slope for Race at LGA level (likely singular fit). Showing statewide effect only."
        ),
        caption = "Per-LGA race disparity"
      )

    } else {

      race_fix <- lme4::fixef(m_best)["RaceAfrican"]
      nLGA <- dim(post)[3]
      se   <- sqrt(vapply(seq_len(nLGA), function(i) post[j, j, i], FUN.VALUE = double(1)))

      re_df <- tibble::tibble(
        LGA       = rownames(re_LGA),
        RaceSlope = re_LGA[, j, drop = TRUE]
      ) %>%
        dplyr::mutate(
          logOR_total = race_fix + RaceSlope,
          OR          = exp(logOR_total),
          OR_lower    = exp(logOR_total - 1.96 * se),
          OR_upper    = exp(logOR_total + 1.96 * se),
          sig_flag    = dplyr::case_when(
            OR_lower > 1 ~ "Higher hit rate (Afr > White)",
            OR_upper < 1 ~ "Lower hit rate (Afr < White)",
            TRUE         ~ "No significant difference"
          )
        ) %>%
        dplyr::arrange(OR)

      # (Optional preview)
      # kbl_print(head(re_df, 10), caption = "Per-LGA race disparity (top 10 by OR)")
    }
  }
}


```

The model indicates that, after accounting for clustering by LGA and Year, searches of African-appearance individuals are **significantly less likely to find contraband** than searches of White-appearance individuals (odds ratio ≈ 0.77, p = 0.036).  

A likelihood-ratio test comparing this model to a simpler random-intercept version showed **no improvement in fit when allowing the race effect to vary across LGAs** (χ² = 0.005, p = 0.998).  

This suggests that the racial disparity is **consistent statewide**, rather than concentrated in particular areas.  
Per-LGA “shrunken” estimates confirm that almost all LGAs show odds ratios below 1, indicating uniformly lower hit rates for African-appearance individuals across Victoria.

Overall, the multilevel results demonstrate that the racial disparity in search hit rates remains robust after adjusting for both geographic and temporal clustering, with no evidence that the magnitude of the disparity differs meaningfully between LGAs.

## Model 2: LGA × Year random-intercept model

Model 1 used separate random intercepts for LGA and Year. As a robustness check, we replace those with a single random intercept for each LGA–Year combination, allowing every area in every year to have its own baseline hit rate. If this structure captures additional variance, we should see an improvement in model fit (lower AIC and a significant LRT).

```{r racial_hit_multilevel_nested no rate, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
# --------- Helpers ----------
if (!exists("fmt_p")) fmt_p <- function(p)
  ifelse(is.na(p), "", ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

if (!exists("is_yes")) is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes", "y", "true", "1")
}

# --------- Data prep ----------
if (!exists("dat")) {
  dat <- ddatr |>
    dplyr::filter(
      !is.na(Local.Government.Area),
      Racial.appearance.transformed %in% c("White", "African"),
      !is.na(Found)
    ) |>
    dplyr::mutate(
      Found_yes = as.integer(is_yes(Found)),
      Race = factor(Racial.appearance.transformed, levels = c("White", "African")),
      LGA  = factor(Local.Government.Area),
      Year = factor(Year)
    )
}

if (nrow(dat) < 2L || dplyr::n_distinct(dat$Found_yes) < 2L) {
  knitr::kable(
    data.frame(Message = "Not enough variation to fit Model 1b."),
    caption = "Model 1b could not be fitted"
  ) |> print()
} else {

  ctrl <- lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))

  # --- Model A: additive LGA + Year random intercepts
  m_additive <- lme4::glmer(
    Found_yes ~ Race + (1 | LGA) + (1 | Year),
    family = binomial, data = dat, control = ctrl
  )

  # --- Model B: combined LGA×Year random intercept
  m_lgayear <- lme4::glmer(
    Found_yes ~ Race + (1 | LGA:Year),
    family = binomial, data = dat, control = ctrl
  )

  # --- Compare fits (LRT + AIC)
  cmp <- suppressWarnings(anova(m_additive, m_lgayear, test = "LRT"))
  cmp_df <- as.data.frame(cmp)

  chisq_val <- cmp_df$Chisq[2]
  chidf_val <- cmp_df$`Chi Df`[2]
  pval_val  <- cmp_df$`Pr(>Chisq)`[2]

results <- dplyr::tibble(
  Model   = c("(1|LGA)+(1|Year)", "(1|LGA:Year)"),
  logLik  = sprintf("%.1f", cmp_df$logLik),
  AIC     = sprintf("%.1f", cmp_df$AIC),
  BIC     = sprintf("%.1f", cmp_df$BIC),
  Delta_Chi2 = c("-", sprintf("%.3f", chisq_val)),
  Delta_df   = c("-", as.character(chidf_val)),
  p_value    = c("-", fmt_p(pval_val))
)


  knitr::kable(
    results,
    caption = "Model 1b: Comparing baseline structures (separate LGA+Year vs combined LGA×Year)",
    align = "lrrrrrr"
  ) |> print()

  # --- Fixed effect under preferred baseline
  m_pref <- if (AIC(m_lgayear) < AIC(m_additive)) m_lgayear else m_additive
  fx <- broom.mixed::tidy(m_pref, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
    dplyr::filter(term %in% c("(Intercept)", "RaceAfrican")) |>
    dplyr::transmute(
      Term = term,
      Odds_Ratio = sprintf("%.3f", estimate),
      CI_L = sprintf("%.3f", conf.low),
      CI_U = sprintf("%.3f", conf.high),
      p_value = fmt_p(p.value)
    )

  knitr::kable(
    fx,
    caption = "Race (fixed effect) under the preferred baseline structure",
    align = "lrrrr"
  ) |> print()
}


```
Introducing the local drug-crime rate does not materially alter the estimated racial disparity. The coefficient for Race = African remains statistically significant (OR ≈ 0.77, p = 0.036), while the crime-rate covariate itself is non-significant (p = 0.41). This suggests that area-level variation in drug offences does not account for the lower hit rate among African-appearance individuals. The modest random-effect variance at the LGA–Year level indicates that most of the remaining variability lies within rather than between area–year contexts.

## Model 3: LGA × Year random-intercept model with local crime context  

To allow for sharper local and temporal variation, Model 3 introduces a **random intercept for each LGA–Year combination**, allowing each LGA–Year combination its own baseline hit rate.

We also include the **local drug-crime rate** as a continuous covariate to capture contextual differences in policing intensity and local offending environments.   

This model therefore isolates racial disparities within the same place, year, and crime-context level.


```{r racial_hit_multilevel_nested year, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}

# ----------------- Helpers ------------------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
to_utf8 <- function(df) {
  dplyr::mutate(df, dplyr::across(where(is.character),
                                  ~ enc2utf8(replace(., is.na(.), ""))))
}

# ----------------- Prep data ----------------
dat <- ddatr |>
  dplyr::filter(
    !is.na(Local.Government.Area),
    !is.na(Found),
    Racial.appearance.transformed %in% c("White","African")
  ) |>
  dplyr::transmute(
    Found_yes = as.integer(is_yes(Found)),
    Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
    LGA       = factor(Local.Government.Area),
    Year      = factor(Year),
    Drug.crime.rate = suppressWarnings(as.numeric(Drug.crime.rate)) # keep it!
  )

stopifnot(dplyr::n_distinct(dat$Found_yes, na.rm = TRUE) == 2L)

# Decide whether we can use Drug.crime.rate
use_drug <- ("Drug.crime.rate" %in% names(dat)) && !all(is.na(dat$Drug.crime.rate))

form <- if (use_drug) {
  Found_yes ~ Race + scale(Drug.crime.rate) + (1 | LGA:Year)
} else {
  message("Note: Drug.crime.rate missing or all NA; fitting without it.")
  Found_yes ~ Race + (1 | LGA:Year)
}

# --------------- Fit model -----------------
m_lgayear <- lme4::glmer(
  form, family = binomial, data = dat,
  control = lme4::glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

# --------------- Fixed effects (OR) -----------------
fe <- broom.mixed::tidy(m_lgayear, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::transmute(
    Term        = term,
    `Odds Ratio`= sprintf("%.3f", estimate),
    `95% CI (L)`= sprintf("%.3f", conf.low),
    `95% CI (U)`= sprintf("%.3f", conf.high),
    `p-value`   = dplyr::case_when(is.na(p.value) ~ "",
                                   p.value < 0.001 ~ "<0.001",
                                   TRUE ~ sprintf("%.3f", p.value))
  ) |>
  to_utf8()

cap_fe <- "LGA × Year random-intercept model: Fixed effects (odds ratios, 95% CI, p-value)"

if (knitr::is_html_output() || knitr::is_latex_output()) {
  fe |>
    knitr::kable(caption = cap_fe, align = c("l","r","r","r","r")) |>
    kableExtra::kable_paper(full_width = FALSE) |>
    kableExtra::kable_styling(bootstrap_options = c("striped","hover"), position = "left") |>
    print()
} else {
  print(knitr::kable(fe, caption = cap_fe, align = c("l","r","r","r","r")))
}

# --------------- Fit stats -----------------
fit_summary <- data.frame(
  AIC = AIC(m_lgayear),
  BIC = BIC(m_lgayear),
  LogLik = as.numeric(logLik(m_lgayear)),
  Deviance = stats::deviance(m_lgayear),
  check.names = FALSE
) |>
  to_utf8()

cap_fit <- "Model fit: LGA × Year random-intercept model"

if (knitr::is_html_output() || knitr::is_latex_output()) {
  fit_summary |>
    knitr::kable(caption = cap_fit, align = c("r","r","r","r")) |>
    kableExtra::kable_paper(full_width = FALSE) |>
    kableExtra::kable_styling(bootstrap_options = c("condensed","hover"), position = "left") |>
    print()
} else {
  print(knitr::kable(fit_summary, caption = cap_fit, align = c("r","r","r","r")))
}

# --------------- Random effects -----------------
re_var <- as.data.frame(lme4::VarCorr(m_lgayear)) |>
  dplyr::transmute(
    Group = grp,
    Variance = round(vcov, 4),
    SD = round(sdcor, 4)
  ) |>
  to_utf8()

cap_re <- "Random-effects variance components (LGA × Year model)"

if (knitr::is_html_output() || knitr::is_latex_output()) {
  re_var |>
    knitr::kable(caption = cap_re, align = c("l","r","r")) |>
    kableExtra::kable_paper(full_width = FALSE) |>
    kableExtra::kable_styling(bootstrap_options = c("condensed","hover"), position = "left") |>
    print()
} else {
  print(knitr::kable(re_var, caption = cap_re, align = c("l","r","r")))
}

```

Introducing the local drug-crime rate does not materially alter the estimated racial disparity. The coefficient for Race = African remains statistically significant (OR ≈ 0.77, p = 0.036), while the crime-rate covariate itself is non-significant (p = 0.41). This suggests that area-level variation in drug offences does not account for the lower hit rate among African-appearance individuals. The modest random-effect variance at the LGA–Year level indicates that most of the remaining variability lies within rather than between area–year contexts.

## Model 4: LGA × Year + Station random-intercept model  

While Model 3 controls for geographic and temporal clustering at the LGA–Year level, it does not yet account for differences between police stations operating within the same area. 

Stations vary in staffing, priorities, and local enforcement practices, all of which can affect the likelihood that a search finds contraband.

To capture this layer of variation, we extend the model to include a **random intercept for each Station**, allowing baseline hit rates to differ between stations within the same LGA–Year combination. 

The specification is:

\[
\text{logit}(P(\text{Hit})) =
\beta_0 + \beta_1(\text{RaceAfrican}) + \beta_2(\text{scale(Drug.crime.rate)}) +
u_{\text{LGA:Year}} + v_{\text{Station}},
\]
where \(u_{\text{LGA:Year}}\) represents unobserved area–year differences, 
and \(v_{\text{Station}}\) captures station-level deviations from those baselines.

This structure makes comparisons even more conservative: it contrasts searches of African- and White-appearance individuals **within the same year, same LGA, and same station context**, holding local crime rates constant.


```{r racial_hit_multilevel_station, echo=FALSE, warning=FALSE, message=FALSE}
# ================= Packages (quiet) =================

suppressPackageStartupMessages({
library(dplyr)
library(lme4)
library(broom.mixed)
library(knitr)
have_kx <- requireNamespace("kableExtra", quietly = TRUE)
library(rlang)   # for .data pronoun
})

# ================= Helpers =================

is_yes <- function(x) {
if (is.logical(x)) return(!is.na(x) & x)
tolower(as.character(x)) %in% c("yes", "y", "true", "1")
}

to_utf8 <- function(df) {
dplyr::mutate(
df,
dplyr::across(
where(is.character),
~ enc2utf8(replace(., is.na(.), ""))
)
)
}

# ================= Build analysis frame (with Station) =================

station_candidates <- c("Station.uniform", "Station", "Station_Name", "Station.name")
station_col <- station_candidates[station_candidates %in% names(ddatr)]
if (length(station_col) == 0L) {
stop("No station column found. Tried: ", paste(station_candidates, collapse = ", "))
}
station_col <- station_col[[1]]

dat_station <- ddatr |>
dplyr::filter(
!is.na(Local.Government.Area),
!is.na(Found),
Racial.appearance.transformed %in% c("White", "African")
) |>
dplyr::transmute(
Found_yes        = as.integer(is_yes(Found)),
Race             = factor(Racial.appearance.transformed, levels = c("White", "African")),
LGA              = factor(Local.Government.Area),
Year             = factor(Year),
Station          = factor(.data[[station_col]]),
Drug.crime.rate  = suppressWarnings(as.numeric(Drug.crime.rate))
)

# Sanity checks

stopifnot(dplyr::n_distinct(dat_station$Found_yes, na.rm = TRUE) == 2L)
stopifnot(nlevels(dat_station$Race) == 2L)

# ================= Model formula =================

use_drug <- ("Drug.crime.rate" %in% names(dat_station)) && !all(is.na(dat_station$Drug.crime.rate))

form_station <- if (use_drug) {
Found_yes ~ Race + scale(Drug.crime.rate) + (1 | LGA:Year) + (1 | Station)
} else {
message("Note: Drug.crime.rate missing or all NA; fitting without it.")
Found_yes ~ Race + (1 | LGA:Year) + (1 | Station)
}

# ================= Fit model =================

m_station <- lme4::glmer(
form_station,
family  = binomial,
data    = dat_station,
control = lme4::glmerControl(
optimizer = "bobyqa",
optCtrl   = list(maxfun = 2e5)
)
)

# ================= Fixed effects (odds ratios) =================

fe_station <- broom.mixed::tidy(
m_station,
effects      = "fixed",
conf.int     = TRUE,
exponentiate = TRUE
) |>
dplyr::transmute(
Term          = term,
`Odds Ratio`  = sprintf("%.3f", estimate),
`95% CI (L)`  = sprintf("%.3f", conf.low),
`95% CI (U)`  = sprintf("%.3f", conf.high),
`p-value`     = dplyr::case_when(
is.na(p.value)   ~ "",
p.value < 0.001  ~ "<0.001",
TRUE             ~ sprintf("%.3f", p.value)
)
) |>
to_utf8()

cap_fe <- "Model 3 – LGA×Year + Station random intercepts: Fixed effects (odds ratios, 95% CI, p-value)"

if (knitr::is_html_output() || knitr::is_latex_output()) {
tbl <- knitr::kable(fe_station, caption = cap_fe, align = c("l", "r", "r", "r", "r"))
if (have_kx) {
tbl |>
kableExtra::kable_paper(full_width = FALSE) |>
kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), position = "left") |>
print()
} else {
print(tbl)
}
} else {
print(knitr::kable(fe_station, caption = cap_fe, align = c("l", "r", "r", "r", "r")))
}

# ================= Fit statistics =================

fit_station <- data.frame(
AIC       = AIC(m_station),
BIC       = BIC(m_station),
LogLik    = as.numeric(logLik(m_station)),
Deviance  = stats::deviance(m_station),
check.names = FALSE
) |>
to_utf8()

cap_fit <- "Model 3 – Fit statistics (random intercepts for LGA×Year and Station)"

if (knitr::is_html_output() || knitr::is_latex_output()) {
tbl <- knitr::kable(fit_station, caption = cap_fit, align = c("r", "r", "r", "r"))
if (have_kx) {
tbl |>
kableExtra::kable_paper(full_width = FALSE) |>
kableExtra::kable_styling(bootstrap_options = c("condensed", "hover"), position = "left") |>
print()
} else {
print(tbl)
}
} else {
print(knitr::kable(fit_station, caption = cap_fit, align = c("r", "r", "r", "r")))
}

# ================= Random-effects variance & SD =================

re_var_station <- as.data.frame(lme4::VarCorr(m_station)) |>
dplyr::transmute(
Group    = grp,
Term     = var1,
Variance = round(vcov, 4),
SD       = round(sdcor, 4)
) |>
to_utf8()

cap_re <- "Random-effects variance and SD by group (LGA×Year and Station)"

if (knitr::is_html_output() || knitr::is_latex_output()) {
tbl <- knitr::kable(re_var_station, caption = cap_re, align = c("l", "l", "r", "r"))
if (have_kx) {
tbl |>
kableExtra::kable_paper(full_width = FALSE) |>
kableExtra::kable_styling(bootstrap_options = c("condensed", "hover"), position = "left") |>
print()
} else {
print(tbl)
}
} else {
print(knitr::kable(re_var_station, caption = cap_re, align = c("l", "l", "r", "r")))
}

```

Even after adjusting for local crime rates, LGA–Year differences, and between-station variation, the odds ratio for Race = African remains around 0.75 (p ≈ 0.02).  

This means that, within the same local and organisational context, searches of African-appearance individuals are about **25 % less likely** to yield contraband than those of White-appearance individuals.  

Adding the station-level random effect improves model fit slightly (AIC ≈ 5987 vs 6008 for Model 2) but does not change the substantive conclusion:   the racial disparity in hit rates is **consistent and robust** to every adjustment for geography, time, and operational clustering.

```{r prop_test_division_station_sig, message=FALSE, warning=FALSE, echo = FALSE}
alpha <- 0.05

div_tbl_all <- make_hit_table(
  data = ddatr,
  group_var = "Division",
  alpha = alpha,
  min_african_total = 1,      # keep any division with ≥1 African search
  filter_sig_only = FALSE,    # set TRUE if you want to show only p < 0.05
  add_fdr = TRUE
)

print_hit_table(
  div_tbl_all,
  caption = sprintf("Divisions — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
  show_bh = TRUE
)
```
---

## Station-Level Findings  

- **Werribee Station** also displayed a significant disparity:  
  - African: **5/57 (8.8%)** vs White: **22/103 (21.4%)**  
  - Risk difference = –12.6%, Risk ratio = 0.41, *p* = 0.042  
- Werribee is in the same outer-western growth corridor as Brimbank/Melton, with a **substantial African community** and a history of **public concern about over-policing**.  

```{r stations-hitrate-with-prop, echo=FALSE, message=FALSE, warning=FALSE}
alpha <- 0.05

sta_tbl <- make_hit_table(
  data = ddatr,
  group_var = "Station.uniform",
  alpha = alpha,
  min_african_total = 5,      # inclusion rule you requested
  filter_sig_only = TRUE,     # show only p < 0.05; set FALSE to see all included
  add_fdr = TRUE
)

print_hit_table(
  sta_tbl,
  caption = sprintf("Stations (≥ 5 African searches) — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
  show_bh = TRUE
)

```

## Interpretation  

The proportion tests suggest that disparities are **geographically concentrated** in Melbourne’s **outer west**, where disadvantage and demographic change intersect with intensive policing.  
Other Divisions and Stations did not show statistically significant differences, but many had small African sample sizes. Absence of significance does not mean absence of disparity. 

These results highlight the need for multivariate modelling to control for confounders such as **gender, year, and local policing context**.  

# Statistical Modelling  


Logistic regression estimates the odds that a search will result in contraband being found, while adjusting for other variables. This allows us to move beyond descriptive comparisons and ask whether observed differences reflect systematic disparities in how suspicion is applied.  

## Division-Level Logistic Regression Models  

The proportion tests showed that disparities in hit rates were not uniform across Melbourne, with some divisions and stations suggesting stronger evidence of racial disparity than others.  

To explore this further, we estimated **logistic regression models separately for each Local Government Area.**  

The outcome is whether **any contraband was found** (`Found = Yes`). Each model includes predictors for:  

- **Race** (African vs White)  
- **Gender** (Female vs Male)   
- An **interaction between race and gender**  
- Local Government Area - recognising that 
- Year: recognising that search intensity may vary from year to year.

Summary tables models where the coefficient of racial appearance was lower than 0.05 appear below.  



```{r lga-models-optionA-minAfr, echo=FALSE, warning=FALSE, message=FALSE}


# ---------------- params ----------------
MIN_AFRICAN <- 20   # adjust if needed

# -------------- helpers -----------------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --------- prepare data ----------
dat_lga <- ddatr %>%
  filter(
    !is.na(Local.Government.Area),
    Racial.appearance.transformed %in% c("African","White"),
    !is.na(Found),
    Gender %in% c("M","F")
  ) %>%
  transmute(
    LGA     = factor(Local.Government.Area),
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F")),
    Year    = Year
  )

# Context for captions (counts by LGA & race)
lga_counts <- dat_lga %>%
  group_by(LGA, Race) %>%
  summarise(succ = sum(Success), tot = n(), .groups = "drop") %>%
  pivot_wider(names_from = Race, values_from = c(succ, tot), values_fill = 0) %>%
  mutate(n_total = tot_African + tot_White)

# Eligibility: enforce minimum African sample size
eligible_lgas <- dat_lga %>%
  count(LGA, Race) %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(LGA) %>%
  as.character()

# --------- fit per-LGA models ----------
data_by_lga <- split(dat_lga, dat_lga$LGA)

walk(eligible_lgas, function(lga_name) {
  df <- data_by_lga[[lga_name]]

  # skip if only one outcome class
  if (n_distinct(df$Success) < 2L) return(invisible(NULL))

  # Model: Race * Gender + Year (fixed effects)
  mdl <- tryCatch(
    glm(Success ~ Race * Gender + factor(Year), data = df, family = binomial),
    error = function(e) NULL
  )
  if (is.null(mdl)) return(invisible(NULL))

  td_raw <- tryCatch(
    tidy(mdl, exponentiate = TRUE, conf.int = TRUE),
    error = function(e) {
      tidy(mdl, exponentiate = TRUE, conf.int = FALSE) %>%
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    }
  ) %>%
    select(term, estimate, conf.low, conf.high, p.value)

  # show only if any non-intercept term significant
  keep_if_sig <- any(td_raw$term != "(Intercept)" & td_raw$p.value < 0.05, na.rm = TRUE)
  if (!keep_if_sig) return(invisible(NULL))

  # Friendly labels
  td <- td_raw %>%
    mutate(
      Term = case_when(
        term == "(Intercept)" ~ "(Intercept)",
        term == "RaceAfrican" ~ "Race: African vs White (Male ref.)",
        term == "GenderF" ~ "Gender: Female vs Male",
        term == "RaceAfrican:GenderF" ~ "Interaction: African × Female",
        str_detect(term, "^factor\\(Year\\).+") ~
          paste0("Year: ", str_replace(term, "^factor\\(Year\\)", ""), " vs ref."),
        TRUE ~ term
      ),
      `Odds Ratio`      = estimate,
      `95% CI (Lower)`  = conf.low,
      `95% CI (Upper)`  = conf.high,
      `p-value`         = p.value
    ) %>%
    select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
    mutate(
      `p-value` = fmt_p(`p-value`),
      `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
      `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
      `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`))
    )

  # Caption: n and years
  cx <- filter(lga_counts, LGA == lga_name)
  yrs <- sort(unique(stats::na.omit(df$Year)))
  years_txt <- if (length(yrs) == 0) NULL else if (length(yrs) == 1) {
    paste0("Year: ", yrs)
  } else {
    paste0("Years: ", min(yrs), "–", max(yrs))
  }

  cap_bits <- c(
    paste0("n=", ifelse(nrow(cx) == 1, cx$n_total, NA_integer_)),
    if (nrow(cx) == 1) paste0("Afr: ", cx$succ_African, "/", cx$tot_African) else NULL,
    if (nrow(cx) == 1) paste0("Wh: ",  cx$succ_White,   "/", cx$tot_White)   else NULL,
    years_txt
  )
  cap_ctx <- paste(cap_bits[!is.na(cap_bits)], collapse = " · ")

  # Print table
  td %>%
    kbl(
      caption = paste0("LGA: ", lga_name,
                       " — Logistic regression of search success (any contraband). ",
                       cap_ctx),
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","condensed","hover"),
                  position = "left") %>%
    column_spec(1, width = "22em") %>%
    add_footnote(
      label = paste0(
        "Model: Success ~ Race * Gender + Year (fixed effects). ",
        "Odds ratios shown (95% CI). Reference group = White males and earliest Year. ",
        "Eligibility: African count per LGA ≥ ", MIN_AFRICAN, "."
      ),
      notation = "symbol"
    ) %>%
    print()
})
```


```{r station-models-optionA-minAfr, echo=FALSE, warning=FALSE, message=FALSE}



# ---------------- params ----------------
MIN_AFRICAN <- 20   # adjust if needed

# -------------- helpers -----------------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# --------- prepare data ----------
dat_sta <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("African","White"),
    !is.na(Found),
    Gender %in% c("M","F")
  ) %>%
  transmute(
    Station = factor(Station.uniform),
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F")),
    Year    = Year
  )

# Context for captions (counts by Station & race)
sta_counts <- dat_sta %>%
  group_by(Station, Race) %>%
  summarise(succ = sum(Success), tot = n(), .groups = "drop") %>%
  pivot_wider(names_from = Race, values_from = c(succ, tot), values_fill = 0) %>%
  mutate(n_total = tot_African + tot_White)

# Eligibility: enforce minimum African sample size per Station
eligible_stations <- dat_sta %>%
  count(Station, Race) %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  filter(African >= MIN_AFRICAN) %>%
  pull(Station) %>%
  as.character()

# --------- fit per-Station models ----------
data_by_sta <- split(dat_sta, dat_sta$Station)

walk(eligible_stations, function(sta_name) {
  df <- data_by_sta[[sta_name]]

  # skip if only one outcome class
  if (n_distinct(df$Success) < 2L) return(invisible(NULL))

  # Model: Race * Gender + Year (fixed effects)
  mdl <- tryCatch(
    glm(Success ~ Race * Gender + factor(Year), data = df, family = binomial),
    error = function(e) NULL
  )
  if (is.null(mdl)) return(invisible(NULL))

  td_raw <- tryCatch(
    tidy(mdl, exponentiate = TRUE, conf.int = TRUE),
    error = function(e) {
      tidy(mdl, exponentiate = TRUE, conf.int = FALSE) %>%
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    }
  ) %>%
    select(term, estimate, conf.low, conf.high, p.value)

  # show only if any non-intercept term significant
  keep_if_sig <- any(td_raw$term != "(Intercept)" & td_raw$p.value < 0.05, na.rm = TRUE)
  if (!keep_if_sig) return(invisible(NULL))

  # Friendly labels
  td <- td_raw %>%
    mutate(
      Term = case_when(
        term == "(Intercept)" ~ "(Intercept)",
        term == "RaceAfrican" ~ "Race: African vs White (Male ref.)",
        term == "GenderF" ~ "Gender: Female vs Male (White ref.)",
        term == "RaceAfrican:GenderF" ~ "Interaction: African × Female",
        str_detect(term, "^factor\\(Year\\).+") ~
          paste0("Year: ", str_replace(term, "^factor\\(Year\\)", ""), " vs ref."),
        TRUE ~ term
      ),
      `Odds Ratio`      = estimate,
      `95% CI (Lower)`  = conf.low,
      `95% CI (Upper)`  = conf.high,
      `p-value`         = p.value
    ) %>%
    select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
    mutate(
      `p-value` = fmt_p(`p-value`),
      `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
      `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
      `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`))
    )

  # Caption: n and years
  cx <- filter(sta_counts, Station == sta_name)
  yrs <- sort(unique(stats::na.omit(df$Year)))
  years_txt <- if (length(yrs) == 0) NULL else if (length(yrs) == 1) {
    paste0("Year: ", yrs)
  } else {
    paste0("Years: ", min(yrs), "–", max(yrs))
  }

  cap_bits <- c(
    paste0("n=", ifelse(nrow(cx) == 1, cx$n_total, NA_integer_)),
    if (nrow(cx) == 1) paste0("Afr: ", cx$succ_African, "/", cx$tot_African) else NULL,
    if (nrow(cx) == 1) paste0("Wh: ",  cx$succ_White,   "/", cx$tot_White)   else NULL,
    years_txt
  )
  cap_ctx <- paste(cap_bits[!is.na(cap_bits)], collapse = " · ")

  # Print table
  td %>%
    kbl(
      caption = paste0("Station: ", sta_name,
                       " — Logistic regression of search success (any contraband). ",
                       cap_ctx),
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","condensed","hover"),
                  position = "left") %>%
    column_spec(1, width = "22em") %>%
    add_footnote(
      label = paste0(
        "Model: Success ~ Race * Gender + Year (fixed effects). ",
        "Odds ratios shown (95% CI). Reference group = White males and earliest Year. ",
        "Eligibility: African count per Station ≥ ", MIN_AFRICAN, "."
      ),
      notation = "symbol"
    ) %>%
    print()
})
```

# Appendix  

**Table X Hit rates by largest racial grouping**

```{r Finds racial all, echo=FALSE, warning=FALSE, results='asis'}
ddats<-ddat %>%
  filter(Racial.appearance.transformed %in% c("Missing", "White","Mediterranean/Middle Eastern – Unusable",
                                              "Asian","African", "Aboriginal"))

ct <- crosstable::crosstable(
  ddats,
  by = "Found",
  cols = "Racial.appearance.transformed",
  total = TRUE,
  test = FALSE,          # we'll add our own robust test
  showNA = "ifany",
  percent_digits = 1
)

ft <- crosstable::as_flextable(ct)

# 3) Safe association test (Fisher for 2x2, else Chi-square with simulated p)
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tres <- safe_test(ddats$Racial.appearance.transformed, ddats$Found)
p_line <- sprintf("Racial.appearance.transformed — %s, p = %s",
                  tres$label, format.pval(tres$p, digits = 4))

# 4) Add footer and a clear title
ft <- flextable::add_footer_lines(ft, values = c("Overall association test:", p_line))
ft <- flextable::set_caption(ft, caption = "Racial appearance vs Found (with totals)")
ft
```

```{r Finds lga effect, echo=FALSE, warning=FALSE, results='asis'}
# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <-  ddat 
names(ddatr)
# Variables to tabulate against Found
vars_for_table <- c("Local.Government.Area")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft
```