---
title: "Racial profiling - drug searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(scales)
library(stringr)
library(tidyr)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

```


```{r hitrate-prop-helpers, echo=FALSE, message=FALSE, warning=FALSE}


is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}


make_hit_table <- function(
  data,
  group_var,                   # e.g. "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    dplyr::filter(!is.na(!!grp), !is.na(Found)) %>%
    dplyr::filter(Racial.appearance.transformed %in% c("African","White")) %>%
    dplyr::mutate(FoundYes = is_yes(Found)) %>%
    dplyr::count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    dplyr::group_by(!!grp, Racial.appearance.transformed) %>%
    dplyr::summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    dplyr::filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble::tibble())

  out <- tab %>%
    dplyr::mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ stats::prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    dplyr::select(-test) %>%
    tidyr::unnest(tidied) %>%
    dplyr::mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = dplyr::if_else(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    # Important: use !!rlang::sym(group_var) on LHS, and avoid Unicode minus in names
    dplyr::transmute(
      !!rlang::sym(group_var) := !!grp,
      `African (succ/total)`   = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`     = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`       = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`         = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr - Wh)`        = scales::percent(diff_est,         accuracy = 0.1),  # <-- ASCII hyphen
      `Risk ratio (Afr/Wh)`    = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                      = sprintf("%.3f", p.value),
      `p (BH)`                 = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                    = stars
    )

  if (filter_sig_only) {
    out <- out %>% dplyr::filter(as.numeric(stringr::str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% dplyr::arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))
}

print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) display <- display %>% dplyr::select(-`p (BH)`)

  display %>%
    knitr::kable(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kableExtra::kable_paper(full_width = FALSE) %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    kableExtra::column_spec(1, width = "24em") %>%
    kableExtra::add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}

```

# Short summary

This analysis of 7,534 warrantless drug searches carried out by **uniform officers on foot patrol in metropolitan Melbourne (2018, 2019, 2022, 2023)** provides clear evidence that individuals perceived as **African** are searched on a **lower evidentiary threshold** compared to those perceived as White.

Key findings include:

-   **Descriptive results** showed lower hit rates for African-appearing individuals across most contexts.\
-   **Proportion tests** identified particularly strong disparities in Brimbank, Melton, and Werribee.\
-   **Division level models** controlling for gender, as well operating context including year and Local Government Area (LGA) revealed disparities in\
-   **Local government area level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in
-   **Statio level models** controlling for gender, year and de facto controlling for Local Government Area revealed disparities in

These results are consistent with **racial profiling**: African-appearing individuals are more frequently subjected to searches that are less likely to yield contraband, implying that suspicion is applied more readily to them.

The magnitude of disparity varies locally, the overall pattern is systematic.

This highlights the need for greater police accountability and independent oversight, and provides an empirical foundation for community advocacy and policy reform aimed at fairer, more equitable policing in Melbourne.

### Melbourne-wide findings

-   Hit rates: White 14.1% (957/6,779) vs African 11.4% (86/755); Fisher’s exact p = 0.040.
-   Pooled logistic regression (controls: Division fixed effects, Year, Gender): OR = 0.771 (95% CI 0.596–0.985), p = 0.042 — African-appearing males had \~23% lower odds of a find than White-appearing males.
-   Multilevel logistic regression (random intercepts for Division): OR = 0.761 (95% CI 0.594–0.975), p = 0.031 — confirms the disparity after accounting for clustering by Division.

### Local hotspots

-   NWMR Division 3 (Brimbank & Melton): African 4.3% (3/70) vs White 14.2% (30/212); risk ratio = 0.30, p = 0.026.
-   Werribee Station: African 8.8% (5/57) vs White 21.4% (22/103); risk ratio = 0.41, p = 0.042.

### Limits of statistical significance

-   Most other divisions and stations did not reach statistical significance, primarily because of small African sample sizes (often only a few dozen searches). Wide confidence intervals and multiple-testing adjustments mean that many apparent disparities cannot be declared significant.
-   Non-significance does not equal absence of disparity — rather, it reflects the limited statistical power in areas with few recorded searches of African-appearing individuals.

### Interpretation

Across Melbourne, African-appearing people are consistently subjected to searches with lower hit rates and lower odds of contraband being found, even after adjusting for year and division. This pattern is strongest in Melbourne’s outer west (Brimbank, Melton, Werribee), but persists at a city-wide level. Together, these results are consistent with racial profiling: African-appearing individuals are more frequently searched on a weaker threshold of suspicion.

# Introduction

This analysis examines **7,534 drug searches conducted by uniformed officers on foot patrols in Melbourne, of white- and African-appearing people**. These encounters represent the most routine form of warrantless public-place searches, where outcomes are highly dependent on officer discretion. The racial appearance categories in scope represent two of the largest and consistently recorded racial groupings in the dataset.

The focus is on whether search outcomes suggest evidence of **racial profiling**. Specifically, we compare the **hit rates** — the proportion of searches that result in contraband being found — for two groups:

-   **White-appearing individuals**, who form the majority baseline\
-   **African-appearing individuals**, a racialised minority consistently highlighted in community concerns about disproportionate policing

Concerns about the over-policing of African-Australian communities are not new. People such as **Haile Michael**, who successfully challenged Victoria Police’s discriminatory stop-and-search practices in the landmark [**Haile Michael**](https://www.abc.net.au/news/2013-03-18/victoria-police-accused-of-racially-profiling-african-men/4578358) case, have documented and contested these patterns for more than a decade.

By restricting the dataset to this context and to these two appearance categories, we provide a clear test of whether African-appearing people are subjected to searches on a **lower evidentiary threshold**. Lower hit rates for one group, despite comparable or higher search intensity, would suggest that police suspicion is being applied unevenly — a hallmark of racial profiling.

# About the dataset (subset definition)

This dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**.

Data from **2020 and 2021** is excluded due to the atypical policing environment during the COVID-19 pandemic.

The dataset was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and covers public-place searches carried out under a range of legal powers.

For this analysis, we restrict to:

-   **Geography:** Melbourne metropolitan area\
-   **Contact type:** Foot patrol only\
-   **Unit type:** Uniform police only\
-   **Search type:** Drug searches conducted under s82 of the *Drugs, Poisons and Controlled Substances Act 1981*\
-   **Station linkage:** Only searches associated with a uniform station\
-   **Racial appearance:** White and African appearance

This restriction enables us to compare searches for the most common drug searches experienced by the public in Melbourne, and restrict the data to reliable racial appearance categories.

## What s82 allows

Under section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.\
This is the most frequently invoked search power in the dataset and represents a context where discretionary judgments by officers play a central role.

## Variables in this subset

The following variables are included in the analysis:

-   **Year** — year of the stop and search\
-   **Region / Division / PSA / Station** — police geographic hierarchy\
-   **Racial.appearance.transformed** — officer-perceived racial appearance (cleaned categories)\
-   **Gender** — officer-perceived gender of the person searched\
-   **Found** — whether any contraband was located (hit outcome)\
-   **Unit.type** — uniform officers only (subset restriction)\
-   **Rank** — officer rank (e.g. Constable, Senior Constable)\
-   **Year** — 2018, 2019, 2022, or 2023

Derived variables include:

-   **Found.Flag** — binary indicator (1 if contraband found, 0 otherwise)\
-   **Race** — restricted to White vs African

------------------------------------------------------------------------

## Data limitations

### Racial appearance categories

The original dataset for drug searches conducted by uniform officers on foot patrol numbered 13,896. Race data was missing from *20% of the records*. Several categories had very small numbers or had been recategorised over the observation period. As a result, disparity testing is restricted to **White vs African** searches, which are sufficiently large and analytically comparable. Together these racial categories account for over half (54%) of the relevant data.

```{r race prop, echo=FALSE, message=FALSE, warning=FALSE}
# Count + percent for all groups (no filtering)
tab_ra <- ddat %>%
  mutate(
    Racial.appearance.transformed =
      fct_explicit_na(Racial.appearance.transformed, "Missing")
  ) %>%
  count(Racial.appearance.transformed, name = "Count") %>%
  arrange(desc(Count)) %>%
  mutate(Percent = Count / sum(Count))

# Add total row
tab_ra_tot <- bind_rows(
  tab_ra,
  tibble(
    Racial.appearance.transformed = "Total",
    Count = sum(tab_ra$Count),
    Percent = 1
  )
)

# Format percent for display
tab_ra_tot <- tab_ra_tot %>%
  mutate(Percent = percent(Percent, accuracy = 0.1))

# Render table
kbl(
  tab_ra_tot,
  caption = "Racial appearance (all groups): count and percentage of searches",
  align = c("l", "r", "r"),
  col.names = c("Racial appearance (transformed)", "Count", "Percent")
) %>%
  kable_classic(full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Searches" = 2))

```

### Age data

Age data was missing for 42.9% of searches in the White vs African subset. Age is therefore excluded from modelling, which limits the ability to adjust for age differences across groups. We expect this would have an effect as African migrant populations in in local government areas tend to be younger than their white-appearing counterparts.

### Scope of coverage

-   The dataset covers only **warrantless stop and search events conducted in public places**. This focus is deliberate: such encounters are entirely reliant on officer discretion, since there is no warrant, prior intelligence, or judicial approval.\
-   The analysis does not include searches with warrants, searches in private settings, or informal/unrecorded encounters.\
-   Findings therefore apply only to **Melbourne metropolitan drug searches by uniform officers on foot patrol**. They do not generalise to specialist units, vehicle stops, or regional policing contexts.\
-   Because modelling is restricted to White vs African, results do not extend to other racialised groups with smaller or analytically unreliable sample sizes.

# Data Overview

## Persons searched

This section summarises the characteristics of people searched in warrantless public-place encounters by uniform officers on foot patrols (2018, 2019, 2022, 2023).\
Records with missing or unusable racial appearance are excluded to allow meaningful comparison.

-   **Racial appearance.** Most searches involved people perceived as White, with African-appearance individuals making up around one in ten of the analytic sample.\
-   **Gender.** Searches were concentrated on men, who accounted for about four-fifths of those stopped.\
-   **Age.** While children and adolescents appear in the dataset, most searches targeted adults aged 18 to 45. People aged 46 and over were searched relatively infrequently.

These patterns highlight that discretionary stop-and-search practices fall unevenly across the community, with young men and racialised groups particularly over-represented.

**Table 2 Characteristics of persons searched**

```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddatr <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(ddat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police undertaking searches

Search numbers were relatively consistent across the four years examined, with a slight peak in 2022. By rank, Constables and Senior Constables were the most frequently involved, reflecting the predominance of frontline officers in this type of activity.

**Table 3 Police and units undertaking searches**

```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Year", "Rank.of.Member", "Region",  "Division")

ddatr <- ddatr %>%
  mutate(
    Rank = case_when(
      Rank.of.Member == "SSGT" ~ "SGT",
      Rank.of.Member == "PSO"  ~ "CONST",
      TRUE           ~ Rank.of.Member
    )
  )

dat.pol <- ddatr %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')


```

# Hit Rate - cross tables

An important way to assess stop-and-search practices is through **hit rates** — the proportion of searches that result in contraband being found.

If searches are consistently based on *reasonable suspicion* in comparable spatial areas, hit rates should be similar across groups.

Where a group has consistently lower hit rates for a particular area, this may suggest that officers may be applying a **lower threshold of suspicion**, a core indicator of potential racial profiling.

This section examines hit rates in the restricted dataset of **7,534 Melbourne metropolitan drug searches by uniform police on foot patrol** (2018, 2019, 2022, 2023).

A “hit” is defined as the discovery of any prohibited item, not just the item specified under the search power.

## Hit rates - cross tables

An initial analysis of comparative hit rates across our sample of 7,453 searches was performed, looking both at:

-   whether hit rates varied by racial appearance to a degree that is statistically significant and

-   Whether hit rates varied by other factors (such as local government area, gender, etc) to a degree that may counfound our results.

-   **By racial appearance.**

    -   White-appearing people had a hit rate of **14.1%** (957 finds out of 6,779 searches).\
    -   African-appearing people had a lower hit rate of **11.4%** (86/755).\
    -   This indicates that African-appearance individuals were **less likely to be found with contraband**, despite being searched in substantial numbers.

-   **By gender.**

    -   Men had a hit rate of **14.1%** (837/5,956), compared with **13.1%** for women (206/1,578).\
    -   Searches were heavily concentrated on men, accounting for four in five cases.

-   **By age.**

    -   Differences in hit rates across ages were marginal, however there does appear to be a slight decrease in hit rates where older people are searched.

-   **By time.** Across 2018, 2019, 2022, and 2023, hit rates varied only modestly.

-   **By rank.** Constables and Senior Constables dominated search activity, and their hit rates were broadly similar.

-   **By geography.** Hit rates were lowest in the **North West Metro Region**, which includes diverse and disadvantaged areas such as Brimbank and Melbourne, compared with other parts of the city.

**Table 5 Hit rates by person characteristics**

```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}




# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddatr %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),                
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop(),
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Rank = dplyr::recode(Rank.of.Member,
                         "SSGT" = "SGT",
                         "PSO"  = "CONST",
                         .default = Rank.of.Member)
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Age group", "Year","Rank.of.Member", "Division")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```

**Interpretation.**\
These preliminary results show that both racial appearance and where a search occurs have a statistically significant effect on hit rates.

## Hit rates and time

We determined to look at whether the disparity in hit rates persisted over the years in our data - 2018, 2019, 2022 and 2023.

```{r race trend, echo=FALSE, message=FALSE, warning=FALSE}

# --- Filter to White & African and prep ---
ddatm <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African")) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1L, 0L),
    Race = factor(Racial.appearance.transformed, levels = c("White","African")),
    Year = suppressWarnings(as.integer(as.character(Year)))
  ) %>%
  filter(!is.na(Year))

# --- Yearly hit rates ---
ddat_rate <- ddatm %>%
  group_by(Race, Year) %>%
  summarise(n = n(),
            hits = sum(Found.Flag, na.rm = TRUE),
            hit_rate = hits / n,
            .groups = "drop")

# Build gap metadata
ddat_rate2 <- ddat_rate %>%
  arrange(Race, Year) %>%
  group_by(Race) %>%
  mutate(
    next_year = lead(Year),
    year_gap  = next_year - Year,
    next_hit  = lead(hit_rate),
    # run id: increments when the year jump is not 1
    runid = cumsum(dplyr::lag(Year, default = first(Year)) != Year - 1)
  ) %>%
  ungroup()

# Keep only runs that have at least 2 points (so a line can be drawn)
solid_data <- ddat_rate2 %>%
  group_by(Race, runid) %>%
  filter(n() >= 2) %>%
  ungroup()

# Dotted gaps (e.g., 2019 -> 2022)
gap_data <- ddat_rate2 %>%
  filter(!is.na(next_year), year_gap > 1)

x_breaks <- seq(min(ddat_rate2$Year), max(ddat_rate2$Year), by = 1)

ggplot(ddat_rate2, aes(x = Year, y = hit_rate, colour = Race)) +
  # solid lines for contiguous sequences (2018–2019 and 2022–2023)
  geom_line(
    data = solid_data,
    aes(group = interaction(Race, runid)),
    linewidth = 1
  ) +
  # dotted neutral bridge across missing years (2019 → 2022)
  geom_segment(
    data = gap_data,
    aes(x = Year, y = hit_rate, xend = next_year, yend = next_hit),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    colour = "grey50"
  ) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +  # no fixed limits
  scale_x_continuous(breaks = x_breaks) +
  labs(
    title = "Drug Search Hit Rate by Year (Metro)",
    subtitle = "White vs African. Dotted grey line spans missing years (2020–2021).",
    x = "Year",
    y = "Hit rate (percent of searches with a find)",
    colour = "Race"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

Hit rates were persistently higher for white-appearing versus African-appearing people searched across all years, however this disparity did vary from year to year, with the most recent observations showing a small disparity.

```{r race trend year, echo=FALSE, message=FALSE, warning=FALSE}
# Summarise into wide format: one column per race
hit_table <- ddat_rate %>%
  select(Race, Year, hit_rate) %>%
  pivot_wider(names_from = Race, values_from = hit_rate) %>%
  mutate(
    Difference = (`African` - `White`) * 100,        # difference in percentage points
    across(c(`White`, `African`), ~ round(.x * 100, 2)),  # convert to percent
    Difference = round(Difference, 2)
  ) %>%
  rename(
    "Year" = Year,
    "White (%)" = `White`,
    "African (%)" = `African`,
    "Difference (African - White, pp)" = Difference
  )

# Pretty table for RMarkdown
kbl(hit_table, caption = "Hit Rate by Year and Race (Metro)") |>
  kable_paper(full_width = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    position = "left"
  )

```


# Modelling racial disparities in search outcomes

To assess whether the lower hit rates observed for African-appearance individuals persist once broader contextual factors are accounted for, we estimated a sequence of multilevel (hierarchical) logistic regression models. Each successive model introduces additional controls for geography, time, and local policing context.

This stepwise approach isolates the component of the racial disparity that remains once variation across **stations**, **years**, and **policing environments** is held constant.

---

## Model 1 - Race, Station and Year

### Model 1a: Baseline multilevel model (random intercepts)

We first fit a model including **random intercepts for Station and Year**:

\[
\text{Found\_yes} \sim \text{Race} + (1 \mid \text{Station}) + (1 \mid \text{Year})
\]

This model accounts for overall differences in hit rates across places and years, while estimating a single statewide Race effect.

The results indicate that, after controlling for clustering by Station and Year, searches of African-appearance individuals were significantly **less likely** to find contraband than those of White-appearance individuals (odds ratio ≈ 0.74, *p* = 0.016).

```{r racial_hit_rate model, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
# ================================================
#  Hit rate ~ Race + Gender + Year + z_Crime + Station (fixed effects)
# ================================================

library(fixest)
library(dplyr)
library(broom)
library(stringr)
library(kableExtra)

# ---------- prepare data ----------
dat <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    !is.na(Found),
    Racial.appearance.transformed %in% c("White","African"),
    Gender %in% c("M","F"),
    !is.na(Year)
  ) %>%
  transmute(
    Success  = as.integer(tolower(as.character(Found)) %in% c("yes","y","true","1")),
    Race     = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender   = factor(Gender, levels = c("M","F")),
    Year     = factor(Year),
    Station  = factor(Station.uniform),
    z_crime  = as.numeric(scale(suppressWarnings(as.numeric(Drug.crime.rate))))
  )

# ---------- fit model ----------
setFixest_nthreads(0)

m_all <- feglm(
  Success ~ Race + Gender + i(Year) + z_crime | Station,
  data = dat,
  family = "binomial"
)

# ---------- tidy & format output ----------
tbl <- broom::tidy(m_all, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    Term = case_when(
      term == "RaceAfrican" ~ "Race: African vs White",
      term == "GenderF"     ~ "Gender: Female vs Male",
      term == "z_crime"     ~ "Drug crime rate (z)",
      str_starts(term, "i(Year)") ~ str_replace(term, "^i\\(Year\\)", "Year: "),
      TRUE ~ term
    ),
    `Odds Ratio` = sprintf("%.3f", estimate),
    `95% CI (Lower)` = sprintf("%.3f", conf.low),
    `95% CI (Upper)` = sprintf("%.3f", conf.high),
    `p-value` = ifelse(is.na(p.value), "",
                       ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)))
  ) %>%
  select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`)

# ---------- nice table ----------
kbl(
  tbl,
  caption = "Hit rate ~ Race + Gender + Year + z_Crime + Station (fixed effects)",
  align = c("l","r","r","r","r")
) %>%
  kable_paper(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  column_spec(1, width = "25em") %>%
  add_footnote(
    label = "Model: Logistic regression (binomial link). Station absorbed as fixed effect. 
             Odds ratios with 95% CIs. Ref = White, Male, earliest Year, baseline Station.",
    notation = "symbol"
  )

```

```{r racial_hit_rate model station year intensity, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
# ======================================================
# Conditional logit (fast) with Station×Year strata
# Force efron method + cluster-robust SE by Station
# ======================================================

library(dplyr)
library(survival)
library(broom)
library(kableExtra)

# ---- build analysis frame ----
dat0 <- ddatr %>%
  filter(
    !is.na(Station.uniform),
    !is.na(Found),
    Racial.appearance.transformed %in% c("White","African"),
    Gender %in% c("M","F"),
    !is.na(Year)
  ) %>%
  transmute(
    Success  = as.integer(tolower(as.character(Found)) %in% c("yes","y","true","1")),
    Race     = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender   = factor(Gender, levels = c("M","F")),
    Station  = droplevels(factor(Station.uniform)),
    Year     = droplevels(factor(Year)),
    z_crime  = suppressWarnings(as.numeric(Drug.crime.rate)) |> scale() |> as.numeric(),
    strataID = interaction(Station, Year, drop = TRUE)  # Station×Year strata
  )

# ---- drop non-informative strata (all 0 or all 1) ----
informative_strata <- dat0 %>%
  group_by(strataID) %>%
  summarise(min_y = min(Success), max_y = max(Success), .groups = "drop") %>%
  filter(min_y == 0 & max_y == 1) %>%               # keep strata with both outcomes
  pull(strataID)

dat <- dat0 %>% filter(strataID %in% informative_strata)

# Safety check: need at least some strata left
if (n_distinct(dat$strataID) < 2) {
  stop("Not enough informative Station×Year strata after filtering to fit a conditional logit.")
}

# ---- fit conditional logit ----
# method = "efron" avoids the exact method and allows robust variance
fit <- clogit(
  Success ~ Race + Gender + z_crime + strata(strataID) + cluster(Station),
  data   = dat,
  method = "efron"
)

# ---- tidy to pretty table (odds ratios) ----
tbl <- broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
  mutate(
    term = dplyr::recode(term,
      "RaceAfrican" = "Race: African vs White",
      "GenderF"     = "Gender: Female vs Male",
      "z_crime"     = "Drug crime rate (z)"
    ),
    `Odds Ratio`      = sprintf("%.3f", estimate),
    `95% CI (Lower)`  = sprintf("%.3f", conf.low),
    `95% CI (Upper)`  = sprintf("%.3f", conf.high),
    `p-value`         = ifelse(p.value < 0.001, "<0.001", sprintf("%.3f", p.value))
  ) |>
  select(Term = term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`)

kbl(
  tbl,
  caption = "Conditional logit (efron): Hit rate ~ Race + Gender + z_crime, stratified by Station×Year (robust SEs clustered by Station)",
  align = c("l","r","r","r","r")
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed")) |>
  add_footnote(
    label = "Station×Year strata condition out intensity; efron method enables robust (clustered) variance. Odds ratios with 95% CIs.",
    notation = "symbol"
  )


```

```{r station tables, echo=FALSE, warning=FALSE, message=FALSE}



library(dplyr)
library(tidyr)
library(broom)
library(kableExtra)
library(scales)
library(stringr)

# ---- parameters ----
MIN_AFRICAN <- 20
MIN_WHITE   <- 10

# ---- auto-detect Station column in `ddatr` ----
station_col <- names(ddatr)[stringr::str_detect(names(ddatr), regex("station", ignore_case = TRUE))]
if (length(station_col) == 0) stop("No column containing 'station' found in `ddatr`. Check names(ddatr).")
if (length(station_col) > 1) warning(paste("Multiple station-like columns detected:",
                                           paste(station_col, collapse = ", "),
                                           "\nUsing the first one:", station_col[1]))
STATION_COLNAME <- station_col[1]

# ---- build analysis frame from `ddatr` ----
dat <- ddatr %>%
  filter(
    !is.na(Found),
    Racial.appearance.transformed %in% c("White", "African")
  ) %>%
  transmute(
    Station   = as.factor(.data[[STATION_COLNAME]]),
    Race      = factor(Racial.appearance.transformed, levels = c("White","African")),
    Found_yes = ifelse(tolower(Found) %in% c("yes","y","true","1"), 1L, 0L)
  )

# ---- summarise counts by Station x Race ----
dat_station <- dat %>%
  group_by(Station, Race) %>%
  summarise(
    n = n(),
    found_yes = sum(Found_yes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(found_no = n - found_yes) %>%
  pivot_wider(
    names_from = Race,
    values_from = c(found_yes, found_no, n),
    values_fill = 0
  ) %>%
  filter(n_African >= MIN_AFRICAN, n_White >= MIN_WHITE)

# ---- compute proportions + prop.test per station ----
station_tests <- dat_station %>%
  mutate(
    prop_African = found_yes_African / n_African,
    prop_White   = found_yes_White   / n_White
  ) %>%
  rowwise() %>%
  mutate(
    test = list(prop.test(
      x = c(found_yes_African, found_yes_White),
      n = c(n_African, n_White)
    )),
    tidy = list(broom::tidy(test))
  ) %>%
  unnest(tidy) %>%
  mutate(
    risk_ratio = prop_African / prop_White,
    p_val = p.value,
    sig = case_when(
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  arrange(p_val)  # sort before pretty-printing

# ---- format columns for display ----
station_tests_fmt <- station_tests %>%
  transmute(
    Station,
    `African n`   = n_African,
    `White n`     = n_White,
    `African %`   = percent(prop_African, accuracy = 0.1),
    `White %`     = percent(prop_White,   accuracy = 0.1),
    `Risk ratio`  = round(risk_ratio, 2),
    `95% CI (L)`  = round(conf.low, 2),
    `95% CI (U)`  = round(conf.high, 2),
    `p-value`     = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val)),
    sig
  )

# ---- pretty table ----
kbl(
  station_tests_fmt,
  caption = paste0(
    "Hit-rate disparity by Station (African vs White; Station column = '",
    STATION_COLNAME,
    "'; only stations with ≥", MIN_AFRICAN, " African & ≥", MIN_WHITE, " White searches)"
  ),
  align = "lrrrrrrrrr"
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    font_size = 12,
    position = "left"
  ) |>
  add_header_above(
    c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2)
  )



```

```{r division tables, echo=FALSE, message=FALSE, warning=FALSE}



# ============================
# Division-level rate ratios (African vs White)
# ============================

library(dplyr)
library(tidyr)
library(broom)
library(kableExtra)
library(scales)
library(stringr)

# ---- parameters ----
MIN_AFRICAN <- 20
MIN_WHITE   <- 10

# ---- prepare dataset from ddatr ----
dat <- ddatr %>%
  filter(
    !is.na(Found),
    Racial.appearance.transformed %in% c("White", "African")
  ) %>%
  transmute(
    Division = as.factor(Division),  # change if your column differs
    Race      = factor(Racial.appearance.transformed, levels = c("White", "African")),
    Found_yes = ifelse(tolower(Found) %in% c("yes", "y", "true", "1"), 1L, 0L)
  )

# ---- summarise counts by Division and Race ----
dat_div <- dat %>%
  group_by(Division, Race) %>%
  summarise(
    n = n(),
    found_yes = sum(Found_yes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(found_no = n - found_yes) %>%
  pivot_wider(
    names_from = Race,
    values_from = c(found_yes, found_no, n),
    values_fill = 0
  ) %>%
  filter(n_African >= MIN_AFRICAN, n_White >= MIN_WHITE)

# ---- compute proportions + run prop.test ----
div_tests <- dat_div %>%
  mutate(
    prop_African = found_yes_African / n_African,
    prop_White   = found_yes_White   / n_White
  ) %>%
  rowwise() %>%
  mutate(
    test = list(prop.test(
      x = c(found_yes_African, found_yes_White),
      n = c(n_African, n_White)
    )),
    tidy = list(broom::tidy(test))
  ) %>%
  unnest(tidy) %>%
  mutate(
    risk_ratio = prop_African / prop_White,
    p_val = p.value,
    sig = case_when(
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  transmute(
    Division,
    `African n`   = n_African,
    `White n`     = n_White,
    `African %`   = percent(prop_African, accuracy = 0.1),
    `White %`     = percent(prop_White, accuracy = 0.1),
    `Risk ratio`  = round(risk_ratio, 2),
    `95% CI (L)`  = round(conf.low, 2),
    `95% CI (U)`  = round(conf.high, 2),
    `p-value`     = ifelse(p_val < 0.001, "<0.001", sprintf("%.3f", p_val)),
    sig
  ) %>%
  arrange(as.numeric(gsub("<|>", "", `p-value`)))

# ---- pretty table ----
kbl(
  div_tests,
  caption = "Hit-rate disparity by Division (African vs White; only Divisions with ≥20 African & ≥10 White searches)",
  align = "lrrrrrrrrr"
) |>
  kable_paper(full_width = FALSE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    font_size = 12,
    position = "left"
  ) |>
  add_header_above(
    c(" " = 1, "Counts" = 2, "Hit Rates" = 2, "Ratio & CI" = 3, "Significance" = 2)
  )


```
### Racial Disparities in Search Outcomes by Station

The random-slope model  
\[
\text{Hit (contraband found)} \sim \text{Race} + \text{Gender} + \text{factor(Year)} + z_\text{crime} + (1 + \text{Race} \mid \text{Station})
\]
was fitted to all search records between 2018–2023.  
This specification controls for gender, year of search, and local drug-crime rates, while allowing both the overall search success rate and the *race effect* to vary across police stations.

#### Overall (population-level) effect
Across all 83 stations, African-appearance individuals were **less likely to be found with contraband** than White-appearance individuals searched under similar conditions.  
The average within-station odds ratio was **0.73 (95% CI 0.57 – 0.93, p = 0.012)**, meaning that—holding other factors constant—the odds of a “hit” were about 27 % lower for African-appearance people.  
Gender, year, and local crime rate were not significantly associated with hit rates after adjusting for station-level variation.

#### Variation across stations
The random-slope term for Race indicates that this disparity is **not uniform**: some stations show stronger racial differences in outcomes than others.  

Figure X (below) plots the estimated *station-specific* odds ratios (African vs White). Each dot represents a station; lines show approximate 95 % confidence intervals, and the dashed vertical line at 1 represents parity (no difference).

Stations with odds ratios **below 1** display a *lower* hit rate for African-appearance individuals (stronger evidence of profiling), while stations near 1 show minimal or no difference.  

Across most stations, the estimated race effect lies below 1, suggesting that the overall disparity reflects a pattern present at many sites rather than being driven by a few outliers.  

However, several stations—including **Frankston, Flemington, and Forest Hill**—exhibit particularly low odds ratios (≈ 0.65–0.70), indicating **stronger racial disparity in search outcomes**.  

A smaller number of stations (e.g. **Richmond, Greensborough, Sunshine**) show near-equal hit rates across racial groups.

#### Interpretation
These results imply that:
- Racial disparities in search outcomes are **systematic** but **vary in magnitude** between stations.
- The average disparity (OR ≈ 0.73) remains statistically significant even after accounting for gender, year, crime context, and station-level heterogeneity.
- Some local contexts may have stronger or weaker patterns of differential search success, warranting further qualitative or operational investigation.

Together, these findings provide quantitative evidence consistent with **racially disparate policing outcomes**, with station-level variability pointing to potential differences in local practices or community demographics.

# Appendix

**Table X Hit rates by largest racial grouping**

```{r Finds racial all, echo=FALSE, warning=FALSE, results='asis'}
ddats<-ddat %>%
  filter(Racial.appearance.transformed %in% c("Missing", "White","Mediterranean/Middle Eastern – Unusable",
                                              "Asian","African", "Aboriginal"))

ct <- crosstable::crosstable(
  ddats,
  by = "Found",
  cols = "Racial.appearance.transformed",
  total = TRUE,
  test = FALSE,          # we'll add our own robust test
  showNA = "ifany",
  percent_digits = 1
)

ft <- crosstable::as_flextable(ct)

# 3) Safe association test (Fisher for 2x2, else Chi-square with simulated p)
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tres <- safe_test(ddats$Racial.appearance.transformed, ddats$Found)
p_line <- sprintf("Racial.appearance.transformed — %s, p = %s",
                  tres$label, format.pval(tres$p, digits = 4))

# 4) Add footer and a clear title
ft <- flextable::add_footer_lines(ft, values = c("Overall association test:", p_line))
ft <- flextable::set_caption(ft, caption = "Racial appearance vs Found (with totals)")
ft
```

```{r Finds lga effect, echo=FALSE, warning=FALSE, results='asis'}
# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <-  ddat 
names(ddatr)
# Variables to tabulate against Found
vars_for_table <- c("Local.Government.Area")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft
```
