---
title: "Racial profiling - drug searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(dplyr)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

ddat <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
  #  !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

```

```{r hitrate-prop-helpers, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(kableExtra)
library(scales)
library(stringr)

is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}

# Build a tidy hit-rate + prop.test table for a grouping var
make_hit_table <- function(
  data,
  group_var,                   # e.g., "Division" or "Station.uniform"
  alpha = 0.05,
  min_african_total = 1,       # set to 5 for stations
  filter_sig_only = FALSE,     # TRUE to keep only p < alpha
  add_fdr = TRUE               # add Benjamini–Hochberg adjusted p
) {
  grp <- rlang::sym(group_var)

  tab <- data %>%
    filter(!is.na(!!grp), !is.na(Found)) %>%
    # guard (your 'ddat' is already Afr/White, but keep it robust)
    filter(Racial.appearance.transformed %in% c("African","White")) %>%
    mutate(FoundYes = is_yes(Found)) %>%
    count(!!grp, Racial.appearance.transformed, FoundYes, name = "n") %>%
    group_by(!!grp, Racial.appearance.transformed) %>%
    summarise(successes = sum(n[FoundYes], na.rm = TRUE),
              total     = sum(n,          na.rm = TRUE),
              .groups   = "drop") %>%
    pivot_wider(
      names_from  = Racial.appearance.transformed,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    # Inclusion thresholds
    filter(total_African >= min_african_total, total_White >= 1)

  if (nrow(tab) == 0L) return(tibble())

  out <- tab %>%
    mutate(
      test = pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = map(test, broom::tidy)
    ) %>%
    select(-test) %>%
    unnest(tidied) %>%
    mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = ifelse(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_fdr) p.adjust(p.value, method = "BH") else NA_real_,
      stars            = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    transmute(
      !!group_var := !!grp,
      `African (succ/total)` = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`   = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`     = percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`       = percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr − Wh)`      = percent(diff_est,         accuracy = 0.1),
      `Risk ratio (Afr/Wh)`  = ifelse(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                    = sprintf("%.3f", p.value),
      `p (BH)`               = if (add_fdr) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                  = stars
    )

  if (filter_sig_only) {
    out <- out %>% filter(as.numeric(str_replace(`p`, ",", ".")) < alpha)
  }

  out %>% arrange(as.numeric(str_replace(`p`, ",", ".")))
}

# Pretty printer with consistent styling
print_hit_table <- function(df, caption, show_bh = TRUE) {
  if (nrow(df) == 0L) {
    cat("No groups meet the criteria or show significance.\n")
    return(invisible(NULL))
  }
  display <- df
  if (!show_bh) {
    display <- display %>% select(-`p (BH)`)
  }
  display %>%
    kbl(caption = caption, align = "lrrrrrrrr", escape = TRUE) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "hover"), position = "left") %>%
    column_spec(1, width = "24em") %>%
    add_footnote(
      label = "Hit rate = proportion of searches with any contraband found (Found == \"Yes\"). Stars: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "symbol"
    ) %>%
    print()
}
```

# Short summary

This analysis examines 7,534 drug searches conducted by uniformed officers on foot patrols in Melbourne, focusing on people perceived as White or African. These encounters are the most routine form of warrantless public-place searches and depend heavily on officer discretion.


### Melbourne-wide findings
- Hit rates: White 14.1% (957/6,779) vs African 11.4% (86/755); Fisher’s exact p = 0.040.
- Pooled logistic regression (controls: Division fixed effects, Year, Gender): OR = 0.771 (95% CI 0.596–0.985), p = 0.042 — African-appearing males had ~23% lower odds of a find than White males.
- Multilevel logistic regression (random intercepts for Division): OR = 0.761 (95% CI 0.594–0.975), p = 0.031 — confirms the disparity after accounting for clustering by Division.


### Local hotspots
- NWMR Division 3 (Brimbank & Melton): African 4.3% (3/70) vs White 14.2% (30/212); risk ratio = 0.30, p = 0.026.
- Werribee Station: African 8.8% (5/57) vs White 21.4% (22/103); risk ratio = 0.41, p = 0.042.


### Limits of statistical significance
- Most other divisions and stations did not reach statistical significance, primarily because of small African sample sizes (often only a few dozen searches). Wide confidence intervals and multiple-testing adjustments mean that many apparent disparities cannot be declared significant.
- Non-significance does not equal absence of disparity — rather, it reflects the limited statistical power in areas with few recorded searches of African-appearing individuals.


### Interpretation
Across Melbourne, African-appearing people are consistently subjected to searches with lower hit rates and lower odds of contraband being found, even after adjusting for year and division. This pattern is strongest in Melbourne’s outer west (Brimbank, Melton, Werribee), but persists at a city-wide level. Together, these results are consistent with racial profiling: African-appearing individuals are more frequently searched on a weaker threshold of suspicion.

# Introduction  

This analysis examines **7,534 drug searches conducted by uniformed officers on foot patrols in Melbourne, of white and African appearing people**. These encounters represent the most routine form of warrantless public-place searches, where outcomes are highly dependent on officer discretion. The racial appearance categories in scope represent the two largest recorded racial groupings in the dataset.

The focus is on whether search outcomes suggest evidence of **racial profiling**. Specifically, we compare the **hit rates** — the proportion of searches that result in contraband being found — for two groups:  

- **White-appearing individuals**, who form the majority baseline  
- **African-appearing individuals**, a racialised minority consistently highlighted in community concerns about disproportionate policing  

By restricting the dataset to this context and to these two appearance categories, we provide a clear test of whether African-appearing people are subjected to searches on a **lower evidentiary threshold**. Lower hit rates for one group, despite comparable or higher search intensity, would suggest that police suspicion is being applied unevenly — a hallmark of racial profiling.  
  
  
# About the dataset (subset definition)  

This dataset contains records of **7,534 warrantless searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**.  

Data from **2020 and 2021** is excluded due to the atypical policing environment during the COVID-19 pandemic.  

The dataset was released through a Freedom of Information (FOI) request lodged by the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and covers public-place searches carried out under a range of legal powers.  

For this analysis, we restrict to:  

- **Geography:** Melbourne metropolitan area  
- **Contact type:** Foot patrol only  
- **Unit type:** Uniform police only  
- **Search type:** Drug searches conducted under s82 of the *Drugs, Poisons and Controlled Substances Act 1981*  
- **Station linkage:** Only searches associated with a uniform station  
- **Racial appearance:** White and African appearance

This restriction enables us to compare searches for the most common drug searches experienced by the public in Melbourne, and restrict the data to reliable racial appearance categories.

## What s82 allows  

Under section 82 of the *Drugs, Poisons and Controlled Substances Act 1981*, a police officer may search a person **without a warrant** if they have *reasonable grounds to suspect* possession of a drug of dependence.  
This is the most frequently invoked search power in the dataset and represents a context where discretionary judgments by officers play a central role.  

## Variables in this subset  

The following variables are included in the analysis:  

- **Year** — year of the stop and search  
- **Region / Division / PSA / Station** — police geographic hierarchy  
- **Racial.appearance.transformed** — officer-perceived racial appearance (cleaned categories)  
- **Gender** — officer-perceived gender of the person searched  
- **Found** — whether any contraband was located (hit outcome)  
- **Unit.type** — uniform officers only (subset restriction)  
- **Rank** — officer rank (e.g. Constable, Senior Constable)  
- **Year** — 2018, 2019, 2022, or 2023  

Derived variables include:  

- **Found.Flag** — binary indicator (1 if contraband found, 0 otherwise)  
- **Race** — restricted to White vs African  

---

## Data limitations  

### Racial appearance categories  
The original dataset for drug searches conducted by uniform officers on foot patrol numbered 13,896. Race data was missing from 20% of the records.  Several categories had very small numbers or had been recategorised over the observation period.  As a result, disparity testing is restricted to **White vs African** searches, which are sufficiently large and analytically comparable.  These are the two largest race categories and account for over half (54%) of the relevant data.

### Age data  
Age data was missing for 42.9% of searches in the White vs African subset. Age is therefore excluded from modelling, which limits the ability to adjust for age differences across groups. We expect this would have an effect as African migrant populations in in local government areas tend to be younger than their white-appearing counterparts. 

### Scope of coverage  
- The dataset covers only **warrantless stop and search events conducted in public places**. This focus is deliberate: such encounters are entirely reliant on officer discretion, since there is no warrant, prior intelligence, or judicial approval.  
- The analysis does not include searches with warrants, searches in private settings, or informal/unrecorded encounters.  
- Findings therefore apply only to **Melbourne metropolitan drug searches by uniform officers on foot patrol**. They do not generalise to specialist units, vehicle stops, or regional policing contexts.  
- Because modelling is restricted to White vs African, results do not extend to other racialised groups with smaller or analytically unreliable sample sizes.  


# Data Overview  

## Persons searched  

This section summarises the characteristics of people searched in warrantless public-place encounters by uniform officers on foot patrols (2018, 2019, 2022, 2023).  
Records with missing or unusable racial appearance are excluded to allow meaningful comparison.  

- **Racial appearance.** Most searches involved people perceived as White, with African-appearance individuals making up around one in ten of the analytic sample.  
- **Gender.** Searches were concentrated on men, who accounted for about four-fifths of those stopped.  
- **Age.** While children and adolescents appear in the dataset, most searches targeted young adults in their late teens and twenties, followed by those in their thirties.  

These patterns highlight that discretionary stop-and-search practices fall unevenly across the community, with young men and racialised groups particularly over-represented.  


**Table 2 Characteristics of persons searched**
```{r summary persons, echo = FALSE, warning = FALSE}
of.interest <- c("Racial.appearance.transformed","Gender","Age group")

#ddat$`Age group`[ddat$`Age group` == "Missing"] <- NA  
ddat <- ddat %>%
  filter(Racial.appearance.transformed %in% c("White", "African"))

ddat.per <- ddat%>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

ddat.per <- ddat.per %>% 
         mutate(`Age group` = fct_relevel(`Age group`, c("Under 12", "12 to 17", "18 to 25","26 to 35", "36 to 45", "46 and over")))
 
print(dfSummary(ddat.per, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

## Police undertaking searches  

Search numbers were relatively consistent across the four years examined, with a slight peak in 2022.  By rank, Constables and Senior Constables were the most frequently involved, reflecting the predominance of frontline officers in this type of activity.  


**Table 3 Police and units undertaking searches**
```{r summary police, echo = FALSE, warning = FALSE}
of.interest <- c("Year", "Rank.of.Member", "Region",  "Division")

ddat <- ddat %>%
  mutate(
    Rank = case_when(
      Rank.of.Member == "SSGT" ~ "SGT",
      Rank.of.Member == "PSO"  ~ "CONST",
      TRUE           ~ Rank.of.Member
    )
  )

dat.pol <- ddat %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))

dat.pol <- dat.pol %>% 
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))
 
print(dfSummary(dat.pol, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

# Hit Rate Patterns  

An important way to assess stop-and-search practices is through **hit rates** — the proportion of searches that result in contraband being found.  

If searches are consistently based on *reasonable suspicion*, hit rates should be similar across groups.  

Where a group has consistently lower hit rates, this suggests that officers may be applying a **lower threshold of suspicion**, a core indicator of potential racial profiling.  

This section examines hit rates in the restricted dataset of **7,534 Melbourne metropolitan drug searches by uniform police on foot patrol** (2018, 2019, 2022, 2023).  

A “hit” is defined as the discovery of any prohibited item, not just the item specified under the search power.  

## Hit rates by person searched  

Hit rates varied by race, gender, and age:  

- **By racial appearance.**  
  - White-appearing people had a hit rate of **14.1%** (957 finds out of 6,779 searches).  
  - African-appearing people had a lower hit rate of **11.4%** (86/755).  
  - This indicates that African-appearance individuals were **less likely to be found with contraband**, despite being searched in substantial numbers.  

- **By gender.**  
  - Men had a hit rate of **14.1%** (837/5,956), compared with **13.1%** for women (206/1,578).  
  - Searches were heavily concentrated on men, accounting for four in five cases.  

- **By age.**  
  - Younger people, particularly those under 18, had lower hit rates compared with adults.  

- **By year.**  
  - Hit rates were lowest in **2018 (12.1%)**, rose in **2019 (14.5%)**, and were fairly stable in **2022 (14.0%)** and **2023 (14.4%)**.  

**Table 5 Hit rates by person characteristics**

```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}

# Recode and reorder factors (ddat assumed pre-filtered to your analysis cohort)
dat2 <- ddat %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),                 # hits first
    Year  = fct_relevel(as.character(Year), "2018","2019","2022","2023"),
    Gender = fct_infreq(Gender) |> fct_drop(),
    Racial.appearance.transformed =
      fct_relevel(Racial.appearance.transformed, "White", "African") |> fct_drop()
  )

# Variables to tabulate against Found
vars_for_table <- c("Racial.appearance.transformed", "Gender", "Year")

# Crosstable (no internal tests; cleaner & avoids small-cell errors)
ct <- crosstable(
  dat2[, c(vars_for_table, "Found")],
  by = "Found",
  total = TRUE,
  test = FALSE,
  showNA = "ifany",
  percent_digits = 1
)

ft <- as_flextable(ct)

# Safe association test per variable: Fisher for 2x2 else Chi-square with simulated p
safe_test <- function(x, y) {
  tbl <- table(x, y, useNA = "no")
  tbl <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]
  if (all(dim(tbl) == c(2, 2))) {
    res <- fisher.test(tbl, workspace = 2e8)
    list(label = "Fisher's exact", p = res$p.value)
  } else {
    res <- chisq.test(tbl, simulate.p.value = TRUE, B = 100000)
    list(label = "Chi-square (simulated p)", p = res$p.value)
  }
}

tests <- map(vars_for_table, ~ safe_test(dat2[[.x]], dat2$Found)) |>
  set_names(vars_for_table) |>
  imap_dfr(~ tibble(variable = .y, test = .x$label, p = .x$p)) |>
  mutate(p_str = sprintf("%s — %s, p = %s",
                         variable, test, format.pval(p, digits = 4)))

# Append p-values to the table footer
ft <- add_footer_lines(ft, values = c("Overall association tests:", tests$p_str))

ft

```
**Interpretation.**  
Hit rates were consistently lower for African-appearing individuals compared with White individuals, suggesting a lower evidentiary threshold for initiating searches. Gender differences were small, and year-to-year variation modest, indicating that the racial disparity is not explained by shifts in overall effectiveness.  

## Hit rates by police member and operating context  

Hit rates varied by rank and division. 

- **By rank.** Constables and Senior Constables dominated search activity, and their hit rates were broadly similar.  
- **By geography.** Hit rates were lower in the **North West Metro Region**, which includes diverse and disadvantaged areas such as Brimbank and Melbourne, compared with other parts of the city.  
- **By time.** Across 2018, 2019, 2022, and 2023, hit rates varied only modestly.  

**Interpretation.**  
These results show that **who conducts the search** and **where it occurs** both shape outcomes. However, the consistent gap between African- and White-appearing individuals cuts across these contexts, reinforcing concerns that discretion is exercised differently by race.  


**Table 6 Hit rates by police member and operating context**
```{r Finds pol, echo=FALSE, warning=FALSE, results='asis'}

### Summary of restricted sample (hit rates by rank, region, division, and year)

library(dplyr)
library(knitr)

# Clean up ranks (SSGT -> SGT, PSO -> CONST)
ddat_clean <- ddat %>%
  mutate(
    Rank = dplyr::recode(Rank.of.Member,
                         "SSGT" = "SGT",
                         "PSO"  = "CONST",
                         .default = Rank.of.Member)
  )

# Function to summarise hit rates by variable
summarise_hitrate <- function(data, group_var, label) {
  data %>%
    group_by({{ group_var }}) %>%
    summarise(
      Searches = n(),
      Hits = sum(Found == "Yes", na.rm = TRUE),
      Hit_rate = round(Hits / Searches * 100, 1),
      .groups = "drop"
    ) %>%
    rename(Category = {{ group_var }}) %>%
    mutate(Dimension = label)
}

# Summaries
rank_summary     <- summarise_hitrate(ddat_clean, Rank, "Rank")
region_summary   <- summarise_hitrate(ddat_clean, Region, "Region")
division_summary <- summarise_hitrate(ddat_clean, Division, "Division")
year_summary     <- summarise_hitrate(ddat_clean, Year, "Year")

# Combine into one table
combined_summary <- bind_rows(rank_summary, region_summary, division_summary, year_summary) %>%
  select(Dimension, Category, Searches, Hits, Hit_rate)

# Print
kable(combined_summary,
      caption = "Hit rates by rank, region, division, and year (restricted sample).")


```

# Division- and Station-Level Proportion Tests  

To move beyond descriptive comparisons, we tested whether differences in hit rates between **African** and **White** individuals were statistically significant within each Division and Station. We used two-sample proportion tests (`prop.test`).

---

## Division-Level Findings  

- **North West Metro Region (NWMR) Division 3 – Brimbank and Melton** showed a clear disparity:  
  - African: **3/70 (4.3%)** vs White: **30/212 (14.2%)**  
  - Risk difference = –9.9%, Risk ratio = 0.30, *p* = 0.026  
- This division is marked by **socio-economic disadvantage and large  diaspora communities**.  


```{r prop_test_division_station_sig, message=FALSE, warning=FALSE, echo = FALSE}
alpha <- 0.05

div_tbl_all <- make_hit_table(
  data = ddat,
  group_var = "Division",
  alpha = alpha,
  min_african_total = 1,      # keep any division with ≥1 African search
  filter_sig_only = FALSE,    # set TRUE if you want to show only p < 0.05
  add_fdr = TRUE
)

print_hit_table(
  div_tbl_all,
  caption = sprintf("Divisions — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
  show_bh = TRUE
)
```
---

## Station-Level Findings  

- **Werribee Station** also displayed a significant disparity:  
  - African: **5/57 (8.8%)** vs White: **22/103 (21.4%)**  
  - Risk difference = –12.6%, Risk ratio = 0.41, *p* = 0.042  
- Werribee is in the same outer-western growth corridor as Brimbank/Melton, with a **substantial African community** and a history of **public concern about over-policing**.  

```{r stations-hitrate-with-prop, echo=FALSE, message=FALSE, warning=FALSE}
alpha <- 0.05

sta_tbl <- make_hit_table(
  data = ddat,
  group_var = "Station.uniform",
  alpha = alpha,
  min_african_total = 5,      # inclusion rule you requested
  filter_sig_only = TRUE,     # show only p < 0.05; set FALSE to see all included
  add_fdr = TRUE
)

print_hit_table(
  sta_tbl,
  caption = sprintf("Stations (≥ 5 African searches) — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
  show_bh = TRUE
)

```

## Interpretation  

The proportion tests suggest that disparities are **geographically concentrated** in Melbourne’s **outer west**, where disadvantage and demographic change intersect with intensive policing.  
Other Divisions and Stations did not show statistically significant differences, but many had small African sample sizes. Absence of significance does not mean absence of disparity. 

These results highlight the need for multivariate modelling to control for confounders such as **gender, year, and local policing context**.  

# Statistical Modelling  


Logistic regression estimates the odds that a search will result in contraband being found, while adjusting for other variables. This allows us to move beyond descriptive comparisons and ask whether observed differences reflect systematic disparities in how suspicion is applied.  

We apply three complementary modelling strategies:  

1. **Division-level models**  
   - Separate logistic regressions are estimated for each police Division.  
   - Predictors: race (African vs White), gender (female vs male), and an interaction between race and gender.  
   - Purpose: to test whether disparities are **geographically concentrated**, identifying divisions where African-appearing individuals face a significantly lower chance of a successful search.  

2. **Pooled model with Division fixed effects**  
   - A single logistic regression pooling all Divisions.  
   - Predictors: race, gender, year, and Division fixed effects.  
   - Purpose: to estimate whether disparities persist **across Melbourne as a whole**, while controlling for both local baseline differences (Divisions) and shifts across time (years).  

3. **Multilevel (hierarchical) model**  
   - A logistic regression with Division included as a **random effect**.  
   - Predictors: race, gender, year; Divisions allowed to vary in their baseline hit rates.  
   - Purpose: to provide a more robust Melbourne-wide estimate by stabilising results in Divisions with small sample sizes and assessing whether disparities are **systematic across the city**.  

Together, these models move from **localised analysis** (division-level) to **city-wide estimation** (pooled and multilevel). This layered approach ensures we can identify both **geographically specific disparities** (e.g. Brimbank/Melton, Werribee) and assess whether those disparities reflect a broader **systematic pattern of racial profiling** in Melbourne.  

## Division- and Station-Level Logistic Regression Models  

The proportion tests showed that disparities in hit rates were not uniform across Melbourne, with some divisions and stations suggesting stronger evidence of racial disparity than others.  

To explore this further, we estimated **logistic regression models separately for each Division and for stations where there were at least twenty searches of people of African appearance in the observations.**  

The outcome is whether **any contraband was found** (`Found = Yes`). Each model includes predictors for:  

- **Race** (African vs White)  
- **Gender** (Female vs Male)  
- An **interaction between race and gender**  

Running separate models at both the **division** and **station** level allows us to:  

1. Assess whether disparities are **geographically concentrated** in particular areas.  
2. Avoid masking local differences that would be hidden in a pooled analysis.  
3. Provide results that are **directly interpretable for community advocacy and policy discussions** in specific local contexts.  

Because sample sizes for African-appearance searches are limited, especially at the station level, we restricted analysis to:  
- Divisions with both White and African searches recorded.  
- Stations with at least **10 searches of African-appearing individuals**, to avoid unstable estimates.  

As with the proportion tests, a finding that African-appearing individuals have significantly lower odds of contraband being found indicates they are being searched on a **lower evidentiary threshold.**  

Summary tables models where the coefficient of racial appearance was lower than 0.05 appear below.  

```{r division-models-optionA-noYear, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(kableExtra)
library(stringr)

# --------- helpers ----------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}

fmt_p <- function(p) {
  ifelse(is.na(p), NA_character_,
         ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
}

# --------- prepare data ----------
dat_div <- ddat %>%
  filter(!is.na(Division),
         Racial.appearance.transformed %in% c("African","White"),
         !is.na(Found),
         Gender %in% c("M","F")) %>%
  transmute(
    Division = Division,
    Success  = as.integer(is_yes(Found)),
    Race     = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender   = factor(Gender, levels = c("M","F"))
  )

# Precompute per-division context for captions
div_context <- ddat %>%
  filter(!is.na(Division),
         Racial.appearance.transformed %in% c("African","White"),
         !is.na(Found)) %>%
  mutate(Success = as.integer(is_yes(Found))) %>%
  group_by(Division, Racial.appearance.transformed) %>%
  summarise(succ = sum(Success), tot = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Racial.appearance.transformed,
                     values_from = c(succ, tot),
                     values_fill = 0) %>%
  group_by(Division) %>%
  mutate(n_total = tot_African + tot_White) %>%
  ungroup()

# --------- fit per-division models & print tables ----------
data_by_div <- split(dat_div, dat_div$Division)

walk(names(data_by_div), function(div_name) {
  df <- data_by_div[[div_name]]

  # skip if only one outcome class
  if (dplyr::n_distinct(df$Success) < 2L) return(invisible(NULL))

  # fit model with Race * Gender (no Year)
  mdl <- tryCatch(
    glm(Success ~ Race * Gender, data = df, family = binomial),
    error = function(e) NULL
  )
  if (is.null(mdl)) return(invisible(NULL))

  # tidy (raw) to inspect p-values before formatting
  td_raw <- tryCatch(
    tidy(mdl, exponentiate = TRUE, conf.int = TRUE),
    error = function(e) {
      tidy(mdl, exponentiate = TRUE, conf.int = FALSE) %>%
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    }
  ) %>%
    select(term, estimate, conf.low, conf.high, p.value)

  # --- show model only if ANY non-intercept term is significant at 0.05 ---
  keep_if_sig <- any(td_raw$term != "(Intercept)" & td_raw$p.value < 0.05, na.rm = TRUE)
  if (!keep_if_sig) return(invisible(NULL))
  # -------------------------------------------------------------------------

  td <- td_raw %>%
    mutate(
      Term = dplyr::case_when(
        term == "RaceAfrican" ~ "Race: African vs White (Male ref.)",
        term == "GenderF" ~ "Gender: Female vs Male",
        term == "RaceAfrican:GenderF" ~ "Interaction: African × Female",
        TRUE ~ term
      ),
      `Odds Ratio`      = estimate,
      `95% CI (Lower)`  = conf.low,
      `95% CI (Upper)`  = conf.high,
      `p-value`         = p.value
    ) %>%
    select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
    mutate(
      `p-value` = fmt_p(`p-value`),
      `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
      `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
      `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`))
    )

  # caption context
  cx <- div_context %>% dplyr::filter(Division == div_name)
  cap_bits <- c(
    paste0("n=", ifelse(nrow(cx) == 1, cx$n_total, NA_integer_)),
    if (nrow(cx) == 1) paste0("Afr: ", cx$succ_African, "/", cx$tot_African) else NULL,
    if (nrow(cx) == 1) paste0("Wh: ",  cx$succ_White,   "/", cx$tot_White)   else NULL
  )
  cap_ctx <- paste(cap_bits[!is.na(cap_bits)], collapse = " · ")

  # print table
  td %>%
    kbl(
      caption = paste0("Division: ", div_name,
                       " — Logistic regression of search success (any contraband).  ",
                       cap_ctx),
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","condensed","hover"),
                  position = "left") %>%
    column_spec(1, width = "22em") %>%
    add_footnote(
      label = "Model: Success ~ Race * Gender. Odds ratios shown (95% CI). Reference group = White males.",
      notation = "symbol"
    ) %>%
    print()
})  # closes walk()


```

```{r station-models-optionA-noYear, echo=FALSE, warning=FALSE, message=FALSE}


# -------- helpers ----------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# -------- prepare data ----------
dat_sta <- ddat %>%
  filter(!is.na(Station.uniform),
         Racial.appearance.transformed %in% c("African","White"),
         !is.na(Found),
         Gender %in% c("M","F")) %>%
  transmute(
    Station = Station.uniform,
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F"))
  )

# Inclusion: stations with at least 10 African searches
sta_context <- dat_sta %>%
  count(Station, Race) %>%
  pivot_wider(names_from = Race, values_from = n, values_fill = 0) %>%
  filter(African >= 20)

# Add success counts for captions
sta_success <- dat_sta %>%
  group_by(Station, Race) %>%
  summarise(succ = sum(Success), tot = n(), .groups = "drop") %>%
  pivot_wider(names_from = Race, values_from = c(succ, tot), values_fill = 0) %>%
  mutate(n_total = tot_African + tot_White)

# -------- fit per-station models ----------
data_by_sta <- split(dat_sta, dat_sta$Station)

walk(sta_context$Station, function(sta_name) {
  df <- data_by_sta[[sta_name]]

  if (dplyr::n_distinct(df$Success) < 2L) return(invisible(NULL))

  mdl <- tryCatch(
    glm(Success ~ Race * Gender, data = df, family = binomial),
    error = function(e) NULL
  )
  if (is.null(mdl)) return(invisible(NULL))

  td <- tryCatch(
    tidy(mdl, exponentiate = TRUE, conf.int = TRUE),
    error = function(e) {
      tidy(mdl, exponentiate = TRUE, conf.int = FALSE) %>%
        mutate(conf.low = NA_real_, conf.high = NA_real_)
    }
  ) %>%
    select(term, estimate, conf.low, conf.high, p.value) %>%
    mutate(
      Term = case_when(
        term == "RaceAfrican" ~ "Race: African vs White (Male ref.)",
        term == "GenderF" ~ "Gender: Female vs Male (White ref.)",
        term == "RaceAfrican:GenderF" ~ "Interaction: African × Female",
        TRUE ~ term
      ),
      `Odds Ratio`      = estimate,
      `95% CI (Lower)`  = conf.low,
      `95% CI (Upper)`  = conf.high,
      `p-value`         = p.value
    ) %>%
    select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
    mutate(
      `p-value` = fmt_p(`p-value`),
      `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
      `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
      `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`))
    )

  # Only print if Race: African vs White is significant at < 0.05
  race_row <- td %>% filter(str_detect(Term, "Race: African vs White"))
  if (nrow(race_row) == 1 && suppressWarnings(as.numeric(race_row$`p-value`)) >= 0.05) {
    return(invisible(NULL))
  }

  cx <- sta_success %>% filter(Station == sta_name)
  cap_bits <- c(
    paste0("n=", ifelse(nrow(cx)==1, cx$n_total, NA_integer_)),
    if (nrow(cx)==1) paste0("Afr: ", cx$succ_African, "/", cx$tot_African) else NULL,
    if (nrow(cx)==1) paste0("Wh: ",  cx$succ_White,   "/", cx$tot_White)   else NULL
  )
  cap_ctx <- paste(cap_bits[!is.na(cap_bits)], collapse = " · ")

  td %>%
    kbl(
      caption = paste0("Station: ", sta_name, " — Logistic regression of search success (any contraband). ", cap_ctx),
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("striped","condensed","hover"),
                  position = "left") %>%
    column_spec(1, width = "22em") %>%
    add_footnote(
      label = "Model: Success ~ Race * Gender. Odds ratios shown (95% CI). Reference group = White males.",
      notation = "symbol"
    ) %>%
    print()
})
```

### Interpretation  

Both the station-level and division-level logistic regressions show evidence of racial disparity in Melbourne’s **outer west**.  

At the **divisional level (NWMR Division 3 – Brimbank & Melton)**, the pattern was even more pronounced. African-appearing males had just **one-fifth the odds** of a successful search compared to White males (*OR = 0.213, p = 0.013*). White females also showed lower odds than White males, but the racial disparity was clear and statistically robust.  

At the **station level**, African-appearing males had only about **one-third the odds** of a successful search compared to White males (*OR = 0.319, p = 0.037*). This suggests that African individuals were more frequently searched on a **lower evidentiary threshold**, with police suspicion less likely to result in the discovery of contraband.  

Interaction terms for African females could not be reliably estimated because there were virtually **no successful searches of African females** recorded in the dataset. The resulting “0.000 … Inf” values are statistical artefacts of these sparse data and should be interpreted as a lack of sufficient cases rather than evidence of no effect.  

Taken together, these results point to **geographically concentrated racial disparities** in stop-and-search practices, particularly in Brimbank, Melton, and associated stations. These are areas marked by **socio-economic disadvantage, larger African diaspora communities, and longstanding tension between residents and police**. The findings align with **community concerns** that African-appearing individuals, especially young men, are being subjected to searches on a weaker threshold of suspicion.  


## Pooled Logistic Regression with Division Fixed Effects 

The division- and station-level models highlight that disparities are **most visible in Melbourne’s outer west**, but their scope is necessarily local and limited by small sample sizes.  

To understand whether these disparities reflect isolated hotspots or a **systematic Melbourne-wide pattern**, we next estimate a **pooled logistic regression**.  

This model combines all divisions into a single analysis, while adjusting for:  

- **Division fixed effects** — to absorb baseline geographic differences  
- **Year** — to account for temporal variation  
- **Gender** — to ensure disparities are not conflated with gender differences  

The pooled model provides a headline estimate of whether, across metropolitan Melbourne, African-appearing individuals are consistently subjected to searches with a **lower evidentiary threshold** than their White counterparts.  



```{r pooled-model-optionB-year, echo=FALSE, warning=FALSE, message=FALSE}
library(broom)
library(kableExtra)
library(forcats)
library(stringr)

# ---------- helpers ----------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_, ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# ---------- prepare data ----------
dat_pool <- ddat %>%
  filter(
    Racial.appearance.transformed %in% c("African","White"),
    Gender %in% c("M","F"),
    !is.na(Found), !is.na(Division), !is.na(Year)
  ) %>%
  transmute(
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F")),
    Year    = factor(Year) %>% fct_relevel("2022","2018","2019","2023"),  # ref = 2022
    Division = factor(Division)  # fixed effects
  )

# ---------- fit pooled model ----------
mdl_pool <- glm(Success ~ Race * Gender + Year + Division, data = dat_pool, family = binomial)

# ---------- tidy to Odds Ratios with 95% CI ----------
td <- tryCatch(
  tidy(mdl_pool, exponentiate = TRUE, conf.int = TRUE),
  error = function(e) {
    tidy(mdl_pool, exponentiate = TRUE, conf.int = FALSE) %>%
      mutate(conf.low = NA_real_, conf.high = NA_real_)
  }
) %>%
  mutate(
    Term = case_when(
      term == "(Intercept)"          ~ "(Intercept) — White male, Year 2022, Division ref.",
      term == "RaceAfrican"          ~ "Race: African vs White (Male, 2022, Division ref.)",
      term == "GenderF"              ~ "Gender: Female vs Male (White, 2022, Division ref.)",
      str_detect(term, "^Year")      ~ str_replace(term, "^Year", "Year: "),
      str_detect(term, "^Division")  ~ str_replace(term, "^Division", "Division: "),
      term == "RaceAfrican:GenderF"  ~ "Interaction: African × Female",
      TRUE                           ~ term
    ),
    `Odds Ratio`      = estimate,
    `95% CI (Lower)`  = conf.low,
    `95% CI (Upper)`  = conf.high,
    `p-value`         = p.value
  ) %>%
  select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
  mutate(
    `Odds Ratio`      = sprintf("%.3f", `Odds Ratio`),
    `95% CI (Lower)`  = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
    `95% CI (Upper)`  = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`)),
    `p-value`         = fmt_p(`p-value`)
  )

# ---------- pretty table ----------
td %>%
  kbl(
    caption = "Pooled logistic regression of search success (any contraband): Success ~ Race * Gender + Year + Division. Reference = White males in 2022, Division reference = first level.",
    align = c("l","r","r","r","r"),
    escape = TRUE
  ) %>%
  kable_paper(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","condensed","hover"), position = "left") %>%
  column_spec(1, width = "28em") %>%
  add_footnote(
    label = "Odds ratios with 95% CI. Division entered as fixed effects to absorb baseline differences across areas.",
    notation = "symbol"
  ) %>%
  print()

# ---------- optional: compact 'headline' effects ----------
headline <- td %>%
  filter(Term %in% c("Race: African vs White (Male, 2022, Division ref.)",
                     "Interaction: African × Female")) %>%
  rename(`Term (key effects)` = Term)

if (nrow(headline) > 0) {
  headline %>%
    kbl(
      caption = "Key effects (pooled model): Race African vs White (male baseline) and African × Female interaction",
      align = c("l","r","r","r","r"),
      escape = TRUE
    ) %>%
    kable_paper(full_width = FALSE) %>%
    kable_styling(bootstrap_options = c("condensed","hover"), position = "left") %>%
    print()
}
```

### Interpretation  

The pooled logistic regression indicates that, across metropolitan Melbourne, **African-appearing males had significantly lower odds of contraband being found than White males**, even after controlling for gender, year, and division fixed effects.  Specifically, their odds of a successful search were about **23% lower** (*OR = 0.771, p = 0.042*).  

Female gender was associated with slightly lower odds of a find, but this was not statistically significant.  
The African × Female interaction was also non-significant, likely reflecting the very small number of African females in the dataset.  

Year-to-year variation was modest and generally non-significant once other factors were accounted for, indicating that disparities are not driven by particular years.  

Several divisions, especially in the **north-western metropolitan region**, showed significantly lower baseline hit rates, consistent with earlier descriptive results.  

Overall, these results confirm that racial disparities in search outcomes are **not limited to isolated divisions or stations**, but persist at a **Melbourne-wide level**.  

This strengthens the conclusion that African-appearing individuals are being searched on a **lower evidentiary threshold**, consistent with patterns of racial profiling.  

## Multilevel Logistic Regression with Division Random Effects  

The pooled model provides a Melbourne-wide estimate while controlling for gender, year, and division differences.  
However, treating divisions as fixed effects assumes each division is **independent** and estimated with equal certainty, even though some divisions have far more searches than others.  

To address this, we use **multilevel (hierarchical) logistic regression**, which accounts for clustering of searches within divisions and “borrows strength” across areas with small samples.

We present two specifications:

1. **Model 1 — Random intercepts only:** each Division has its own baseline hit rate.  
2. **Model 2 — Random intercepts + random slopes for Race:** allows the **African–White disparity** itself to vary by Division.

### Model 1: Random Intercepts for Division

This model estimates the odds of a successful search while allowing each Division to have its own baseline (intercept):

- Outcome: `Found = Yes` (any contraband found)  
- Predictors: **Race** (African vs White), **Gender** (F vs M), **Year** (2018, 2019, 2022, 2023)  
- Random effects: **(1 | Division)** 

```{r multilevel-model-glmer, echo=FALSE, warning=FALSE, message=FALSE}
library(lme4)
library(broom.mixed)
library(dplyr)
library(forcats)
library(kableExtra)

# -------- helpers ----------
is_yes <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  tolower(as.character(x)) %in% c("yes","y","true","1")
}
fmt_p <- function(p) ifelse(is.na(p), NA_character_,
                            ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))

# -------- prepare data ----------
dat_glmer <- ddat %>%
  filter(Racial.appearance.transformed %in% c("African","White"),
         Gender %in% c("M","F"),
         !is.na(Found), !is.na(Division), !is.na(Year)) %>%
  transmute(
    Success = as.integer(is_yes(Found)),
    Race    = factor(Racial.appearance.transformed, levels = c("White","African")),
    Gender  = factor(Gender, levels = c("M","F")),
    Year    = factor(Year) %>% fct_relevel("2022","2018","2019","2023"),
    Division = factor(Division)
  )

# -------- fit multilevel model ----------
mdl_glmer <- glmer(Success ~ Race * Gender + Year + (1 | Division),
                   data = dat_glmer, family = binomial,
                   control = glmerControl(optimizer = "bobyqa"))

# -------- tidy to Odds Ratios with 95% CI ----------
td_glmer <- tidy(mdl_glmer, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    Term = case_when(
      effect == "fixed" & term == "(Intercept)"          ~ "(Intercept) — White male, Year 2022 (Division random intercept)",
      effect == "fixed" & term == "RaceAfrican"          ~ "Race: African vs White (Male, 2022 ref.)",
      effect == "fixed" & term == "GenderF"              ~ "Gender: Female vs Male (White, 2022 ref.)",
      effect == "fixed" & grepl("^Year", term)           ~ sub("^Year", "Year: ", term),
      effect == "fixed" & term == "RaceAfrican:GenderF"  ~ "Interaction: African × Female",
      TRUE ~ term
    ),
    `Odds Ratio`      = estimate,
    `95% CI (Lower)`  = conf.low,
    `95% CI (Upper)`  = conf.high,
    `p-value`         = p.value
  ) %>%
  select(Term, `Odds Ratio`, `95% CI (Lower)`, `95% CI (Upper)`, `p-value`) %>%
  mutate(
    `Odds Ratio` = sprintf("%.3f", `Odds Ratio`),
    `95% CI (Lower)` = ifelse(is.na(`95% CI (Lower)`), "—", sprintf("%.3f", `95% CI (Lower)`)),
    `95% CI (Upper)` = ifelse(is.na(`95% CI (Upper)`), "—", sprintf("%.3f", `95% CI (Upper)`)),
    `p-value` = fmt_p(`p-value`)
  )

# -------- pretty table ----------
td_glmer %>%
  kbl(
    caption = "Multilevel logistic regression of search success (any contraband): Success ~ Race * Gender + Year + (1 | Division). Reference group = White males in 2022.",
    align = c("l","r","r","r","r"),
    escape = TRUE
  ) %>%
  kable_paper(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","condensed","hover"),
                position = "left") %>%
  column_spec(1, width = "28em") %>%
  add_footnote(
    label = "Odds ratios with 95% CI. Random intercepts allow each Division to have its own baseline hit rate, while estimating an overall racial effect across Melbourne.",
    notation = "symbol"
  ) %>%
  print()
```


### **Interpretation**

The model shows that African-appearing males had about 24% lower odds of a successful search than White males (*OR ≈ 0.761, p = 0.031*), controlling for gender and year.  

Gender and year effects were not statistically significant once other variables were included.  

The **random intercept variance** indicates modest baseline differences across divisions, suggesting the disparity is **systematic across Melbourne**, not confined to a few local areas.


### **Model 2: Random Intercepts + Random Slopes for Race**

This extension tests whether the **size of the African–White disparity** varies by Division:  

- **Outcome:** `Found = Yes`  
- **Predictors:** Race, Gender, Year  
- **Random effects:** `(Race | Division)` — random intercepts **and** random slopes for Race  

```{r multilevel-model-glmer-2, echo=FALSE, warning=FALSE, message=FALSE}
mdl_glmer_slopes <- glmer(Success ~ Race * Gender + Year + (Race | Division),
                          data = dat_glmer, family = binomial,
                          control = glmerControl(optimizer = "bobyqa"))


# Tidy model output
mdl_tbl <- broom.mixed::tidy(mdl_glmer_slopes, effects = "fixed", conf.int = TRUE)

# Clean up column names for readability
mdl_tbl <- mdl_tbl[, c("term", "estimate", "conf.low", "conf.high", "p.value")]
names(mdl_tbl) <- c("Term", "Estimate (log-odds)", "95% CI Low", "95% CI High", "p-value")

# Print table
knitr::kable(
  mdl_tbl,
  digits = 3,
  caption = "Fixed effects from multilevel logistic regression (Success ~ Race * Gender + Year + (Race | Division))."
)
```
### **Interpretation**

Allowing the Race effect to vary by Division yielded **very small slope variance** and a **singular (boundary) fit**, indicating **no meaningful between-division variation** in the disparity.  

In practical terms, the **African–White gap in hit rates appears consistent across divisions**, reinforcing the conclusion from Model 1 that **African-appearing individuals face a lower evidentiary threshold city-wide**.  


# **Modelling Analysis: Overall Insights**

Across the different modelling strategies — division-level regressions, pooled logistic regression, and multilevel models — the evidence consistently points to **racial disparities in search outcomes**.  

- **Division and station models** showed that disparities were especially pronounced in Melbourne’s **outer western suburbs** (Brimbank, Melton, Werribee), aligning with areas of socio-economic disadvantage and strong community concern about over-policing.  
- The **pooled model** confirmed that these disparities are not just local anomalies: African-appearing males had **significantly lower odds of contraband being found** across Melbourne, even after adjusting for gender, year, and division differences.  
- The **multilevel models** demonstrated that the disparity is **systematic across the city**, not meaningfully varying by division, which strengthens the conclusion that this is a broad structural pattern rather than an isolated hotspot effect.  

Together, these models provide a **robust and layered picture**: while raw hit rates highlight where disparities look most obvious, statistical modelling shows that the issue is both **geographically concentrated** in some areas and **systematic city-wide**.

---

# **Conclusion**

This analysis of **uniform foot patrol drug searches in metropolitan Melbourne (2018, 2019, 2022, 2023)** provides clear evidence that individuals perceived as **African** are searched on a **lower evidentiary threshold** compared to those perceived as White.  

Key findings include:  

- **Descriptive results** showed lower hit rates for African-appearing individuals across most contexts.  
- **Proportion tests** and **division/station-level models** identified particularly strong disparities in Brimbank, Melton, and Werribee.  
- **Pooled and multilevel models** demonstrated that the disparity persists after accounting for gender, year, and division, and is **consistent across Melbourne** rather than confined to a few local areas.  

These results are consistent with **racial profiling**: African-appearing individuals are more frequently subjected to searches that are less likely to yield contraband, implying that suspicion is applied more readily to them.  

While the magnitude of disparity varies locally, the overall pattern is systematic.  

This highlights the need for greater police accountability and independent oversight, and provides an empirical foundation for community advocacy and policy reform aimed at fairer, more equitable policing in Melbourne.  
