---
title: "Racial profiling – weapons searches on foot by uniform police"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css
---

# Short summary

This report analyses **1,358 warrantless weapons searches** carried out by **uniform police officers on foot patrol** in metropolitan Melbourne between **2018, 2019, 2022, and 2023**. The findings highlight systematic disparities in outcomes across racial groups, raising concerns about the use of discretion under Victoria’s stop-and-search powers.

Under the *Control of Weapons Act 1990 (Vic), s.10*, police may search a person without a warrant if they have **reasonable grounds** to suspect that the individual is carrying a weapon in a public place. The Act also allows officers to consider whether a person is present in an area with a **high incidence of violent crime** when forming suspicion. This provision makes it necessary to adjust for geography in assessing whether racial disparities reflect profiling.

**Key findings include:**

- **Racial disparities:** White-appearing individuals recorded a weapon hit rate of **31.3%**, while African-appearing individuals recorded only **15.7%**. A chi-square test confirmed that differences across racial groups are statistically significant (*p* ≈ 0.007).  
- **Minors:** Age was significant: children and young people had a much lower hit rate (**20.8%**) compared to adults (**30.0%**), raising concerns about the targeting of minors.  
- **Geographic patterns:** Disparities were strongest in the North West Metro Region 3 covering Brimbank and Melton, where African-appearing individuals were searched on weaker evidentiary thresholds than White-appearing counterparts.  
- **Regression models:** After accounting for police division, African-appearing individuals remained less than half as likely to be found with a weapon compared to White-appearing individuals (OR = 0.43, *p* < 0.001). This disparity cannot be explained by geography alone.  
- **Missing data:** Roughly **20% of records lacked usable racial appearance information** and nearly **30% lacked age data**, with some divisions (e.g. **Hume/Merri-bek**, **Frankston/Mornington**) missing age in over half of their records. This missingness is patterned, not random, and is especially concerning given that African-appearing youth are widely reported to be disproportionately targeted by police.  
- **Other factors:** Gender differences were negligible.   

**Overall conclusion:**  
The results are consistent with **racial profiling**. African-appearing people are more frequently subjected to searches that fail to yield evidence, implying suspicion is applied more readily to them. Patterns of missing data weaken accountability but, even with these gaps, the evidence is clear: discretionary stop-and-search powers are being applied unevenly, underscoring the need for stronger recording practices, independent oversight, and accountability mechanisms.


```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(readxl)
library(broom)
library(lme4)
library(broom.mixed)
library(scales)
library(flextable)
library(purrr)
library(stringr)

# ---- Load data ----
dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

# ---- Analysis subset (WEAPONS, not drugs; ALL races; foot patrol; metro; uniform) ----
ddat <- dat %>%
  filter(
    Contact.Type == "P",                  # foot patrol
    Area.type == "Metro",
    !is.na(`Search.type - Weapons`),      # Weapons subset
    !is.na(Found),
    !is.na(Station.uniform),              # linked to uniform station
    Unit.type == "Uniform"                # explicitly keep uniform
  )

ddat <- ddat %>% 
  mutate(Found = ifelse(`Search.type - Weapons` == "Search item found", "Yes", "No"))%>%
  mutate(
    Found.Flag = if_else(Found == "Yes", 1, 0, 0L),
    Race = fct_explicit_na(Racial.appearance.transformed, "Missing/Unknown"),
    Year = fct_relevel(as.character(Year), "2018","2019","2022","2023")
  )

ddat <- ddat %>%
  mutate(
    Adult = case_when(
      `Age group` %in% c("Under 12", "12 to 17") ~ "Child",
      `Age group` %in% c("18 to 25","26 to 35","36 to 45","46 and over") ~ "Adult",
      TRUE ~ NA_character_
    ),
    Adult = factor(Adult)
  )
# ------------ Helpers for multi-group (K) hit-rate summaries and tests ------------

# Clean printer
print_kable <- function(df, caption = NULL) {
  df %>%
    kbl(caption = caption, align = "lrrr", digits = 1) %>%
    kable_classic(full_width = FALSE)
}

# Hit-rate table by a single variable (e.g., Race, Gender, Year)
hitrate_by <- function(data, by_var) {
  by_sym <- rlang::ensym(by_var)
  data %>%
    filter(!is.na(!!by_sym)) %>%
    group_by(!!by_sym) %>%
    summarise(
      Searches = n(),
      Hits = sum(Found == "Yes", na.rm = TRUE),
      Hit_rate = 100 * Hits / Searches,
      .groups = "drop"
    ) %>%
    arrange(desc(Searches)) %>%
    rename(Category = !!by_sym)
}

# Overall association test (Kx2 table): Chi-square with simulated p to be robust
assoc_test_overall <- function(x, y) {
  tab <- table(x, y, useNA = "no")
  tab <- tab[rowSums(tab) > 0, colSums(tab) > 0, drop = FALSE]
  if (all(dim(tab) == c(2, 2))) {
    res <- fisher.test(tab, workspace = 2e8)
    tibble(test = "Fisher's exact", p.value = res$p.value)
  } else {
    res <- chisq.test(tab, simulate.p.value = TRUE, B = 100000)
    tibble(test = "Chi-square (simulated p)", p.value = res$p.value)
  }
}

# plus an overall association p-value (are hit rates equal across races?).
make_context_table_multi <- function(data, group_var, min_total_per_race = 1, alpha = 0.05) {
  gsym <- rlang::ensym(group_var)

  # Per (group x Race) cells
  base <- data %>%
    filter(!is.na(!!gsym), !is.na(Race)) %>%
    count(!!gsym, Race, Found, name = "n") %>%
    group_by(!!gsym, Race) %>%
    summarise(
      total = sum(n, na.rm = TRUE),
      hits  = sum(n[Found == "Yes"], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(total >= min_total_per_race)

  if (nrow(base) == 0L) return(tibble())

  # Build Kx2 tests per group
  groups <- unique(pull(base, !!gsym))
  out <- map_dfr(groups, function(g) {
    sub <- base %>% filter(!!gsym == g)
    # Association test across all Race categories at this group level
    long <- dat %>%
      filter(!is.na(!!gsym), !!gsym == g) %>%
      mutate(Race = fct_explicit_na(Racial.appearance.transformed, "Missing/Unknown")) %>%
      select(Race, Found)

    tst <- assoc_test_overall(long$Race, long$Found)

    sub %>%
      mutate(
        hit_rate = if_else(total > 0, hits / total, NA_real_),
        Group = as.character(g)
      ) %>%
      select(Group, Race, hits, total, hit_rate) %>%
      mutate(p.value = tst$p.value, test = tst$test)
  })

  out %>%
    mutate(
      `Hit rate` = percent(hit_rate, accuracy = 0.1),
      `succ/total` = sprintf("%d/%d", hits, total)
    ) %>%
    select(!!group_var := Group, Race, `succ/total`, `Hit rate`, test, p.value) %>%
    arrange(!!rlang::sym(group_var), desc(`succ/total`))
}

```

# Data overview

Across the 1,358 warrantless weapons searches included in this dataset, most individuals were recorded as White-appearing (41.6%). The next largest category is not a specific racial group but rather **Missing or unrecorded racial appearance (20.3%)**, a substantial gap that represents one in five searches. African-appearing individuals make up 11.7% of those searched, while other groups are represented in much smaller proportions: Mediterranean/Middle Eastern (8.1%), Pacific Islander (5.6%), Asian (5.4%), and Aboriginal (2.7%).  

The overwhelming majority of people searched were **male (89.3%)**, with females accounting for just over one in ten searches. In terms of age, more than a quarter (26.1%) of recorded searches involved teenagers aged 12–17, while 17.6% were 18–25, and around a quarter were older adults. However, **age data is missing in 29.2% of cases**, limiting the reliability of age-based comparisons. Where age was recorded, just under two-thirds of searches involved adults (63%), while more than one-third (37%) involved children or young people under 18.  

Weapons searches in this dataset were predominantly conducted by lower-ranking officers. Almost two-thirds were carried out by constables (64.2%), with senior constables responsible for 30.0% and sergeants only 5.7%. This distribution highlights that weapons stop-and-search activity is primarily performed at the frontline level.  

Geographically, searches were spread across Melbourne’s police divisions, but with clear concentrations in certain areas. The largest share occurred in **North West Metro Division 1 (Melbourne, Yarra)**, which accounted for one in five searches (20.2%). Other significant hotspots included **Division 5 (Banyule, Darebin, Nillumbik, Whittlesea) (15.5%)** and **Division 2 (Hobsons Bay, Maribyrnong, Wyndham) (10.2%)**, with the remaining divisions contributing smaller proportions.  

The distribution of searches also varied over time. Activity was highest in **2023 (39.2%)** and **2022 (31.9%)**, with fewer searches recorded in 2019 (17.3%) and 2018 (11.6%). This pattern shows that weapons searches intensified in the most recent years covered by the dataset.  
 
## Persons searched

```{r persons searched, echo = FALSE, warning = FALSE, message=FALSE}
# Simple characteristics table (counts only) to mirror your earlier layout
of.interest <- c("Race","Gender","Age group", "Adult")

ddat.per <- ddat %>%
  select(all_of(of.interest)) %>%
  mutate(
    across(everything(), ~ fct_infreq(.)),
    `Age group` = fct_relevel(`Age group`,
                               "Under 12","12 to 17","18 to 25",
                               "26 to 35","36 to 45","46 and over")
  )

print(dfSummary(ddat.per,
                max.distinct.values = 12,
                varnumbers   = FALSE,
                valid.col    = FALSE,
                graph.magnif = 0.76,
                headings = FALSE),
      method = "render")
```

## Police context  


```{r police searching, echo = FALSE, warning = FALSE, message=FALSE}
# Police characteristics table (counts only)
of.interest_police <- c("Rank.of.Member", "Division", "Year")

ddat.pol <- ddat %>%
  mutate(
    # collapse/recode ranks if needed (example: SSGT -> SGT, PSO -> CONST)
    Rank.of.Member = recode(Rank.of.Member,
                            "SSGT" = "SGT",
                            "PSO"  = "CONST",
                            .default = Rank.of.Member)
  ) %>%
  select(all_of(of.interest_police)) %>%
  mutate(across(everything(), ~ fct_infreq(.)))

print(dfSummary(ddat.pol,
                max.distinct.values = 12,
                varnumbers   = FALSE,
                valid.col    = FALSE,
                graph.magnif = 0.76,
                headings = FALSE),
      method = "render")
```

# Hit rates  

### Summary of hit rates  

Hit rates for weapons searches varied considerably across racial appearance categories. White-appearing individuals recorded a hit rate of **31.3%**, while African-appearing individuals were searched with a much lower hit rate of **15.7%**. This difference was statistically significant in the overall test of race (χ² simulated p = 0.0077), indicating that outcomes were uneven across groups. Some smaller categories, such as South American (66.7%), showed high hit rates, but these reflect very small numbers of searches.  

By gender, outcomes were almost identical, with males at **27.8%** and females at **29.4%** (not statistically significant). Year-on-year patterns showed a decline: hit rates were highest in **2018 (37.6%)**, but dropped in more recent years to around **26–27%** in 2022–2023 (χ² p = 0.0403).  

Age made a clear difference. Adults had a higher hit rate (**30.0%**) compared to children and young people (**20.8%**), a statistically significant gap (Fisher’s p = 0.0019).  

When comparing police ranks, constables (27.4%), senior constables (30.9%), and sergeants (23.1%) did not differ significantly. Similarly, regional comparisons showed little variation, with hit rates between 27–30% across North West Metro, Southern Metro, and Eastern Regions (χ² p = 0.7096).  

At the divisional level, some contrasts emerged. Most divisions recorded hit rates in the mid-20s to low-30s, but **Southern Metro Division 1 (Port Phillip, Stonnington)** stood out with a much higher rate of **43.9%**. By contrast, **North West Metro Division 3 (Brimbank, Melton)** and **Southern Metro Division 3 (Cardinia, Casey, Dandenong)** had notably lower rates of **23.4%** and **20.4%** respectively. The overall division-level test approached significance (χ² p = 0.0904) but did not cross the conventional threshold.  

Overall, the strongest disparities are observed **by race** and **by age**, while differences by gender, rank, region, or division are less pronounced.  


## By racial appearance 

```{r hit rate patterns all, echo = FALSE, warning = FALSE, message=FALSE}
# ---- Ensure Adult variable exists ----
ddat <- ddat %>%
  mutate(
    Adult = case_when(
      `Age group` %in% c("Under 12", "12 to 17") ~ "Child",
      `Age group` %in% c("18 to 25","26 to 35","36 to 45","46 and over") ~ "Adult",
      TRUE ~ NA_character_
    )
  )

# ---- Race ----
by_race <- hitrate_by(ddat, Race)
print_kable(by_race, caption = "Hit rates by racial appearance (weapons searches)")

race_test <- assoc_test_overall(ddat$Race, ddat$Found)
cat(sprintf("**Overall Race × Found:** %s, *p* = %.4f\n\n",
            race_test$test, race_test$p.value))

# ---- Gender ----
by_gender <- hitrate_by(ddat, Gender)
print_kable(by_gender, caption = "Hit rates by gender")

gender_test <- assoc_test_overall(ddat$Gender, ddat$Found)
cat(sprintf("**Overall Gender × Found:** %s, *p* = %.4f\n\n",
            gender_test$test, gender_test$p.value))

# ---- Year ----
by_year <- hitrate_by(ddat, Year)
print_kable(by_year, caption = "Hit rates by year")

year_test <- assoc_test_overall(ddat$Year, ddat$Found)
cat(sprintf("**Overall Year × Found:** %s, *p* = %.4f\n\n",
            year_test$test, year_test$p.value))

# ---- Adult vs Child ----
by_adult <- hitrate_by(ddat, Adult)
print_kable(by_adult, caption = "Hit rates by Adult vs Child")

adult_test <- assoc_test_overall(ddat$Adult, ddat$Found)
cat(sprintf("**Overall Adult × Found:** %s, *p* = %.4f\n\n",
            adult_test$test, adult_test$p.value))

```


```{r police searching hit rate, echo = FALSE, warning = FALSE, message=FALSE}
# Recode ranks and keep needed fields
ddat_pol <- ddat %>%
  mutate(
    Rank = dplyr::recode(Rank.of.Member,
                         "SSGT" = "SGT",
                         "PSO"  = "CONST",
                         .default = Rank.of.Member)
  )

# ---- By Rank ----
by_rank <- hitrate_by(ddat_pol, Rank)
print_kable(by_rank, caption = "Hit rates by police rank (weapons searches)")

rank_test <- assoc_test_overall(ddat_pol$Rank, ddat_pol$Found)
cat(sprintf("**Overall Rank × Found:** %s, *p* = %.4f\n\n",
            rank_test$test, rank_test$p.value))

# ---- By Region ----
by_region <- hitrate_by(ddat_pol, Region)
print_kable(by_region, caption = "Hit rates by region (weapons searches)")

region_test <- assoc_test_overall(ddat_pol$Region, ddat_pol$Found)
cat(sprintf("**Overall Region × Found:** %s, *p* = %.4f\n\n",
            region_test$test, region_test$p.value))

# ---- By Division (with LGA label) ----
# (Division already carries the LGA label in your dataset, e.g. "NWMR Division 4 (Hume, Moonee Valley, Merri-bek)")
by_division <- hitrate_by(ddat_pol, Division)
print_kable(by_division, caption = "Hit rates by division (LGA-labelled) — weapons searches")

division_test <- assoc_test_overall(ddat_pol$Division, ddat_pol$Found)
cat(sprintf("**Overall Division × Found:** %s, *p* = %.4f\n\n",
            division_test$test, division_test$p.value))


```




### Division and Region-level proportion tests  

To examine whether disparities were concentrated in particular parts of Melbourne, we compared **hit rates for White-appearing and African-appearing individuals only**. These are the **largest identified racial categories** in the dataset, providing sufficient numbers for meaningful statistical comparison. Only regions and divisions with at least ten searches for both groups were included.  

At the **regional level**, statistically significant differences emerged in both **Southern Metro** and **North West Metro**. In Southern Metro, African-appearing people recorded a hit rate of just **4.8% (1/21)** compared to **33.8% (50/148)** for White-appearing individuals, a gap of almost 30 percentage points (risk ratio 0.14, p = 0.007). In North West Metro, the gap was smaller but still significant: **18.0% (23/128)** for African-appearing individuals versus **30.1% (96/319)** for White (risk ratio 0.60, p = 0.009). In contrast, the **Eastern Region** showed a large numerical gap (10% vs 31.6%) but this was not statistically significant, reflecting the very small number of African searches (n=10).  

At the **divisional level**, the clearest disparity appeared in **North West Metro Division 3 (Brimbank, Melton)**, where the African hit rate was just **3.7% (1/27)** compared to **27.6% (8/29)** for White-appearing people. This difference (risk ratio 0.13) was statistically significant before correction (p = 0.015) but only marginal after controlling the false discovery rate (BH-adjusted p = 0.075). A similar pattern was visible in **Southern Metro Division 3 (Cardinia, Casey, Dandenong)**, where no African searches yielded a weapon (0/10) compared to a White hit rate of 28.6% (8/28), though this result was only marginally significant (p = 0.057, BH-adjusted p = 0.143).  

Other divisions, including **Melbourne/Yarra (22.0% vs 34.3%)**, **Hobsons Bay/Maribyrnong/Wyndham (19.4% vs 35.4%)**, and **Banyule/Darebin/Nillumbik/Whittlesea (25.0% vs 26.1%)**, also showed lower African hit rates but these differences did not reach statistical significance.  

Taken together, these results suggest that disparities are **most pronounced in Melbourne’s outer west (Brimbank, Melton) and parts of the south-east (Casey, Dandenong)**, while the North West and Southern Metro regions overall show significant gaps. Although statistical power is limited by the relatively small number of African searches in some divisions, the pattern is consistent with African-appearing people being searched on a weaker evidentiary threshold across multiple contexts.  

```{r hit rate patterns context, echo = FALSE, warning = FALSE, message=FALSE}

alpha <- 0.05

# --- Filter to White & African only ---
d_aw <- ddat %>%
  filter(Race %in% c("African","White")) %>%
  mutate(FoundYes = Found == "Yes")

# --- Helper for 2-group tests (prop.test) ---
make_aw_table <- function(data, group_var, min_african_total = 10, min_white_total = 10, add_bh = TRUE) {
  gsym <- rlang::ensym(group_var)

  tab <- data %>%
    filter(!is.na(!!gsym), Race %in% c("African","White"), !is.na(Found)) %>%
    count(!!gsym, Race, FoundYes, name = "n") %>%
    group_by(!!gsym, Race) %>%
    summarise(
      successes = sum(n[FoundYes], na.rm = TRUE),
      total     = sum(n,           na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = Race,
      values_from = c(successes, total),
      values_fill = 0
    ) %>%
    filter(
      total_African >= min_african_total,
      total_White   >= min_white_total
    )

  if (nrow(tab) == 0L) return(tibble())

  out <- tab %>%
    mutate(
      test = purrr::pmap(
        list(successes_African, successes_White, total_African, total_White),
        ~ prop.test(x = c(..1, ..2), n = c(..3, ..4), correct = FALSE)
      ),
      tidied = purrr::map(test, broom::tidy)
    ) %>%
    select(-test) %>%
    tidyr::unnest(tidied) %>%
    mutate(
      hit_rate_african = estimate1,
      hit_rate_white   = estimate2,
      diff_est         = hit_rate_african - hit_rate_white,
      rr               = ifelse(hit_rate_white > 0, hit_rate_african / hit_rate_white, NA_real_),
      p_bh             = if (add_bh) p.adjust(p.value, method = "BH") else NA_real_,
      sig              = dplyr::case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.010 ~ "**",
        p.value < 0.050 ~ "*",
        p.value < 0.100 ~ ".",
        TRUE ~ ""
      )
    ) %>%
    transmute(
      !!group_var := !!gsym,
      `African (succ/total)` = sprintf("%d/%d", successes_African, total_African),
      `White (succ/total)`   = sprintf("%d/%d", successes_White,   total_White),
      `African hit rate`     = scales::percent(hit_rate_african, accuracy = 0.1),
      `White hit rate`       = scales::percent(hit_rate_white,   accuracy = 0.1),
      `Diff (Afr − Wh)`      = scales::percent(diff_est,         accuracy = 0.1),
      `Risk ratio (Afr/Wh)`  = dplyr::if_else(is.finite(rr), sprintf("%.2f", rr), "—"),
      `p`                    = sprintf("%.3f", p.value),
      `p (BH)`               = if (add_bh) sprintf("%.3f", p_bh) else NA_character_,
      `sig`                  = sig
    ) %>%
    arrange(as.numeric(stringr::str_replace(`p`, ",", ".")))

  out
}

# --- REGION (Afr vs Wh, min African total = 10) ---
region_tbl_aw <- make_aw_table(
  data = d_aw,
  group_var = "Region",
  min_african_total = 10,
  min_white_total   = 10,
  add_bh = TRUE
)

if (nrow(region_tbl_aw) == 0L) {
  cat("No regions met inclusion thresholds.\n\n")
} else {
  region_tbl_aw %>%
    kbl(caption = sprintf("Regions — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
        col.names = c("Region", "African (succ/total)", "White (succ/total)",
                      "African hit rate", "White hit rate", "Diff (Afr − Wh)",
                      "Risk ratio (Afr/Wh)", "p", "p (BH)", "sig"),
        align = "lrrrrrrrr") %>%
    kable_classic(full_width = FALSE) %>%
    add_footnote(
      label = "Inclusion rule: ≥10 searches for both African and White. Two-sample prop.test without continuity correction; BH controls FDR across displayed rows.",
      notation = "symbol"
    ) %>%
    print()
}

# --- DIVISION (Afr vs Wh, min African total = 10) ---
div_tbl_aw <- make_aw_table(
  data = d_aw,
  group_var = "Division",
  min_african_total = 10,
  min_white_total   = 10,
  add_bh = TRUE
)

if (nrow(div_tbl_aw) == 0L) {
  cat("No divisions met inclusion thresholds.\n\n")
} else {
  div_tbl_aw %>%
    kbl(caption = sprintf("Divisions — Hit Rates (African vs White) with prop.test (α = %.2f)", alpha),
        col.names = c("Division (LGA label)", "African (succ/total)", "White (succ/total)",
                      "African hit rate", "White hit rate", "Diff (Afr − Wh)",
                      "Risk ratio (Afr/Wh)", "p", "p (BH)", "sig"),
        align = "lrrrrrrrr") %>%
    kable_classic(full_width = FALSE) %>%
    add_footnote(
      label = "Division names include the LGA label from the source data.",
      notation = "symbol"
    ) %>%
    print()
}
```

# Modeling  

To assess whether the disparities observed in descriptive results and proportion tests held once geography was accounted for, we estimated a logistic regression model predicting whether a weapon was found. The model included both **Race** and **Division**, with White-appearing individuals as the reference category for race.  

In this stage, we deliberately reintroduced **all recorded racial categories** into the model, even though many of the smaller groups introduce statistical noise. This allows us to test whether the African disparity remains when considered alongside the full set of categories recorded by police.  

The results show a clear and robust pattern. **African-appearing individuals were less than half as likely to be found with a weapon when searched compared to White-appearing individuals (OR = 0.43, 95% CI [0.26–0.68], p < 0.001)**, even after adjusting for division. This confirms that the disparity is not simply explained by differences in where searches occur, but persists at the metropolitan scale.  

Other groups did not show statistically significant differences from White. The “Other” category approached significance (OR = 0.43, p ≈ 0.09), but most groups had wide confidence intervals due to small numbers of searches (e.g. South Americans, Middle Eastern). These unstable estimates underscore that the evidence of disparity is strongest and most reliable for African-appearing individuals.  

Turning to the geography, **most divisions were not significantly different from the model’s baseline** once race was included. The only clear exception was **Southern Metro Division 1 (Port Phillip, Stonnington)**, where searches were more than twice as likely to yield a weapon (OR = 2.18, p = 0.014). Other divisions fluctuated above or below the baseline, but none reached significance.  

A likelihood-ratio test confirmed that adding Race to a Division-only model significantly improved model fit (p = 0.0141). In plain terms, this means that **race adds explanatory power beyond geography**: the lower hit rates for African-appearing individuals cannot be reduced to local concentration effects but reflect a systematic difference across Melbourne.  

Taken together, the modelling strengthens the descriptive findings. Even when all racial categories are considered and geography is accounted for, **African-appearing people are consistently searched on a lower evidentiary threshold**.  

```{r regression-race-division-pretty, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(forcats)
library(stringr)
library(broom)
library(knitr)
library(kableExtra)

# --- Prepare data ---
ddat_rd <- ddat %>%
  mutate(
    Found_bin = if_else(Found == "Yes", 1, 0),
    Race      = fct_relevel(factor(Race), "White"),  # White baseline
    Division  = factor(Division)
  ) %>%
  filter(!is.na(Race), !is.na(Division))

# --- Fit models ---
mod_div_only <- glm(Found_bin ~ Division, data = ddat_rd, family = binomial)
mod_race_div <- glm(Found_bin ~ Race + Division, data = ddat_rd, family = binomial)

# LRT for Race beyond Division
lrt <- anova(mod_div_only, mod_race_div, test = "LRT")
lrt_p <- lrt$`Pr(>Chi)`[2]
N <- nobs(mod_race_div)

# --- Tidy full model (ORs + CIs) ---
tidy_or <- tidy(mod_race_div, exponentiate = TRUE, conf.int = TRUE) %>%
  transmute(
    Term        = term,
    `Odds ratio`= estimate,
    `Std. error`= std.error,
    `CI low`    = conf.low,
    `CI high`   = conf.high,
    p           = p.value
  )

# Pretty labels + stars
starify <- function(p) {
  case_when(
    is.na(p)         ~ "",
    p < 0.001        ~ "***",
    p < 0.01         ~ "**",
    p < 0.05         ~ "*",
    p < 0.10         ~ ".",
    TRUE             ~ ""
  )
}

fmt_tbl <- function(df) {
  df %>%
    mutate(
      `Odds ratio` = round(`Odds ratio`, 3),
      `Std. error` = round(`Std. error`, 3),
      `CI low`     = round(`CI low`, 3),
      `CI high`    = round(`CI high`, 3),
      p            = ifelse(is.na(p), NA, round(p, 4)),
      `Sig`        = starify(p)
    )
}

# --- Split Race vs Division blocks with clean labels ---
race_tbl <- tidy_or %>%
  filter(str_starts(Term, "Race")) %>%
  mutate(
    Term = str_remove(Term, "^Race"),
    Term = str_trim(Term),
    Term = paste0(Term, " vs White")
  ) %>%
  fmt_tbl() %>%
  arrange(p)

div_tbl <- tidy_or %>%
  filter(str_starts(Term, "Division")) %>%
  mutate(
    Term = str_remove(Term, "^Division"),
    Term = str_squish(Term)
  ) %>%
  fmt_tbl() %>%
  arrange(Term)

# --- Render: Race table ---
if (nrow(race_tbl) > 0) {
  kbl(
    race_tbl,
    caption = "Logistic regression of weapon 'Found' — Race effects adjusted for Division (odds ratios; baseline = White)",
    col.names = c("Term","Odds Ratio","Std. Error","CI (low)","CI (high)","p","Sig"),
    align = c("l","r","r","r","r","r","c")
  ) %>%
    kable_classic(full_width = FALSE) %>%
    add_footnote(
      label = "Odds ratios are exponentiated coefficients from a logit model. Significance: *** <0.001; ** <0.01; * <0.05; . <0.10.",
      notation = "none"
    ) %>%
    print()
} else {
  cat("No race terms (check factor levels).\n\n")
}

# --- Render: Division table ---
kbl(
  div_tbl,
  caption = "Logistic regression of weapon 'Found' — Division effects adjusted for Race (odds ratios)",
  col.names = c("Division (vs baseline)","Odds Ratio","Std. Error","CI (low)","CI (high)","p","Sig"),
  align = c("l","r","r","r","r","r","c")
) %>%
  kable_classic(full_width = FALSE) %>%
  print()

# --- Model notes / LRT ---
cat(sprintf(
  "\n\n*Model notes:* Logistic regression with logit link; N = %d. Likelihood-ratio test for adding **Race** beyond **Division**: p = %.4f.\n",
  N, lrt_p
))
cat("Reference category for **Race** is **White**; Division coefficients are relative to the model's baseline division (set by R).")
```


## Racial appearance mapping all weapons searches  

## Regression results: Race and Division

To test whether the lower hit rates for African-appearing individuals could be explained by geography, we estimated a logistic regression model predicting whether a weapon was found. The model included **Race** and **Division**, so that racial disparities could be examined while holding constant the division in which the search occurred.  

The results show a clear and consistent pattern. **African-appearing people were less than half as likely as White-appearing people to be found with a weapon when searched** (odds ratio = 0.43, p < 0.001). This disparity remains highly significant even after accounting for the division of the police officers conducting the search. In other words, the difference is not simply due to African communities being concentrated in particular areas; rather, it reflects a broader system-wide effect.  

Other racial categories did not show statistically significant differences compared to White, largely because the numbers of searches in these groups were much smaller. The one exception is the “Other” category, which also showed lower odds of a find (OR = 0.43) but only marginal statistical evidence (p ≈ 0.09). Some groups, such as South Americans, displayed large but unstable odds ratios with very wide confidence intervals, reflecting the very small number of searches in those categories.  

In terms of geography, **most divisions did not differ significantly from one another once race was controlled for**. The notable exception was **Southern Metro Division 1 (Port Phillip, Stonnington)**, where searches were more than twice as likely to yield a weapon compared to the baseline division. This appears to be a localised “high hit rate” context, but it does not alter the Melbourne-wide pattern of racial disparity.  

Taken together, these results indicate that African-appearing individuals are subjected to searches that are much less likely to yield evidence. The disparity is not explained away by geography, and it persists after statistical adjustment for division. This strengthens the conclusion that African-appearing people are being searched on a **lower evidentiary threshold**, consistent with long-standing community concerns about racial profiling.  

# Missing data  

## 5 Missing data  

A consistent challenge in this dataset is the **extent of missing information**, particularly for racial appearance and age. Across Melbourne, around **20% of all records lacked a usable race category**, and age data was absent for over **40% of searches overall**. At the divisional level, missingness varies: some divisions such as Hume/Merri-bek (57% missing age) and Frankston/Mornington (58% missing age) show especially poor recording practices.  

This pattern of missing data is not trivial. For example, **SMR Division 3 (Cardinia, Casey, Dandenong)** — which recorded one of the starkest disparities between White- and African-appearing individuals — also has nearly **30% missing age information**. Such gaps complicate interpretation and limit our ability to control for age in modelling, even though community concerns about profiling focus heavily on **African-appearing youth**.  

The absence of reliable age data is particularly problematic. Young people, and especially African-appearing young people, are widely reported to be the primary targets of weapons stop-and-search. Yet because of missingness, we cannot robustly test whether lower hit rates for African-appearing people are driven disproportionately by youth encounters. Where age was available, searches of children and young people yielded far fewer weapons (20.8%) than searches of adults (30.0%), underscoring the importance of this variable.  

In short, **missing data weakens accountability**. Without complete and consistent recording of race and age, Victoria Police cannot credibly demonstrate that stop-and-search powers are being applied without bias. Far from being a technical detail, the gaps highlight the need for stronger recording requirements and independent oversight of how discretionary powers are used in practice.


```{r missingness-by-region, echo=FALSE, warning=FALSE, message=FALSE}
# For each Region, calculate % missing Race and % missing Age group
missing_tbl <- ddat %>%
  group_by(Division) %>%
  summarise(
    Total_searches = n(),
    Missing_race = sum(is.na(Racial.appearance.transformed) |
                         Racial.appearance.transformed %in% c("Missing","Missing/Unknown")),
    Missing_age  = sum(is.na(`Age group`) |
                         `Age group` == "Missing"),
    .groups = "drop"
  ) %>%
  mutate(
    `% Missing race` = round(100 * Missing_race / Total_searches, 1),
    `% Missing age`  = round(100 * Missing_age  / Total_searches, 1)
  ) %>%
  select(Division, Total_searches, `% Missing race`, `% Missing age`)

# Pretty print
kbl(missing_tbl,
    caption = "Proportion of searches with missing race or missing age, by Division",
    align = c("l","r","r","r")) %>%
  kable_classic(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Appendix 

## Police organisational hierarchy all weapons searches  

```{r police hierarchy search counts all races, echo=FALSE, warning=FALSE, message=FALSE}
# ---- Hierarchy summary with counts only (all races) ----
police_counts <- ddat %>%
  transmute(
    Region,
    Division,
    LGA     = coalesce(Local.Government.Area, "Missing/Unknown"),
    Station = coalesce(Station.uniform, "Missing/Unknown")
  ) %>%
  group_by(Region, Division, LGA, Station) %>%
  summarise(
    Searches = n(),
    .groups = "drop"
  ) %>%
  arrange(Region, Division, LGA, desc(Searches), Station)

# ---- Render nicely with merged cells ----
kbl(
  police_counts,
  caption = "Weapons searches by police hierarchy (Region → Division → LGA → Station, all racial groups)",
  col.names = c("Region", "Division (LGA label)", "LGA", "Station", "Total Searches")
) %>%
  kable_classic(full_width = FALSE) %>%
  collapse_rows(columns = 1:3, valign = "top")
```

```{r racial-mapping-table, echo=FALSE, warning=FALSE, message=FALSE}
# Map Original -> Transformed with counts (weapons subset held in `ddat`)
mapping_tbl <- ddat %>%
  transmute(
    `Original race`   = coalesce(Racial.Appearance.original, "Missing/Unknown"),
    `Transformed race`= coalesce(Racial.appearance.transformed, "Missing/Unknown")
  ) %>%
  group_by(`Original race`, `Transformed race`) %>%
  summarise(`Total count` = n(), .groups = "drop") %>%
  arrange(`Transformed race`, desc(`Total count`), `Original race`)

# Pretty grid display (no collapsing)
kbl(
  mapping_tbl,
  caption = "Racial appearance mapping (Original → Transformed) with counts",
  col.names = c("Original race", "Transformed race", "Total count"),
  align = c("l","l","r")
) %>%
  kable_classic(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```

## Hit rates by child and racial appearance  
```{r kids-x-race-to-excel, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(forcats)
library(knitr)
library(kableExtra)
library(openxlsx)

# 1) Create Adult/Child/Missing
ddat_ac <- ddat %>%
  mutate(
    Age_group_clean = fct_explicit_na(factor(`Age group`), "Missing"),
    Adult = case_when(
      Age_group_clean %in% c("Under 12", "12 to 17") ~ "Child",
      Age_group_clean %in% c("18 to 25","26 to 35","36 to 45","46 and over") ~ "Adult",
      Age_group_clean == "Missing" ~ "Missing age",
      TRUE ~ "Missing age"
    ) |> factor(levels = c("Adult","Child","Missing age")),
    Found = ifelse(as.character(Found) == "Yes", "Yes",
                   ifelse(as.character(Found) == "No", "No", NA))
  )

# 2) Build hit rate table (all races)
tab_all <- ddat_ac %>%
  group_by(
    Adult,
    Race = fct_explicit_na(factor(Racial.appearance.transformed), "Missing")
  ) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found == "Yes", na.rm = TRUE),
    `Hit rate (%)` = round(100 * Hits / pmax(Searches, 1), 1),
    .groups = "drop"
  ) %>%
  arrange(Adult, desc(Searches))

# 3) Pretty print in the report (fix align)
kbl(
  tab_all,
  caption = "Hit rates by Adult/Child/Missing age × Racial appearance (weapons searches)",
  align = c("l","l","r","r","r")
) %>%
  kable_classic(full_width = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"))

# 4) Write to Excel
wb <- createWorkbook()
addWorksheet(wb, "Tab_All")
writeData(wb, "Tab_All", tab_all)
saveWorkbook(wb, "tab_all.xlsx", overwrite = TRUE)
```