---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---


```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(strayr)
library(sf)
library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(viridis)

dat <- readRDS("../Output.data/Clean.search.data.RDS")

```

# Data overview 

This report summarises over 13,000 warrantless drug searches conducted by uniformed Victoria Police officers while patrolling on foot in public places. The data covers the years 2018, 2019, 2022, and 2023, and is sourced from  [**The Racial Profiling Data Monitoring Project**](https://khaki-pepper-b3rf.squarespace.com/). Data from 2020 and 2021 is excluded due to the extraordinary circumstances of the COVID-19 pandemic.

Under section 82 of the [*Drugs Poisons and Controlled Substances Act 1981*](http://classic.austlii.edu.au/au/legis/vic/consol_act/dpacsa1981422/s82.html), a police officer may stop and search a person in a public place if they reasonably suspect the person is in possession of illegal drugs. Drug searches are the most common type of warrantless search carried out by police in Victoria.

## Racial appearance categories  

Victoria Police records the perceived racial appearance of individuals subjected to stop and search. These categories are based on the officer’s perception, not the individual's self-reported ancestry or identity. 

***Missing data***

Racial appearance was not recorded in 2,860 search records — approximately 21% of all records — despite being a mandatory field for stop and search data collection since 2019.

## Melbourne Drug Searches – Evidence of Racial Profiling by Station  

Find rates — the percentage of searches resulting in the discovery of illegal drugs — for people of different recorded racial appearances at stations across Melbourne are presented below.

Where fewer than 20 individuals from a particular racial appearance group were searched at a station, find rates are not reported, as small sample sizes limit meaningful analysis.


```{r read geometry, echo=FALSE, warning=FALSE, message = FALSE}

melbs <- read_absmap("lga2022")

melbs <- melbs %>%
  filter(state_name_2021 == "Victoria") #


melbs %>%   # let's just look Melbourne
  ggplot() +
  geom_sf(aes(geometry = geometry))  # use the geometry variable
```


```{r Station, echo=FALSE, warning=FALSE, message = FALSE}
loc.racial <- dat %>%
  filter(!is.na(Station.uniform),Contact.Type == "P", 
         !is.na(`Search.type - Drugs`),
        Area.type == "Metro")%>%
  group_by(Division, Station.uniform)%>%
  mutate(`Finds all (%)` = round(sum(Found == "Yes", na.rm = T)/n()*100, digits = 0), 
         Total.searches = n())%>%
  mutate(Racial.appearance.abridged = fct_infreq(Racial.appearance.abridged))%>%
  group_by(Division, Station.uniform, Total.searches, `Finds all (%)`, Racial.appearance.abridged)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), Searches = n())%>%
  ungroup()%>%
  mutate(`Hit rate` = round(Finds/Searches*100, digits = 1))%>%
  # Limit to 20 in group
  mutate(`Hit rate` = ifelse(Searches <= 20, NA, `Hit rate`))%>%
  select(-Finds)%>%
  pivot_wider(names_from = Racial.appearance.abridged,
    values_from = c(`Hit rate`, Searches)
  )%>%
  #Add disproportionality ratio
  #Add disproportionality ratio
  mutate(`Relative likelihood: African` = 
           ifelse(
             `Hit rate_African`>0, 
             round(((`Hit rate_White`-`Hit rate_African`)/`Hit rate_African`), digits = 1 ), 
             round((`Hit rate_White`-`Hit rate_African`)/(`Hit rate_African` + 1), digits = 1 )
                  )
         )
         
names(loc.racial)#         

loc.racial.data <- loc.racial %>%
 select(
   Division,
    Station = Station.uniform, 
   `Total drug searches` = Total.searches,
    `Finds all (%)`,  
    
    `Searches: White` = `Searches_White`, 
    `Finds: White (%)` = `Hit rate_White`,
    
    `Searches: African` = `Searches_African`,
    `Finds: African (%)` = `Hit rate_African`,
   
    `Searches: Asian` = `Searches_Asian`, 
    `Finds: Asian (%)` = `Hit rate_Asian`, 
   
    `Searches: South Asian` = `Searches_South Asian`,
    `Finds: South Asian (%)` = `Hit rate_South Asian`,
   
    `Searches: Aboriginal` = Searches_Aboriginal, 
    `Finds: Aboriginal (%)` = `Hit rate_Aboriginal`, 
   
    `Searches: Middle Eastern` = `Searches_Middle Eastern`, 
    `Finds: Middle Eastern (%)`= `Hit rate_Middle Eastern` , 
   
    `Searches: Mediterranean/Mid` = `Searches_Mediterarranean/Mid`,
    `Finds: Middle Eastern/Mid (%)`=`Hit rate_Mediterarranean/Mid`,
   
    `Searches: Pacific Islander` = `Searches_Pacific Islander`,
    `Finds: Pacific Islander (%)` = `Hit rate_Pacific Islander`, 
   
    `Searches: Other racialised` = `Searches_Other racialised`,
    `Finds: Other racialised (%)` = `Hit rate_Other racialised`,
    `Searches: Missing` = Searches_Missing,
    `Finds: Missing (%)` = `Hit rate_Missing`,
    `Relative likelihood: African`)%>%
  arrange(desc(`Relative likelihood: African`), is.na(`Relative likelihood: African`))

loc.racial.display <- loc.racial.data %>%
 select(
    Station,
     `Search success likelihood: White compared with African` = `Relative likelihood: African`,
     `Finds: White (%)` , 
    `Finds: African (%)` , 
    `Finds: Asian (%)` , 
    `Finds: South Asian (%)` , 
    `Finds: Aboriginal (%)` , 
    `Finds: Middle Eastern (%)` , 
    `Finds: Pacific Islander (%)`, 
    `Finds all (%)`,
    `Total drug searches` 
  
   )

```

```{r display, echo=FALSE, warning=FALSE, message = FALSE}


# Apply the custom cell background color and bold text to the 'Ratio: African to White appearance finds' column

#display data 
cols_to_colour <- names(loc.racial.data)[c(4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)]

# Build the base table
dt <- datatable(
  loc.racial.data,
  rownames = FALSE,
  extensions = 'Buttons',
  class = 'cell-border',  # no 'stripe' class
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'excel', 'print'),
    pageLength = 20,
    autoWidth = TRUE
  )
)

# Apply a consistent background color (you can alternate or vary this if desired)
for (col in cols_to_colour) {
  dt <- dt %>% formatStyle(col, backgroundColor = '#f2f2f2')
}

dt


```


## Melbourne Drug Searches – Evidence of Racial Profiling by Local Government Area  

Find rates — the percentage of searches resulting in the discovery of illegal drugs — for people of different recorded racial appearances at stations across Melbourne are presented below.

Where fewer than 20 individuals from a particular racial appearance group were searched at a station, find rates are not reported, as small sample sizes limit meaningful analysis.


```{r LGA, echo=FALSE, warning=FALSE, message = FALSE}

loc.racial <- dat %>%
  filter(!is.na(Station.uniform),Contact.Type == "P", 
         !is.na(`Search.type - Drugs`),
        Area.type == "Metro")%>%
  group_by(Local.Government.Area)%>%
  mutate(`Finds all (%)` = round(sum(Found == "Yes", na.rm = T)/n()*100, digits = 1), 
         Total.searches = n())%>%
  mutate(Racial.appearance.abridged = fct_infreq(Racial.appearance.abridged))%>%
  group_by(Local.Government.Area, Total.searches, `Finds all (%)`, Racial.appearance.abridged)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), Searches = n())%>%
  ungroup()%>%
  mutate(`Hit rate` = round(Finds/Searches*100, digits = 1))%>%
  # Limit to 20 in group
  mutate(`Hit rate` = ifelse(Searches <= 10, NA, `Hit rate`))%>%
  select(-Finds)%>%
  pivot_wider(names_from = Racial.appearance.abridged,
    values_from = c(`Hit rate`, Searches)
  )%>%
  #Add disproportionality ratio
  mutate(`Relative likelihood: African` = round(((`Hit rate_White`-`Hit rate_African`)/`Hit rate_African`)*100, 
         digits = 1 ))
         

loc.racial.data <- loc.racial%>%
 select(
   LGA = Local.Government.Area,
   `Total drug searches` = Total.searches,
    `Finds all (%)`,  
    
    `Searches: White` = `Searches_White`, 
    `Finds: White (%)` = `Hit rate_White`,
    
    `Searches: African` = `Searches_African`,
    `Finds: African (%)` = `Hit rate_African`,
   
    `Searches: Asian` = `Searches_Asian`, 
    `Finds: Asian (%)` = `Hit rate_Asian`, 
   
    `Searches: South Asian` = `Searches_South Asian`,
    `Finds: South Asian (%)` = `Hit rate_South Asian`,
   
    `Searches: Aboriginal` = Searches_Aboriginal, 
    `Finds: Aboriginal (%)` = `Hit rate_Aboriginal`, 
   
    `Searches: Middle Eastern` = `Searches_Middle Eastern`, 
    `Finds: Middle Eastern (%)`= `Hit rate_Middle Eastern` , 
   
    `Searches: Mediterranean/Mid` = `Searches_Mediterarranean/Mid`,
    `Finds: Middle Eastern/Mid (%)`=`Hit rate_Mediterarranean/Mid`,
   
    `Searches: Pacific Islander` = `Searches_Pacific Islander`,
    `Finds: Pacific Islander (%)` = `Hit rate_Pacific Islander`, 
   
    `Searches: Other racialised` = `Searches_Other racialised`,
    `Finds: Other racialised (%)` = `Hit rate_Other racialised`,
    `Searches: Missing` = Searches_Missing,
    `Finds: Missing (%)` = `Hit rate_Missing`,
    `Relative likelihood: African`)%>%
  arrange(desc(`Relative likelihood: African`), is.na(`Relative likelihood: African`))
```


```{r choropleth, echo=FALSE, warning=FALSE, message = FALSE}

melbs <- melbs %>%
  mutate(LGA = lga_name_2022)

melbs2 <- melbs %>%
  select(LGA, geometry, cent_long, cent_lat)%>%
  mutate(LGA = case_when(
    LGA %in% c("Moreland") ~ "Merri-bek",
    LGA == "Kingston (Vic.)" ~ "Kingston",
    LGA == "Bayside (Vic.)"~ "Bayside",
    TRUE ~ LGA
  ))

melbs2 <- melbs2 %>%
  filter(LGA %in% c(unique(loc.racial.data$LGA)))
  


datm <- melbs2 %>%
  left_join(loc.racial.data, by = "LGA")%>%
  mutate(Relative.likelihood = case_when(
    is.na(`Relative likelihood: African`) ~ "Insufficient data",
    `Relative likelihood: African` >300 ~ "Over 300%",
    `Relative likelihood: African` >100 ~ "Over 100%",
    `Relative likelihood: African` >50 ~ "Over 50%",
    `Relative likelihood: African` >25 ~ "Over 25%",
    `Relative likelihood: African` >10 ~ "Over 10%",
    `Relative likelihood: African` >-10 ~ "Within 10%",
    `Relative likelihood: African` >-100 ~ "Half as likely"))

datm <- datm %>%
   mutate(Colour = case_when(
    is.na(`Relative likelihood: African`) ~ "#D3D3D3",
    `Relative likelihood: African` >300 ~ "#7f0000",
    `Relative likelihood: African` >100 ~ "#b30000",
    `Relative likelihood: African` >50 ~ "#d7301f",
    `Relative likelihood: African` >25 ~ "#ef6548",
    `Relative likelihood: African` >10 ~ "#fdbb84",
    `Relative likelihood: African` >-10 ~ "#ffffbf",
    `Relative likelihood: African` >- 100~ "#4292c6"))#
    
    


# 1. Set your geo layout (no invalid fields!)
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'mercator')  # Or just use default
)

# 2. Generate your base plot (choropleth)
p <- plot_ly() %>%
  add_sf(
    data = datm,
    split = ~LGA,
    color = ~I(Colour),
    text = ~paste0(
      "LGA: ", LGA, "<br>",
      "Relative likelihood: ", `Relative likelihood: African`, "<br>",
      "Searches: ", `Total drug searches`
    ),
    hoveron = "fills",
    hoverinfo = "text",
    showlegend = FALSE  # hide map legend, we build it manually
  )

# 3. Custom legend entries — use type = "scatter" NOT "scattergeo"
legend_items <- datm %>%
  distinct(Relative.likelihood, Colour) %>%
  arrange(match(Relative.likelihood, c(
    "Over 300%", "Over 100%", "Over 50%", "Over 25%", "Over 10%",
    "Within 10%", "Half as likely%", "Insufficent data"
  )))

# Add fake points off-map that don’t mess with projection
for (i in seq_len(nrow(legend_items))) {
  p <- p %>%
    add_trace(
      type = "scatter",
      mode = "markers",
      x = list(NA),  # No actual data
      y = list(NA),
      marker = list(color = legend_items$Colour[i], size = 10),
      name = legend_items$Relative.likelihood[i],
      hoverinfo = "none",
      showlegend = TRUE,
      inherit = FALSE
    )
}

# 4. Apply layout (correct geo and legend positioning)
p <- p %>%
  layout(
    geo = g,
    legend = list(
      title = list(text = "Relative likelihood"),
      x = 1.02,
      y = 0.95
    )
  )

p
```



```{r display 2, echo=FALSE, warning=FALSE, message = FALSE}




# Apply the custom cell background color and bold text to the 'Ratio: African to White appearance finds' column

#display data 
cols_to_colour <- names(loc.racial.data)[c(3, 5, 7, 9, 11, 13, 15, 17, 19, 21, 23)]

# Build the base table
dt <- datatable(
  loc.racial.data,
  rownames = FALSE,
  extensions = 'Buttons',
  class = 'cell-border',  # no 'stripe' class
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'excel', 'print'),
    pageLength = 20,
    autoWidth = TRUE
  )
)

# Apply a consistent background color (you can alternate or vary this if desired)
for (col in cols_to_colour) {
  dt <- dt %>% formatStyle(col, backgroundColor = '#f2f2f2')
}

dt


```

# Appendix - ethnic appearance mapping  

Ethnic appearance categories for different years were transformed to enable comparison across the four years. A table of transformations undertaken, along with the number of searches in each category, appears below,


```{r Map 1, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping1 <-dat %>%
  filter(Year %in% c(2022, 2023))%>%
  count(Racial.Appearance.original, Racial.appearance.abridged)

ethnic.mapping1 %>%
  kable()
```

```{r Map 2, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping2 <- dat %>%
  filter(Year %in% c(2018, 2019))%>%
  count(Racial.Appearance.original, Racial.appearance.abridged) 


ethnic.mapping2 %>%
  kable()
```