---
title: "Victoria Police Search Data Overview and Hit Rates"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

```

# Victoria Police Warrantless Search Data (Melbourne Searches)

> **TL;DR:**  
> This dataset includes over **50,000 warrantless searches** conducted by Victoria Police in **2018, 2019, 2022, and 2023**, obtained through the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/). It captures key details such as **search type**, **search outcomes**, and **perceived racial appearance**.  
> 
> The analysis provides **summary statistics** and **exploratory hit rate comparisons** for over 25,000 drug searches undertaken by uniform police in Melbourne to help identify patterns that warrant further investigation. Findings should be interpreted with caution — **formal significance testing has not been applied** to most of the descriptive summaries presented on this page.


## About the Dataset

This dataset contains records of over **50,000 warrantless searches** conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**. 

Data for the years **2020 and 2021** is excluded due to the atypical policing environment associated with the COVID-19 pandemic.

The dataset comprises records of **warrantless stop and search events** conducted in public places across Victoria. It was released through a Freedom of Information (FOI) request lodged by the[Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/) and includes detailed information about each search, including location, officer-perceived characteristics of the person searched, search type, and outcome.

For a warrantless search to be lawful, police must **suspect on reasonable grounds** that the person is carrying a prohibited item, such as drugs, weapons, or graffiti implements.

Each search is recorded under the relevant **legal power** (e.g. *Drugs, Poisons and Controlled Substances Act 1981*, s82) and includes one of three possible **outcomes**:

- **Nothing found**  
- **Search item found**: an item matching the type of contraband suspected (e.g. drugs found during a drug search)  
- **Non-search item found**: a different type of contraband was found (e.g. a weapon found during a drug search)

This dataset has been cleaned and transformed to enable analysis of **hit rates**, with particular focus on **racial disparities in drug searches**.

### Key Variables

The following key variables are are available in this dataset. 

- `Date`: Date of the stop and search  
- `Area.type`: Metropolitan or regional classification  
- `Region`, `Division`, `PSA`, `Station`: Police geographic hierarchy  
- `Racial.appearance.original`: The original officer-perceived racial appearance category as recorded by Victoria Police  
- `Racial.appearance.transformed`: A harmonised version of racial appearance, created to adjust for multiple changes in category definitions across the reference period  
- `Gender`: Officer-perceived gender of the person searched  
- `Age`: Age group of the person searched  
- `Search.type - Drugs`, `Search.type - Weapons`, `Search.type - Graffiti`: Indicates the type(s) of prohibited item being searched for  
- `Found`: Whether **any** prohibited item was found — regardless of whether it was the item searched for under the power invoked (i.e. both drug power invoked, and drugs found, *and* drug power invoked, but weapons found instead count as a 'hit') 
- `Contact.Type`: Whether the search was conducted on foot (`P`) or during a vehicle stop (`V`)  
- `Unit.type`: Broad category of the police unit conducting the search (e.g. uniform, highway patrol)  
- `Station.uniform`: The police station linked to the uniformed officer conducting the search (available for uniform police only)
- `Rank`: Rank of the officer responsible for conducting the search (e.g. Constable, Senior Constable, Sergeant) 

## Data limitations  

### Racial Appearance Categories

Victoria Police records the **perceived racial appearance** of individuals at the time of the search. These categories reflect how the officer perceives the person’s racial appearance — they are *not* based on self-identification or ancestry and should be understood as indicators of how race is *perceived* in the policing context.

### Missing Data

A substantial proportion of records are missing demographic information about the person who was searched.

In particular, **racial appearance information is either missing or undetermined in a significant number of cases**. Specifically, **42% of all searches (22,117 records)** have no recorded racial appearance — despite this being a mandatory field since 2019. An additional **1.5% of searches (785 records)** are recorded as **‘Other’**, a category sometimes used when officers are unable or unwilling to assign a racial appearance. Together, these issues raise concerns about data quality and the consistency with which demographic details are recorded.

In addition, **2,097 records** are classified under the merged **‘Mediterranean/Middle Eastern’** category. Since 2020, Victoria Police has combined these two appearance labels, limiting the ability to meaningfully assess outcomes for people of Middle Eastern appearance. While individuals of Mediterranean descent are often perceived as non-racialised or White, those of Middle Eastern appearance may be racialised differently in police interactions. The merger obscures this distinction, rendering these records **unusable for racial disparity analysis** involving Middle Eastern individuals.

Altogether, these limitations affect over **45% of all records**, substantially impacting the completeness and reliability of race-based analysis.

📁 **Download the cleaned dataset:**  
[**VicPol Search Data Clean.xlsx**](https://github.com/monikasarder/VictoriaPoliceRacialSearch/raw/main/VicPol%20Search%20Data%20Clean/VicPol%20Search%20Data%20Clean.xlsx)

## About the Analysis

This analysis is **exploratory** and provides a high-level overview of Victoria Police stop and search practices, with the aim of informing further research, advocacy, and policy reform.

Although the original dataset includes over 50,000 records of warrantless searches conducted across Victoria between 2018 and 2023, this analysis is **limited to drug searches** carried out by **uniform police officers** on **foot patrol** within the **Melbourne metropolitan area**. This scope was chosen for several key reasons:

- **Drug searches** represent the vast majority of all search records — approximately **87%** — and are the most consistently recorded type of search.
- Different search types (e.g. for weapons, graffiti implements, or volatile substances) have markedly different hit rates, reflecting differing legal thresholds and suspect profiles. Focusing only on drug searches reduces this variation and enhances comparability.
- **Uniform police** are both the largest cohort and less likely to be engaged in specialist or covert operations. They are also the officers that members of the public are most likely to encounter during routine interactions in public places.
- Restricting the analysis to **foot patrols** improves data quality, as these records have significantly less missing information — especially for **racial appearance**, which is a key variable of interest.
- Limiting the analysis to **Melbourne** helps control for geographic differences in demographics, policing strategies, and enforcement practices that may confound the results.

By focusing on a more homogeneous subset of the data — **drug searches by uniform police on foot patrol in metropolitan Melbourne** — we aim to reduce contextual variation and provide a more reliable basis for examining potential racial disparities in stop and search outcomes.


### Hit Rate Analysis

**Hit rate analysis is a standard method used to detect possible racial profiling**. It involves comparing the proportion of searches that result in the discovery of prohibited items — known as a “find” — across different racial groups. This approach is based on the principle that if all individuals are being searched under the same legal threshold, then hit rates should be broadly similar across groups. A consistently lower hit rate for a particular group may suggest that police are applying a lower threshold of suspicion, raising concerns about potential bias ([Gelman, Fagan & Kiss, 2007](https://www.jstor.org/stable/40059996); [Epp, Maynard-Moody & Haider-Markel, 2014](https://press.uchicago.edu/ucp/books/book/chicago/P/bo18061052.html)).

This analysis focuses specifically on **drug searches conducted under Section 82 of the *Drugs, Poisons and Controlled Substances Act 1981***. Under this provision, police are permitted to search a person without a warrant if they form a reasonable belief that the person is carrying illicit drugs.

Hit rates may be influenced by a range of factors such as the **type of search**, the **police unit conducting it**, or the **local area** in which the search occurred. For example, some suburbs may have higher or lower hit rates due to differences in enforcement practices, officer discretion, or local demographics.

Therefore, to determine whether **race** is independently associated with lower hit rates, it is important to explore how these other factors may be influencing the results — and whether they are **confounding** the apparent impact of race. This section provides a preliminary exploration of hit rate patterns to help guide more robust, controlled statistical analysis in subsequent sections.

# Data Overview  

## Data summary  

This section provides a summary of the **13,896** records included in the analysis. These represent **warrantless drug searches** conducted by **uniform police on foot patrol** in **Melbourne metropolitan areas** between **2018 and 2023**.

The summary outlines the distribution of key variables such as year, racial appearance, gender, search outcomes, and other relevant characteristics. Notably, only 20.6% of the records are missing racial appearance data and only 4.5% of records are mssing gender data.

**Table 1 Characteristics of searches**
```{r summary, echo = FALSE, warning = FALSE}
of.interest <- c("Year","Racial.appearance.transformed","Gender","Rank.of.Member", "Division","Found",
                 "Search.type - Drugs")

datsum <- dat %>%
  filter(!is.na(`Search.type - Drugs`),
        Area.type == "Metro",
         Unit.type == "Uniform",
        Contact.Type == "P")

datsum$Racial.appearance.transformed[datsum$Racial.appearance.transformed == "Missing"] <- NA  

datsum <- datsum %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))%>%
  mutate(Year = fct_relevel(as.character(Year), c("2018", "2019", "2022", "2023")))

print(dfSummary(datsum, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```



## Hit Rate Patterns

This section summarises whether the likelihood of a “find” (i.e. discovery of a prohibited item) varied across key demographic and operational variables. The analysis includes **13,896 warrantless drug searches** conducted by **uniform police on foot patrol** in **Melbourne** between **2018 and 2023**. The overall **average hit rate was 13.5%**.


### Racial Appearance 

Hit rates varied significantly across racial appearance categories (*p = 0.0216*). Individuals perceived as **White** had a hit rate of **14.1%**, higher than most racialised groups. Lower hit rates were observed for:

- **South Asian**: 9.7%  
- **African**: 11.4%  
- **Middle Eastern**: 10.4%  
- **Other**: 10.6%

### Year of Search

There were statistically significant differences in hit rates across years (*p = 0.0039*). The highest hit rate occurred in **2019 (15.3%)**, followed by **2022 and 2023 (13.3%)**, and the lowest in **2018 (12.1%)**. This may reflect changing operational priorities or enforcement practices over time.


These disparities suggest that individuals from certain racial groups may be subject to more frequent searches under a lower threshold of suspicion — a key indicator of potential racial profiling.

### Gender 

Gender was significantly associated with hit rate (*p = 0.0005*), although the differences were relatively modest: 

- **Male**: 13.5%  
- **Female**: 12.4%  
- **Missing gender**: 19.0%  
- **Unknown (U)**: 0%


### Rank of Police Member  

Rank was not significantly associated with differences in hit rates (*p = 0.0991*). While **Sergeants (SGT)** had a slightly higher hit rate (16.8%), the difference was not statistically robust. Constables and Senior Constables, who conducted the majority of searches, had similar rates (13.1% and 14.1% respectively).

### Police Division  

Hit rates also varied significantly by police division (*p < 0.0001*), suggesting that local enforcement practices or operational focus may influence search outcomes. Divisions with the lowest hit rates included:

- **North West Metro Region (NWMR) Division 1 (Melbourne, Yarra)**: 10.3%  
- **NWMR Division 3 (Brimbank, Melton)**: 11.4%  
- **NWMR Division 4  (Hume, Moonee Valley, Merri-bek)**: 11.6%  

In contrast, divisions with the highest hit rates included:

- **Southern Metro Region (SMR) Division 2 (Glen Eira, Kingston)**: 18.6%  
- **SMR Division 4 (Frankston, Mornington)**: 18.0%  
- **Eastern Region Division 1 (Boroondara, Manningham, Monash, Whitehorse)**: 18.2%  

Divisions with lower hit rates tended to cover areas with more culturally diverse populations and greater socioeconomic disadvantage, while those with higher hit rates were generally less diverse and more advantaged. These patterns suggest that racial disparities in search outcomes depend on a range of local factors.


**Table 3 Hit rates by variables of interest - Metro**
```{r Finds racial, echo=FALSE, warning=FALSE, results='asis'}


# Recode and reorder factors
datsum <- datsum %>%
  mutate(
    Found = factor(Found, levels = c("Yes", "No")),  # Yes before No
    Racial.appearance.transformed = fct_explicit_na(Racial.appearance.transformed, na_level = "Missing"),
    Gender = fct_explicit_na(Gender, na_level = "Missing"),
    Year = fct_relevel(as.character(Year), "2018", "2019", "2022", "2023"),
    across(c("Year", "Racial.appearance.transformed", "Gender", "Rank.of.Member"), ~ fct_infreq(.))
  )

# Select only variables of interest + Found
vars_for_table <- c("Year", "Racial.appearance.transformed", "Gender", "Rank.of.Member", "Found", "Division")

# Run crosstable across multiple rows
crosstable(datsum[, vars_for_table], 
           by = "Found", 
           total = TRUE, 
           test = TRUE, 
           showNA = "ifany", 
           percent_digits = 1) %>%
  as_flextable()

```

### Implications for Missing Data and Confounding Variables

It was determined that, for modelling purposes, the dataset would be filtered to exclude unusable or ambiguous records.

The following decisions were made regarding the treatment of potential confounding variables:

- **Missing racial appearance** records will be excluded from the remainder of the analysis. While still present in the dataset (20.6%), their exclusion ensures that comparisons are based on interpretable and distinct racial appearance categories. Given their hit rate is similar to that of White-appearing individuals, their removal is unlikely to meaningfully bias the analysis.

- The **Mediterranean/Middle Eastern** category will also be excluded. This merged label combines individuals who may be racialised (e.g. Middle Eastern appearance) with those who are not (e.g. Southern European descent), reducing its analytic value and potentially obscuring meaningful differences.

- **Year** will be retained and explored as a potential confounding variable in the next section. The observed variation in hit rates across years suggests that operational focus or discretionary targeting may have changed over time. If certain racial groups were subject to increased scrutiny in specific years, this could influence hit rates independently of race and must be accounted for.

- **Gender** and **police rank** will not be retained as confounders. While minor differences in hit rates were observed, these variables are **not expected to be meaningfully associated with race** in the context of police search decision-making. Their inclusion would add complexity without significantly strengthening the validity of the analysis.

- **Police division** will require further attention. As shown in the previous section, there are large differences in hit rates between divisions, which may reflect differences in local policing approaches, demographics, or community relationships. To isolate the effect of race, we will either need to **control for division in the modelling** or **analyse racial disparities within individual divisions**. This will ensure that any observed disparities are not confounded by geographic variation in policing outcomes.

These decisions ensure that the remaining analysis is focused on clear, interpretable comparisons while reducing the risk of bias from poorly defined or inconsistently applied categories.

## Hit Rates Cross Tabs - Unit Type and LGA  

### Hit Rates by Racial Appearance and Unit Type

Racialised hit rates across different unit types were compared. As described above, statistics should be interpreted with caution as further significance testing is necessary to ensure that any observed differences in hit rates are not due to random variation and can be meaningfully interpreted.


### Hit Rates by Racial Appearance and Local Government Area

Searches that were undertaken by a uniform officer based at a local Police station were able to be mapped to their Local Government Area.   As described above, statistics should be interpreted with caution as further significance testing is necessary to ensure that any observed differences in hit rates are not due to random variation and can be meaningfully interpreted.

The hit rates for persons of different racial appearances for Local Government Areas appears below.   

## Appendix – racial Appearance Mapping

To enable consistent comparison across the four years of data, racial appearance categories were standardised. This step was necessary because Victoria Police modified the racial appearance categories multiple times during the period under review. The table below outlines the transformations applied, along with the corresponding number of searches in each category.


**Table 8: Mapping of original racial appearance categories to standardised categories**
```{r Map 1, echo=FALSE, warning=FALSE, message = FALSE}

racial.mapping1 <-dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed)

racial.mapping1 %>%
  kable()
```

