---
title: "Victoria Police Search Data Overview and Hit Rates"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)

dat <- readRDS("../VicPol Search Data Clean/Clean.search.data.RDS")

```

# Victoria Police Search Data

> **TL;DR:**  
> This dataset includes over **50,000 warrantless searches** conducted by Victoria Police in **2018, 2019, 2022, and 2023**, obtained through the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/). It includes **search type**, **outcomes**, and **perceived racial appearance**.  
> The analysis provides **summary statistics** and **exploratory hit rate comparisons** to identify patterns for further investigation. Interpret findings with caution — this is not a causal or statistical inference study.

This dataset contains information on over 50,000 warrantless searches conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**. It was obtained through the [**Racial Profiling Data Monitoring Project**](https://khaki-pepper-b3rf.squarespace.com/), an initiative of the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/).  Data for 2020 and 2021 is not included due to the extraordinary and atypical policing context of the COVID-19 pandemic. These years are excluded to avoid drawing misleading inferences about standard police practices.

You can download the cleaned dataset used in this analysis as an Excel file from the GitHub repository:  
[**Download VicPol Search Data Clean.xlsx**](https://github.com/monikasarder/VictoriaPoliceRacialSearch/raw/main/VicPol%20Search%20Data%20Clean/VicPol%20Search%20Data%20Clean.xlsx)

This analysis is **exploratory** and designed to provide an overview of the data to assist researchers, advocates, and policymakers in identifying areas for further work. It includes:

- **Summary statistics** of search volumes, types, and outcomes;  
- **Hit rate comparisons** by racial appearance, unit type, search context, area, and gender.

The hit rate analysis uses basic proportions to assess the likelihood that a search resulted in a “find.” While this can reveal preliminary patterns, these comparisons should be interpreted with caution. Hit rates may be influenced by underlying contextual or operational differences — such as search type, officer discretion, or the policing environment — and do **not** control for confounding variables.

To draw valid conclusions about potential racial profiling or systemic bias, further statistical testing is needed — such as **multivariate or panel data modelling**, or **focused subgroup analysis** (e.g. uniform police foot patrol drug searches in metropolitan Melbourne).

## Search Types in Scope

The dataset includes records of five types of **warrantless searches**, each of which requires that a police officer has *reasonable grounds* to suspect the person is in possession of a prohibited item relevant to that search power:

- **Drugs**: Section 82, *Drugs, Poisons and Controlled Substances Act 1981*  
- **Volatile substances for inhalation** (under 18s): Sections 60E–60F, *Drugs, Poisons and Controlled Substances Act 1981*  
- **Weapons or dangerous articles**: Section 10, *Control of Weapons Act 1990*  
- **Firearms**: Section 149, *Firearms Act 1996*  
- **Graffiti implements** (at designated places): Section 13, *Graffiti Prevention Act*  

Nearly **90%** of searches were conducted under drug search powers. Some individuals were searched under multiple powers (e.g. both drug and weapon suspicions), and this is reflected in the dataset.

## Racial Appearance Categories

Victoria Police records the **perceived racial appearance** of individuals searched, based on officer observation. These categories reflect *perceived raciality*, not self-identification or ancestry data.

### Missing Data

A significant portion of records — **22,117 searches or 42%** — lack racial appearance data, despite this field being mandatory for stop and search events since 2019. In cases where raciality could not be determined, officers used the category **‘Other’**.

### Middle Eastern Category Limitation

In 2020, Victoria Police merged the **‘Middle Eastern’** and **‘Mediterranean’** appearance categories into one. This limits the ability to meaningfully assess searches involving people of Middle Eastern appearance. Typically, people of Mediterranean descent might be categorised as *non-racialised* or *White*, while people of Middle Eastern appearance are more likely to be viewed as *racialised*. The aggregation obscures this distinction and hampers analysis of potential racial profiling.

# Summary Statistics

The table below summarises key variables in the dataset, including demographic fields, search types, and outcomes.



**Table 1 Characteristics of searches**
```{r summary 2, echo = FALSE, warning = FALSE}

of.interest <- c("Year", "Racial.appearance.transformed","Gender", "Area.type", "Unit.type", "Contact.Type" ,"Rank.of.Member", "Found",
                 "Search.type - Drugs","Search.type - Weapons","Search.type - Firearms",          
                  "Search.type - Graffiti","Search.type - Volatile.sub.U18","Search.type - Volatile.sub.adult")

`VicPol station data` <- dat %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))
  
print(dfSummary(`VicPol station data`, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

Table 2 below shows the breakdown of searches by search type. This summary is presented separately from the main dataset because some individual searches involved multiple search powers — for example, when an officer believed they had reasonable grounds to suspect a person was carrying both drugs and weapons.


**Table 2 Legislative basis for searches**
```{r summary search type, echo = FALSE, warning = FALSE}

search.types <- c("Search.type - Drugs", "Search.type - Weapons","Search.type - Firearms",          
                  "Search.type - Graffiti" , "Search.type - Volatile.sub.U18","Search.type - Volatile.sub.adult")

Search.dat <- dat %>%
  pivot_longer(cols =search.types,
               names_to = "Search.type", values_to = "Search.outcome")%>%
  filter(!is.na(Search.outcome))%>%
  mutate(Search.type = str_remove(Search.type, "Search.type - "))%>%
  mutate(Search.type = ifelse(str_detect(Search.type, "Vol"), "Volatile substance", Search.type))

`Search type` <- Search.dat %>%
  select(Search.type)%>%
  mutate(Search.type = fct_infreq(Search.type))

print(dfSummary(`Search type`, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

# Hit Rate Analysis

The **hit rate** — the proportion of searches that result in the discovery of contraband — is a key measure of how effectively the legal threshold of *reasonable suspicion* is applied. Across all warrantless searches in the dataset, the average hit rate was **17.2%**. When broken down by perceived racial appearance, the data suggests that some groups are searched under looser thresholds than others, pointing to the possibility of racial bias. Communities subjected to disproportionate or less effective searches often experience this as racial targeting.

## Hit Rates by Racial Appearance

Compared with people perceived to be **White** (hit rate: **18.0%**), significantly lower hit rates were observed for:

- **African appearance**: 12.9%  
- **Middle Eastern/Mediterranean**: 13.8%  
- **Asian appearance**: 13.7%  
- **South Asian appearance**: 13.9%  
- **Aboriginal and Torres Strait Islander**: 15.6%  

The only group with a higher hit rate than White-appearing individuals was:

- **Pacific Islander appearance**: 18.8%

These differences were found to be statistically significant using **Pearson’s chi-squared test**. Table 3 below summarises these racial appearance hit rates.

These findings aggregate **all search types**, **all police units**, and **all locations**. Search thresholds likely vary between units and contexts, and there are differences between **foot searches and vehicle searches** that are not separated here. 

To address these limitations, further analysis was conducted focusing solely on **drug searches by uniform police on foot in metropolitan Melbourne**, where conditions are more homogeneous. More rigorous statistical modelling — controlling for search type, location, operational unit, and search context — would be needed to isolate the effect of perceived racial appearance on search outcomes.  

Also, over **40%** of records are missing racial appearance data. While these are excluded from the breakdown, their absence limits the completeness of the comparison and may affect the generalisability of the findings. The missing data may also reflect systemic issues in police data collection and reporting.


**Table 3 Hit rates by variables of interest**
```{r Finds racial, echo=FALSE, warning=FALSE}
dat <- dat %>%
  mutate(Found = factor(Found, levels = c("Yes", "No")))

dat %>% 
  crosstable(c(Racial.appearance.transformed), by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()
```

### Change Over Time  

Figure 1 below illustrates how hit rates varied by perceived racial appearance across the four-year period. A red reference line indicates the overall average hit rate across all searches during this time.

Key observations include:

- Individuals perceived to be **White** consistently experienced higher-than-average hit rates in most years, except in **2022**, which may reflect the lingering effects of pandemic-related disruptions.  
- Individuals perceived to be **African**, **Middle Eastern**, and **Asian** consistently recorded **lower hit rates** than the overall average across all years.  
- Hit rates for individuals perceived as **Aboriginal and Torres Strait Islander** increased significantly over time — from **11% in 2018** to **18.6% in 2023** — which may suggest a reduction in unjustified or racially biased searches affecting this group.

> **Caution:**  
> This is an **exploratory analysis**, and the findings should be interpreted with caution. This summary does not account for important covariates such as search type, policing unit, location, or search context (e.g. on foot vs in vehicle).  
>  
> A more rigorous approach would involve **panel data modelling**, which would enable us to test whether the observed disparities persist once differences in time, geography, unit culture, and operational context are taken into account.


**Figure 1 Changes in hit rates by racial appearance over time and compared to the average**
```{r Finds race year, echo=FALSE, warning=FALSE, message = FALSE}
trend <- dat %>%
  filter(!Racial.appearance.transformed %in% c("Missing", "Mediterranean AND Middle Eastern - DO NOT USE", "Other"))%>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed))%>%
  group_by(Year, `Racial.appearance`)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), `Total searches`= n())%>%
  mutate(`Hit rate` = round(Finds/`Total searches`*100, digits = 1)) %>%
  ungroup()%>%
  ggplot(aes(x = Year, y = `Hit rate`)) +geom_col(fill = "darkblue")+
  facet_wrap(~`Racial.appearance`)+ 
  geom_hline(yintercept = 17.2,linetype="dotted", color = "red") +theme(legend.position = "none")

ggplotly(trend)
```


## Hit Rates by Other Characteristics

Hit rates by search type vary across a number of other characteristics that may influence the likelihood of a “find.” These include the type of police unit conducting the search, the gender of the person searched, and whether the search occurred in a regional or metropolitan area. Variations across these dimensions are summarised in the tables below, reinforcing the importance of controlling for contextual factors when analysing racial disparities in policing outcomes.


**Table 4 Hit rates by Search Type**

```{r Finds search, echo=FALSE, warning=FALSE}

Search.dat %>%
  mutate(Found = factor(Found, levels = c("Yes", "No")))%>%
  crosstable(Search.type, by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()

```

**Table 5 Hit rates by Other Search Characteristics**
```{r Finds all other, echo=FALSE, warning=FALSE}

dat %>%
  crosstable(c(Year, Unit.type, Contact.Type, Gender, Area.type), by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()
```

## Hit Rates Cross Tabs - Unit Type and LGA  

### Hit Rates by Racial Appearance and Unit Type

Racialised hit rates across different unit types were compared. As described above, statistics should be interpreted with caution as further significance testing is necessary to ensure that any observed differences in hit rates are not due to random variation and can be meaningfully interpreted.

**Table 6 racial appearance hit rates by Unit type**

```{r unit.type, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(forcats)
library(DT)

# Step 1: Summarise search and hit counts by unit and race
unit_racial_long <- dat %>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed)) %>%
  group_by(Unit.type, Racial.appearance) %>%
  summarise(
    Searches = n(),
    Finds = sum(Found == "Yes", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `Hit rate` = round(Finds / Searches * 100, 1),
    Searches = ifelse(Searches <= 5, NA, Searches),
    `Hit rate` = ifelse(is.na(Searches), NA, `Hit rate`)
  ) %>%
  select(Unit.type, Racial.appearance, Searches, `Hit rate`) %>%
  pivot_longer(cols = c(Searches, `Hit rate`), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = paste(Racial.appearance, Metric, sep = " – ")) %>%
  select(-Racial.appearance)

# Step 2: Pivot wide to get one row per Unit.type
unit_racial_wide <- unit_racial_long %>%
  pivot_wider(names_from = Metric, values_from = Value)

# Step 3: Calculate total searches and overall hit rate
unit_summary <- dat %>%
  group_by(Unit.type) %>%
  summarise(
    `Total searches` = n(),
    `Overall %` = round(sum(Found == "Yes", na.rm = TRUE) / n() * 100, 1),
    .groups = "drop"
  )

# Step 4: Join and arrange
unit_racial_final <- unit_summary %>%
  left_join(unit_racial_wide, by = "Unit.type") %>%
  arrange(desc(`Total searches`))

# Step 5: Display table
datatable(unit_racial_final,
          rownames = FALSE,
          extensions = 'Buttons',
          options = list(
            dom = 'Blfrtip',
            buttons = c('copy', 'excel', 'print'),
            pageLength = 10,
            scrollX = TRUE,
            autoWidth = TRUE
          ))


```

### Hit Rates by Racial Appearance and Local Government Area

Searches that were undertaken by a uniform officer based at a local Police station were able to be mapped to their Local Government Area.   As described above, statistics should be interpreted with caution as further significance testing is necessary to ensure that any observed differences in hit rates are not due to random variation and can be meaningfully interpreted.

The hit rates for persons of different racial appearances for Local Government Areas appears below.   

**Table 7 racial appearance hit rates by LGA**
```{r final_lga_table, echo=FALSE, warning=FALSE, message=FALSE}


# Step 1: Filter to uniform police and valid LGA
filtered <- dat %>%
  filter(Unit.type == "Uniform", !is.na(Local.Government.Area))

# Step 2: Compute searches and hits by LGA and racial appearance
lga_race_stats <- filtered %>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed)) %>%
  group_by(Local.Government.Area, Racial.appearance) %>%
  summarise(
    Searches = n(),
    Finds = sum(Found == "Yes", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    `Hit Rate` = round(100 * Finds / Searches, 1),
    Searches = ifelse(Searches <= 5, NA, Searches),
    `Hit Rate` = ifelse(is.na(Searches), NA, `Hit Rate`)
  )

# Step 3: Pivot longer then wider for proper interleaving
lga_long <- lga_race_stats %>%
  pivot_longer(cols = c(Searches, `Hit Rate`), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = paste(Racial.appearance, Metric, sep = " – ")) %>%
  select(Local.Government.Area, Metric, Value)

lga_wide <- lga_long %>%
  pivot_wider(names_from = Metric, values_from = Value)

# Step 4: Overall stats by LGA
lga_overall <- filtered %>%
  group_by(Local.Government.Area) %>%
  summarise(
    `Total Searches` = n(),
    `Overall %` = round(100 * sum(Found == "Yes", na.rm = TRUE) / n(), 1),
    .groups = "drop"
  )

# Step 5: Join and arrange final table
lga_final <- lga_overall %>%
  left_join(lga_wide, by = "Local.Government.Area") %>%
  arrange(desc(`Total Searches`))

# Step 6: Display
datatable(lga_final,
          rownames = FALSE,
          extensions = 'Buttons',
          options = list(
            dom = 'Blfrtip',
            buttons = c('copy', 'excel', 'print'),
            pageLength = 10,
            scrollX = TRUE,
            autoWidth = TRUE
          ))
```


## Appendix – racial Appearance Mapping

To enable consistent comparison across the four years of data, racial appearance categories were standardised. This step was necessary because Victoria Police modified the racial appearance categories multiple times during the period under review. The table below outlines the transformations applied, along with the corresponding number of searches in each category.


**Table 8: Mapping of original racial appearance categories to standardised categories**
```{r Map 1, echo=FALSE, warning=FALSE, message = FALSE}

racial.mapping1 <-dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed)

racial.mapping1 %>%
  kable()
```

