---
title: "Victoria Police search data overview and hit rates"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---


```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)

dat <- readRDS("../VicPol Search Data Clean/Clean.search.data.RDS")


```

# Victoria Police Search Data

This dataset contains information on over 50,000 warrantless searches conducted by Victoria Police in the years **2018, 2019, 2022, and 2023**. It was obtained through the [**Racial Profiling Data Monitoring Project**](https://khaki-pepper-b3rf.squarespace.com/), an initiative of the [Centre Against Racial Profiling](https://www.centreagainstracialprofiling.au/).

You can download the cleaned dataset used in this analysis as an Excel file from the GitHub repository:  
[**Download VicPol Search Data Clean.xlsx**](https://github.com/monikasarder/VictoriaPoliceRacialSearch/raw/main/VicPol%20Search%20Data%20Clean/VicPol%20Search%20Data%20Clean.xlsx)


> **Note:**  
> Data for 2020 and 2021 is not included due to the extraordinary and atypical policing context of the COVID-19 pandemic. These years are excluded to avoid drawing misleading inferences about standard police practices.

## Search Types in Scope

The dataset includes records of five types of **warrantless searches**, each of which requires that a police officer has *reasonable grounds* to suspect the person is in possession of a prohibited item relevant to that search power:

- **Drugs**: Section 82, *Drugs, Poisons and Controlled Substances Act 1981*  
- **Volatile substances for inhalation** (under 18s): Sections 60E–60F, *Drugs, Poisons and Controlled Substances Act 1981*  
- **Weapons or dangerous articles**: Section 10, *Control of Weapons Act 1990*  
- **Firearms**: Section 149, *Firearms Act 1996*  
- **Graffiti implements** (at designated places): Section 13, *Graffiti Prevention Act*  

Nearly **90%** of searches were conducted under drug search powers. Some individuals were searched under multiple powers (e.g. both drug and weapon suspicions), and this is reflected in the dataset.

## Ethnic Appearance Categories

Victoria Police records the **perceived ethnic appearance** of individuals searched, based on officer observation. These categories reflect *perceived ethnicity*, not self-identification or ancestry data.

### Missing Data

A significant portion of records — **22,117 searches or 42%** — lack ethnic appearance data, despite this field being mandatory for stop and search events since 2019. In cases where ethnicity could not be determined, officers used the category **‘Other’**.

### Middle Eastern Category Limitation

In 2020, Victoria Police merged the **‘Middle Eastern’** and **‘Mediterranean’** appearance categories into one. This limits the ability to meaningfully assess searches involving people of Middle Eastern appearance. Typically, people of Mediterranean descent might be categorised as *non-racialised* or *White*, while people of Middle Eastern appearance are more likely to be viewed as *racialised*. The aggregation obscures this distinction and hampers analysis of potential racial profiling.

# Summary Statistics

The table below summarises key variables in the dataset, including demographic fields, search types, and outcomes.


**Table 1 Characteristics of searches**
```{r summary 2, echo = FALSE, warning = FALSE}

of.interest <- c("Year", "Racial.appearance.transformed","Gender", "Area.type", "Unit.type", "Contact.Type" ,"Rank.of.Member", "Found",
                 "Search.type - Drugs","Search.type - Weapons","Search.type - Firearms",          
                  "Search.type - Graffiti","Search.type - Volatile.sub.U18","Search.type - Volatile.sub.adult")

`VicPol station data` <- dat %>%
  select(all_of(of.interest))%>%
  mutate(across(everything(), ~ fct_infreq(.)))
  
print(dfSummary(`VicPol station data`, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

Table 2 below shows the breakdown of searches by search type. This summary is presented separately from the main dataset because some individual searches involved multiple search powers — for example, when an officer believed they had reasonable grounds to suspect a person was carrying both drugs and weapons.


**Table 2 Legislative basis for searches**
```{r summary search type, echo = FALSE, warning = FALSE}

search.types <- c("Search.type - Drugs", "Search.type - Weapons","Search.type - Firearms",          
                  "Search.type - Graffiti" , "Search.type - Volatile.sub.U18","Search.type - Volatile.sub.adult")

Search.dat <- dat %>%
  pivot_longer(cols =search.types,
               names_to = "Search.type", values_to = "Search.outcome")%>%
  filter(!is.na(Search.outcome))%>%
  mutate(Search.type = str_remove(Search.type, "Search.type - "))%>%
  mutate(Search.type = ifelse(str_detect(Search.type, "Vol"), "Volatile substance", Search.type))

`Search type` <- Search.dat %>%
  select(Search.type)%>%
  mutate(Search.type = fct_infreq(Search.type))

print(dfSummary(`Search type`, 
          max.distinct.values = 10, 
           varnumbers   = FALSE, 
                valid.col    = FALSE, 
                graph.magnif = 0.76,
          headings = FALSE),
      method = 'render')
```

# Hit Rate Analysis

The **hit rate** refers to the percentage of searches that resulted in the discovery of contraband. Across all warrantless searches in the dataset, the average hit rate was **17.2%**.

A lower hit rate indicates that police are less likely to find contraband when searching individuals from that group. This suggests that the legal threshold of *reasonable suspicion* may be applied more loosely to some groups — particularly those perceived as racialised. In practice, this can reflect *conscious or unconscious racial bias* and is experienced by affected communities as racial targeting.

Compared with people perceived to be **White** (hit rate: **18.0%**), significantly lower hit rates were observed for:

- **African appearance**: 12.9%  
- **Middle Eastern/Mediterranean**: 13.8%  
- **Asian appearance**: 13.7%  
- **South Asian appearance**: 13.9%  
- **Aboriginal and Torres Strait Islander**: 15.6%  

The only group with a higher hit rate than White-appearing individuals was:

- **Pacific Islander appearance**: 18.8%

These differences were found to be statistically significant using **Pearson’s chi-squared test**.

## Summary of Hit Rates by Ethnic Appearance

Table 3 below presents these hit rates in summary form. It shows the percentage of searches that resulted in the discovery of contraband for each ethnic appearance category as perceived and recorded by Victoria Police officers.

> **Caution:**  
> A substantial portion of records — over **40%** — are missing ethnic appearance data. While these records are excluded from the breakdown, their absence limits the completeness of this comparison and may affect the generalisability of the findings. The missing data may also reflect systemic issues in data collection and reporting practices.

> **Interpretation note:**  
> These findings should be interpreted with caution. This summary aggregates **all search types**, across **all Victoria Police units** and **all locations**. In reality, thresholds for applying *reasonable suspicion* may differ between units or operational contexts. Victoria is also **ethnically diverse and geographically varied**, with different crime patterns and population distributions across regions. Differences may also exist between **foot searches and vehicle searches**, which are not distinguished here.

Because of these limitations, a more **focused and robust analysis** was also undertaken, examining only **drug searches conducted by police on foot in metropolitan Melbourne**. This provides a clearer view of potential disparities under more controlled conditions.

A further area of exploration is the use of **statistical modelling** to control for the factors listed above — including search type, policing unit, location, mode of search (e.g. on foot vs in vehicle), and local demographic or crime context. Such models can help isolate the effect of perceived racial appearance on search outcomes while accounting for confounding variables, and would offer a more rigorous basis for identifying patterns of potential racial bias in police decision-making.


**Table 3 Hit rates by variables of interest**
```{r Finds racial, echo=FALSE, warning=FALSE}
dat <- dat %>%
  mutate(Found = factor(Found, levels = c("Yes", "No")))

dat %>% 
  crosstable(c(Racial.appearance.transformed), by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()
```

Figure 1 below shows how the hit rate for different racial appearances changed over the four year period, with a reference line showing the average hit rate for all searches over four years in red.  The following observations are made:  

* people who appear White experienced higher hit-rates compared with the overall population in most years, except for 2022. This may be due to the effect of pandemic-related measures.  
* people of African, Middle Eastern and Asian appearance experienced persistently lower hit rates than the average.  
* hit-rates for people appearing to be Aboriginal and Torres Strait Islander have increased over the four year period from 11% in 2018 to 18.6% in 2023 - suggesting fewer racially biased searches.


**Figure 1 Changes in hit rates by racial appearance over time and compared to the average**
```{r Finds race year, echo=FALSE, warning=FALSE, message = FALSE}
trend <- dat %>%
  filter(!Racial.appearance.transformed %in% c("Missing", "Mediterranean AND Middle Eastern - DO NOT USE", "Other"))%>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed))%>%
  group_by(Year, `Racial.appearance`)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), `Total searches`= n())%>%
  mutate(`Hit rate` = round(Finds/`Total searches`*100, digits = 1)) %>%
  ungroup()%>%
  ggplot(aes(x = Year, y = `Hit rate`)) +geom_col(fill = "darkblue")+
  facet_wrap(~`Racial.appearance`)+ 
  geom_hline(yintercept = 17.2,linetype="dotted", color = "red") +theme(legend.position = "none")

ggplotly(trend)
```


Figure 1 below shows how the hit rate for different racial appearances changed over the four year period, with a reference line showing the average hit rate for all searches over four years in red.  The following observations are made:  

* people who appear White experienced higher hit-rates compared with the overall population in most years, except for 2022. This may be due to the effect of pandemic-related measures.  
* people of African, Middle Eastern/Mediterranean and Asian appearance experienced persistently lower hit rates than the average.  
* hit-rates for people appearing to be Aboriginal and Torres Strait Islander have reduced over the four year period from 11% in 2018 to 18.6% in 2023.

## Hit rates by other search characteristics  

Other characteristics which may have influenced hit rates are summarised in Table 4 below.  Different search types had widely different hit rates, with drug-related searches having the lowest hit-rate (15.2%) and graffiti-related searches at designated places, such as train stations, having the highest hit-rate (50.1%), followed by weapons-related searches (32.8%).

**Table 4 Hit rates by legislative power**

```{r Finds search, echo=FALSE, warning=FALSE}

Search.dat %>%
  mutate(Found = factor(Found, levels = c("Yes", "No")))%>%
  crosstable(Search.type, by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()

```

**Table 4 Hit rates by other search characteristics**
```{r Finds all other, echo=FALSE, warning=FALSE}

dat %>%
  crosstable(c(Year, Unit.type, Gender, Area.type), by=Found, total=TRUE, 
           test=TRUE, showNA='ifany', percent_digits=1) %>% 
  as_flextable()
```

# Racialised hit within police units and local areas  

Racialised hit rates across different unit types were compared.  Where there were less than five searches for a given group in a Local Government Areas hit rates are not shown, as statistical results may be misleading.

## Unit types

**Table 6 Ethnic appearance hit rates by Unit type**
```{r unit.type, echo=FALSE, warning=FALSE, message = FALSE}
unit.racial <- dat %>%
  group_by(Unit.type)%>%
  mutate(`Overall %` = round(sum(Found == "Yes", na.rm = T)/n()*100, digits = 1), Total.searches = n())%>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed))%>%
  group_by(Unit.type, Total.searches, `Overall %`, Racial.appearance)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), Searches = n())%>%
  ungroup()%>%
  mutate(Searches = ifelse(Searches <= 5, NA, Searches))%>%
  mutate(`Hit rate` = round(Finds/Searches*100, digits = 1))%>%
  select(-Finds, -Searches)%>%
  pivot_wider(names_from = Racial.appearance,
    values_from = c(`Hit rate`)
  )%>%
  arrange(desc(Total.searches))%>%
  select(Unit.type,
         `Total searches` = Total.searches,
         `Overall %`, `White %` = White, `Aboriginal %` = Aboriginal, `Middle Eastern %` = `Middle Eastern`, 
         `African %`= African, `Asian %` = Asian,`South Asian %` = `South Asian`, `Pacific Islander %` = `Pacific Islander`)

unit.racial %>%
  datatable(rownames = F, extensions = 'Buttons',
            options = list(
              dom = 'Blfrtip',
              buttons = c('copy', 'excel','print'),
  pageLength = 10, autoWidth = TRUE
))
```

## Local Government Areas

Searches that were undertaken by a uniform officer based at a local Police station were able to be mapped to their Local Government Area.  

The hit rates for persons of different ethnic appearances for Local Government Areas appears below.   

**Table 7 Ethnic appearance hit rates by LGA**
```{r LGA, echo=FALSE, warning=FALSE, message = FALSE}

lga.psa <- dat %>%
  select(Police.Service.Area, Local.Government.Area)%>%
  unique()

lga.racial <- dat %>%
  filter(!is.na(Police.Service.Area), Unit.type == "Uniform")%>%
  group_by(Police.Service.Area)%>%
  mutate(`Overall %` = round(sum(Found == "Yes", na.rm = T)/n()*100, digits = 1), Total.searches = n())%>%
  mutate(Racial.appearance = fct_infreq(Racial.appearance.transformed))%>%
  group_by(Area.type, Police.Service.Area, Total.searches, `Overall %`, Racial.appearance)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), Searches = n())%>%
  ungroup()%>%
  mutate(Searches = ifelse(Searches <= 5, NA, Searches))%>%
  mutate(`Hit rate` = round(Finds/Searches*100, digits = 1))%>%
  select(-Finds, -Searches)%>%
  pivot_wider(names_from = Racial.appearance,
    values_from = c(`Hit rate`)
  )%>%
  left_join(lga.psa, by = "Police.Service.Area")%>%
  select(`Area` = Area.type, `Local Government Area` = Local.Government.Area, `Police Service Area` = Police.Service.Area,
         `Total searches` = Total.searches,
         `Overall %`, `White %` = White, `Aboriginal %` = Aboriginal, `Middle Eastern %` = `Middle Eastern`, 
         `African %`= African, `Asian %` = Asian,`South Asian %` = `South Asian`, `Pacific Islander %` = `Pacific Islander`)

lga.racial %>%
  datatable(rownames = F, extensions = 'Buttons',
            options = list(
              dom = 'Blfrtip',
              buttons = c('copy', 'excel','print'),
  pageLength = 10, autoWidth = TRUE
))



```


# Appendix - ethnic appearance mapping  

Ethnic appearance categories for different years were transformed to enable comparison across the four years. A table of transformations undertaken, along with the number of searches in each category, appears below,

# 2022 and 2023 mapping  

```{r Map 1, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping1 <-dat %>%
  filter(Year %in% c(2022, 2023))%>%
  count(Racial.Appearance.original, Racial.appearance.transformed)

ethnic.mapping1 %>%
  kable()
```

# 2018 and 2019 mapping  

```{r Map 2, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping2 <- dat %>%
  filter(Year %in% c(2018, 2019))%>%
  count(Racial.Appearance.original, Racial.appearance.transformed) 


ethnic.mapping2 %>%
  kable()
```