---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

# Introduction

This report analyses whether individuals perceived by police to belong to different racial appearance groups were subjected to drug-related stop and search practices under different thresholds of suspicion. It draws on over **13,000 records of warrantless drug searches** conducted by **uniformed Victoria Police officers on foot patrol in public places across metropolitan Melbourne** during the years **2018, 2019, 2022, and 2023**.

The analysis uses the **hit rate method**—the proportion of searches that result in the discovery of drugs—to assess whether stop and search powers are being applied equitably across racial groups. A consistently lower hit rate for a particular group suggests that police may be applying the legal threshold of *reasonable suspicion* less stringently for that group, raising concerns about racial profiling.

This report compares the hit rates for **six racial appearance groups**—**African, Asian, Aboriginal, South Asian, Middle Eastern**, and **Pacific Islander**—against those perceived to be **White**. These comparisons aim to detect **disparities in the likelihood of contraband being found**, which may indicate racially biased decision-making in discretionary policing.

Because **policing practices and public movement patterns can vary by year**, especially in light of COVID-19 disruptions in 2020 and 2021, the analysis controls for **year of search** to ensure that observed differences are not confounded by broader temporal shifts. By statistically adjusting for year, this analysis isolates the effect of racial appearance on search outcomes, providing a clearer picture of potential bias in police decision-making.

# About the Data

The data comes from *The Racial Profiling Data Monitoring Project*, which collated records of discretionary drug searches conducted by Victoria Police across **metropolitan Melbourne**. Searches from **2020 and 2021** were excluded due to the abnormal conditions of the COVID-19 pandemic, which significantly altered police activity and public mobility.

Under **section 82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)**, officers may stop and search people in public places if they have *reasonable grounds* to suspect drug possession. Drug searches are the most common form of warrantless search in Victoria Police operations.

# Racial Appearance Categories

Victoria Police policy requires officers to record the **perceived racial appearance** of individuals subject to stop and search. These categories reflect the officer’s visual perception and not the person’s actual ethnicity or self-identified background.

Of the **13,896 available records**, **2,860 (21%)** were excluded due to missing racial appearance data, despite the field being mandatory since 2019. In **2022 and 2023**, Victoria Police also merged *Middle Eastern* and *Mediterranean* categories into a single unusable grouping. These, along with all records from low-frequency race categories (under 100 observations) or indeterminate classifications, were excluded from this analysis.

The final sample includes **9,584 records**. This means over **31% of records were excluded**, highlighting a major concern: Victoria Police must urgently improve the **quality and consistency of racial appearance data collection**, particularly when exercising discretionary powers such as stop and search. Without reliable data, transparency and accountability are severely undermined.

# Methodology

This analysis uses **hit rate benchmarking**, a method widely adopted in racial profiling research (see Gelman, Fagan & Kiss, 2007; Open Justice Project, 2023), to evaluate whether **reasonable suspicion** is applied equitably.

If individuals of African appearance are consistently found with drugs at lower rates than White individuals, this may indicate a lower evidentiary threshold for conducting searches—suggesting racially biased policing.

To assess where such disparities are most pronounced, the report compares hit rates between these two groups at five geographic levels:

- **Whole of Melbourne (metro aggregate)**  
- **Police Region**  
- **Police Division**  
- **Police Service Area (PSA)**  
- **Individual Police Stations**  

This multi-level approach allows us to identify whether disparities are widespread or driven by specific police units or local cultures, providing a more granular picture of racial bias in discretionary policing practices.
```{r load-packages-and-data, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(strayr)
library(sf)
library(broom)
library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(viridis)

dat <- readRDS("../VicPol Search Data Clean/VicPol Search Data Clean.RDS")

division_lga_lookup <- dat %>%
  filter(
    Area.type == "Metro",
    !is.na(Division),
    !is.na(Local.Government.Area)
  ) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )
```

## Adjusted Comparison of Hit Rates by Racial Appearance

We use a binomial logistic regression to model the probability of a "hit" (drugs found), controlling for **year of search**. The reference group for race is **White**, and all coefficients are interpreted relative to this group.

```{r logistic-regression-hitrate, message=FALSE, warning=FALSE, echo = FALSE}
# Filter and prepare data
cleaned_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Filter dataset
model_data <- cleaned_data %>%
  filter(
    Race %in% c("White", "African", "Asian", "Aboriginal", "South Asian", "Middle Eastern", "Pacific Islander"),
    !is.na(Year)
  ) %>%
  mutate(
    Race = factor(Race, levels = c("White", "African", "Asian", "Aboriginal", "South Asian", "Middle Eastern", "Pacific Islander")),
    Year = as.factor(Year)
  )

# Logistic regression model
model <- glm(Found.Flag ~ Race + Year, data = model_data, family = binomial)

# Tidy model with odds ratios and CIs
results <- tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(grepl("^Race", term)) %>%
  mutate(
    Race = gsub("Race", "", term),
    p.value = round(p.value, 4),
    OR = round(estimate, 2),
    CI = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")"),
    Sig = ifelse(p.value < 0.05, "*", "")
  ) %>%
  select(Race, OR, CI, p.value, Sig)

# Add reference group
white_row <- data.frame(
  Race = "White (ref)",
  OR = 1.00,
  CI = "(ref)",
  p.value = NA,
  Sig = ""
)

final_table <- bind_rows(white_row, results)

# Display as nice markdown table
kable(final_table,
      caption = "Adjusted Odds Ratios for Drug Search Hits by Racial Appearance (Reference: White)",
      align = "lccrc") %>%
  kable_styling(full_width = FALSE, position = "left")
```

### Interpretation of Results

The table above presents **adjusted odds ratios** comparing the likelihood that a drug search resulted in a **"hit"** (drugs found) for each racial appearance group, relative to individuals perceived by police as **White**. The model controls for **year of search**, helping to isolate the effect of racial appearance from broader shifts in policing practices or public behaviour across time.

- An **odds ratio (OR) below 1** means the group was **less likely** than White individuals to have drugs found during a search.
- An OR of **1** indicates **no difference**.
- An asterisk (*) marks results that are **statistically significant** (*p* < 0.05).

**Key findings:**

- **South Asian** individuals were **36% less likely** to have a drug search result in a hit compared to White individuals (*OR = 0.64, p = 0.0275*). This statistically significant result suggests a **lower threshold of suspicion** may be applied when stopping and searching South Asian people.
- **African** individuals were **22% less likely** to yield a drug find (*OR = 0.78, p = 0.0381*), also a statistically significant disparity.
- Other groups—**Asian**, **Aboriginal**, **Middle Eastern**, and **Pacific Islander**—also showed lower odds of a drug find compared to White individuals, but these differences were **not statistically significant** in this model.

These findings indicate that **South Asian and African** individuals are more likely to be searched **without sufficient grounds**, reinforcing long-standing concerns about **racial profiling** in Victoria Police’s use of discretionary stop and search powers. The disparities persisted even after controlling for **year of search**, suggesting they are not explained by shifting policing patterns over time.


## Search Counts by Race and Police Region

The following table presents a **descriptive summary** of drug searches by racial appearance and Victoria Police Region. This analysis is **exploratory** and aims to illustrate the geographic distribution of search activity across major operational areas.

No statistical testing has been conducted in this section. Therefore, the figures should be interpreted with **caution**. Apparent differences between regions or racial groups may be due to a range of factors, and no conclusions should be drawn about disparities or bias without further statistical analysis to control for confounding variables.

# Introduction

This report examines evidence of racial profiling in drug-related stop and search practices conducted by Victoria Police in Melbourne. It draws on over **13,000 records of warrantless drug searches** conducted by **uniformed officers on foot patrol in public places** during the years **2018, 2019, 2022, and 2023**. By comparing *hit rates*—the proportion of searches that result in a drug find—for people perceived by police as of **African appearance** versus those perceived as **White**, the analysis seeks to identify disparities that may indicate racially biased policing practices.

# About the Data

The data used in this report comes from *The Racial Profiling Data Monitoring Project*, which compiled records of discretionary drug searches carried out by Victoria Police across **metropolitan Melbourne**. Searches from **2020 and 2021** were excluded due to the exceptional conditions of the COVID-19 pandemic, which significantly altered policing patterns and public mobility.

Under **section 82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)**, police may stop and search individuals in public places if they have *reasonable grounds* to suspect drug possession. Drug searches are the most frequently used form of warrantless search conducted by Victoria Police.

# Racial Appearance Categories

Victoria Police officers are required to record the **perceived racial appearance** of individuals during stop and search encounters. These categories are based on the officer’s visual assessment, not the individual's self-identification or background.

Of the **13,896 records** initially available, **2,860 (approximately 21%)** were excluded from the analysis due to missing racial appearance data—despite this being a mandatory field since 2019. In **2022 and 2023**, Victoria Police also amalgamated the *Middle Eastern* and *Mediterranean* categories, rendering them analytically unusable; these records were also filtered out.

To ensure statistical validity, racial appearance groups with **fewer than 100 observations**, or where police were **unable to determine a racial category**, were relabelled as **"Other"** and excluded from the primary analysis.

As a result, only **4,312 of the original 13,896 records** were retained, meaning **over 31% of records were excluded**. This significant data loss reflects a broader issue: **Victoria Police must urgently improve the quality, consistency, and completeness of its racial data collection practices**—especially in the exercise of discretionary powers. Without reliable and usable data, efforts to identify and address racial disparities in policing remain compromised.

# Methodology

This analysis focuses on detecting **racial profiling against people recorded as being of African appearance**, using the **hit rate method**—a widely recognised tool in policing research for assessing the fairness and effectiveness of search practices. A *hit rate* is the percentage of searches that result in the discovery of contraband.

Disproportionately low hit rates for a particular group suggest that the legal standard of *reasonable suspicion* is being applied more leniently, raising concerns of discrimination and misuse of police discretion.

This report specifically compares **hit rates for people perceived as African** with those perceived as **White**, across five levels of geography:

- Whole of Melbourne (metropolitan aggregate)  
- Police Region  
- Police Division  
- Police Service Area (PSA)  
- Individual Police Stations  

This approach aligns with established frameworks in racial profiling research, including **Gelman, Fagan & Kiss (2007)** and the **Open Justice Project (2023)**, which emphasise geographic disaggregation and benchmarking as essential tools for accountability.


### Data summary - Melbourne 

```{r hit-rate-all-races-final, echo=FALSE}

# Filter and prepare data
cleaned_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Summarise by race
race_hit_rates <- cleaned_data %>%
  group_by(Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    `Hit Rate` = round(mean(Found.Flag), 3),
    .groups = "drop"
  )

# Create total row
total_row <- cleaned_data %>%
  summarise(
    Race = "Total",
    Searches = n(),
    Hits = sum(Found.Flag),
    `Hit Rate` = round(mean(Found.Flag), 3)
  )

# Bind and order: descending by Searches, Total last
race_hit_rates <- bind_rows(race_hit_rates, total_row) %>%
  mutate(Race = factor(Race, levels = c(
    race_hit_rates %>% arrange(desc(Searches)) %>% pull(Race),
    "Total"
  ))) %>%
  arrange(Race)

# Display table
kable(
  race_hit_rates,
  caption = "Drug Searches, Hits, and Hit Rates by Racial Appearance (Metro Victoria, Uniform Police at Stations)",
  align = "lrrr"
)



```

### Hit rate analysis - Melbourne 








```{r prop-test-all,  echo = FALSE}


# Filter and prepare data
vic_test_data <- dat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Station.uniform)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0))

# Summarise by group
melb_summary <- vic_test_data %>%
  group_by(Racial.appearance.transformed) %>%
  summarise(
    Searches = n(),
    Hits = sum(found_flag),
    Misses = Searches - Hits,
    HitRate = mean(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Searches, Hits, Misses, HitRate),
    names_sep = "."
  )

# Compute prop.test, odds, odds ratio, interpretation
melb_results <- melb_summary %>%
  rowwise() %>%
  mutate(
    Odds_African = Hits.African / Misses.African,
    Odds_White = Hits.White / Misses.White,
    Odds_Ratio = round(Odds_African / Odds_White, 3),
    prop_result = list(prop.test(
      x = c(Hits.African, Hits.White),
      n = c(Searches.African, Searches.White),
      correct = FALSE
    )),
    tidy_result = list(tidy(prop_result))
  ) %>%
  unnest(tidy_result) %>%
  mutate(
    `Difference in Hit Rate` = round(estimate2 - estimate1, 3),
    `95% CI Lower` = round(conf.low, 3),
    `95% CI Upper` = round(conf.high, 3),
    `P-value` = round(p.value, 3),
    `Odds Ratio (African vs White)` = Odds_Ratio,
    Interpretation = case_when(
      `P-value` >= 0.05 ~ "No statistically significant difference",
      Odds_Ratio < 1 ~ "African appearing people searched on weaker suspicion",
      Odds_Ratio > 1 ~ "African appearing people searched on stronger suspicion",
      TRUE ~ "—"
    )
  ) %>%
  select(
    `Searches (African)` = Searches.African,
    `Searches (White)` = Searches.White,
    `Hits (African)` = Hits.African,
    `Hits (White)` = Hits.White,
    `Hit Rate (African)` = HitRate.African,
    `Hit Rate (White)` = HitRate.White,
    `Difference in Hit Rate`,
    `95% CI Lower`,
    `95% CI Upper`,
    `P-value`,
    `Odds Ratio (African vs White)`,
    Interpretation
  )

# Display result in table
kable(
  melb_results,
  caption = "Drug Search Hit Rate Comparison – Whole of Melbourne (African vs White)",
  digits = 3,
  align = "rrrrrrrrrrrl"
)
```


### Hit rate analysis - Region  

```{r prop-test-region, echo = FALSE}

# Summarise hits and searches by Region and Race
region_summary <- vic_test_data %>%
  group_by(Region, Racial.appearance.transformed) %>%
  summarise(
    Searches = n(),
    Hits = sum(found_flag),
    Misses = Searches - Hits,
    HitRate = round(mean(found_flag), 3),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Searches, Hits, Misses, HitRate),
    names_sep = "."
  )



region_results <- region_summary %>%
  rowwise() %>%
  mutate(
    can_test = all(!is.na(c(Hits.African, Hits.White, Searches.African, Searches.White))) &&
               Misses.African > 0 && Misses.White > 0,

    Odds_African = if (can_test) Hits.African / Misses.African else NA_real_,
    Odds_White   = if (can_test) Hits.White / Misses.White else NA_real_,
    Odds_Ratio   = if (can_test) round(Odds_African / Odds_White, 3) else NA_real_,

    # Run test and safely wrap result as list
    tidy_result = list(
      if (can_test) {
        result <- prop.test(
          x = c(Hits.African, Hits.White),
          n = c(Searches.African, Searches.White),
          correct = FALSE
        )
        tidy(result)
      } else {
        tibble(
          estimate1 = NA_real_,
          estimate2 = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          p.value = NA_real_
        )
      }
    )
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    `Difference in Hit Rate` = round(estimate2 - estimate1, 3),
    `95% CI Lower` = round(conf.low, 3),
    `95% CI Upper` = round(conf.high, 3),
    `P-value` = round(p.value, 3),
    Interpretation = case_when(
      is.na(Odds_Ratio) ~ "Insufficient data for comparison",
      `P-value` >= 0.05 ~ "No statistically significant difference",
      Odds_Ratio < 1 ~ "African appearing people searched on weaker suspicion",
      Odds_Ratio > 1 ~ "African appearing people searched on stronger suspicion",
      TRUE ~ "—"
    )
  ) %>%
  ungroup() %>%
    arrange(`P-value`) %>%
  select(
    Region,
    `Searches (African)` = Searches.African,
    `Searches (White)` = Searches.White,
    `Hits (African)` = Hits.African,
    `Hits (White)` = Hits.White,
    `Hit Rate (African)` = HitRate.African,
    `Hit Rate (White)` = HitRate.White,
    `Difference in Hit Rate`,
    `95% CI Lower`,
    `95% CI Upper`,
    `P-value`,
    `Odds Ratio (African vs White)` = Odds_Ratio,
    Interpretation
  )


# Display interactive scrollable DT table
kable(
  region_results,
  caption = "Drug Search Hit Rate Comparison by Region (African vs White)",
  digits = 3,
  align = "lrrrrrrrrrrl"
)
```



### Hit rate analysis - Division  


```{r prop-test-psa, echo = FALSE, warning = FALSE, message = FALSE}

# Step 1: Summarise data by Division and Race
division_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Division)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Division, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = "."
  )

# Step 2: Run prop.test for African vs White at Division level
division_results <- division_summary %>%
  rowwise() %>%
  mutate(
    African_HitRate = ifelse(Total_Searches.African > 0,
                             Total_Hits.African / Total_Searches.African, NA_real_),
    White_HitRate = ifelse(Total_Searches.White > 0,
                           Total_Hits.White / Total_Searches.White, NA_real_),

    African_Miss = Total_Searches.African - Total_Hits.African,
    White_Miss = Total_Searches.White - Total_Hits.White,

    Odds_African = ifelse(African_Miss > 0, Total_Hits.African / African_Miss, NA_real_),
    Odds_White = ifelse(White_Miss > 0, Total_Hits.White / White_Miss, NA_real_),
    Odds_Ratio = ifelse(!is.na(Odds_African) & !is.na(Odds_White), Odds_African / Odds_White, NA_real_),

    prop_test = if (
      all(!is.na(c(Total_Hits.African, Total_Hits.White,
                   Total_Searches.African, Total_Searches.White)))
    ) {
      list(tryCatch(
        prop.test(
          x = c(Total_Hits.African, Total_Hits.White),
          n = c(Total_Searches.African, Total_Searches.White),
          correct = TRUE
        ),
        error = function(e) NULL
      ))
    } else {
      list(NULL)
    },

    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & African_HitRate < White_HitRate ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & African_HitRate > White_HitRate ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%
  select(
    Division,
    African_Searches = Total_Searches.African,
    White_Searches = Total_Searches.White,
    African_Hits = Total_Hits.African,
    White_Hits = Total_Hits.White,
    African_HitRate,
    White_HitRate,
    Odds_Ratio,
    ConfLow = conf.low,
    ConfHigh = conf.high,
    P_Value = p.value,
    Interpretation
  ) %>%
  arrange(P_Value)

# Step 3: Format for display
division_results_clean <- division_results %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  ) %>%
  rename(
    `Searches (African)` = African_Searches,
    `Searches (White)` = White_Searches,
    `Hits (African)` = African_Hits,
    `Hits (White)` = White_Hits,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `Odds Ratio (African/White)` = Odds_Ratio,
    `95% CI Lower` = ConfLow,
    `95% CI Upper` = ConfHigh,
    `P-value` = P_Value
  )

# Step 4: Create Division → LGA lookup and join
division_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Division), !is.na(Local.Government.Area)) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

division_results_clean <- division_results_clean %>%
  left_join(division_lga_lookup, by = "Division") %>%
  relocate(LGAs, .after = Division)

# Step 5: Export to Excel
write_xlsx(division_results_clean, "division_drug_search_proptest_results.xlsx")

# Step 6: Display in DT
datatable(
  division_results_clean,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 5,  # Set default rows per page to 5
    lengthMenu = c(5, 10, 20, 50),  # Include 5 in the page size options
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '300px', targets = 1)
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Search Hit Rate Comparison by Division (African vs White)'
  )
)



```
### Hit rate analysis - PSA

```{r hit rate psa, echo = FALSE, message = FALSE, warning = FALSE}


# Step 1: Summarise data by PSA and Race
psa_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Police.Service.Area)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Police.Service.Area, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = "."
  )

# Step 2: Run prop.test for African vs White per PSA
psa_results <- psa_summary %>%
  rowwise() %>%
  mutate(
    prop_test = if (
      all(!is.na(c(Total_Hits.African, Total_Hits.White,
                   Total_Searches.African, Total_Searches.White)))
    ) {
      list(prop.test(
        x = c(Total_Hits.African, Total_Hits.White),
        n = c(Total_Searches.African, Total_Searches.White),
        correct = FALSE
      ))
    } else {
      list(NULL)
    },
    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & estimate1 < estimate2 ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & estimate1 > estimate2 ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%
  select(
    Police.Service.Area,
    African_Searches = Total_Searches.African,
    White_Searches = Total_Searches.White,
    African_Hits = Total_Hits.African,
    White_Hits = Total_Hits.White,
    African_HitRate = estimate1,
    White_HitRate = estimate2,
    ConfLow = conf.low,
    ConfHigh = conf.high,
    P_Value = p.value,
    Interpretation
  ) %>%
  arrange(P_Value)

# Step 3: Format for display
psa_results_clean <- psa_results %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  ) %>%
  rename(
    `Searches (African)` = African_Searches,
    `Searches (White)` = White_Searches,
    `Hits (African)` = African_Hits,
    `Hits (White)` = White_Hits,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `95% CI Lower` = ConfLow,
    `95% CI Upper` = ConfHigh,
    `P-value` = P_Value
  )

# Step 4: Export to Excel
write_xlsx(psa_results_clean, "psa_drug_search_proptest_results.xlsx")

# Step 5: Display interactive table
datatable(
  psa_results_clean,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 5,  # Set default rows per page
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '250px', targets = 0)
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Search Hit Rate Comparison by Police Service Area (African vs White)'
  )
)


```

### Hit rate analysis - Station

```{r prop-test-station, warning=FALSE, message=FALSE, echo = FALSE}
# Step 1: Create station_summary
station_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("African", "White")
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Station.uniform, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = ".",
    values_fill = 0
  )

# Step 2: Run prop.test and compute odds ratio
station_results <- station_summary %>%
  rowwise() %>%
  mutate(
    African_HitRate = ifelse(Total_Searches.African > 0,
                             Total_Hits.African / Total_Searches.African, NA_real_),
    White_HitRate = ifelse(Total_Searches.White > 0,
                           Total_Hits.White / Total_Searches.White, NA_real_),

    African_Miss = Total_Searches.African - Total_Hits.African,
    White_Miss = Total_Searches.White - Total_Hits.White,

    Odds_African = ifelse(African_Miss > 0, Total_Hits.African / African_Miss, NA_real_),
    Odds_White = ifelse(White_Miss > 0, Total_Hits.White / White_Miss, NA_real_),
    Odds_Ratio = ifelse(!is.na(Odds_African) & !is.na(Odds_White), Odds_African / Odds_White, NA_real_),

    can_test = Total_Searches.African > 0 & Total_Searches.White > 0,
    prop_test = if (can_test) {
      list(tryCatch(
        prop.test(
          x = c(Total_Hits.African, Total_Hits.White),
          n = c(Total_Searches.African, Total_Searches.White),
          correct = TRUE
        ),
        error = function(e) NULL
      ))
    } else {
      list(NULL)
    },

    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%

  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & African_HitRate < White_HitRate ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & African_HitRate > White_HitRate ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%  select(
    Station = Station.uniform,
    `Searches (African)` = Total_Searches.African,
    `Searches (White)` = Total_Searches.White,
    `Hits (African)` = Total_Hits.African,
    `Hits (White)` = Total_Hits.White,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `Odds Ratio (African/White)` = Odds_Ratio,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `P-value` = p.value,
    Interpretation
  ) %>%
  arrange(`P-value`)

DT::datatable(
  station_results,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 5,
    autoWidth = TRUE
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Search Hit Rate Comparison by Police Station (African vs White)'
  )
)

```


# Appendix – Ethnic Appearance Mapping

```{r mapping-categories-2022-2023, echo = FALSE}
dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```
