---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---


```{r setup, echo = FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)

dat <- readRDS("../Output.data/Clean.search.data.RDS")

```

# Data overview 

This report summarises over 13,000 warrantless drug searches conducted by uniformed Victoria Police officers while patrolling on foot in public places. The data covers the years 2018, 2019, 2022, and 2023, and is sourced from  [**The Racial Profiling Data Monitoring Project**](https://khaki-pepper-b3rf.squarespace.com/). Data from 2020 and 2021 is excluded due to the extraordinary circumstances of the COVID-19 pandemic.

Under section 82 of the [*Drugs Poisons and Controlled Substances Act 1981*](http://classic.austlii.edu.au/au/legis/vic/consol_act/dpacsa1981422/s82.html), a police officer may stop and search a person in a public place if they reasonably suspect the person is in possession of illegal drugs. Drug searches are the most common type of warrantless search carried out by police in Victoria.

## Racial appearance categories  

Victoria Police records the perceived racial appearance of individuals subjected to stop and search. These categories are based on the officer’s perception, not the individual's self-reported ancestry or identity. 

***Missing data***

Racial appearance was not recorded in 2,860 search records — approximately 21% of all records — despite being a mandatory field for stop and search data collection since 2019.

## Melbourne Drug Searches – Evidence of Racial Profiling by Station  

Find rates — the percentage of searches resulting in the discovery of illegal drugs — for people of different recorded racial appearances at stations across Melbourne are presented below.

Where fewer than 20 individuals from a particular racial appearance group were searched at a station, find rates are not reported, as small sample sizes limit meaningful analysis.


```{r Station, echo=FALSE, warning=FALSE, message = FALSE}

loc.racial <- dat %>%
  filter(!is.na(Station.uniform),Contact.Type == "P", 
         !is.na(`Search.type - Drugs`),
        Area.type == "Metro")%>%
  group_by(Division, Station.uniform)%>%
  mutate(`Finds all (%)` = round(sum(Found == "Yes", na.rm = T)/n()*100, digits = 0), 
         Total.searches = n())%>%
  mutate(Racial.appearance.abridged = fct_infreq(Racial.appearance.abridged))%>%
  group_by(Division, Station.uniform, Total.searches, `Finds all (%)`, Racial.appearance.abridged)%>%
  summarise(Finds = sum(Found == "Yes", na.rm = T), Searches = n())%>%
  ungroup()%>%
  mutate(`Hit rate` = round(Finds/Searches*100, digits = 1))%>%
  mutate(`Hit rate` = ifelse(Searches <= 20, NA, `Hit rate`))%>%
  select(-Finds)%>%
  pivot_wider(names_from = Racial.appearance.abridged,
    values_from = c(`Hit rate`, Searches)
  )%>%
 select(
   Division,
    Station = Station.uniform, 
   `Total drug searches` = Total.searches, 
    `Finds all (%)`,  
     `Finds: White (%)` = `Hit rate_White`, 
    `Finds: African (%)` = `Hit rate_African`, 
    `Finds: Asian (%)` = `Hit rate_Asian`, 
    `Finds: South Asian (%)` = `Hit rate_South Asian`, 
    `Finds: Aboriginal (%)` = `Hit rate_Aboriginal`, 
    `Finds: Middle Eastern (%)`= `Hit rate_Middle Eastern` , 
    `Finds: Pacific Islander (%)` = `Hit rate_Pacific Islander`, 
    `Finds: Other non-White (%)` = `Hit rate_Other racialised`,
    `Finds: Missing (%)` = `Hit rate_Missing`,
    
  #  `Searches: White` = `Searches_White`, 
  #  `Searches: African` = `Searches_African`
  #  `Searches: Asian` = `Searches_Asian`, 
  #  `Searches: South Asian` = `Searches_South Asian`,
  #  `Searches: Aboriginal` = Searches_Aboriginal, 
  #  `Searches: Middle Eastern` = `Searches_Middle Eastern`, 
  #  `Searches: Pacific Islander` = `Searches_Pacific Islander`,
  #  `Searches: Mediterranean/Mid` = `Searches_Mediterarranean/Mid`,
  #  `Searches: Other non-White` = `Searches_Other racialised`,
  #  `Searches: Missing` = Searches_Missing
  )


loc.racial %>%
#  filter(Division %in% c("NWMR Division 2", "NWMR Division 3", "NWMR Division 4"))%>%
#  filter(!is.na(`Finds: African (%)`))%>%
  datatable(rownames = F, extensions = 'Buttons',
            options = list(
              dom = 'Blfrtip',
              buttons = c('copy', 'excel','print'),
  pageLength = 20, autoWidth = TRUE
))


```


# Appendix - ethnic appearance mapping  

Ethnic appearance categories for different years were transformed to enable comparison across the four years. A table of transformations undertaken, along with the number of searches in each category, appears below,


```{r Map 1, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping1 <-dat %>%
  filter(Year %in% c(2022, 2023))%>%
  count(Racial.Appearance.original, Racial.appearance.abridged)

ethnic.mapping1 %>%
  kable()
```

```{r Map 2, echo=FALSE, warning=FALSE, message = FALSE}

ethnic.mapping2 <- dat %>%
  filter(Year %in% c(2018, 2019))%>%
  count(Racial.Appearance.original, Racial.appearance.abridged) 


ethnic.mapping2 %>%
  kable()
```