---
title: "Victoria Police Uniform Drug Searches: Adjusted Hit‑Rate Analysis"
author: "Monika Sarder"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
params:
  # If you already have an object called `datsum` in the environment, it will be used.
  data_path: "data/datsum.csv"   # or an .rds/.qs if you prefer
  place_control: "station"       # one of: "station", "lga"
  cluster_level: "station"       # one of: "station", "shift", "person" (if available)
  uniform_regex: "uniform"       # case-insensitive pattern that flags uniform police rows
  year_min: 2018
  year_max: 2025
---

```{r setup, message=FALSE, warning=FALSE}
# Packages ---------------------------------------------------------------
library(tidyverse)
library(janitor)
library(lubridate)
library(glue)
library(broom)
library(broom.helpers)
library(modelsummary)
library(fixest)          # fast GLMs with fixed effects
library(marginaleffects) # average marginal effects, supports fixest
library(metafor)         # meta-analysis of division effects
library(gt)              # pretty tables

options(dplyr.summarise.inform = FALSE, width = 100)
set.seed(42)
```

# 1. Scope & Design

To assess racial profiling credibly, we compare like with like by restricting the analysis to **drug searches conducted by uniform police**. These searches make up the vast majority of ordinary public‑place encounters and provide the most comparable context for decision‑making at the point of search.

We further limit the racial comparison to people perceived as **White** and **African**. Pooling "Middle Eastern" and "Mediterranean" into a single broad category introduces excess heterogeneity, so these groups are excluded from the racial comparison.

Within this scope, we estimate **adjusted hit‑rate differences at the division level**. For each division we fit a logistic regression of the hit indicator on an **African (vs White)** indicator, controlling for age (teenage vs adult), gender, time (year/month), and place (station *or* LGA). We use **cluster‑robust SEs** at the chosen cluster level (station by default) to account for peer effects. We then summarise across divisions with a random‑effects meta‑analysis. Sensitivity checks address missing racial appearance and alternative place/cluster controls; where repeated individuals can be identified, we can also cluster by person.

Results are presented as **odds ratios (ORs)** and **average marginal effects (AMEs)** to convey both statistical and substantive significance. Findings generalise to **uniform drug searches**.

> **Tip:** Configure `params` at the top if you want `place_control` to be by LGA, or to change the clustering level.

# 2. Data import & column mapping

```{r import, message=FALSE, warning=FALSE}
# Helper: read data if needed -------------------------------------------------
read_any <- function(path) {
  stopifnot(length(path) == 1)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("rds")) readRDS(path)
  else if (ext %in% c("qs")) {
    if (!requireNamespace("qs", quietly = TRUE)) stop("File has .qs extension but the 'qs' package is not installed.")
    qs::qread(path)
  } else if (ext %in% c("csv")) readr::read_csv(path, guess_max = 100000)
  else stop("Unsupported file extension: ", ext)
}

# Prefer an existing object called `datsum` if present; otherwise read from params
raw <- if (exists("datsum", inherits = TRUE)) datsum else read_any(params$data_path)
raw <- janitor::clean_names(raw)

nms <- names(raw)

# Helper to pick the first existing candidate column, else error
pick <- function(cands, label) {
  hit <- cands[cands %in% nms]
  if (length(hit) == 0) stop(glue("Missing a required column for {label}. Tried: {paste(cands, collapse = ', ')}"))
  hit[[1]]
}

# Flexible boolean parser for various encodings
as_bool <- function(x) {
  if (is.logical(x)) return(x)
  if (is.numeric(x)) return(!is.na(x) & x > 0)
  if (is.character(x)) return(str_detect(str_to_lower(x), "^(y(es)?|true|t|1|found|positive|pos|drug|drugs)"))
  rep(NA, length(x))
}

# Column mapping (edit if your names differ) ----------------------------------
cols <- list(
  date     = pick(c("search_date", "date", "search_date_time"), "date"),
  division = pick(c("division", "police_division"), "division"),
  station  = pick(c("station", "police_station", "location_station"), "station"),
  lga      = pick(c("lga", "local_government_area"), "lga"),
  unit     = pick(c("unit", "business_unit", "workgroup", "unit_type"), "unit"),
  gender   = pick(c("gender", "sex"), "gender"),
  age      = pick(c("age", "person_age"), "age"),
  # Racial appearance, free text or coded
  race_raw = pick(c("ethnic_appearance", "appearance", "perceived_ethnicity", "ethnicity"), "race"),
  # Indicators for search type & outcome
  search_type_drugs = pick(c("search_type_drugs", "search_type", "search_reason"), "search type / drugs"),
  found_col = pick(c("finds_drugs", "found", "search_result", "outcome"), "found / result")
)

# Build a standardised dataset -------------------------------------------------
dat0 <- raw %>%
  transmute(
    date = as.Date(!!sym(cols$date)),
    year = year(date),
    month = month(date),
    ym = sprintf("%04d-%02d", year, month),
    division = as.character(!!sym(cols$division)),
    station  = as.character(!!sym(cols$station)),
    lga      = as.character(!!sym(cols$lga)),
    unit     = as.character(!!sym(cols$unit)),
    gender   = forcats::fct_explicit_na(as.factor(!!sym(cols$gender)), na_level = "Missing"),
    age      = suppressWarnings(as.numeric(!!sym(cols$age))),
    race_raw = as.character(!!sym(cols$race_raw)),
    search_type_drugs = !!sym(cols$search_type_drugs),
    found_col = !!sym(cols$found_col)
  )

# Coerce drugs search type to boolean
if (is.character(dat0$search_type_drugs)) {
  dat0 <- dat0 %>% mutate(is_drug_search = str_detect(str_to_lower(search_type_drugs), "drug"))
} else {
  dat0 <- dat0 %>% mutate(is_drug_search = as_bool(search_type_drugs))
}

# Coerce 'found' to boolean "hit" (for drug searches)
dat0 <- dat0 %>% mutate(hit = as_bool(found_col))

# Uniform police flag
uniform_regex <- stringr::regex(params$uniform_regex, ignore_case = TRUE)
dat0 <- dat0 %>% mutate(is_uniform = str_detect(unit, uniform_regex))

# Race: collapse to White vs African; others set to NA for the main analysis
race_map <- function(x) {
  z <- case_when(
    str_detect(str_to_lower(x), "white")   ~ "White",
    str_detect(str_to_lower(x), "afric")   ~ "African",
    TRUE                                    ~ NA_character_
  )
  factor(z, levels = c("White", "African"))
}

dat0 <- dat0 %>% mutate(race2 = race_map(race_raw))

# Teenage (<=17) vs Adult (>=18)
dat0 <- dat0 %>% mutate(age_group = case_when(
  !is.na(age) & age <= 17 ~ "Teenage",
  !is.na(age) & age >= 18 ~ "Adult",
  TRUE ~ NA_character_
))

# Keep years within range
dat0 <- dat0 %>% filter(between(year, params$year_min, params$year_max))

# Final analysis subset: uniform + drug searches + White/African, non-missing outcome
analytical <- dat0 %>%
  filter(is_uniform, is_drug_search, !is.na(race2), !is.na(hit)) %>%
  mutate(
    african = if_else(race2 == "African", 1, 0),
    teenage = if_else(age_group == "Teenage", 1, 0),
    ym = factor(ym),
    station = forcats::fct_lump_n(factor(station), n = Inf),
    lga = forcats::fct_lump_n(factor(lga), n = Inf),
    division = forcats::fct_drop(factor(division))
  ) %>%
  tidyr::drop_na(division)

# Choose place FE and cluster level
place_fe <- switch(params$place_control,
                   station = ~ station + ym,
                   lga     = ~ lga + ym,
                   { warning("Unknown place_control, defaulting to station"); ~ station + ym })

cluster_var <- switch(params$cluster_level,
                      shift  = if ("shift_id" %in% nms) "shift_id" else "station",
                      person = if ("person_id" %in% nms) "person_id" else "station",
                      station = "station",
                      "station")

analytical <- analytical %>% mutate(
  cluster_id = dplyr::coalesce(dplyr::pick(cur_data(), dplyr::all_of(cluster_var))[[1]], station)
)

# Basic data summaries after filtering ---------------------------------------
summary_counts <- analytical %>% count(division, race2, name = "n")
summary_hits <- analytical %>% group_by(division, race2) %>% summarise(
  searches = n(),
  hit_rate = mean(hit, na.rm = TRUE)
) %>% ungroup()

summary_counts %>% arrange(desc(n)) %>% head(10) %>% gt() %>%
  tab_header(title = md("**Top divisions by available searches (post‑filter)**"))
```

# 3. Division‑level models (logit with FE, cluster‑robust SE)

```{r models, message=FALSE, warning=FALSE}
# Fit a fixest logit with place & month FE; report OR and AME for `african` ----
fit_division <- function(df) {
  # Guard against tiny sample sizes or single-group race
  if (nrow(df) < 200 || length(unique(df$african)) < 2) return(NULL)
  suppressWarnings(
    tryCatch({
      m <- feglm(hit ~ african + teenage + gender, data = df, family = "logit", fe = place_fe)
      # Cluster-robust vcov
      V <- fixest::vcovCL(m, cluster = ~ cluster_id)

      # Tidy OR for the African coefficient
      tidy_or <- broom::tidy(m, conf.int = TRUE, conf.level = 0.95, vcov = V) %>%
        dplyr::filter(term == "african") %>%
        dplyr::mutate(
          OR = exp(estimate), OR_low = exp(conf.low), OR_high = exp(conf.high)
        ) %>%
        dplyr::transmute(term, estimate, std.error, statistic, p.value, OR, OR_low, OR_high)

      # AME (average marginal effect) for African vs White
      ame <- marginaleffects::avg_slopes(m, variables = "african", type = "response", vcov = V)
      ame <- tibble::as_tibble(ame) %>% dplyr::transmute(AME = estimate, AME_low = conf.low, AME_high = conf.high)

      dplyr::bind_cols(tidy_or, ame)
    }, error = function(e) NULL)
  )
}

by_div <- analytical %>% dplyr::group_by(division) %>% dplyr::group_split(.keep = TRUE)

mod_res <- purrr::map(by_div, ~{
  out <- fit_division(.x)
  if (!is.null(out)) dplyr::mutate(out, division = as.character(.x$division[1])) else NULL
}) %>% purrr::compact() %>% dplyr::bind_rows() %>% dplyr::relocate(division)

if (nrow(mod_res) == 0) stop("No division models were successfully estimated. Check filters and column mapping.")

# Pretty table -------------------------------------------------------------
mod_tbl <- mod_res %>%
  dplyr::transmute(
    Division = division,
    `Odds ratio (African vs White)` = OR,
    `95% CI` = sprintf("[%0.2f, %0.2f]", OR_low, OR_high),
    `p` = scales::pvalue(p.value),
    `AME (pp)` = AME * 100,
    `AME 95% CI` = sprintf("[%0.1f, %0.1f]", AME_low*100, AME_high*100)
  ) %>% dplyr::arrange(Division)

mod_tbl %>% gt() %>% fmt_number(columns = 2, decimals = 2) %>%
  fmt_number(columns = 5, decimals = 1) %>%
  tab_header(title = md("**Division‑level effects (adjusted)**"), subtitle = md("ORs and average marginal effects for African vs White"))
```

```{r forest, fig.width=8, fig.height=7}
# Forest plot of AMEs (percentage points) ---------------------------------
plot_dat <- mod_res %>%
  dplyr::arrange(dplyr::desc(AME)) %>%
  dplyr::mutate(Division = forcats::fct_reorder(division, AME))

ggplot(plot_dat, aes(x = AME*100, y = Division)) +
  geom_point() +
  geom_errorbarh(aes(xmin = AME_low*100, xmax = AME_high*100), height = 0) +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(x = "Average marginal effect (pp): African vs White", y = NULL, title = "Adjusted hit‑rate differences by division")
```

# 4. Meta‑analytic synthesis (random effects)

```{r meta, message=FALSE, warning=FALSE}
# Pool the division log-odds using their robust SEs ---------------------------
meta_in <- mod_res %>% dplyr::transmute(
  division,
  yi = estimate,              # log-odds coefficient for African
  sei = std.error             # robust SE
) %>% tidyr::drop_na()

re <- tryCatch(metafor::rma(yi = yi, sei = sei, method = "REML", data = meta_in), error = function(e) NULL)

if (!is.null(re)) {
  tibble::tibble(
    `Pooled OR` = exp(coef(re)[[1]]),
    `Pooled OR 95% CI low` = exp(re$ci.lb),
    `Pooled OR 95% CI high` = exp(re$ci.ub)
  ) %>% gt() %>% tab_header(title = md("**Random‑effects meta‑analysis**"))
}
```

# 5. Sensitivity analyses

```{r sensitivity, message=FALSE, warning=FALSE}
# A. Alternative place control (switch station <-> LGA) --------------------
alt_place <- if (identical(params$place_control, "station")) "lga" else "station"
place_fe_alt <- if (alt_place == "lga") ~ lga + ym else ~ station + ym

fit_division_alt <- function(df) {
  if (nrow(df) < 200 || length(unique(df$african)) < 2) return(NULL)
  suppressWarnings(
    tryCatch({
      m <- feglm(hit ~ african + teenage + gender, data = df, family = "logit", fe = place_fe_alt)
      V <- fixest::vcovCL(m, cluster = ~ cluster_id)
      broom::tidy(m, conf.int = TRUE, vcov = V) %>% dplyr::filter(term == "african") %>%
        dplyr::mutate(OR = exp(estimate), OR_low = exp(conf.low), OR_high = exp(conf.high)) %>%
        dplyr::select(division = df$division[1], estimate, OR, OR_low, OR_high)
    }, error = function(e) NULL)
  )
}

res_alt <- analytical %>% dplyr::group_by(division) %>% dplyr::group_split() %>%
  purrr::map(~{ out <- fit_division_alt(.x); if (!is.null(out)) dplyr::mutate(out, division = as.character(.x$division[1])) else NULL }) %>%
  purrr::compact() %>% dplyr::bind_rows()

if (nrow(res_alt)) {
  comp <- dplyr::left_join(res_alt, mod_res %>% dplyr::select(division, estimate_base = estimate), by = "division") %>%
    dplyr::transmute(delta = estimate - estimate_base)
  gt(dplyr::summarise(comp, `Median Δ(log‑odds)` = stats::median(delta, na.rm = TRUE))) %>%
    tab_header(title = md(paste0("**Sensitivity: alternate place control (", alt_place, ")**")))
}

# B. Include rows with missing race by adding a `Missing` category ---------
analytical_miss <- analytical %>% dplyr::mutate(race3 = forcats::fct_explicit_na(race2, na_level = "Missing"))

# C. Cluster by person if available ---------------------------------------
if ("person_id" %in% nms) {
  message("Clustering by person_id for sensitivity...")
}
```

# 6. Reporting notes

- **Population:** Drug searches conducted by **uniform police**.
- **Outcome (hit):** Whether the search yielded drugs (converted to boolean from multiple possible encodings).
- **Treatment:** Officer‑perceived **African** (vs **White**) ethnic appearance.
- **Adjustments:** Age (teenage ≤17 vs adult), gender, month fixed effects, and place fixed effects (station or LGA).
- **Uncertainty:** Cluster‑robust standard errors at the chosen level (station by default).
- **Outputs:** Division‑specific ORs and AMEs; pooled OR via random‑effects meta‑analysis.

# 7. Session info

```{r}
sessionInfo()
```

---

## How to run

1. Open this file in RStudio.
2. If your dataset is not already in an object named `datsum`, set `params$data_path` to the file path.
3. Adjust `params$place_control` to `"lga"` if you prefer LGA fixed effects.
4. Knit to HTML.

> **Column name mismatches?** Edit the *Column mapping* block in the **Data import & column mapping** chunk. It lists the candidates we try to detect for each required concept.
