---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
---

```{r load-packages-and-data, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(strayr)
library(sf)
library(broom)
library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(viridis)

dat <- readRDS("../Output.data/Clean.search.data.RDS")

dat <- dat %>%
  mutate(Racial.appearance.transformed = case_when(
      Racial.appearance.transformed == "Mediterranean AND Middle Eastern - DO NOT USE" ~ "Mediterranean/Middle Eastern – Unusable",
      TRUE ~ Racial.appearance.transformed
    ))

division_lga_lookup <- dat %>%
  filter(
    Area.type == "Metro",
    !is.na(Division),
    !is.na(Local.Government.Area)
  ) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )
```

# About the data  

This report summarises over 13,000 warrantless drug searches conducted by uniformed Victoria Police officers while patrolling on foot in public places in Melbourne. The data covers the years 2018, 2019, 2022, and 2023, and is sourced from [**The Racial Profiling Data Monitoring Project**](https://khaki-pepper-b3rf.squarespace.com/). Data from 2020 and 2021 is excluded due to the extraordinary circumstances of the COVID-19 pandemic.

Under section 82 of the [*Drugs, Poisons and Controlled Substances Act 1981*](http://classic.austlii.edu.au/au/legis/vic/consol_act/dpacsa1981422/s82.html), a police officer may stop and search a person in a public place if they reasonably suspect the person is in possession of illegal drugs. Drug searches are the most common type of warrantless search carried out by police in Victoria.

## Racial appearance categories

Victoria Police records the perceived racial appearance of individuals subjected to stop and search. These categories are based on the officer’s perception, not the individual's self-reported ancestry or identity.

**Missing data:**  
Racial appearance was not recorded in 2,860 search records — approximately 21% of all records — despite being a mandatory field for stop and search data collection since 2019.

# Methodology  

The following sections present **hit rate analyses**—the proportion of stop and searches that resulted in the discovery of illegal drugs—for different racial appearance groups across metropolitan Melbourne. A **hit rate** is widely used in policing research to evaluate whether stop and search practices are effectively identifying contraband, and to assess whether they are applied equitably across racial groups.

This analysis specifically compares hit rates for people recorded by police as being of **African appearance** and those of **White appearance**. Comparing these two groups helps assess whether individuals of African appearance are being subjected to searches based on weaker suspicion. If hit rates are consistently lower for African appearing individuals, it may indicate that racial profiling is influencing police decision-making.

Hit rate comparisons are presented at five geographic levels:
- **Whole of Melbourne** (Metro area aggregate),
- **Police Region**,
- **Police Division**,
- **Police Service Area (PSA)**, and
- **Individual Police Stations**.

## Whole of Melbourne

### Data summary - Melbourne 

```{r hit-rate-all-races-final,  echo=FALSE}
# Filter and prepare data
cleaned_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# List categories to appear last
special_groups <- c("Mediterranean/Middle Eastern – Unusable", "Missing")

# Summarise by race
race_hit_rates <- cleaned_data %>%
  group_by(Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    `Hit Rate` = round(mean(Found.Flag), 3),
    .groups = "drop"
  )

# Reorder by descending Searches, with special groups last
ordered_non_special <- race_hit_rates %>%
  filter(!Race %in% special_groups) %>%
  arrange(desc(Searches)) %>%
  pull(Race)

final_order <- c(ordered_non_special, special_groups)

# Apply ordering
race_hit_rates <- race_hit_rates %>%
  mutate(Race = factor(Race, levels = final_order)) %>%
  arrange(Race)

# Display table
kable(
  race_hit_rates,
  caption = "Drug Searches, Hits, and Hit Rates by Racial Appearance (Metro Victoria, Uniform police at Stations)",
  align = "lrrr"
)

```

### Hit rate analysis - Melbourne 


```{r prop-test-all,  echo = FALSE}


# Filter and prepare data
vic_test_data <- dat %>%
  filter(
    Racial.appearance.transformed %in% c("African", "White"),
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Station.uniform)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0))

# Summarise by group
melb_summary <- vic_test_data %>%
  group_by(Racial.appearance.transformed) %>%
  summarise(
    Searches = n(),
    Hits = sum(found_flag),
    Misses = Searches - Hits,
    HitRate = mean(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Searches, Hits, Misses, HitRate),
    names_sep = "."
  )

# Compute prop.test, odds, odds ratio, interpretation
melb_results <- melb_summary %>%
  rowwise() %>%
  mutate(
    Odds_African = Hits.African / Misses.African,
    Odds_White = Hits.White / Misses.White,
    Odds_Ratio = round(Odds_African / Odds_White, 3),
    prop_result = list(prop.test(
      x = c(Hits.African, Hits.White),
      n = c(Searches.African, Searches.White),
      correct = FALSE
    )),
    tidy_result = list(tidy(prop_result))
  ) %>%
  unnest(tidy_result) %>%
  mutate(
    `Difference in Hit Rate` = round(estimate2 - estimate1, 3),
    `95% CI Lower` = round(conf.low, 3),
    `95% CI Upper` = round(conf.high, 3),
    `P-value` = round(p.value, 3),
    `Odds Ratio (African vs White)` = Odds_Ratio,
    Interpretation = case_when(
      `P-value` >= 0.05 ~ "No statistically significant difference",
      Odds_Ratio < 1 ~ "African appearing people searched on weaker suspicion",
      Odds_Ratio > 1 ~ "African appearing people searched on stronger suspicion",
      TRUE ~ "—"
    )
  ) %>%
  select(
    `Searches (African)` = Searches.African,
    `Searches (White)` = Searches.White,
    `Hits (African)` = Hits.African,
    `Hits (White)` = Hits.White,
    `Hit Rate (African)` = HitRate.African,
    `Hit Rate (White)` = HitRate.White,
    `Difference in Hit Rate`,
    `95% CI Lower`,
    `95% CI Upper`,
    `P-value`,
    `Odds Ratio (African vs White)`,
    Interpretation
  )

# Display result in table
kable(
  melb_results,
  caption = "Drug Search Hit Rate Comparison – Whole of Melbourne (African vs White)",
  digits = 3,
  align = "rrrrrrrrrrrl"
)
```


## Police region level 

### Data overview - Region

```{r race-tally-by-region-wide,  echo=FALSE}


# Step 1: Define ordered race levels
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Prepare data
region_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !is.na(Racial.appearance.transformed),
    !is.na(Region)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Step 3: Summarise
region_summary_long <- region_data %>%
  group_by(Region, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Region, Race, label)

# Step 4: Pivot and reorder columns explicitly
region_table <- region_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  select(Region, all_of(ordered_races))  # Force column order

write_xlsx(region_table, "region_race_table_summary.xlsx")

# Step 5: Display
kable(
  region_table,
  caption = "Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Region)",
  align = "l"
)

```

### Hit rate analysis - Region  

```{r prop-test-region, echo = FALSE}

# Summarise hits and searches by Region and Race
region_summary <- vic_test_data %>%
  group_by(Region, Racial.appearance.transformed) %>%
  summarise(
    Searches = n(),
    Hits = sum(found_flag),
    Misses = Searches - Hits,
    HitRate = round(mean(found_flag), 3),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Searches, Hits, Misses, HitRate),
    names_sep = "."
  )



region_results <- region_summary %>%
  rowwise() %>%
  mutate(
    can_test = all(!is.na(c(Hits.African, Hits.White, Searches.African, Searches.White))) &&
               Misses.African > 0 && Misses.White > 0,

    Odds_African = if (can_test) Hits.African / Misses.African else NA_real_,
    Odds_White   = if (can_test) Hits.White / Misses.White else NA_real_,
    Odds_Ratio   = if (can_test) round(Odds_African / Odds_White, 3) else NA_real_,

    # Run test and safely wrap result as list
    tidy_result = list(
      if (can_test) {
        result <- prop.test(
          x = c(Hits.African, Hits.White),
          n = c(Searches.African, Searches.White),
          correct = FALSE
        )
        tidy(result)
      } else {
        tibble(
          estimate1 = NA_real_,
          estimate2 = NA_real_,
          conf.low = NA_real_,
          conf.high = NA_real_,
          p.value = NA_real_
        )
      }
    )
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    `Difference in Hit Rate` = round(estimate2 - estimate1, 3),
    `95% CI Lower` = round(conf.low, 3),
    `95% CI Upper` = round(conf.high, 3),
    `P-value` = round(p.value, 3),
    Interpretation = case_when(
      is.na(Odds_Ratio) ~ "Insufficient data for comparison",
      `P-value` >= 0.05 ~ "No statistically significant difference",
      Odds_Ratio < 1 ~ "African appearing people searched on weaker suspicion",
      Odds_Ratio > 1 ~ "African appearing people searched on stronger suspicion",
      TRUE ~ "—"
    )
  ) %>%
  ungroup() %>%
    arrange(`P-value`) %>%
  select(
    Region,
    `Searches (African)` = Searches.African,
    `Searches (White)` = Searches.White,
    `Hits (African)` = Hits.African,
    `Hits (White)` = Hits.White,
    `Hit Rate (African)` = HitRate.African,
    `Hit Rate (White)` = HitRate.White,
    `Difference in Hit Rate`,
    `95% CI Lower`,
    `95% CI Upper`,
    `P-value`,
    `Odds Ratio (African vs White)` = Odds_Ratio,
    Interpretation
  )


# Display interactive scrollable DT table
kable(
  region_results,
  caption = "Drug Search Hit Rate Comparison by Region (African vs White)",
  digits = 3,
  align = "lrrrrrrrrrrl"
)
```



## Police Division level  

### Data overview - Division 


```{r division-hit-rate-summary, echo=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

# Step 1: Define ordered race levels
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Create Division → LGA lookup
division_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Division), !is.na(Local.Government.Area)) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

# Step 3: Prepare Division-level data
division_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !is.na(Racial.appearance.transformed),
    !is.na(Division)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Step 4: Summarise by Division and Race
division_summary_long <- division_data %>%
  group_by(Division, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Division, Race, label)

# Step 5: Pivot wide and join LGAs
division_table <- division_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  left_join(division_lga_lookup, by = "Division") %>%
  relocate(LGAs, .after = Division) %>%
  select(Division, LGAs, all_of(ordered_races))

write_xlsx(division_table, "division_race_search_summary.xlsx")

# Step 6: Display table (plain)
kable(
  division_table,
  caption = "Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Division)",
  align = "ll"
)
```

### Hit rate analysis - Division  


```{r prop-test-psa, echo = FALSE, warning = FALSE, message = FALSE}

# Step 1: Summarise data by Division and Race
division_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Division)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Division, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = "."
  )

# Step 2: Run prop.test for African vs White at Division level
division_results <- division_summary %>%
  rowwise() %>%
  mutate(
    African_HitRate = ifelse(Total_Searches.African > 0,
                             Total_Hits.African / Total_Searches.African, NA_real_),
    White_HitRate = ifelse(Total_Searches.White > 0,
                           Total_Hits.White / Total_Searches.White, NA_real_),

    African_Miss = Total_Searches.African - Total_Hits.African,
    White_Miss = Total_Searches.White - Total_Hits.White,

    Odds_African = ifelse(African_Miss > 0, Total_Hits.African / African_Miss, NA_real_),
    Odds_White = ifelse(White_Miss > 0, Total_Hits.White / White_Miss, NA_real_),
    Odds_Ratio = ifelse(!is.na(Odds_African) & !is.na(Odds_White), Odds_African / Odds_White, NA_real_),

    prop_test = if (
      all(!is.na(c(Total_Hits.African, Total_Hits.White,
                   Total_Searches.African, Total_Searches.White)))
    ) {
      list(tryCatch(
        prop.test(
          x = c(Total_Hits.African, Total_Hits.White),
          n = c(Total_Searches.African, Total_Searches.White),
          correct = TRUE
        ),
        error = function(e) NULL
      ))
    } else {
      list(NULL)
    },

    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & African_HitRate < White_HitRate ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & African_HitRate > White_HitRate ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%
  select(
    Division,
    African_Searches = Total_Searches.African,
    White_Searches = Total_Searches.White,
    African_Hits = Total_Hits.African,
    White_Hits = Total_Hits.White,
    African_HitRate,
    White_HitRate,
    Odds_Ratio,
    ConfLow = conf.low,
    ConfHigh = conf.high,
    P_Value = p.value,
    Interpretation
  ) %>%
  arrange(P_Value)

# Step 3: Format for display
division_results_clean <- division_results %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  ) %>%
  rename(
    `Searches (African)` = African_Searches,
    `Searches (White)` = White_Searches,
    `Hits (African)` = African_Hits,
    `Hits (White)` = White_Hits,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `Odds Ratio (African/White)` = Odds_Ratio,
    `95% CI Lower` = ConfLow,
    `95% CI Upper` = ConfHigh,
    `P-value` = P_Value
  )

# Step 4: Create Division → LGA lookup and join
division_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Division), !is.na(Local.Government.Area)) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

division_results_clean <- division_results_clean %>%
  left_join(division_lga_lookup, by = "Division") %>%
  relocate(LGAs, .after = Division)

# Step 5: Export to Excel
write_xlsx(division_results_clean, "division_drug_search_proptest_results.xlsx")

# Step 6: Display in DT
datatable(
  division_results_clean,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 10,
    lengthMenu = c(10, 20, 50),
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '300px', targets = 1)
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Search Hit Rate Comparison by Division (African vs White)'
  )
)


```

## Police Servic Area level  

### Data overview - PSA

```{r summary PSA, echo= FALSE, warning = FALSE, message = FALSE}


# Step 1: Define racial appearance ordering
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Create PSA → LGA lookup table
psa_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Station.uniform), !is.na(Local.Government.Area)) %>%
  distinct(Police.Service.Area, Local.Government.Area) %>%
  group_by(Police.Service.Area) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

# Step 3: Prepare PSA-level summary
psa_summary_long <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  ) %>%
  group_by(Police.Service.Area, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Police.Service.Area, Race, label)

# Step 4: Pivot wide and join LGAs
psa_table <- psa_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  left_join(psa_lga_lookup, by = "Police.Service.Area") %>%
  relocate(LGAs, .after = Police.Service.Area) %>%
  select(`Police Service Area` = Police.Service.Area, LGAs, all_of(ordered_races))

# Step 5: Export to Excel
write_xlsx(psa_table, "psa_racial_search_summary.xlsx")

# Step 6: Display interactive table with download button
datatable(
  psa_table,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 15,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '300px', targets = 1)  # Widen LGA column
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by PSA)'
  )
)

```

### Hit rate analysis - PSA

```{r hit rate psa, echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(broom)
library(DT)
library(writexl)

# Step 1: Summarise data by PSA and Race
psa_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Police.Service.Area)
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Police.Service.Area, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = "."
  )

# Step 2: Run prop.test for African vs White per PSA
psa_results <- psa_summary %>%
  rowwise() %>%
  mutate(
    prop_test = if (
      all(!is.na(c(Total_Hits.African, Total_Hits.White,
                   Total_Searches.African, Total_Searches.White)))
    ) {
      list(prop.test(
        x = c(Total_Hits.African, Total_Hits.White),
        n = c(Total_Searches.African, Total_Searches.White),
        correct = FALSE
      ))
    } else {
      list(NULL)
    },
    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%
  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & estimate1 < estimate2 ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & estimate1 > estimate2 ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%
  select(
    Police.Service.Area,
    African_Searches = Total_Searches.African,
    White_Searches = Total_Searches.White,
    African_Hits = Total_Hits.African,
    White_Hits = Total_Hits.White,
    African_HitRate = estimate1,
    White_HitRate = estimate2,
    ConfLow = conf.low,
    ConfHigh = conf.high,
    P_Value = p.value,
    Interpretation
  ) %>%
  arrange(P_Value)

# Step 3: Format for display
psa_results_clean <- psa_results %>%
  mutate(
    across(where(is.numeric), ~ round(.x, 3))
  ) %>%
  rename(
    `Searches (African)` = African_Searches,
    `Searches (White)` = White_Searches,
    `Hits (African)` = African_Hits,
    `Hits (White)` = White_Hits,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `95% CI Lower` = ConfLow,
    `95% CI Upper` = ConfHigh,
    `P-value` = P_Value
  )

# Step 4: Export to Excel
write_xlsx(psa_results_clean, "psa_drug_search_proptest_results.xlsx")

# Step 5: Display interactive table
datatable(
  psa_results_clean,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 15,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '250px', targets = 0)
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Search Hit Rate Comparison by Police Service Area (African vs White)'
  )
)

```
## Station level  

### Data overview - Station 



```{r summarise-station-level-hit-rates,  echo=FALSE, warning=FALSE, message = FALSE}



# Step 1: Define racial ordering
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Summarise searches and hit rates by Station
station_summary_long <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  ) %>%
  group_by(Station.uniform, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Station.uniform, Race, label)

# Step 3: Pivot wide
station_table <- station_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  select(Station = Station.uniform, any_of(ordered_races))

# Step 4: Save Excel file
write_xlsx(station_table, "station_racial_search_summary.xlsx")

# Step 5: Display interactive datatable
datatable(
  station_table,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 15,
    autoWidth = TRUE
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Police Station)'
  )
)

```


### Hit rate analysis - Station

```{r prop-test-station, warning=FALSE, message=FALSE, echo = FALSE}
# Step 1: Create station_summary
station_summary <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform),
    Racial.appearance.transformed %in% c("African", "White")
  ) %>%
  mutate(found_flag = ifelse(Found == "Yes", 1, 0)) %>%
  group_by(Station.uniform, Racial.appearance.transformed) %>%
  summarise(
    Total_Searches = n(),
    Total_Hits = sum(found_flag),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Racial.appearance.transformed,
    values_from = c(Total_Searches, Total_Hits),
    names_sep = ".",
    values_fill = 0
  )

# Step 2: Run prop.test and compute odds ratio
station_results <- station_summary %>%
  rowwise() %>%
  mutate(
    African_HitRate = ifelse(Total_Searches.African > 0,
                             Total_Hits.African / Total_Searches.African, NA_real_),
    White_HitRate = ifelse(Total_Searches.White > 0,
                           Total_Hits.White / Total_Searches.White, NA_real_),

    African_Miss = Total_Searches.African - Total_Hits.African,
    White_Miss = Total_Searches.White - Total_Hits.White,

    Odds_African = ifelse(African_Miss > 0, Total_Hits.African / African_Miss, NA_real_),
    Odds_White = ifelse(White_Miss > 0, Total_Hits.White / White_Miss, NA_real_),
    Odds_Ratio = ifelse(!is.na(Odds_African) & !is.na(Odds_White), Odds_African / Odds_White, NA_real_),

    can_test = Total_Searches.African > 0 & Total_Searches.White > 0,
    prop_test = if (can_test) {
      list(tryCatch(
        prop.test(
          x = c(Total_Hits.African, Total_Hits.White),
          n = c(Total_Searches.African, Total_Searches.White),
          correct = TRUE
        ),
        error = function(e) NULL
      ))
    } else {
      list(NULL)
    },

    tidy_result = if (!is.null(prop_test)) list(broom::tidy(prop_test)) else list(NULL)
  ) %>%
  unnest(tidy_result, keep_empty = TRUE) %>%

  mutate(
    Interpretation = case_when(
      is.na(p.value) ~ "Insufficient data",
      p.value < 0.05 & African_HitRate < White_HitRate ~ "African-appearing people searched on weaker suspicion",
      p.value < 0.05 & African_HitRate > White_HitRate ~ "African-appearing people searched on stronger suspicion",
      TRUE ~ "No statistically significant difference"
    )
  ) %>%

  select(
    Station = Station.uniform,
    `Searches (African)` = Total_Searches.African,
    `Searches (White)` = Total_Searches.White,
    `Hits (African)` = Total_Hits.African,
    `Hits (White)` = Total_Hits.White,
    `Hit Rate (African)` = African_HitRate,
    `Hit Rate (White)` = White_HitRate,
    `Odds Ratio (African/White)` = Odds_Ratio,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `P-value` = p.value,
    Interpretation
  ) %>%
  arrange(`P-value`)



```


# Appendix – Ethnic Appearance Mapping

```{r mapping-categories-2022-2023, echo = FALSE}
dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```
