---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: readable
    highlight: zenburn
    fig_caption: true
    df_print: paged
    css: styles.css  # Optional: include this file if you want to customise styles
---

```{r load-packages-and-data, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results = "asis")

library(strayr)
library(sf)
library(broom)
library(tidyverse)
library(plotly)
library(summarytools)
library(kableExtra)
library(magrittr)
library(rmarkdown)
library(forcats)
library(gmodels)
library(crosstable)
library(janitor)
library(DT)
library(writexl)
library(viridis)

dat <- readRDS("../VicPol Search Data Clean/Clean.search.data.RDS")

division_lga_lookup <- dat %>%
  filter(
    Area.type == "Metro",
    !is.na(Division),
    !is.na(Local.Government.Area)
  ) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )
```

---
title: "Victoria Police Drug Stop and Search Data"
author: "Monika Sarder"
date: "2025-07-27"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction

This report presents **descriptive tallies** of discretionary drug-related stop and search activities conducted by Victoria Police. The dataset includes over **13,000 records** of **warrantless searches** carried out by **uniformed officers on foot patrol in public places** across **metropolitan Melbourne** in the years **2018, 2019, 2022, and 2023**.

The tables below summarise the number of searches conducted by **perceived racial appearance**, as well as across key policing geographies—**Police Regions**, **Divisions**, **Service Areas (PSAs)**, and **Stations**.

> **Important note:** These counts are provided for **exploratory purposes only**. No statistical testing has been conducted, and no conclusions should be drawn about disparities or discrimination based on this data alone. Further analysis, including appropriate statistical controls, is required to assess whether any observed differences are meaningful or attributable to bias.

# About the Data

This dataset is sourced from the *Racial Profiling Data Monitoring Project*, which compiled records of discretionary drug searches conducted by Victoria Police. Records from **2020 and 2021** are excluded due to the extraordinary conditions of the COVID-19 pandemic, which significantly altered patterns of police activity and public mobility.

Under **section 82 of the Drugs, Poisons and Controlled Substances Act 1981 (Vic)**, police officers may search individuals in public places if they form a **reasonable suspicion** that the person is in possession of illegal drugs. Drug searches are the most commonly used form of **warrantless search** by Victoria Police.

# Racial Appearance Categories

Victoria Police policy requires officers to record the **perceived racial appearance** of each individual subject to a stop and search. These categories are based on the officer’s observation and do not reflect the individual’s self-identification or cultural background.

Of the original **13,896 records**, **2,860 (21%)** were missing racial appearance data, despite this being a mandatory field since 2019. In 2022 and 2023, Victoria Police also merged *Middle Eastern* and *Mediterranean* appearance categories into a single label that could not be reliably analysed. These, along with low-frequency or indeterminate race classifications, have been excluded from the counts.

The resulting dataset includes **9,584 records** with usable racial appearance data. The high rate of missing or unusable entries highlights a significant data quality concern and underlines the need for **improved recording practices**.

# Purpose of This Page

The tables that follow present basic **counts of searches by race and geography**, without further interpretation or modelling. These tallies are intended to support transparency, facilitate further investigation, and assist stakeholders in identifying patterns that may warrant deeper statistical analysis.



### Data summary - Melbourne 

```{r hit-rate-all-races-final, echo=FALSE}

# Filter and prepare data
cleaned_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !Racial.appearance.transformed %in% c("Missing", "Other", "South American", "Mediterranean/Middle Eastern – Unusable"),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Summarise by race
race_hit_rates <- cleaned_data %>%
  group_by(Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    `Hit Rate` = round(mean(Found.Flag), 3),
    .groups = "drop"
  )

# Create total row
total_row <- cleaned_data %>%
  summarise(
    Race = "Total",
    Searches = n(),
    Hits = sum(Found.Flag),
    `Hit Rate` = round(mean(Found.Flag), 3)
  )

# Bind and order: descending by Searches, Total last
race_hit_rates <- bind_rows(race_hit_rates, total_row) %>%
  mutate(Race = factor(Race, levels = c(
    race_hit_rates %>% arrange(desc(Searches)) %>% pull(Race),
    "Total"
  ))) %>%
  arrange(Race)

# Display table
kable(
  race_hit_rates,
  caption = "Drug Searches, Hits, and Hit Rates by Racial Appearance (Metro Victoria, Uniform Police at Stations)",
  align = "lrrr"
)

```
## Adjusted Comparison of Hit Rates by Racial Appearance

We use a binomial logistic regression to model the probability of a "hit" (drugs found), controlling for **year of search**. The reference group for race is **White**, and all coefficients are interpreted relative to this group.

```{r logistic-regression-hitrate, message=FALSE, warning=FALSE, echo = FALSE}

# Filter dataset
model_data <- cleaned_data %>%
  filter(
    Race %in% c("White", "African", "Asian", "Aboriginal", "South Asian", "Middle Eastern", "Pacific Islander"),
    !is.na(Year)
  ) %>%
  mutate(
    Race = factor(Race, levels = c("White", "African", "Asian", "Aboriginal", "South Asian", "Middle Eastern", "Pacific Islander")),
    Year = as.factor(Year)
  )

# Logistic regression model
model <- glm(Found.Flag ~ Race + Year, data = model_data, family = binomial)

# Tidy model with odds ratios and CIs
results <- tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(grepl("^Race", term)) %>%
  mutate(
    Race = gsub("Race", "", term),
    p.value = round(p.value, 4),
    OR = round(estimate, 2),
    CI = paste0("(", round(conf.low, 2), ", ", round(conf.high, 2), ")"),
    Sig = ifelse(p.value < 0.05, "*", "")
  ) %>%
  select(Race, OR, CI, p.value, Sig)

# Add reference group
white_row <- data.frame(
  Race = "White (ref)",
  OR = 1.00,
  CI = "(ref)",
  p.value = NA,
  Sig = ""
)

final_table <- bind_rows(white_row, results)

# Display as nice markdown table
kable(final_table,
      caption = "Adjusted Odds Ratios for Drug Search Hits by Racial Appearance (Reference: White)",
      align = "lccrc") %>%
  kable_styling(full_width = FALSE, position = "left")
```

### Interpretation of Results

The table above presents **adjusted odds ratios** comparing the likelihood that a drug search resulted in a **"hit"** (drugs found) for each racial appearance group, relative to individuals perceived by police as **White**. The model controls for **year of search**, helping to isolate the effect of racial appearance from broader shifts in policing practices or public behaviour across time.

- An **odds ratio (OR) below 1** means the group was **less likely** than White individuals to have drugs found during a search.
- An OR of **1** indicates **no difference**.
- An asterisk (*) marks results that are **statistically significant** (*p* < 0.05).

**Key findings:**

- **South Asian** individuals were **36% less likely** to have a drug search result in a hit compared to White individuals (*OR = 0.64, p = 0.0275*). This statistically significant result suggests a **lower threshold of suspicion** may be applied when stopping and searching South Asian people.
- **African** individuals were **22% less likely** to yield a drug find (*OR = 0.78, p = 0.0381*), also a statistically significant disparity.
- Other groups—**Asian**, **Aboriginal**, **Middle Eastern**, and **Pacific Islander**—also showed lower odds of a drug find compared to White individuals, but these differences were **not statistically significant** in this model.

These findings indicate that **South Asian and African** individuals are more likely to be searched **without sufficient grounds**, reinforcing long-standing concerns about **racial profiling** in Victoria Police’s use of discretionary stop and search powers. The disparities persisted even after controlling for **year of search**, suggesting they are not explained by shifting policing patterns over time.


## Search Counts by Race and Police Region

The following table presents a **descriptive summary** of drug searches by racial appearance and Victoria Police Region. This analysis is **exploratory** and aims to illustrate the geographic distribution of search activity across major operational areas.

No statistical testing has been conducted in this section. Therefore, the figures should be interpreted with **caution**. Apparent differences between regions or racial groups may be due to a range of factors, and no conclusions should be drawn about disparities or bias without further statistical analysis to control for confounding variables.

```{r race-tally-by-region-wide,  echo=FALSE}


# Step 1: Define ordered race levels
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Prepare data
region_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !is.na(Racial.appearance.transformed),
    !is.na(Region)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Step 3: Summarise
region_summary_long <- region_data %>%
  group_by(Region, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Region, Race, label)

# Step 4: Pivot and reorder columns explicitly
region_table <- region_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  select(Region, all_of(ordered_races))  # Force column order

write_xlsx(region_table, "region_race_table_summary.xlsx")

# Step 5: Display
kable(
  region_table,
  caption = "Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Region)",
  align = "l"
)

```

## Search Counts by Race and Police Division

The following table presents a **descriptive summary** of drug searches by racial appearance and Victoria Police Region. This analysis is **exploratory** and aims to illustrate the geographic distribution of search activity across major operational areas.

No statistical testing has been conducted in this section. Therefore, the figures should be interpreted with **caution**. Apparent differences between regions or racial groups may be due to a range of factors, and no conclusions should be drawn about disparities or bias without further statistical analysis to control for confounding variables.



```{r division-hit-rate-summary, echo=FALSE}

# Step 1: Define ordered race levels
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Create Division → LGA lookup
division_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Division), !is.na(Local.Government.Area)) %>%
  distinct(Division, Local.Government.Area) %>%
  group_by(Division) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

# Step 3: Prepare Division-level data
division_data <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(`Search.type - Drugs`),
    !is.na(Found),
    !is.na(Racial.appearance.transformed),
    !is.na(Division)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  )

# Step 4: Summarise by Division and Race
division_summary_long <- division_data %>%
  group_by(Division, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Division, Race, label)

# Step 5: Pivot wide and join LGAs
division_table <- division_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  left_join(division_lga_lookup, by = "Division") %>%
  relocate(LGAs, .after = Division) %>%
  select(Division, LGAs, all_of(ordered_races))

write_xlsx(division_table, "division_race_search_summary.xlsx")

# Step 6: Display table (plain)
kable(
  division_table,
  caption = "Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Division)",
  align = "ll"
)
```


## Search Counts by Race and Police Service Area (PSA)

The following table presents a **descriptive tally** of drug searches by racial appearance and Police Service Area (PSA). This analysis is **exploratory** and designed to provide an overview of the distribution of search activity across different locations and groups.

No statistical testing has been conducted in this section. As such, the figures should be interpreted with **caution**, and no firm conclusions should be drawn about disparities or significance. Further modelling and statistical analysis would be required to determine whether any observed differences reflect patterns of racial bias or are attributable to other factors.


```{r summary PSA, echo= FALSE, warning = FALSE, message = FALSE}


# Step 1: Define racial appearance ordering
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Create PSA → LGA lookup table
psa_lga_lookup <- dat %>%
  filter(Area.type == "Metro", !is.na(Station.uniform), !is.na(Local.Government.Area)) %>%
  distinct(Police.Service.Area, Local.Government.Area) %>%
  group_by(Police.Service.Area) %>%
  summarise(
    LGAs = paste(sort(unique(Local.Government.Area)), collapse = ", "),
    .groups = "drop"
  )

# Step 3: Prepare PSA-level summary
psa_summary_long <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  ) %>%
  group_by(Police.Service.Area, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Police.Service.Area, Race, label)

# Step 4: Pivot wide and join LGAs
psa_table <- psa_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  left_join(psa_lga_lookup, by = "Police.Service.Area") %>%
  relocate(LGAs, .after = Police.Service.Area) %>%
  select(`Police Service Area` = Police.Service.Area, LGAs, all_of(ordered_races))

# Step 5: Export to Excel
write_xlsx(psa_table, "psa_racial_search_summary.xlsx")

# Step 6: Display interactive table with download button
datatable(
  psa_table,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 15,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = '300px', targets = 1)  # Widen LGA column
    )
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by PSA)'
  )
)

```


## Search Counts by Race and Station

The following table presents a **descriptive summary** of drug searches by racial appearance and Victoria Police Station. This analysis is **exploratory** and aims to illustrate the geographic distribution of search activity across major operational areas.

No statistical testing has been conducted in this section. Therefore, the figures should be interpreted with **caution**. Apparent differences between regions or racial groups may be due to a range of factors, and no conclusions should be drawn about disparities or bias without further statistical analysis to control for confounding variables.



```{r summarise-station-level-hit-rates,  echo=FALSE, warning=FALSE, message = FALSE}



# Step 1: Define racial ordering
special_races <- c("Mediterranean/Middle Eastern – Unusable", "Missing")
first_race <- "White"

ordered_races <- dat %>%
  filter(!is.na(Racial.appearance.transformed)) %>%
  distinct(Racial.appearance.transformed) %>%
  pull() %>%
  unique() %>%
  {c(first_race, sort(setdiff(., c(first_race, special_races))), special_races)}

# Step 2: Summarise searches and hit rates by Station
station_summary_long <- dat %>%
  filter(
    Contact.Type == "P",
    Area.type == "Metro",
    !is.na(Found),
    !is.na(`Search.type - Drugs`),
    !is.na(Racial.appearance.transformed),
    !is.na(Station.uniform)
  ) %>%
  mutate(
    Found.Flag = ifelse(Found == "Yes", 1, 0),
    Race = Racial.appearance.transformed
  ) %>%
  group_by(Station.uniform, Race) %>%
  summarise(
    Searches = n(),
    Hits = sum(Found.Flag),
    HitRate = round(mean(Found.Flag), 3),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0(Searches, " (", sprintf("%.1f", HitRate * 100), "%)"),
    Race = factor(Race, levels = ordered_races)
  ) %>%
  select(Station.uniform, Race, label)

# Step 3: Pivot wide
station_table <- station_summary_long %>%
  pivot_wider(names_from = Race, values_from = label) %>%
  select(Station = Station.uniform, any_of(ordered_races))

# Step 4: Save Excel file
write_xlsx(station_table, "station_racial_search_summary.xlsx")

# Step 5: Display interactive datatable
datatable(
  station_table,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = list('copy', 'excel'),
    scrollX = TRUE,
    pageLength = 15,
    autoWidth = TRUE
  ),
  rownames = FALSE,
  class = 'compact stripe hover',
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Drug Searches and Hit Rates by Racial Appearance (Metro Victoria, by Police Station)'
  )
)

```


# Appendix – Ethnic Appearance Mapping

```{r mapping-categories-2022-2023, echo = FALSE}
dat %>%
  count(Racial.Appearance.original, Racial.appearance.transformed) %>%
  kable()
```
